\chapter{Large Language Models: Architecture and Practice}
\label{ch:llm-foundations}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Derive the vanishing-gradient bound for RNNs and explain how the LSTM cell
        state resolves it.
  \item State the scaled dot-product attention formula, prove the $\sqrt{d_k}$
        variance normalisation, and describe multi-head attention.
  \item Explain temperature sampling, top-$p$ nucleus sampling, and beam search,
        and choose the right scheme for a given financial NLP task.
  \item Describe constrained decoding and implement structured JSON extraction
        from an LLM using Python.
  \item Build a retrieval-augmented generation pipeline for financial documents.
  \item Explain knowledge distillation, LoRA fine-tuning, and quantisation.
  \item Apply at least three state-of-the-art hallucination-mitigation techniques
        to a financial analysis task.
\end{enumerate}
\end{remark}


% --- moved from ch01 ---
\section{Document-Level Representations}
\label{sec:doc-repr}

Word embeddings assign vectors to individual tokens.  In most
financial NLP applications the natural unit of analysis is longer: an
earnings call transcript (several thousand words), a 10-K filing
(tens of thousands of words), a news headline (ten words), or a
research note (hundreds of words).  To use embedding-based features
at the document level we need an aggregation scheme that maps a
sequence of token embeddings to a single fixed-size vector.

% ----------------------------------------------------------------
\subsection{Averaging Word Vectors}
\label{sec:doc-repr-avg}

The simplest aggregation is the unweighted arithmetic mean.

\begin{definition}[Mean Embedding]
\label{def:mean-embedding}
Let $d = (w_1, w_2, \dots, w_n)$ be a document of $n$ tokens and let
$\varphi : \mathcal{V} \to \mathbb{R}^d$ be a word embedding.
The \emph{mean embedding} of $d$ is
\begin{equation}
  \bar{\boldsymbol{\varphi}}(d)
  \;=\;
  \frac{1}{n} \sum_{i=1}^{n} \varphi(w_i)
  \;\in\; \mathbb{R}^d.
  \label{eq:mean-embedding}
\end{equation}
\end{definition}

Mean embedding is algebraically equivalent to multiplying the
Bag-of-Words term-frequency vector $\mathbf{tf}(d) \in \mathbb{R}^V$
by the (column-normalised) embedding matrix.  Specifically, let
$\mathbf{tf}(d)_j = \text{count}(w_j, d)/n$ be the relative frequency
of word $w_j$ in document $d$.  Then
\begin{equation}
  \bar{\boldsymbol{\varphi}}(d)
  \;=\;
  \mathbf{W}^{\top} \mathbf{tf}(d)
  \;\in\; \mathbb{R}^d,
  \label{eq:mean-embedding-bow}
\end{equation}
where $\mathbf{W} \in \mathbb{R}^{V \times d}$ is the embedding matrix
from Definition~\ref{def:word-embedding}.  This shows that mean
embedding is a linear operation on the standard BoW representation,
projecting from $V$-dimensional one-hot/count space into the dense
$d$-dimensional embedding space via the weight matrix $\mathbf{W}$.

\begin{proposition}
\label{thm:mean-embed-is-bow}
Mean embedding is a Bag-of-Words model.  Specifically, if
$\hat{\mathbf{W}} = [\varphi(w_1) \;\cdots\; \varphi(w_V)]^{\top}
\in \mathbb{R}^{V \times d}$ is the full embedding matrix, then
\[
  \bar{\boldsymbol{\varphi}}(d)
  \;=\;
  \hat{\mathbf{W}}^{\top} \mathbf{tf}(d).
\]
In particular, the mean embedding is invariant to permutations of the
token sequence; word order is discarded.
\end{proposition}
\begin{proof}
  Expand \eqref{eq:mean-embedding} using the row-selection
  interpretation $\varphi(w_i) = \hat{\mathbf{W}}^{\top}\mathbf{e}_{w_i}$,
  then use linearity:
  \[
    \frac{1}{n}\sum_{i=1}^n \hat{\mathbf{W}}^{\top}\mathbf{e}_{w_i}
    \;=\;
    \hat{\mathbf{W}}^{\top}
    \!\underbrace{\left(\frac{1}{n}\sum_{i=1}^n \mathbf{e}_{w_i}\right)}_{\mathbf{tf}(d)}
    \;=\;
    \hat{\mathbf{W}}^{\top}\mathbf{tf}(d). \qedhere
  \]
\end{proof}

Despite its simplicity and computational efficiency, mean embedding
has several well-documented weaknesses:
\begin{enumerate}
  \item \textbf{Equal weighting.}  Common words such as \emph{the},
        \emph{is}, and \emph{of} contribute as much to the mean as
        domain-specific terms such as \emph{default} or
        \emph{liquidity}.  Stop-word removal partially addresses this,
        but introduces a pre-processing dependency.
  \item \textbf{No word order.}  The mean embedding of
        ``\emph{the company did not meet earnings expectations}'' is
        identical to that of ``\emph{earnings expectations met the
        company not did}'' — a garbled sentence that contains the same
        words.  Syntactic structure, negation, and conditionals are
        invisible.
  \item \textbf{Poor performance on long documents.}  As $n$ grows,
        the mean converges (by the law of large numbers) toward the
        corpus-wide average embedding, losing document-specific
        signal.  Empirically, mean embedding performs well for short
        texts (headlines, short paragraphs) but degrades for filings
        or call transcripts \cite{reimers2019}. % [CITE: Reimers & Gurevych (2019). Sentence-BERT: Sentence Embeddings using Siamese BERT-Networks. EMNLP]
  \item \textbf{Out-of-vocabulary words.}  Words absent from
        $\mathcal{V}$ (e.g.\ rare tickers, neologisms) must be mapped
        to a special \texttt{<UNK>} vector or dropped, both of which
        introduce noise.
\end{enumerate}

% ----------------------------------------------------------------
\subsection{Weighted Aggregation and Sentence Transformers}
\label{sec:doc-repr-sbert}

\subsubsection{TF-IDF Weighted Averaging}

A natural refinement is to weight each word's embedding by its
TF-IDF score, up-weighting terms that are rare in the corpus but
frequent in the document and down-weighting common uninformative words.

\begin{definition}[TF-IDF Weighted Embedding]
\label{def:tfidf-embedding}
Let $\text{tfidf}(w_i, d)$ denote the TF-IDF weight of token $w_i$
in document $d$ (see Section~\ref{sec:classical-tfidf}).  The
\emph{TF-IDF weighted embedding} is
\begin{equation}
  \boldsymbol{\varphi}_{\mathrm{tfidf}}(d)
  \;=\;
  \frac{\displaystyle\sum_{i=1}^{n} \text{tfidf}(w_i, d)\; \varphi(w_i)}
       {\displaystyle\sum_{i=1}^{n} \text{tfidf}(w_i, d)}.
  \label{eq:tfidf-embedding}
\end{equation}
\end{definition}

This is a weighted arithmetic mean with weights $\alpha_i =
\text{tfidf}(w_i, d) / \sum_j \text{tfidf}(w_j, d) > 0$.  By
design, $\sum_i \alpha_i = 1$, so $\boldsymbol{\varphi}_{\mathrm{tfidf}}(d)$
lies in the convex hull of the token embeddings.  TF-IDF weighting
consistently outperforms uniform averaging on document classification
and similarity tasks when the embedding quality is held fixed
\cite{arora2017}. % [CITE: Arora et al. (2017). A Simple but Tough-to-Beat Baseline for Sentence Embeddings. ICLR]

\begin{remark}
\label{rem:sif}
A closely related approach, \emph{Smooth Inverse Frequency} (SIF)
\cite{arora2017}, % [CITE: same as above]
replaces TF-IDF weights with
$\alpha_i = a / (a + p(w_i))$,
where $p(w_i)$ is the unigram probability of word $w_i$ in the
training corpus and $a \approx 10^{-3}$ is a smoothing constant.
After computing the weighted sum, SIF removes the first principal
component, which typically captures uninformative ``common discourse''
variance.  SIF embeddings match or exceed supervised methods on
several semantic textual similarity benchmarks at negligible
computational cost.
\end{remark}

\subsubsection{The Limitation of Static Embeddings: Polysemy}

Both Word2Vec and GloVe assign a \emph{single, fixed} vector to each
vocabulary item.  This is adequate for words with a single dominant
sense, but fails for polysemous words — words with multiple distinct
meanings depending on context.  In finance, this problem is pervasive:
\begin{itemize}
  \item \emph{Bank} can mean a financial institution or the riverbank.
  \item \emph{Yield} can mean bond yield, crop yield, or to give way.
  \item \emph{Position} can mean a financial position (long/short) or
        a physical location.
  \item \emph{Default} can mean failure to repay a debt or the
        default setting of a software parameter.
\end{itemize}
Under a static embedding, the vector for \emph{bank} is a single
point in $\mathbb{R}^d$ that is a weighted average of both senses,
equidistant from financial and geographic contexts.  This conflation
can harm downstream tasks: a sentiment model that associates
\emph{bank} with positive sentiment in financial reporting will
incorrectly transfer that sentiment to sentences about river flooding.

\subsubsection{Contextual Embeddings and Sentence-BERT}

The solution to polysemy is to produce embeddings that depend on the full
surrounding context.  ELMo \cite{peters2018} % [CITE: Peters et al. (2018). Deep contextualized word representations. NAACL]
introduced deep contextual embeddings from a bidirectional language
model; BERT \cite{devlin2019} % [CITE: Devlin et al. (2019). BERT: Pre-training of Deep Bidirectional Transformers for Language Understanding. NAACL]
later generalised this with a transformer architecture.  In a BERT
encoder, the representation of the token \emph{bank} at layer $\ell$
is a function of the entire input sentence, so it will differ in
``\emph{The Federal Reserve Bank raised rates}'' versus
``\emph{The river bank flooded after the storm}.''

Sentence-BERT (SBERT) addresses the gap between contextual word representations and sentence-level embeddings. Vanilla BERT was designed for token-level tasks (named entity
recognition, question answering) rather than for producing
sentence- or document-level representations.  Naively averaging the
BERT token representations (mean pooling) or using the \texttt{[CLS]}
token produces suboptimal sentence embeddings for semantic similarity
tasks \cite{reimers2019}. % [CITE: Reimers & Gurevych (2019). Sentence-BERT. EMNLP]

Sentence-BERT (SBERT) \cite{reimers2019} % [CITE: same]
addresses this by fine-tuning a BERT encoder on a Natural Language
Inference (NLI) objective using a \emph{Siamese} architecture:
two weight-sharing BERT encoders process a sentence pair
$(s_a, s_b)$, producing pooled representations
$\mathbf{u} = \text{POOL}(\text{BERT}(s_a))$ and
$\mathbf{v} = \text{POOL}(\text{BERT}(s_b))$.
The model is trained to predict whether $s_a$ and $s_b$ are
semantically equivalent (entailment), contradictory, or unrelated,
using a classification head over the concatenation
$[\mathbf{u};\, \mathbf{v};\, |\mathbf{u} - \mathbf{v}|]$.

\begin{definition}[Mean Pooling in SBERT]
\label{def:sbert-pool}
Let $\mathbf{H} = [\mathbf{h}_1, \mathbf{h}_2, \dots, \mathbf{h}_n]
\in \mathbb{R}^{n \times d_\text{model}}$
be the matrix of last-layer hidden states from the BERT encoder for an
$n$-token input sequence (including special tokens \texttt{[CLS]} and
\texttt{[SEP]}).  Under \emph{mean pooling}, the sentence embedding is
\begin{equation}
  \mathbf{s}
  \;=\;
  \frac{1}{n}\sum_{i=1}^{n} \mathbf{h}_i \cdot m_i,
  \label{eq:sbert-mean-pool}
\end{equation}
where $m_i \in \{0, 1\}$ is an attention mask excluding padding tokens.
\end{definition}

BERT's \texttt{[CLS]} token is prepended to every input sequence and
its final hidden state $\mathbf{h}_{\texttt{[CLS]}}$ is used in
classification tasks; one might expect it to encode sentence-level
meaning.  In practice, without explicit fine-tuning on semantic
similarity tasks, the \texttt{[CLS]} representation performs
surprisingly poorly as a general-purpose sentence embedding
\cite{reimers2019}. % [CITE: same]
Mean pooling over all token representations consistently outperforms
\texttt{[CLS]} pooling on standard semantic textual similarity (STS)
benchmarks.  After SBERT fine-tuning, both pooling strategies converge
to high-quality representations, with mean pooling retaining a slight
advantage.

A key practical advantage of SBERT is that sentence embeddings can be
\emph{pre-computed and indexed}: once the corpus of $N$ documents has
been encoded, semantic search reduces to a nearest-neighbour query in
$\mathbb{R}^{d_{\text{model}}}$, which can be accelerated with
approximate nearest-neighbour (ANN) data structures (e.g.\ FAISS
\cite{johnson2019}, % [CITE: Johnson et al. (2019). Billion-scale similarity search with GPUs. IEEE TPAMI]
HNSW).  Cross-encoding a query against all $N$ documents with vanilla
BERT would require $N$ separate forward passes; SBERT bi-encoding
requires only one forward pass per document, reducing inference time
from $O(N)$ to $O(1)$ at query time.


\begin{example}
\label{ex:doc-similarity-finance}
The following applications of contextual document embeddings are
documented in the finance literature:
\begin{enumerate}
  \item \textbf{Earnings call similarity.}
        Encode all earnings call transcripts in a corpus using SBERT
        (or a finance-tuned variant such as FinBERT-SBERT).  At each
        reporting period, retrieve the $k$ most similar past calls
        using cosine similarity.  The similarity score can serve as a
        feature for predicting short-window abnormal returns
        \cite{ko2020}. % [CITE: Ko et al. (2020). or relevant FinNLP earnings similarity paper]
  \item \textbf{Merger target identification.}
        Encode 10-K business description sections for all public firms.
        High cosine similarity between two firms' descriptions predicts
        product-market proximity, which is a strong predictor of
        acquisition likelihood \cite{frattaroli2019}. % [CITE: relevant M&A text paper]
  \item \textbf{Regulatory document search.}
        A compliance team encodes a large corpus of regulatory guidance
        documents (Basel III, IFRS 9, MiFID II).  When a new
        regulation is published, SBERT-based retrieval identifies
        the most relevant precedents, replacing keyword search with
        semantic search.
  \item \textbf{Factor model narrative alignment.}
        Embed analyst research notes and quarterly earnings guidance.
        Measure the cosine similarity between the guidance embedding
        and subsequent actual results narratives as a proxy for
        management credibility or guidance quality.
\end{enumerate}
\end{example}

\begin{remark}
\label{rem:context-window-limit}
BERT-based models impose a maximum context length (typically 512
tokens for standard BERT, 4096 or more for long-context variants such
as Longformer \cite{beltagy2020}). % [CITE: Beltagy et al. (2020). Longformer: The Long-Document Transformer. arXiv:2004.05150]
Long financial documents (full 10-K filings, lengthy earnings call
transcripts) must be handled by one of three strategies:
(i) \emph{truncation} — only the first 512 tokens are encoded, which
may miss material information in later sections;
(ii) \emph{chunking} — the document is split into overlapping windows,
each encoded independently, and the chunk embeddings are aggregated
(e.g.\ by mean pooling or weighted by position); or
(iii) \emph{hierarchical encoding} — sentences are encoded individually
and a second-level encoder (another transformer or an LSTM) aggregates
sentence embeddings into a document embedding.  The choice among these
strategies introduces a design trade-off between completeness and
computational cost that has no universal solution; practitioners should
benchmark all three on their specific downstream task.
\end{remark}

% ---------------------------------------------------------------
\section{Sequential Models: From RNNs to Attention}
\label{sec:sequential}

The representations developed in Sections~\ref{sec:embeddings} and~\ref{sec:doc-repr}
treat documents as \emph{bags} of tokens: order is either ignored or captured only
crudely through $n$-gram features.
Financial text, however, is inherently sequential.
An earnings-call transcript unfolds over time; a regulatory filing builds an argument
across hundreds of sentences; the sentiment of \textit{the company did not miss its
targets} depends on the negation appearing \emph{before} the noun phrase.
This section traces the evolution of architectures designed to model sequences:
vanilla recurrent networks, gated variants that address their chief failure mode,
and the attention mechanism that ultimately superseded them.

% ---
\subsection{Recurrent Neural Networks and the Vanishing Gradient Problem}
\label{sec:sequential-rnn}

\begin{definition}[Vanilla Recurrent Neural Network]
\label{def:rnn}
Let $\mathbf{x}_t \in \mathbb{R}^{d_x}$ be the input vector at time step $t$
and $\mathbf{h}_t \in \mathbb{R}^{d_h}$ be the hidden (state) vector.
A \emph{vanilla recurrent neural network} (RNN) is defined by the recursion
\begin{align}
  \mathbf{h}_t &= \sigma\!\left(W_h\,\mathbf{h}_{t-1} + W_x\,\mathbf{x}_t
    + \mathbf{b}\right), \label{eq:rnn-hidden}\\
  \hat{\mathbf{y}}_t &= \mathrm{softmax}\!\left(W_y\,\mathbf{h}_t
    + \mathbf{b}_y\right), \label{eq:rnn-output}
\end{align}
where $W_h \in \mathbb{R}^{d_h \times d_h}$, $W_x \in \mathbb{R}^{d_h \times d_x}$,
$W_y \in \mathbb{R}^{d_y \times d_h}$ are learnable weight matrices,
$\mathbf{b} \in \mathbb{R}^{d_h}$ and $\mathbf{b}_y \in \mathbb{R}^{d_y}$ are
bias vectors, and $\sigma$ is an element-wise nonlinearity (sigmoid or $\tanh$).
The initial hidden state $\mathbf{h}_0$ is either $\mathbf{0}$ or a learnable
parameter.
\end{definition}

The recurrence in~\eqref{eq:rnn-hidden} allows information to propagate across time
steps: $\mathbf{h}_t$ summarises the history $\mathbf{x}_1, \ldots, \mathbf{x}_t$.
All parameters are \emph{shared} across time steps, giving the RNN its weight tying.

Training requires $\partial \mathcal{L} / \partial \theta$ for each parameter in
$\{W_h, W_x, W_y, \mathbf{b}, \mathbf{b}_y\}$,
where $\mathcal{L} = \sum_{t=1}^{T} \ell_t$.
Applying the chain rule gives the \emph{backpropagation through time} (BPTT)
formula \cite{werbos1990bptt}: % [CITE: Werbos 1990, Backpropagation Through Time: What It Does and How to Do It]
\begin{equation}
  \frac{\partial \mathcal{L}}{\partial \mathbf{h}_t}
  = \frac{\partial \mathcal{L}}{\partial \mathbf{h}_T}
    \prod_{k=t}^{T-1} \frac{\partial \mathbf{h}_{k+1}}{\partial \mathbf{h}_k},
  \label{eq:bptt-chain}
\end{equation}
where each Jacobian factor is
\begin{equation}
  \frac{\partial \mathbf{h}_{k+1}}{\partial \mathbf{h}_k}
  = \mathrm{diag}\!\left(\sigma'\!\left(W_h\,\mathbf{h}_k +
    W_x\,\mathbf{x}_{k+1} + \mathbf{b}\right)\right) W_h,
  \label{eq:rnn-jacobian}
\end{equation}
and $\mathrm{diag}(\mathbf{v})$ is the diagonal matrix with $\mathbf{v}$ on its
main diagonal.

\begin{proposition}[Vanishing and Exploding Gradients in RNNs]
\label{thm:vanish-explode}
Let $\rho = \|W_h\|_2$ (spectral norm) and $M_\sigma = \sup_z |\sigma'(z)|$.
Then
\begin{equation}
  \left\|\prod_{k=t}^{T-1}
    \frac{\partial \mathbf{h}_{k+1}}{\partial \mathbf{h}_k}\right\|_2
  \;\leq\; \left(M_\sigma\,\rho\right)^{T-t}.
  \label{eq:vanish-bound}
\end{equation}
In particular: (i) if $M_\sigma\,\rho < 1$, gradients vanish exponentially with
$T-t$; (ii) if $\rho > 1/M_\sigma$, they can grow exponentially
(\emph{exploding gradients}).
\end{proposition}

\begin{proof}
By sub-multiplicativity and~\eqref{eq:rnn-jacobian},
$\|\partial \mathbf{h}_{k+1}/\partial \mathbf{h}_k\|_2 \leq M_\sigma\,\rho$.
Applying this bound $T-t$ times yields~\eqref{eq:vanish-bound}.
For the logistic sigmoid $M_\sigma = 1/4$; for $\tanh$, $M_\sigma = 1$.
If $M_\sigma\,\rho < 1$ the right-hand side decays geometrically; the lower
explosive bound follows by considering Jacobians with a shared dominant eigenvector.
\end{proof}

\begin{remark}
Gradient clipping \cite{pascanu2013gradient} % [CITE: Pascanu et al. 2013, On the difficulty of training recurrent neural networks]
is the standard remedy for explosion, but there is no analogous in-architecture fix
for vanishing gradients in the vanilla RNN.
\end{remark}

These theoretical constraints have concrete consequences for financial text. An S\&P~500 earnings-call transcript is typically 8,000--12,000 tokens.
Management guidance on page~20 may negate a claim on page~3.
Proposition~\ref{thm:vanish-explode} implies the gradient from position $t = 100$
contributes essentially nothing to $W_h$ after $T - t \approx 9{,}900$ BPTT
steps --- making long-range financial dependencies practically unlearnable.

% ---
\subsection{LSTMs and GRUs}
\label{sec:sequential-lstm}

\subsubsection{Long Short-Term Memory}

The Long Short-Term Memory (LSTM) unit was introduced by \citet{hochreiter1997long} % [CITE: Hochreiter & Schmidhuber 1997, Long Short-Term Memory]
precisely to overcome the vanishing-gradient problem.
The key innovation is a \emph{cell state} $\mathbf{c}_t$ with an \emph{additive}
update path, gated by three sigmoid-activated vectors.

\begin{definition}[LSTM Equations]
\label{def:lstm}
Let $[\mathbf{h}_{t-1},\,\mathbf{x}_t] \in \mathbb{R}^{d_h + d_x}$ denote
concatenation. The LSTM update at step $t$ is:
\begin{align}
  \mathbf{f}_t &= \sigma\!\left(W_f\,[\mathbf{h}_{t-1},\,\mathbf{x}_t]
    + \mathbf{b}_f\right), \label{eq:lstm-forget}\\
  \mathbf{i}_t &= \sigma\!\left(W_i\,[\mathbf{h}_{t-1},\,\mathbf{x}_t]
    + \mathbf{b}_i\right), \label{eq:lstm-input}\\
  \mathbf{g}_t &= \tanh\!\left(W_g\,[\mathbf{h}_{t-1},\,\mathbf{x}_t]
    + \mathbf{b}_g\right), \label{eq:lstm-candidate}\\
  \mathbf{c}_t &= \mathbf{f}_t \odot \mathbf{c}_{t-1}
    + \mathbf{i}_t \odot \mathbf{g}_t, \label{eq:lstm-cell}\\
  \mathbf{o}_t &= \sigma\!\left(W_o\,[\mathbf{h}_{t-1},\,\mathbf{x}_t]
    + \mathbf{b}_o\right), \label{eq:lstm-output}\\
  \mathbf{h}_t &= \mathbf{o}_t \odot \tanh(\mathbf{c}_t), \label{eq:lstm-hidden}
\end{align}
where $\odot$ denotes element-wise product and
$W_f, W_i, W_g, W_o \in \mathbb{R}^{d_h \times (d_h + d_x)}$ are learnable
weight matrices.
\end{definition}

Intuitive interpretations:
\begin{itemize}
  \item \textbf{Forget gate} $\mathbf{f}_t \in (0,1)^{d_h}$:
        fraction of $\mathbf{c}_{t-1}$ to erase ($f_{t,j}\approx 0$: discard;
        $f_{t,j}\approx 1$: retain). In finance, clears irrelevant preamble
        sentiment as the transcript shifts to core earnings figures.
  \item \textbf{Input gate} $\mathbf{i}_t \in (0,1)^{d_h}$:
        fraction of candidate $\mathbf{g}_t$ to write; filters boilerplate
        language in 10-K filings.
  \item \textbf{Candidate} $\mathbf{g}_t \in (-1,1)^{d_h}$:
        new information from current input and recent context.
  \item \textbf{Output gate} $\mathbf{o}_t \in (0,1)^{d_h}$:
        which cell-state components are exposed as $\mathbf{h}_t$.
\end{itemize}

The reason the LSTM cell state avoids vanishing gradients lies in the additive structure of its update. The update~\eqref{eq:lstm-cell} is \emph{additive} in $\mathbf{g}_t$, unlike the
purely multiplicative vanilla RNN update~\eqref{eq:rnn-hidden}.
The cell-state gradient satisfies
\begin{equation}
  \frac{\partial \mathcal{L}}{\partial \mathbf{c}_t}
  = \frac{\partial \mathcal{L}}{\partial \mathbf{c}_{t+1}} \odot \mathbf{f}_{t+1}
    + \text{(terms via } \mathbf{h}_t\text{)}.
  \label{eq:lstm-cellgrad}
\end{equation}
When $\mathbf{f}_{t+1} \approx \mathbf{1}$, the gradient is transmitted without
decay --- the \emph{constant error carousel} of \citeauthor{hochreiter1997long}. % [CITE: Hochreiter & Schmidhuber 1997]

\subsubsection{Gated Recurrent Units}

\citet{chung2014empirical} % [CITE: Cho et al. 2014, Learning Phrase Representations using RNN Encoder-Decoder for Statistical Machine Translation]
proposed the GRU, which merges cell state and hidden state, reducing parameters
while retaining essential gating.

\begin{definition}[GRU Equations]
\label{def:gru}
\begin{align}
  \mathbf{z}_t &= \sigma\!\left(W_z\,[\mathbf{h}_{t-1},\,\mathbf{x}_t]\right),
  \label{eq:gru-update}\\
  \mathbf{r}_t &= \sigma\!\left(W_r\,[\mathbf{h}_{t-1},\,\mathbf{x}_t]\right),
  \label{eq:gru-reset}\\
  \tilde{\mathbf{h}}_t &= \tanh\!\left(
    W\,[\mathbf{r}_t \odot \mathbf{h}_{t-1},\,\mathbf{x}_t]\right),
  \label{eq:gru-candidate}\\
  \mathbf{h}_t &= (1 - \mathbf{z}_t) \odot \mathbf{h}_{t-1}
    + \mathbf{z}_t \odot \tilde{\mathbf{h}}_t,
  \label{eq:gru-hidden}
\end{align}
where $\mathbf{z}_t$ is the \emph{update gate}, $\mathbf{r}_t$ is the
\emph{reset gate}, and $W_z, W_r, W \in \mathbb{R}^{d_h \times (d_h + d_x)}$.
\end{definition}

A GRU has $\approx 3(d_h+d_x)d_h$ parameters versus $4(d_h+d_x)d_h$ for an LSTM.
LSTMs outperform GRUs on tasks with very long-range dependencies; GRUs are preferred
under data or compute constraints \cite{chung2014empirical}. % [CITE: Chung et al. 2014, Empirical Evaluation of Gated Recurrent Neural Networks on Sequence Modeling]

\subsubsection{Bidirectional RNNs}

A standard LSTM sees only the prefix $\mathbf{x}_1,\ldots,\mathbf{x}_t$.
A \emph{bidirectional} LSTM \cite{schuster1997bilstm} % [CITE: Schuster & Paliwal 1997, Bidirectional Recurrent Neural Networks]
runs one forward and one backward pass, concatenating results:
\begin{equation}
  \mathbf{h}_t =
  [\,\overrightarrow{\mathbf{h}}_t;\,\overleftarrow{\mathbf{h}}_t\,],
  \quad
  \overrightarrow{\mathbf{h}}_t =
    \mathrm{LSTM}_{\mathrm{fwd}}(\mathbf{x}_1,\ldots,\mathbf{x}_t),
  \quad
  \overleftarrow{\mathbf{h}}_t =
    \mathrm{LSTM}_{\mathrm{bwd}}(\mathbf{x}_T,\ldots,\mathbf{x}_t).
  \label{eq:bilstm}
\end{equation}
BiLSTMs were the dominant architecture for financial NER and sentiment classification
before Transformers; applications include credit-event extraction from news
\cite{araci2019finbert} % [CITE: Araci 2019, FinBERT: Financial Sentiment Analysis with Pre-trained Language Models]
and earnings-surprise prediction from analyst reports.

% ---
\subsection{The Attention Mechanism}
\label{sec:sequential-attention}

Even with gated architectures, an RNN encoder must compress the input into a
fixed-length vector before the decoder can generate output.
\citet{bahdanau2015neural} % [CITE: Bahdanau, Cho & Bengio 2015, Neural Machine Translation by Jointly Learning to Align and Translate]
proposed \emph{attention}, allowing the decoder to consult all encoder hidden states
at each step.

Let $\mathbf{h}_1, \ldots, \mathbf{h}_T \in \mathbb{R}^{d_h}$ be encoder hidden
states and $\mathbf{s}_t \in \mathbb{R}^{d_s}$ the decoder state at step $t$.

\begin{definition}[Additive (Bahdanau) Attention]
\label{def:bahdanau-attention}
The \emph{alignment score} between $\mathbf{s}_t$ and $\mathbf{h}_i$ is
\begin{equation}
  e_{t,i} = \mathbf{v}^{\top}
    \tanh\!\left(W_s\,\mathbf{s}_t + W_h\,\mathbf{h}_i\right),
  \label{eq:alignment-score}
\end{equation}
with $W_s \in \mathbb{R}^{d_a \times d_s}$, $W_h \in \mathbb{R}^{d_a \times d_h}$,
and $\mathbf{v} \in \mathbb{R}^{d_a}$ learnable.
The \emph{attention weights} and \emph{context vector} are:
\begin{align}
  \alpha_{t,i} &= \frac{\exp(e_{t,i})}{\sum_{j=1}^{T} \exp(e_{t,j})},
  \label{eq:attention-weights}\\
  \mathbf{c}_t &= \sum_{i=1}^{T} \alpha_{t,i}\,\mathbf{h}_i.
  \label{eq:context-vector}
\end{align}
\end{definition}

\begin{proposition}[Attention Weights Form a Probability Distribution]
\label{thm:attention-prob}
For any $e_{t,1},\ldots,e_{t,T} \in \mathbb{R}$, the weights
$\{\alpha_{t,i}\}_{i=1}^T$ in~\eqref{eq:attention-weights}
satisfy $\alpha_{t,i} > 0$ for all $i$ and $\sum_{i=1}^{T} \alpha_{t,i} = 1$.
\end{proposition}

\begin{proof}
Positivity: $\exp(\cdot) > 0$.
Summation: $\sum_i \alpha_{t,i} =
\sum_i \exp(e_{t,i}) / \sum_j \exp(e_{t,j}) = 1$. \qedhere
\end{proof}

The context vector $\mathbf{c}_t$ is a convex combination of encoder states.
In a financial summariser, the decoder generating a sentence about net income assigns
high $\alpha_{t,i}$ to tokens where income figures appear in the source.

Applying attention within a \emph{single} sequence --- each token attending to all
others --- yields \emph{self-attention}, the core primitive of the Transformer
(Section~\ref{sec:transformer}).
The dot-product attention of \citet{vaswani2017attention} % [CITE: Vaswani et al. 2017, Attention Is All You Need]
replaces the additive score~\eqref{eq:alignment-score} with a scaled inner product,
reducing cost from $O(Td_a)$ to $O(Td_k)$.

% ---------------------------------------------------------------
\section{The Transformer Architecture}
\label{sec:transformer}

The Transformer, introduced by \citet{vaswani2017attention}, % [CITE: Vaswani et al. 2017, Attention Is All You Need]
dispenses with recurrence and relies exclusively on attention to model dependencies.
Three major advantages motivate it for large-scale financial NLP\@:
(i) all token pairs are connected in $O(1)$ layers, eliminating the
vanishing-gradient bottleneck;
(ii) the attention computation is highly parallelisable;
and (iii) learned attention patterns are interpretable.

% ---
\subsection{Self-Attention (Mathematics)}
\label{sec:transformer-selfattn}

\begin{definition}[Scaled Dot-Product Attention]
\label{def:scaled-dot-product-attn}
Let $X \in \mathbb{R}^{n \times d_{\mathrm{model}}}$ be the $n$-token input matrix.
Given projection matrices
$W^Q, W^K \in \mathbb{R}^{d_{\mathrm{model}} \times d_k}$ and
$W^V \in \mathbb{R}^{d_{\mathrm{model}} \times d_v}$,
define
\begin{equation}
  Q = X W^Q \in \mathbb{R}^{n \times d_k},\quad
  K = X W^K \in \mathbb{R}^{n \times d_k},\quad
  V = X W^V \in \mathbb{R}^{n \times d_v}.
  \label{eq:qkv}
\end{equation}
The \emph{scaled dot-product attention} output is
\begin{equation}
  \mathrm{Attention}(Q, K, V)
  = \mathrm{softmax}\!\left(\frac{Q K^{\top}}{\sqrt{d_k}}\right) V
  \in \mathbb{R}^{n \times d_v},
  \label{eq:scaled-dot-product}
\end{equation}
where $\mathrm{softmax}$ is applied row-wise.
\end{definition}

$QK^{\top} \in \mathbb{R}^{n \times n}$; entry $(i,j)$ is
$\mathbf{q}_i^{\top}\mathbf{k}_j$.
Row-wise softmax gives the attention matrix
$A \in \mathbb{R}^{n \times n}$ with $A_{i,j} > 0$ and $\sum_j A_{i,j} = 1$
(Proposition~\ref{thm:attention-prob}).
The output $AV$ represents token $i$ as $\sum_j A_{i,j}\mathbf{v}_j$, a convex
combination of value vectors.

\begin{proposition}[$\sqrt{d_k}$ Scaling Stabilises Softmax Gradients]
\label{thm:sqrt-dk-scaling}
Suppose query and key components are i.i.d.\ mean-zero unit-variance.
Then
\begin{equation}
  \mathrm{Var}(\mathbf{q}_i^{\top} \mathbf{k}_j) = d_k,
  \label{eq:dot-product-var}
\end{equation}
so dividing by $\sqrt{d_k}$ normalises the pre-softmax variance to one.
\end{proposition}

\begin{proof}
Write $\mathbf{q}_i^{\top}\mathbf{k}_j = \sum_{\ell=1}^{d_k} q_\ell k_\ell$
with components mutually independent, $\mathrm{Var}(q_\ell) = \mathrm{Var}(k_\ell) = 1$.
Since $\mathbb{E}[q_\ell k_\ell] = 0$ and
$\mathbb{E}[(q_\ell k_\ell)^2] = \mathbb{E}[q_\ell^2]\mathbb{E}[k_\ell^2] = 1$,
each term has unit variance, so
$\mathrm{Var}(\mathbf{q}_i^{\top}\mathbf{k}_j) = d_k$.
For large $d_k$ the logits concentrate near $\pm\sqrt{d_k}$, pushing softmax into
its saturated regime where
$\partial\sigma(z)/\partial z = \sigma(z)(1-\sigma(z)) \approx 0$.
Dividing by $\sqrt{d_k}$ normalises variance to $1$, restoring gradient flow.
\end{proof}

In decoder-only models, token $i$ must not attend to $j > i$.
This is enforced by
\begin{equation}
  \tilde{S}_{ij} = \begin{cases}
    \mathbf{q}_i^{\top}\mathbf{k}_j/\sqrt{d_k} & j \leq i,\\
    -\infty & j > i,
  \end{cases}
  \label{eq:causal-mask}
\end{equation}
so that $\mathrm{softmax}(\tilde{S}_{i,:})$ assigns zero weight to future positions.

% ---
\subsection{Multi-Head Attention}
\label{sec:transformer-mha}

\begin{definition}[Multi-Head Attention]
\label{def:mha}
Let $W^Q_i, W^K_i \in \mathbb{R}^{d_{\mathrm{model}} \times d_k}$,
$W^V_i \in \mathbb{R}^{d_{\mathrm{model}} \times d_v}$ for $i = 1,\ldots,h$,
and $W^O \in \mathbb{R}^{h d_v \times d_{\mathrm{model}}}$.
Define $\mathrm{head}_i = \mathrm{Attention}(XW^Q_i, XW^K_i, XW^V_i)$ and
\begin{equation}
  \mathrm{MultiHead}(X)
  = \mathrm{Concat}(\mathrm{head}_1,\ldots,\mathrm{head}_h)\,W^O
  \in \mathbb{R}^{n \times d_{\mathrm{model}}}.
  \label{eq:multihead}
\end{equation}
In the original Transformer: $h = 8$, $d_{\mathrm{model}} = 512$, $d_k = d_v = 64$.
\end{definition}

Summing $O(n^2 d_k)$ per head over $h$ heads gives $O(n^2 d_{\mathrm{model}})$.
For $n = 4{,}096$ tokens, the attention matrix requires $\approx 134\,\mathrm{M}$
values per layer, motivating efficient variants such as FlashAttention
\cite{dao2022flashattention}. % [CITE: Dao et al. 2022, FlashAttention: Fast and Memory-Efficient Exact Attention with IO-Awareness]

Probing studies \cite{clark2019bert} % [CITE: Clark et al. 2019, What Does BERT Look at? An Analysis of BERT's Attention]
show lower-layer heads capture syntactic dependency relations (subject--verb
agreement) and deeper heads capture semantic associations (co-reference, entity
types). In financial models, heads specialise in temporal relations
(``last quarter'' $\to$ figures), entity linking, and negation scope.

\begin{remark}[Connection to the Kernel Trick]
\label{rem:kernel-trick}
The attention output for token $i$ can be written
$\sum_j k(\mathbf{q}_i,\mathbf{k}_j)\mathbf{v}_j / Z_i$,
where $k(\mathbf{q},\mathbf{k}) = \exp(\mathbf{q}^\top\mathbf{k}/\sqrt{d_k})$
is an exponential kernel and $Z_i$ is the partition function.
The kernel analogy \cite{tsai2019transformer} % [CITE: Tsai et al. 2019, Transformer Dissection: A Unified Understanding of Transformer's Attention via the Lens of Kernel]
motivates random-feature approximations (Performers, Random Feature Attention) that
reduce the $O(n^2)$ complexity to $O(n)$.
\end{remark}

% ---
\subsection{Positional Encoding}
\label{sec:transformer-pos}

The attention operation~\eqref{eq:scaled-dot-product} is permutation-equivariant.
Positional information must therefore be injected explicitly.

\begin{definition}[Sinusoidal Positional Encoding]
\label{def:positional-encoding}
For position $\mathrm{pos} \in \{0,\ldots,n-1\}$ and index
$i \in \{0,\ldots,d_{\mathrm{model}}/2 - 1\}$:
\begin{align}
  \mathrm{PE}_{(\mathrm{pos},\,2i)}
    &= \sin\!\left(\tfrac{\mathrm{pos}}{10000^{2i/d_{\mathrm{model}}}}\right),
  \label{eq:pe-sin}\\
  \mathrm{PE}_{(\mathrm{pos},\,2i+1)}
    &= \cos\!\left(\tfrac{\mathrm{pos}}{10000^{2i/d_{\mathrm{model}}}}\right).
  \label{eq:pe-cos}
\end{align}
The positionally-augmented input is
$\tilde{X}_{\mathrm{pos},:} = X_{\mathrm{pos},:} + \mathrm{PE}_{\mathrm{pos},:}$.
\end{definition}

Let $\omega_i = 10000^{-2i/d_{\mathrm{model}}}$.
For small $i$, $\omega_i \approx 1$ and the encoding oscillates rapidly (fine
position); for large $i$, $\omega_i \approx 0$ (coarse position).
This multi-scale design is analogous to a Fourier basis.

\begin{proposition}[Relative-Position Linearity of Sinusoidal Encoding]
\label{thm:pe-relative}
For any fixed offset $k \in \mathbb{Z}$ and index $i$,
\begin{equation}
  \begin{pmatrix}
    \mathrm{PE}_{(\mathrm{pos}+k,\,2i)} \\
    \mathrm{PE}_{(\mathrm{pos}+k,\,2i+1)}
  \end{pmatrix}
  = R(\omega_i k)
  \begin{pmatrix}
    \mathrm{PE}_{(\mathrm{pos},\,2i)} \\
    \mathrm{PE}_{(\mathrm{pos},\,2i+1)}
  \end{pmatrix},
  \label{eq:pe-rotation}
\end{equation}
where $R(\theta) = \bigl(\begin{smallmatrix} \cos\theta & -\sin\theta \\ \sin\theta
& \cos\theta \end{smallmatrix}\bigr)$ depends only on $k$, not on $\mathrm{pos}$.
\end{proposition}

\begin{proof}
Set $s = \sin(\omega_i\,\mathrm{pos})$, $c = \cos(\omega_i\,\mathrm{pos})$.
By the angle-addition identities:
\begin{align*}
  \sin(\omega_i(\mathrm{pos}+k)) &= s\cos(\omega_i k) + c\sin(\omega_i k),\\
  \cos(\omega_i(\mathrm{pos}+k)) &= c\cos(\omega_i k) - s\sin(\omega_i k).
\end{align*}
In matrix form this is
$R(\omega_i k)[\mathrm{PE}_{(\mathrm{pos},2i)},\mathrm{PE}_{(\mathrm{pos},2i+1)}]^\top$,
where $R(\omega_i k)$ depends only on $k$ and $\omega_i$, not $\mathrm{pos}$.
\end{proof}

Proposition~\ref{thm:pe-relative} implies
$\mathrm{PE}_{\mathrm{pos}}^\top\mathrm{PE}_{\mathrm{pos}'}$ depends only on
$|\mathrm{pos}-\mathrm{pos}'|$, making relative position linearly decodable.
This motivated relative positional encodings \cite{shaw2018relative} % [CITE: Shaw et al. 2018, Self-Attention with Relative Position Representations]
and Rotary Position Embeddings (RoPE) \cite{su2024rope}, % [CITE: Su et al. 2024, RoFormer: Enhanced Transformer with Rotary Position Embedding]
which embed the rotation directly into query--key dot products.

% ---
\subsection{Encoder--Decoder Structure}
\label{sec:transformer-encdec}

The encoder layer transforms its input through self-attention followed by a feed-forward network. Let $X^{(\ell)} \in \mathbb{R}^{n \times d_{\mathrm{model}}}$ be the layer-$\ell$ input.
With residual connections \cite{he2016resnet} % [CITE: He et al. 2016, Deep Residual Learning for Image Recognition]
and Layer Normalisation \cite{ba2016layernorm}: % [CITE: Ba et al. 2016, Layer Normalization]
\begin{align}
  Z^{(\ell)}   &= \mathrm{LayerNorm}\!\left(
    X^{(\ell)} +
    \mathrm{MultiHead}(X^{(\ell)}, X^{(\ell)}, X^{(\ell)})\right),
  \label{eq:enc-layer-z}\\
  X^{(\ell+1)} &= \mathrm{LayerNorm}\!\left(
    Z^{(\ell)} + \mathrm{FFN}(Z^{(\ell)})\right),
  \label{eq:enc-layer-out}
\end{align}
with
\begin{equation}
  \mathrm{FFN}(\mathbf{z})
  = \max(0,\, \mathbf{z} W_1 + \mathbf{b}_1)\,W_2 + \mathbf{b}_2,
  \label{eq:ffn}
\end{equation}
where $W_1 \in \mathbb{R}^{d_{\mathrm{model}} \times d_{\mathrm{ff}}}$,
$W_2 \in \mathbb{R}^{d_{\mathrm{ff}} \times d_{\mathrm{model}}}$,
$d_{\mathrm{ff}} = 4\,d_{\mathrm{model}}$.
The $\max(0,\cdot)$ is ReLU; later variants use GELU \cite{hendrycks2016gelu}. % [CITE: Hendrycks & Gimpel 2016, Gaussian Error Linear Units]

Residual connections ensure direct gradient pathways and LayerNorm stabilises
activations across the $d_{\mathrm{model}}$ dimension.

The decoder layer consists of three sub-layers: (i) masked self-attention
(mask~\eqref{eq:causal-mask}), (ii) \emph{cross-attention} with encoder output, and
(iii) FFN.
With $Y^{(\ell)} \in \mathbb{R}^{m \times d_{\mathrm{model}}}$ and
$H = X^{(L_{\mathrm{enc}})}$:
\begin{equation}
  \mathrm{CrossAttn}(Y^{(\ell)}, H)
  = \mathrm{MultiHead}\!\left(Y^{(\ell)} W^Q,\; H W^K,\; H W^V\right),
  \label{eq:cross-attn}
\end{equation}
where $Y^{(\ell)}$ provides queries and $H$ provides keys and values.
This is the direct parallel of the Bahdanau context vector $\mathbf{c}_t$
from~\eqref{eq:context-vector}, computed in a single matrix operation.

\begin{remark}
Figure~\ref{fig:transformer-architecture} shows the architecture schematically.
% [FIGURE: fig:transformer-architecture --- Standard Transformer encoder-decoder
%  diagram with token embedding, positional encoding, $N$ stacked encoder layers,
%  $N$ stacked decoder layers, and output projection.
%  After Vaswani et al.\ (2017) Figure~1.]
The base model uses $L = 6$ layers; the large model uses $L = 24$.
Modern LLMs use $L = 32$--$96$.
\end{remark}

% ---
\subsection{Pre-training, Fine-tuning, and BERT vs.\ GPT}
\label{sec:transformer-pretrain}

\begin{definition}[Pre-training]
\label{def:pretraining}
\emph{Pre-training} is the first phase: training a language model on a large
unlabelled corpus using a self-supervised objective whose labels are derived from
the data itself (predicting masked tokens, or predicting the next token).
The pre-trained weights provide initialisation and inductive bias for fine-tuning.
\end{definition}

\begin{definition}[Fine-tuning]
\label{def:finetuning}
\emph{Fine-tuning} is the second phase: updating pre-trained parameters on a smaller
labelled dataset $\mathcal{D}_{\mathrm{task}}$ for a specific downstream task.
Pre-training substantially reduces labelled-data and compute requirements.
\end{definition}

\subsubsection{Masked Language Modelling: BERT}

BERT \cite{devlin2019bert} % [CITE: Devlin et al. 2019, BERT: Pre-training of Deep Bidirectional Transformers for Language Understanding]
(Bidirectional Encoder Representations from Transformers) is an encoder-only
Transformer trained on two self-supervised objectives.

The Masked Language Model (MLM) objective selects $15\%$ of input tokens; $80\%$ replaced with \texttt{[MASK]},
$10\%$ with a random token, and $10\%$ left unchanged.
The model predicts original tokens at masked positions:
\begin{equation}
  \mathcal{L}_{\mathrm{MLM}}(\theta)
  = -\sum_{i \in \mathcal{M}}
    \log p_\theta\!\left(w_i \mid \mathbf{x}_{\backslash \mathcal{M}}\right),
  \label{eq:mlm-loss}
\end{equation}
where $\mathcal{M}$ is the set of masked positions and
$\mathbf{x}_{\backslash\mathcal{M}}$ the unmasked context.
Bidirectional attention means both left and right context inform each prediction.

The Next Sentence Prediction (NSP) objective uses a \texttt{[CLS]} binary head to predict whether sentence $B$ follows sentence $A$.
\citet{liu2018} % [CITE: Liu et al. 2019, RoBERTa: A Robustly Optimized BERT Pretraining Approach]
later showed NSP is unhelpful and removed it in RoBERTa.

\subsubsection{Causal Language Modelling: GPT}

GPT \cite{radford2018gpt} % [CITE: Radford et al. 2018, Improving Language Understanding by Generative Pre-Training]
and successors (GPT-2, GPT-3, GPT-4) are decoder-only models trained on the causal
language modelling (CLM) objective:
\begin{equation}
  \mathcal{L}_{\mathrm{CLM}}(\theta)
  = -\sum_{t=1}^{T} \log p_\theta(w_t \mid w_1, \ldots, w_{t-1}).
  \label{eq:clm-loss}
\end{equation}
The causal mask~\eqref{eq:causal-mask} makes this a proper autoregressive density
model, naturally supporting text generation.

\subsubsection{Encoder-Only vs.\ Decoder-Only: Task Suitability}

\begin{itemize}
  \item \textbf{Encoder-only (BERT-family):}
        Bidirectional attention conditioned on the full input.
        Ideal for classification (sentiment, credit rating), token-level extraction
        (NER, financial metric extraction), and semantic similarity.

  \item \textbf{Decoder-only (GPT-family):}
        Natural generative engine for summarisation, report drafting, and
        instruction following (RLHF fine-tuning \cite{ouyang2022rlhf}). % [CITE: Ouyang et al. 2022, Training language models to follow instructions with human feedback]
        Unidirectional attention is suboptimal for tasks requiring full-context
        understanding.

  \item \textbf{Encoder--decoder (T5, BART):}
        Well suited to sequence-to-sequence tasks: abstractive summarisation of
        earnings calls, multi-document QA \cite{lewis2020bart}. % [CITE: Lewis et al. 2020, BART: Denoising Sequence-to-Sequence Pre-training for Natural Language Generation, Translation, and Comprehension]
\end{itemize}

\subsubsection{Domain-Adapted Financial Models}

General-purpose models under-perform on financial text due to distinctive vocabulary
(EBITDA, contango, covenants), regulatory legalese, and distribution shift from
web-crawl pre-training data.

\begin{example}[FinBERT]
\label{ex:finbert}
\citet{araci2019finbert} % [CITE: Araci 2019, FinBERT: Financial Sentiment Analysis with Pre-trained Language Models]
released FinBERT, a BERT-base model further pre-trained on Reuters/Bloomberg news,
10-K filings, and earnings call transcripts.
On the Financial PhraseBank dataset \cite{malo2014phrasebank}, % [CITE: Malo et al. 2014, Good Debt or Bad Debt: Detecting Semantic Orientations in Economic Texts]
FinBERT achieves over $97\%$ accuracy on financial phrase sentiment classification,
outperforming general BERT-base by approximately 8 percentage points.
\end{example}

\begin{example}[BloombergGPT]
\label{ex:bloomberggpt}
\citet{wu2023bloomberggpt} % [CITE: Wu et al. 2023, BloombergGPT: A Large Language Model for Finance]
trained BloombergGPT, a 50-billion-parameter decoder-only model, on $363$ billion
tokens from Bloomberg financial archives plus $345$ billion general tokens.
The mixed strategy outperformed purely financial and purely general baselines on
ConvFinQA (numerical reasoning over tables) and FiQA (opinion-based QA).
BloombergGPT illustrates that domain adaptation at scale requires both domain
knowledge and general linguistic competence.
\end{example}

\begin{remark}
Continued pre-training of an open-source base model (Llama, Mistral) on
$10$--$50$ billion domain tokens, followed by instruction fine-tuning on curated
financial datasets, is currently the most cost-effective domain-adaptation strategy
for practitioners.
\end{remark}


\section{The Modern LLM Landscape}
\label{sec:landscape}

The preceding sections established the architectural foundations of the Transformer
and the pre-training paradigm.  We now survey the landscape of deployed large language
models (LLMs) as it stands in the mid-2020s, drawing distinctions that will recur
throughout this book.

\subsection{Generative Models vs.\ Reasoning Models}
\label{sec:landscape-genreason}

\begin{definition}[Autoregressive Generative Language Model]
\label{def:autoregressive}
An \emph{autoregressive language model} is a probability distribution over sequences
of tokens.  Given a vocabulary $\mathcal{V}$, the model factorises the joint
probability of a sequence $\mathbf{x} = (x_1, x_2, \ldots, x_n)$ as a product of
conditional probabilities:
\begin{equation}
  P(x_1, x_2, \ldots, x_n) = \prod_{t=1}^{n} P\!\left(x_t \mid x_1, \ldots, x_{t-1}\right).
  \label{eq:autoregressive}
\end{equation}
Generation proceeds by sampling one token at a time, conditioning on all previously
generated tokens.  Each forward pass through the network produces a probability
distribution over $\mathcal{V}$; the next token is obtained by drawing from this
distribution (or selecting its mode, as discussed in
Section~\ref{sec:api-concepts}).
\end{definition}

This factorisation is not merely a modelling convenience: it imposes a \emph{causal}
(left-to-right) structure that corresponds naturally to how language and, indeed, how
financial disclosures are read.  An earnings call transcript, a research report, or a
regulatory filing all unfold sequentially, and the meaning of a phrase frequently
depends on all that precedes it.

Standard autoregressive generation produces output token by token in a \emph{single
chain of forward passes}: given a prompt of $L$ tokens, the model runs $L$ attention
computations in parallel (using a KV-cache for efficiency) to generate each successive
output token.  For many financial NLP tasks---sentiment labelling, named-entity
recognition, brief summarisation---this single-pass regime is entirely adequate.

However, it is ill-suited to tasks requiring multi-step deliberation: valuing a firm
under a discounted-cash-flow (DCF) framework from a noisy earnings call, reasoning
about whether a merger satisfies regulatory criteria across three jurisdictions, or
constructing a quantitative argument from a numerical table embedded in a 10-K filing.
Such tasks require the model to \emph{plan}, backtrack, and verify intermediate
results before committing to a final answer.  This motivated a second paradigm.

\begin{definition}[Reasoning (Extended-Thinking) Language Model]
\label{def:reasoning}
A \emph{reasoning model} is an autoregressive language model trained (typically via
reinforcement learning with verifiable rewards) to generate a structured
\emph{chain-of-thought} (CoT) before producing its final answer.  Formally, the
model samples a latent reasoning trace $\mathbf{r} = (r_1, \ldots, r_m)$ prior to
the answer tokens $\mathbf{a}$:
\begin{equation}
  P(\mathbf{a} \mid \mathbf{x})
  = \sum_{\mathbf{r}} P(\mathbf{a} \mid \mathbf{x}, \mathbf{r})\,
    P(\mathbf{r} \mid \mathbf{x}).
  \label{eq:cot}
\end{equation}
The trace $\mathbf{r}$ may be exposed to the user (as in OpenAI o1/o3 and
DeepSeek-R1) or kept internal (as in the \emph{extended thinking} mode of
Claude 3.7/4.x).  In either case, the effective compute invested per query is
substantially larger than in standard autoregressive generation.
\end{definition}

\begin{remark}
The distinction in Definition~\ref{def:reasoning} is one of \emph{training
objective} and \emph{inference regime}, not of architecture.  Both paradigms use the
same decoder-only Transformer building block.  What differs is that reasoning models
are rewarded for producing correct final answers, with the chain-of-thought providing
a scratch-pad that can represent thousands of intermediate tokens.  This has been
shown to dramatically improve performance on mathematical proofs, competitive
programming, and, increasingly, on quantitative finance tasks such as numerical
reasoning over financial statements.  It does not, however, eliminate hallucination
(see Section~\ref{sec:limitations-hallucinations}).
\end{remark}

\begin{example}[Merger Valuation from an Earnings Call]
\label{ex:merger-valuation}
Consider the following excerpt from an earnings call (paraphrased for illustration):

\begin{quote}
``We are acquiring TechCo for \$4.2 billion.  TechCo's trailing twelve-month EBITDA
is \$310 million, and we expect \$150 million in synergies within three years.  Our
current weighted average cost of capital is 8.5\%.''
\end{quote}

\noindent\textbf{Standard generative model (GPT-4o, temperature $= 0$):}  The model
reads the excerpt and directly outputs a one-paragraph assessment, typically noting
that the EV/EBITDA multiple is approximately $13.5\times$ and characterising the deal
as \emph{``strategically sound given the synergy guidance.''}  This response is often
plausible but does not verify whether the synergy NPV actually justifies the purchase
price premium.

\noindent\textbf{Reasoning model (o3, DeepSeek-R1):}  The model first computes the
implied enterprise value, then discounts the synergy stream at 8.5\% over three years
(approximated as a growing annuity), compares this to the purchase premium over a
benchmarked stand-alone valuation, and checks internal consistency before concluding
that the deal is modestly value-accretive under the stated assumptions but sensitive
to the synergy realisation timeline.  The reasoning trace makes each step auditable.

This contrast recurs throughout the book.  For tasks with a verifiable numerical or
logical answer, reasoning models offer a meaningful and measurable advantage.  For
primarily generative tasks (summarisation, translation, drafting), the additional
latency and cost of extended thinking may not be justified.
\end{example}

\subsection{Major Model Families}
\label{sec:landscape-families}

Table~\ref{tab:model-families} provides a structured overview of the major model
families referenced throughout this book.

\begin{table}[htbp]
\centering
\caption{Major language model families relevant to financial NLP (as of 2024--2025).
Parameter counts are approximate.  ``Finance'' indicates whether the model was
trained or fine-tuned specifically on financial corpora.}
\label{tab:model-families}
\small
\begin{tabular}{llllll}
\hline
\textbf{Model} & \textbf{Organisation} & \textbf{Arch.} & \textbf{Params} &
\textbf{Training Data} & \textbf{Finance} \\
\hline
GPT-4 / GPT-4o   & OpenAI        & Decoder-only & $\sim$1T (est.) & Web + RLHF       & No  \\
Claude 3.x/4.x   & Anthropic     & Decoder-only & Undisclosed      & Web + RLHF       & No  \\
Llama 3 (405B)   & Meta          & Decoder-only & 405B             & 15T tokens       & No  \\
Gemini 1.5 Pro   & Google DM     & Decoder-only & Undisclosed      & Multimodal web   & No  \\
FinBERT          & Araci (2019)  & Encoder      & 110M             & Financial news   & Yes \\
BloombergGPT     & Bloomberg     & Decoder-only & 50B              & 363B fin.\ tok.  & Yes \\
\hline
\end{tabular}
\end{table}

The most capable models (GPT-4o, Claude 4, Gemini 1.5 Pro) are proprietary: their
weights are not released, and access is mediated through pay-per-token APIs.  This
raises important issues for financial practitioners.  First, there is a
\emph{data-privacy} concern: sending confidential earnings call transcripts, trading
strategies, or client portfolios to a third-party API may conflict with internal
compliance policies, GDPR obligations, or NDA terms.  Second, there is an
\emph{auditability} concern: the model's exact behaviour cannot be reproduced or
independently audited.

Open-weight models such as Llama~3 (Meta, released under a community licence) address
both concerns: they can be deployed on-premises, fine-tuned on proprietary data
without data leaving the organisation, and their inference behaviour can be exactly
reproduced given the same random seed.  The trade-off is that the largest open models
require substantial GPU infrastructure (a 405-billion-parameter model at 16-bit
precision requires approximately 810~GB of GPU memory).
% [CITE: touvron2023llama]

BloombergGPT~\cite{wu2023bloomberggpt} % [CITE: wu2023bloomberggpt Wu et al. 2023]
was trained on 363 billion financial tokens (news, filings, analyst reports) alongside
345 billion general-domain tokens, and represented the first large-scale demonstration
that domain-specific pre-training on financial text was feasible at scale.

Subsequent benchmarks revealed a nuanced picture.  On tasks requiring deep financial
\emph{knowledge} (entity recognition in financial news, financial sentiment on
FinancialPhraseBank), BloombergGPT outperforms same-size general-purpose baselines.
But general-purpose models at significantly larger scale---GPT-4 or Claude 3---often
match or exceed BloombergGPT on the same benchmarks, because scale compensates for
domain specificity.  This is an instance of the \emph{emergent capability} effect:
beyond a threshold of model scale and training compute, models acquire domain knowledge
implicitly, even without targeted domain training. % [CITE: wei2022emergent]
The practical implication is that \emph{task-specific fine-tuning of a general-purpose
model} often outperforms a domain-pre-trained model of the same size, and
\emph{prompt engineering} applied to a large general model often outperforms
both---a theme developed extensively in Chapter~6.

Raw pre-trained models (\emph{base} models) are not directly usable as assistants.
\emph{Instruction tuning} fine-tunes a base model on curated (instruction, response)
pairs, teaching the model to follow natural-language directives.
\emph{Reinforcement Learning from Human Feedback} (RLHF) \cite{ouyang2022training}
% [CITE: ouyang2022training Ouyang et al. 2022, InstructGPT]
further aligns the model with human preferences by training a reward model on pairwise
human rankings and optimising the language model's policy via proximal policy
optimisation (PPO).  The result is the \emph{chat} or \emph{instruct} variant
familiar to practitioners.  For financial NLP, instruction tuning is especially
important: it makes models reliably follow structured output formats (JSON, Markdown
tables) that downstream data pipelines depend on.

\subsection{Financial NLP Benchmarks and Datasets}
\label{sec:landscape-benchmarks}

Progress in financial NLP is measured against a set of standardised benchmarks.
Table~\ref{tab:benchmarks} summarises the major ones; all are used in evaluation
experiments in subsequent chapters.

\begin{table}[htbp]
\centering
\caption{Major financial NLP benchmarks and datasets.  ``Size'' is approximate.
``Metric'' is the primary evaluation metric used in the literature.}
\label{tab:benchmarks}
\small
\begin{tabular}{lllll}
\hline
\textbf{Benchmark} & \textbf{Task} & \textbf{Size} & \textbf{Metric} &
\textbf{Reference} \\
\hline
FinancialPhraseBank & Sentiment (3-class)  & 4,840 sent.       & Accuracy       & \cite{malo2014good} \\
FiQA SA / QA        & Opinion mining; Q\&A & $\sim$1,200 sent. & F1, NDCG       & \cite{fiqa2018} \\
ECTSum              & Earnings call summ.  & 2,425 docs        & ROUGE-L        & \cite{mukherjee2022ectsum} \\
FinQA               & Numerical reasoning  & 8,281 Q\&A pairs  & Execution acc. & \cite{chen2021finqa} \\
FLUE                & Multi-task fin.\ NLU & Multi-task        & Task-specific  & \cite{shah2022flue} \\
\hline
\end{tabular}
\end{table}
% [CITE: malo2014good] Malo et al. (2014), FinancialPhraseBank
% [CITE: fiqa2018] Maia et al. (2018), FiQA challenge
% [CITE: mukherjee2022ectsum] Mukherjee et al. (2022), ECTSum
% [CITE: chen2021finqa] Chen et al. (2021), FinQA
% [CITE: shah2022flue] Shah et al. (2022), FLUE

\textbf{FinancialPhraseBank} (Malo et al.\ 2014) \cite{malo2014good} consists of
4,840 sentences sampled from English-language financial news articles, each annotated
by sixteen domain experts with one of three sentiment labels: \texttt{positive},
\texttt{negative}, or \texttt{neutral}.  The standard evaluation uses the subset where
all annotators agreed (\emph{Agree-all}, 2,264 sentences), yielding a highly reliable
gold standard.  This dataset is used in Chapter~2 as the primary benchmark for
sentiment analysis experiments.

\textbf{FiQA} (2018) \cite{fiqa2018} was released as part of a shared task on
financial opinion mining.  It contains two subtasks: (SA-C) aspect-based sentiment
analysis of financial microblogs and news headlines, and (QA) question answering over
financial documents.  The QA subtask has proven particularly challenging due to the
requirement to retrieve relevant passages before generating an answer---a natural fit
for the retrieval-augmented generation (RAG) systems discussed in Chapter~5.

\textbf{ECTSum} (Mukherjee et al.\ 2022) \cite{mukherjee2022ectsum} provides 2,425
earnings call transcript--summary pairs, where the reference summaries were written by
domain experts and verified for factual consistency.  It is the primary benchmark for
Chapter~4 (Earnings Calls and Transcripts) and is used to evaluate both extractive and
abstractive summarisation methods.

\textbf{FinQA} (Chen et al.\ 2021) \cite{chen2021finqa} requires models to answer
questions that demand \emph{multi-step numerical reasoning} over structured tables and
unstructured prose extracted from SEC 10-K and 10-Q filings.  Correct answers require
executing a sequence of arithmetic operations (addition, division, percentage change)
with values drawn from the document.  This dataset is the primary evaluation target in
Chapter~3 and illustrates clearly why reasoning models
(Section~\ref{sec:landscape-genreason}) outperform standard generative models on
structured financial analysis tasks.

\textbf{FLUE} (Shah et al.\ 2022) \cite{shah2022flue} is a Financial Language
Understanding Evaluation benchmark comprising five tasks: financial sentiment analysis,
news headline classification, named entity recognition in financial documents, structure
boundary detection in filings, and the FiQA-SA subtask.  It is the most comprehensive
single-number evaluation framework for general financial NLP capability and is the
primary evaluation target in Chapter~13.

\begin{remark}
All five benchmarks described above were constructed primarily from English-language,
North American financial documents.  Chapter~13 discusses the implications of this
distributional bias for evaluating models intended for European or Asian financial
markets, and Chapter~12 highlights specific challenges that arise when FLUE-trained
models are applied to MiFID~II or IFRS filings.
\end{remark}

% ---------------------------------------------------------------
\section{Working with LLMs via API}
\label{sec:api}

The transition from understanding how LLMs work to using them in practice requires
familiarity with a small set of engineering concepts.  This section covers those
concepts and provides working code examples for the three ecosystems most relevant to
financial practitioners.

\subsection{Core Concepts: Tokens, Context Windows, and Sampling}
\label{sec:api-concepts}

\begin{definition}[Token]
\label{def:token}
A \emph{token} is the atomic unit of input and output processed by a language model.
Tokens are produced by a \emph{tokeniser}, a deterministic mapping from strings to
integer sequences.  Modern LLMs use sub-word tokenisation schemes such as Byte-Pair
Encoding (BPE) \cite{sennrich2016bpe} % [CITE: sennrich2016bpe Sennrich et al. 2016]
or WordPiece \cite{wu2016wordpiece}, % [CITE: wu2016wordpiece Wu et al. 2016]
in which the vocabulary consists neither of individual characters nor of whole words,
but of frequently occurring sub-word fragments.
\end{definition}

Byte-Pair Encoding builds the vocabulary iteratively:
\begin{enumerate}
  \item Initialise the vocabulary with the set of all individual Unicode characters
    (or bytes) appearing in the training corpus.
  \item Count the frequency of all adjacent symbol pairs in the current tokenised
    corpus.
  \item Merge the most frequent pair into a single new symbol, updating the corpus.
  \item Repeat steps 2--3 until the vocabulary reaches a target size $|V|$
    (typically 32,000--100,000 for LLMs).
\end{enumerate}

\begin{example}[BPE Tokenisation of a Financial Term]
\label{ex:bpe}
Consider the word \texttt{"earnings"}.  A BPE tokeniser with vocabulary size 50,000
(as used by GPT-2 and GPT-3) would likely tokenise this as two sub-words:
\texttt{["earn", "ings"]}.  Rarer financial terms may receive more fragmented
tokenisations; for example, \texttt{"EBITDA"} might tokenise as
\texttt{["EB",\,"IT",\,"DA"]}, and \texttt{"Berkshire"} as
\texttt{["Berk",\,"shire"]}.  As a rule of thumb, one English word corresponds to
approximately 1.3 tokens on average; a typical earnings call transcript of 5,000 words
therefore contains roughly 6,500 tokens.  This conversion factor is used in the cost
analysis in Section~\ref{sec:api-practical}.
\end{example}

\begin{definition}[Context Window]
\label{def:context-window}
The \emph{context window} (or \emph{context length}) of a language model is the
maximum number of tokens $L_{\max}$ that the model can process in a single forward
pass.  Formally, the attention mechanism in a Transformer of context length $L_{\max}$
can only attend to positions $j \leq \min(t, L_{\max})$ when computing the
representation of position $t$.  Any input exceeding $L_{\max}$ tokens must be
truncated or split across multiple calls.
\end{definition}

Context windows grew rapidly in the 2022--2024 period: GPT-3 offered 4,096 tokens;
GPT-4-turbo extended this to 128,000 tokens; Gemini 1.5 Pro supports 1,000,000
tokens; Claude 3/4 supports up to 200,000 tokens with plans to expand further.  For
financial NLP, the context window determines whether an entire 10-K filing (which may
exceed 100,000 tokens) can be processed in a single call---a question with significant
implications for retrieval-augmented generation architectures (Chapter~5).

Three hyperparameters govern sampling. Given the logit vector $\mathbf{z} \in \mathbb{R}^{|\mathcal{V}|}$ produced by the
final Transformer layer, the model converts logits to a probability distribution and
samples the next token.  Three hyperparameters govern this process.

\begin{definition}[Temperature Sampling]
\label{def:temperature}
\emph{Temperature} $\tau > 0$ scales the logit vector before the softmax:
\begin{equation}
  p_i = \frac{\exp(z_i / \tau)}{\sum_{j=1}^{|\mathcal{V}|} \exp(z_j / \tau)},
  \quad i = 1, \ldots, |\mathcal{V}|.
  \label{eq:temperature}
\end{equation}
As $\tau \to 0^+$, all probability mass concentrates on the highest-logit token
(\emph{greedy decoding}).  As $\tau \to \infty$, the distribution approaches uniform
over $\mathcal{V}$.  For financial analysis tasks requiring precision and
reproducibility, $\tau = 0$ (or $\tau = 0.1$) is recommended.  For creative tasks,
$\tau \in [0.7, 1.0]$ is common.
\end{definition}

\begin{definition}[Nucleus (Top-$p$) Sampling]
\label{def:top-p}
\emph{Nucleus sampling} \cite{holtzman2020curious} % [CITE: holtzman2020curious]
draws the next token from the smallest set $\mathcal{S} \subseteq \mathcal{V}$
satisfying
\begin{equation}
  \sum_{i \in \mathcal{S}} p_i \geq p,
  \label{eq:top-p}
\end{equation}
where $p \in (0, 1]$ is a user-specified threshold.  Tokens outside $\mathcal{S}$ are
assigned zero probability and renormalised over $\mathcal{S}$.  Top-$p$ sampling
adapts the effective vocabulary size to the confidence of the model: when the model is
certain, $|\mathcal{S}|$ is small; when uncertain, $|\mathcal{S}|$ grows.  Common
values are $p \in [0.9, 0.95]$.
\end{definition}

\begin{definition}[Top-$k$ Sampling]
\label{def:top-k}
\emph{Top-$k$ sampling} restricts sampling to the $k$ tokens with the highest
probabilities, renormalising over this set.  Unlike top-$p$, the effective vocabulary
size is fixed at $k$ regardless of the model's confidence.  Top-$k$ sampling
(with $k = 50$) was used in GPT-2 generation and remains a common baseline.
\end{definition}

\begin{remark}
In practice, many API providers expose only temperature and top-$p$, not top-$k$, as
user-controllable parameters.  For financial NLP tasks where reproducibility and
factual precision are paramount---such as extracting numerical data from filings or
generating compliance summaries---\emph{greedy decoding} ($\tau = 0$) is strongly
recommended: it is the only sampling regime that is deterministic across identical
inputs (modulo floating-point non-determinism on GPU hardware).
\end{remark}

\subsection{Python Walkthrough: OpenAI, Anthropic, and HuggingFace}
\label{sec:api-python}

We now demonstrate how to call three distinct LLM ecosystems from Python.  All
examples use the same financial NLP task---sentiment analysis of an earnings call
excerpt---to facilitate direct comparison.

The OpenAI Python SDK follows a \emph{chat completion} model: the user submits a list
of messages (with roles \texttt{system}, \texttt{user}, and \texttt{assistant}), and
the API returns a completion.

\begin{lstlisting}[language=Python, caption={OpenAI API: financial sentiment with GPT-4o.}, label={lst:openai}]
from openai import OpenAI

client = OpenAI(api_key="YOUR_API_KEY")  # or set OPENAI_API_KEY env var

EXCERPT = (
    "Revenue increased 12% year-over-year to $4.1 billion, "
    "driven by strong demand in cloud services. However, "
    "operating margins contracted 200 basis points due to "
    "elevated infrastructure investment."
)

response = client.chat.completions.create(
    model="gpt-4o",
    messages=[
        {
            "role": "system",
            "content": (
                "You are a financial analyst. Classify the sentiment "
                "of earnings call excerpts as POSITIVE, NEGATIVE, or "
                "NEUTRAL. Provide a one-sentence justification."
            ),
        },
        {
            "role": "user",
            "content": f"Classify the sentiment: {EXCERPT}",
        },
    ],
    temperature=0.0,    # greedy: maximally deterministic
    max_tokens=128,
)

print(response.choices[0].message.content)
# MIXED/NEUTRAL -- Revenue growth is positive but margin compression
# partially offsets the top-line improvement.

# Token usage (important for cost tracking):
usage = response.usage
print(f"Input tokens:  {usage.prompt_tokens}")
print(f"Output tokens: {usage.completion_tokens}")
\end{lstlisting}

The Anthropic SDK separates the \texttt{system} prompt from the \texttt{messages}
list, passing it as a top-level argument rather than as the first message.  This
structural difference reflects Anthropic's Constitutional AI training approach, where
the system prompt receives special treatment during RLHF.
% [CITE: bai2022constitutional]

\begin{lstlisting}[language=Python, caption={Anthropic API: financial sentiment with claude-sonnet-4-6.}, label={lst:anthropic}]
import anthropic

client = anthropic.Anthropic(api_key="YOUR_API_KEY")  # or ANTHROPIC_API_KEY

message = client.messages.create(
    model="claude-sonnet-4-6",
    max_tokens=128,
    system=(
        "You are a financial analyst. Classify the sentiment "
        "of earnings call excerpts as POSITIVE, NEGATIVE, or "
        "NEUTRAL. Provide a one-sentence justification."
    ),
    messages=[
        {
            "role": "user",
            "content": f"Classify the sentiment: {EXCERPT}",
        }
    ],
    temperature=0.0,
)

print(message.content[0].text)

# Token usage:
print(f"Input tokens:  {message.usage.input_tokens}")
print(f"Output tokens: {message.usage.output_tokens}")
\end{lstlisting}

FinBERT \cite{araci2019finbert} % [CITE: araci2019finbert Araci (2019)]
is a BERT-based encoder fine-tuned on financial news for three-class sentiment
classification.  It runs locally without any API key or per-token cost.

\begin{lstlisting}[language=Python, caption={FinBERT sentiment classification via HuggingFace pipeline.}, label={lst:finbert}]
from transformers import pipeline

# Downloads model weights (~440 MB) on first run; cached thereafter.
nlp = pipeline(
    "text-classification",
    model="ProsusAI/finbert",
    device=0,       # GPU 0; set device=-1 for CPU
)

result = nlp("The company reported record earnings this quarter.")
print(result)
# [{'label': 'positive', 'score': 0.9834}]

# Batch inference (more efficient for large corpora):
excerpts = [
    "Revenue declined 8% amid weakening consumer demand.",
    "The Board declared a special dividend of $2.50 per share.",
    "Management reiterated full-year guidance.",
]
results = nlp(excerpts, batch_size=32)
for text, res in zip(excerpts, results):
    print(f"[{res['label']:8s} {res['score']:.2f}]  {text[:60]}")
\end{lstlisting}

Table~\ref{tab:api-comparison} summarises the key structural differences across the three ecosystems.

\begin{table}[htbp]
\centering
\caption{Comparison of LLM API designs for financial NLP tasks.}
\label{tab:api-comparison}
\small
\begin{tabular}{llll}
\hline
\textbf{Feature} & \textbf{OpenAI (GPT-4o)} & \textbf{Anthropic (Claude)} &
\textbf{HuggingFace (FinBERT)} \\
\hline
System prompt   & First message (\texttt{"system"}) & Top-level \texttt{system=} arg & N/A \\
Output access   & \texttt{choices[0].message}       & \texttt{content[0].text}        & List of dicts \\
Local deploy    & No  & No  & Yes \\
Cost model      & Per input/output token & Per input/output token & Compute only \\
Context window  & 128K tokens & 200K tokens & 512 tokens \\
Task type       & Generative / reasoning & Generative / reasoning & Classification \\
\hline
\end{tabular}
\end{table}

\subsection{Cost, Latency, and Practical Considerations}
\label{sec:api-practical}

API providers charge separately for input tokens (the prompt) and output tokens
(the generated response):
\begin{equation}
  \text{cost} = N_{\text{in}} \cdot c_{\text{in}} + N_{\text{out}} \cdot c_{\text{out}},
  \label{eq:cost-model}
\end{equation}
where $N_{\text{in}}$, $N_{\text{out}}$ are token counts and $c_{\text{in}}$,
$c_{\text{out}}$ are the provider's per-token prices.
Table~\ref{tab:api-costs} gives approximate 2024 prices in USD per million tokens.

\begin{table}[htbp]
\centering
\caption{Approximate 2024 API pricing.  Prices in USD per million tokens (MTok).
Verify current prices on provider websites before budgeting.}
\label{tab:api-costs}
\small
\begin{tabular}{llll}
\hline
\textbf{Model} & \textbf{Input (\$/MTok)} & \textbf{Output (\$/MTok)} &
\textbf{Notes} \\
\hline
GPT-4o                 & \$5.00       & \$15.00       & OpenAI, proprietary \\
Claude 3 Sonnet        & \$3.00       & \$15.00       & Anthropic, proprietary \\
Claude 3 Haiku         & \$0.25       & \$1.25        & Smaller, faster \\
Llama 3 70B (via API)  & \$0.59       & \$0.79        & E.g.\ Together AI \\
Llama 3 (self-hosted)  & $\approx$\$0 & $\approx$\$0  & GPU cost only \\
\hline
\end{tabular}
\end{table}

To make this concrete, suppose a compliance team needs to process $N = 10{,}000$ earnings call transcripts,
each approximately 5,000 words.  At 1.3 tokens per word (see
Example~\ref{ex:bpe}), each transcript is $\approx 6{,}500$ input tokens.  Adding a
500-token system prompt and $\approx 400$ output tokens, and applying
\eqref{eq:cost-model}:

\begin{align}
  \text{cost}_{\text{GPT-4o}}
  &= 10{,}000 \times \left(\frac{7{,}000}{10^6} \times 5.00
    + \frac{400}{10^6} \times 15.00 \right) \nonumber \\
  &= 10{,}000 \times (0.035 + 0.006) \;=\; \$410.
  \label{eq:cost-example}
\end{align}

The same workload on Claude 3 Haiku costs approximately \$20---a twenty-fold
reduction.  This illustrates the core cost-optimisation decision in financial NLP
pipelines: a small, cheap model handles straightforward extraction tasks, while a
large model is reserved for complex reasoning steps.

Several engineering considerations govern the cost-latency trade-off in production pipelines.
\begin{itemize}
  \item \textbf{Batching.}  For offline workloads (historical transcript processing,
    regulatory filing analysis), maximum parallelism reduces wall-clock time without
    increasing cost.  OpenAI's Batch API offers a 50\% price discount in exchange for
    up to 24 hours of processing latency---appropriate for end-of-day compliance runs
    but unsuitable for real-time applications.
  \item \textbf{Latency.}  For real-time applications (live earnings call
    summarisation, intraday news monitoring), latency is a binding constraint.
    Time-to-first-token for GPT-4o is typically 1--3 seconds; for Claude 3 Haiku
    under 500 ms.  Locally-hosted Llama 3 (8B, single A100 GPU) can achieve
    sub-100 ms first-token latency.
  \item \textbf{Model selection heuristic.}  A common two-stage financial NLP
    pipeline: (1) use a cheap, fast model (FinBERT, Claude Haiku) to filter and
    classify large volumes of text; (2) route only the high-signal subset to a large
    model for deep analysis.  This can reduce total API cost by 80--95\% on typical
    workloads.
\end{itemize}

\begin{remark}[Prompt Caching]
\label{rem:prompt-caching}
Anthropic's \emph{prompt caching} API allows users to designate a prefix of the
system prompt as cacheable.  Subsequent requests that share this prefix incur a
reduced input-token price (approximately 10\% of the standard rate for cache hits),
reducing costs by up to 90\% in applications with a fixed, long system prompt---such
as a financial analysis assistant with fixed instructions and a reference document.
OpenAI offers analogous automatic prompt caching for context windows exceeding 1,024
tokens.  When designing pipelines that issue many similar requests (e.g., processing
all filings from a given company under the same analytical framework), caching the
shared system prompt is among the highest-ROI optimisations available.
\end{remark}

% ---------------------------------------------------------------
\section{Limitations and Responsible Use}
\label{sec:limitations}

The preceding sections present LLMs as powerful tools for financial NLP.  This section
provides a necessary counterbalance: a rigorous account of their failure modes and the
ethical and regulatory context in which financial AI systems operate.

\subsection{Hallucinations in Financial Contexts}
\label{sec:limitations-hallucinations}

\begin{definition}[Hallucination]
\label{def:hallucination}
A language model \emph{hallucination} is an output that is (a) presented with apparent
confidence, yet (b) is not supported by---or actively contradicts---the model's input
or the external world.  Following the taxonomy of~\cite{maynez2020faithfulness},
% [CITE: maynez2020faithfulness Maynez et al. 2020]
hallucinations are classified as:
\begin{enumerate}
  \item \textbf{Intrinsic hallucinations:} the output contradicts information
    explicitly present in the input document.
  \item \textbf{Extrinsic hallucinations:} the output introduces information not
    present in the input that cannot be verified from any accessible source.
\end{enumerate}
\end{definition}

Both types are particularly dangerous in finance, where incorrect numerical claims or
fabricated regulatory citations can constitute material misstatements.

\begin{example}[Hallucination in Financial Analysis]
\label{ex:hallucination}
A financial analyst uses an LLM to summarise a 10-K filing.  The filing states that
the company's debt-to-equity ratio is 1.8.  The LLM summarises this as ``a
debt-to-equity ratio of 0.8, indicating a conservatively leveraged balance sheet.''
This is an \emph{intrinsic} hallucination: a direct numerical error that inverts the
true risk profile.  Separately, the LLM notes that ``the company's accounting
practices comply with IFRS 16 as confirmed by the PCAOB in their 2023 inspection
report.''  No such report is mentioned in the filing; this is an \emph{extrinsic}
hallucination.  Both errors could materially mislead an investor relying on the LLM
output without cross-checking the source document.
\end{example}

The root cause of hallucination is not a simple bug but an intrinsic property of
autoregressive generation: the model is trained to produce \emph{plausible} text, and
plausibility correlates with---but is not identical to---factual accuracy.
% [CITE: ji2023survey Ji et al. 2023, survey of hallucination in NLG]

Several mitigation strategies reduce the incidence and impact of hallucination in financial applications.
\begin{itemize}
  \item \textbf{Retrieval-Augmented Generation (RAG).}  Rather than relying on
    parametric knowledge, the model is provided with retrieved passages from a verified
    document store, substantially reducing extrinsic hallucination.  This is the
    primary architecture for financial information extraction in Chapter~5.
  \item \textbf{Grounding with citations.}  Prompting the model to cite specific
    sentences or page numbers from the source document for every factual claim provides
    an auditable trail.  This technique is demonstrated in Chapter~3.
  \item \textbf{Chain-of-thought prompting.}  Requiring the model to show its
    reasoning steps before producing an answer reduces hallucination on numerical and
    multi-hop reasoning tasks. % [CITE: wei2022chain Wei et al. 2022]
  \item \textbf{Calibration and confidence elicitation.}  Asking the model to express
    uncertainty or to generate multiple candidate responses and identify disagreements
    can flag unreliable outputs for human review.
\end{itemize}

\begin{remark}
Reasoning models (Section~\ref{sec:landscape-genreason}) exhibit reduced hallucination
rates on tasks with verifiable answers because the chain-of-thought trace provides a
self-checking mechanism.  However, they are not immune: extrinsic hallucinations
persist, especially on obscure regulatory facts or firm-specific historical data not
well represented in the training corpus.  The mitigation strategies above remain
necessary regardless of model type.
\end{remark}

\subsection{Regulatory and Ethical Considerations}
\label{sec:limitations-ethics}

\begin{remark}
Nothing in this book constitutes legal, regulatory, or investment advice.  The
discussion below describes the regulatory landscape for educational purposes only.
Practitioners deploying AI systems in financial contexts must obtain qualified legal
counsel and ensure compliance with all applicable laws and regulations in their
jurisdiction.
\end{remark}

The European Union's AI Act (Regulation 2024/1689), which entered into force in
August 2024, establishes a risk-based classification for AI systems.
% [CITE: eu-ai-act-2024]
AI systems used for credit scoring, insurance risk assessment, pricing of life
insurance products, and AI-assisted securities pricing fall into the \emph{high-risk}
category (Annex III).  High-risk systems are subject to mandatory conformity
assessments, post-market monitoring, transparency obligations towards regulators, and
human oversight requirements.  An LLM deployed as part of an algorithmic trading
system or a loan-approval pipeline would likely qualify as high-risk under the Act,
requiring documentation of the model's training data, performance metrics, and the
mechanisms ensuring human oversight before market deployment.

In the United States, the Securities and Exchange Commission has issued staff guidance indicating that
investment advisers using AI-generated research must demonstrate that reasonable care
was exercised in evaluating the AI's output, and that AI-generated recommendations are
consistent with the adviser's fiduciary duty to clients.
% [CITE: sec2023ai]
The \emph{fiduciary duty} question is particularly acute: whether an LLM's output
constitutes a ``recommendation'' under the Investment Advisers Act of 1940 has not
been definitively resolved by case law, but the SEC's enforcement posture has been to
hold advisers fully responsible for AI-generated outputs used in client-facing
communications.

Fine-tuning or retrieval-augmented generation on proprietary client data---transaction
histories, portfolio compositions, know-your-customer (KYC) files---raises significant
obligations under the General Data Protection Regulation (GDPR, EU 2016/679).
% [CITE: gdpr2016]
Key issues include: (a) whether training on personal data requires explicit consent;
(b) the right to erasure and whether a model trained on a data subject's information
can comply with an erasure request without full retraining; (c) data-transfer
restrictions if API calls route personal data to non-EU servers.  Chapter~6
(Fine-tuning) and Chapter~9 (Risk and Compliance) discuss technical mitigations
including federated learning, differential privacy, and on-premises deployment.

General-purpose LLMs are trained predominantly on English-language data from North
American and Western European sources.  Financial training data is similarly skewed:
the largest available corpora (SEC filings, English-language Bloomberg terminals,
Reuters news) disproportionately represent US-listed equities and USD-denominated
instruments.  As a consequence:
\begin{itemize}
  \item Sentiment models calibrated on US earnings calls may systematically
    mis-classify sentiment in Japanese, Chinese, or Indian financial contexts.
  \item Named entity recognition models fine-tuned on EDGAR filings may fail on
    corporate structures common in family-controlled firms typical of Asian or Latin
    American markets.
  \item Regulatory knowledge embedded in LLM weights is heavily biased toward US/EU
    rules; extrapolation to other jurisdictions should be treated with extreme caution.
\end{itemize}
Chapter~13 quantifies these biases empirically using cross-lingual evaluation of
FLUE-equivalent tasks in French, German, and Chinese.

The deepest ethical question for financial LLM applications is whether a system that
generates investment recommendations---even as decision-support rather than autonomous
execution---triggers fiduciary obligations and, if so, who bears them.  Current legal
consensus is that the \emph{human adviser} (or firm) deploying the AI retains full
fiduciary responsibility for its outputs, making thorough testing, documentation, and
human oversight essential rather than optional.  Chapter~7 (AI Agents in Finance) and
Chapter~8 (Multi-Agent Systems) return to this question in the context of agentic
systems that take actions---executing trades, sending communications---rather than
merely producing text.

% ---------------------------------------------------------------
\section{Chapter Summary}
\label{sec:ch2-summary}

\subsection*{Summary of Chapter~1}

This chapter established the technical and conceptual foundations for the rest of the
book.  We traced the history of textual analysis in finance from keyword dictionaries
through distributional word vectors to the Transformer architecture, developing the
mathematics of self-attention, positional encoding, and the encoder--decoder structure.
We surveyed the modern LLM landscape, distinguishing standard autoregressive models
from reasoning models trained to produce explicit chains of thought.  We introduced
the major model families (GPT-4o, Claude, Llama, FinBERT, BloombergGPT) and the
principal financial NLP benchmarks (FinancialPhraseBank, FiQA, ECTSum, FinQA, FLUE)
against which models are evaluated throughout the book.  We covered the practical
mechanics of LLM APIs---tokenisation, context windows, sampling parameters, and cost
models---and concluded with an honest account of the limitations: hallucination,
regulatory exposure, data-privacy obligations, and the fiduciary responsibility that
financial AI practitioners inherit.

The remainder of the book proceeds from these foundations, moving from established
tasks (sentiment analysis, information extraction) to frontier applications (multi-agent
trading systems, regulatory technology), and from single-model pipelines to complex
multi-agent workflows.

\subsection*{Chapter-by-Chapter Overview}

\begin{table}[htbp]
\centering
\caption{Overview of book chapters and their relationship to Chapter~1 foundations.}
\label{tab:roadmap}
\small
\begin{tabular}{cll}
\hline
\textbf{Ch.} & \textbf{Title} & \textbf{Key Connection to Ch.~1} \\
\hline
2  & Sentiment Analysis
   & FinancialPhraseBank; temperature; FinBERT vs.\ GPT-4o \\
3  & Information Extraction from Filings
   & FinQA; grounding and citation to reduce hallucination \\
4  & Earnings Calls and Transcripts
   & ECTSum; context-window limits; cost model \\
5  & Retrieval-Augmented Generation
   & RAG as hallucination mitigation; context-window architecture \\
6  & Fine-tuning LLMs on Financial Data
   & Instruction tuning; RLHF; GDPR considerations \\
7  & AI Agents in Finance
   & Reasoning models; agentic loops; fiduciary questions \\
8  & Multi-Agent Systems and Workflows
   & Chain-of-thought; multi-agent coordination protocols \\
9  & Risk and Compliance Applications
   & EU AI Act; hallucination auditing; on-premises deployment \\
10 & Portfolio Construction with LLMs
   & Cost-latency trade-offs; structured output from LLMs \\
11 & Trading Systems and Execution
   & Real-time latency; reasoning vs.\ generative model choice \\
12 & Regulatory Technology
   & Geographic bias; MiFID~II / IFRS; GDPR in practice \\
13 & Evaluation and Benchmarking
   & FLUE; cross-lingual evaluation; bias quantification \\
14 & Frontiers and Future Directions
   & Scaling laws; multimodal models; next-generation reasoning \\
\hline
\end{tabular}
\end{table}

Each chapter is self-contained enough to be read independently, but the sequence is
deliberate.  Chapters~2--4 address the three core financial text sources (news,
filings, transcripts) and establish empirical baselines that later chapters build upon.
Chapters~5--6 introduce the two principal adaptation strategies (retrieval and
fine-tuning) that transform raw LLM capabilities into production-grade financial NLP
systems.  Chapters~7--8 extend the paradigm from single models to agentic systems
capable of multi-step reasoning and action.  Chapters~9--12 turn to the institutional
and regulatory dimensions that govern real-world deployment.  Chapter~13 provides a
unified evaluation framework, and Chapter~14 surveys open research questions at the
frontier.

The thread running through all chapters is the one introduced here: language models
are powerful but imperfect tools, and the financial practitioner's task is to deploy
them in ways that harness their strengths---scale, speed, and breadth of
knowledge---while managing their risks through careful system design, rigorous
evaluation, and meaningful human oversight.


