\chapter{Introduction}
\label{ch:intro}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Describe the evolution of textual analysis in finance from dictionary methods
        to large language models and situate key milestones in that history.
  \item Construct and interpret classical text representations (bag-of-words, one-hot
        encoding, TF-IDF) and explain their limitations.
  \item Explain word embeddings, reproduce the Skip-Gram and GloVe objectives, and
        interpret a PCA visualisation of financial vocabulary.
  \item Derive the vanishing-gradient bound for vanilla RNNs and explain how the LSTM
        cell state addresses it.
  \item State the scaled dot-product attention formula, prove the $\sqrt{d_k}$ variance
        normalisation, and describe the multi-head attention architecture.
  \item Distinguish generative from reasoning language models and identify major model
        families relevant to finance.
  \item Write Python code to query an LLM via the OpenAI, Anthropic, and HuggingFace
        APIs and reason about token costs and latency trade-offs.
  \item Articulate the main limitations of LLMs in financial applications and the
        relevant regulatory landscape.
\end{enumerate}
\end{remark}

% ---------------------------------------------------------------
\section{A Brief History of Textual Analysis in Finance}
\label{sec:history}

\begin{context}
It is the autumn of 2006 and Paul Tetlock, then a doctoral student, is staring
at a spreadsheet.  On one side: a column of numbers representing the daily
fraction of negative words in the \emph{Wall Street Journal}'s \emph{Abreast of
the Market} column, running back to 1984.  On the other: the next-day return on
the Dow Jones Industrial Average.  The correlation is small.  But it is
there---and according to the textbooks that line his shelves, it should not be.
If markets are efficient, prices should already reflect the sentiment of the
country's most widely-read financial column.  Tetlock would publish this finding
in the \emph{Journal of Finance} \cite{tetlock2007giving}, and the paper would
become a founding document of a new field.  But the deeper story began forty
years earlier, in a social science department with no interest in stock prices
whatsoever.

% ---------------------------------------------------------------
\subsection{Early Approaches: Keyword Counting and Dictionary Methods}
\label{sec:history-early}

The modern project of quantifying the meaning of text began in the social sciences.
In the 1950s and 1960s researchers in sociology and communication studies developed
\emph{content analysis} as a formal methodology: a researcher defines a coding scheme
of categories, reads passages, and assigns them to categories according to explicit
rules.  The goal was to turn a fundamentally qualitative object---human language---into
a quantitative variable amenable to statistical inference.

The methodological ambitions grew quickly.  In 1961 Philip Stone and colleagues at
Harvard published the \emph{General Inquirer} system
\cite{stone1966general}, % [CITE: Stone, P.J., Dunphy, D.C., Smith, M.S., Ogilvie, D.M. (1966). The General Inquirer: A Computer Approach to Content Analysis. MIT Press.]
a computer-assisted program that matched words in a document against hand-coded
psychological and semantic dictionaries.  The Harvard IV-4 dictionary, which became one
of its most widely used components, classifies roughly 11\,000 word stems into
categories such as \textsc{Positive}, \textsc{Negative}, \textsc{Strong},
\textsc{Weak}, \textsc{Active}, and \textsc{Passive}.  To compute a sentiment score for
a document one simply counts the number of words belonging to each category and
normalises by document length.  The appeal is transparency: every score can be
explained by pointing to the words that generated it.

Within finance, the landmark empirical application of dictionary-based sentiment analysis
is the paper by Paul Tetlock published in the \emph{Journal of Finance}
\cite{tetlock2007giving}. % [CITE: Tetlock, P.C. (2007). Giving Content to Investor Sentiment: The Role of Media in the Stock Market. Journal of Finance, 62(3), 1139–1168.]
Tetlock extracted the daily ``Abreast of the Market'' column from the \emph{Wall Street
Journal} over the period 1984--1999 and applied the General Inquirer's Harvard IV-4
dictionary to measure the fraction of negative words in each column.  His empirical
findings are striking: (i) high media negativity predicts downward pressure on Dow Jones
Industrial Average returns on the following trading day, followed by a mean-reversion
over the subsequent week; and (ii) unusually high or low negativity predicts high
trading volume.  He interprets the return predictability as consistent with investor
sentiment temporarily depressing prices below fundamental value, and the mean reversion
as the subsequent correction.

The Tetlock study demonstrated three things that shaped the subsequent literature.
First, textual sentiment can be quantified in a reproducible way.  Second, the resulting
signal has economically meaningful predictive content for asset prices even in a world
where professional arbitrageurs are active.  Third, the choice of dictionary matters:
the Harvard IV-4 list was designed for general English prose, not for the specialised
register of financial and legal language.

The inadequacy of general-purpose dictionaries for financial text was made rigorous by
Tim Loughran and Bill McDonald in a highly influential 2011 paper
\cite{loughran2011liability}. % [CITE: Loughran, T. & McDonald, B. (2011). When is a Liability not a Liability? Textual Analysis, Dictionaries, and 10-Ks. Journal of Finance, 66(1), 35–65.]
They examined the textual content of a large sample of 10-K annual filings submitted to
the SEC's EDGAR database and measured the fraction of words classified as negative by
the Harvard IV-4 dictionary.  They then checked, word by word, whether the negative
classification was appropriate for financial text.  Their finding was sobering: a large
share of the ``negative'' words in financial documents---for example \emph{tax},
\emph{liability}, \emph{depreciation}, \emph{cost}---are not negative in the financial
register; they are simply technical vocabulary.  Using the Harvard IV-4 dictionary on
financial filings therefore introduces substantial measurement error.

In response, Loughran and McDonald constructed their own \emph{financial sentiment
dictionary} (often abbreviated LM dictionary) by reading thousands of 10-K filings and
manually assigning financial connotations to words.  The resulting word lists span six
tone categories: Negative, Positive, Uncertainty, Litigious, Strong Modal (words
implying certainty), and Weak Modal (words implying tentativeness).  Their empirical
application showed that the LM Negative list outperformed the Harvard IV-4 list in
predicting initial public offering underperformance, earnings surprise, and subsequent
abnormal returns, among other outcomes.

The LM dictionary remains a standard benchmark in financial NLP.  Most subsequent
studies report results using both the Harvard IV-4 and LM lists to allow comparability.
However, both suffer from the same fundamental limitation: they are static.  Words are
classified once, by human judgment, and the classifications do not adapt to new usages,
new jargon, or sector-specific connotations.  Moreover, dictionary methods score only
the words they recognise, ignoring syntax, negation, context, and word order entirely.
The sentence ``earnings did \emph{not} decline'' receives the same score as ``earnings
\emph{did} decline'' if the word \emph{not} is not explicitly handled, and even
negation-aware heuristics remain fragile.

Before leaving the dictionary era, it is worth noting that scholars also developed
\emph{readability} measures applied to corporate disclosures.  Li (2008)
\cite{li2008annual} % [CITE: Li, F. (2008). Annual Report Readability, Current Earnings, and Earnings Persistence. Journal of Accounting and Economics, 45(2–3), 221–247.]
showed that annual reports that are more difficult to read (as measured by the
Fog Index, a formula based on average sentence length and the fraction of polysyllabic
words) are associated with lower future earnings, suggesting that complexity may serve as
obfuscation.  This line of research extended the textual analysis toolkit to stylistic
features beyond sentiment, opening the door to a broader conception of what ``textual
information'' means in a financial context.

% ---------------------------------------------------------------
\subsection{The Shift to Statistical and Neural Methods}
\label{sec:history-shift}

By the mid-2000s the natural language processing (NLP) community had largely moved beyond
dictionary-based methods toward \emph{statistical} representations that learn structure
from data.  The most influential of these for finance is \emph{Latent Dirichlet
Allocation} (LDA), introduced by Blei, Ng, and Jordan in 2003
\cite{blei2003latent}. % [CITE: Blei, D.M., Ng, A.Y. & Jordan, M.I. (2003). Latent Dirichlet Allocation. Journal of Machine Learning Research, 3, 993–1022.]
LDA is a generative probabilistic model: it posits that each document is a mixture of a
small number of latent \emph{topics}, and each topic is a distribution over vocabulary
words.  The model is fit by approximate Bayesian inference and requires no labelled data.

The application of LDA to financial text proved fertile.  Methodology-wise the key
advantage over dictionary methods is that topics are discovered from the corpus rather
than pre-specified.  Earnings calls, for instance, might reveal latent topics
corresponding to supply-chain concerns, margin pressure, new product introductions, and
macroeconomic commentary, each represented by a characteristic cluster of co-occurring
words.  Researchers such as Boudoukh et al.\ \cite{boudoukh2013news} % [CITE: Boudoukh, J., Feldman, R., Kogan, S. & Richardson, M.P. (2013). Which News Moves Stock Prices? A Textual Analysis. NBER Working Paper 18725.]
and Hansen, McMahon, and Prat \cite{hansen2018transparency} % [CITE: Hansen, S., McMahon, M. & Prat, A. (2018). Transparency and Deliberation within the FOMC: A Computational Linguistics Approach. Quarterly Journal of Economics, 133(2), 801–870.]
used topic models to analyse, respectively, firm-level news and Federal Open Market
Committee transcripts, demonstrating that data-driven topic discovery can reveal
structure invisible to dictionary counting.

Despite these successes, bag-of-words topic models share a fundamental limitation with
dictionary methods: they ignore word order and syntax.  The model treats a document as
an unordered multiset of words, discarding the sequential structure that makes language
meaningful.  Negation, qualification, conditionality, and argumentative structure are
all lost.  This was the central motivation for the neural turn.

A transformative shift came with the introduction of \emph{word2vec} by Mikolov et al.\
in 2013 \cite{mikolov2013efficient}. % [CITE: Mikolov, T., Chen, K., Corrado, G. & Dean, J. (2013). Efficient Estimation of Word Representations in Vector Space. ICLR Workshop.]
Word2vec trains a shallow neural network to predict either a word from its context
(the Continuous Bag-of-Words model) or the context from a word (the Skip-gram model).
As a by-product, each word receives a dense vector representation---an
\emph{embedding}---in $\mathbb{R}^d$ for some moderate $d$ (typically 100--300).
These embeddings capture distributional semantics: words that appear in similar contexts
end up close together in the embedding space.  The famous example is that
$\text{vec}(\textit{king}) - \text{vec}(\textit{man}) + \text{vec}(\textit{woman})
\approx \text{vec}(\textit{queen})$.

For finance, word embeddings offered the prospect of capturing semantic similarity
without hand-crafted dictionaries.  Researchers trained word2vec on large corpora of
earnings call transcripts, news articles, and 10-K filings and used the resulting
embeddings as features for downstream tasks such as sentiment classification and return
prediction.  Manela and Moreira \cite{manela2017news} % [CITE: Manela, A. & Moreira, A. (2017). News Implied Volatility and Disaster Concerns. Journal of Financial Economics, 123(1), 137–162.]
used text-based features including embedding-based similarities to construct a news-based
measure of disaster risk.  Simultaneously, GloVe (Global Vectors for Word
Representation) \cite{pennington2014glove} % [CITE: Pennington, J., Socher, R. & Manning, C.D. (2014). GloVe: Global Vectors for Word Representation. EMNLP 2014.]
offered an alternative factorisation approach that explicitly minimised the
difference between the dot product of word vectors and the logarithm of the words'
co-occurrence count, yielding embeddings with similarly useful geometric properties.

The deep learning era brought a further step: the combination of word embeddings with \emph{recurrent neural networks} (RNNs), and
specifically with Long Short-Term Memory (LSTM) networks
\cite{hochreiter1997long}, % [CITE: Hochreiter, S. & Schmidhuber, J. (1997). Long Short-Term Memory. Neural Computation, 9(8), 1735–1780.]
allowed models to read a document sequentially and maintain a hidden state summarising
everything read so far.  Applied to financial text, LSTM-based models could, in
principle, handle negation and long-range dependencies that defeat bag-of-words
approaches.  They achieved state-of-the-art performance on financial sentiment
benchmarks such as FiQA and Financial Phrase Bank throughout 2016--2018.

The attention mechanism introduced by Bahdanau et al.\ \cite{bahdanau2015neural} % [CITE: Bahdanau, D., Cho, K. & Bengio, Y. (2015). Neural Machine Translation by Jointly Learning to Align and Translate. ICLR 2015.]
and then generalised to the \emph{Transformer} architecture by Vaswani et al.\
\cite{vaswani2017attention} % [CITE: Vaswani, A., Shazeer, N., Parmar, N., et al. (2017). Attention Is All You Need. NeurIPS 2017.]
changed the landscape completely.  The Transformer processes all tokens in a sequence in
parallel via \emph{self-attention}, computing for each token a weighted average of all
other tokens' representations, where the weights are learned and reflect relevance.
This design enabled training on far larger datasets than RNNs could accommodate, and it
enabled \emph{pre-training}: a model could be trained on massive unlabelled corpora to
predict masked words or the next sentence, learning rich contextual representations,
and then \emph{fine-tuned} on a small labelled dataset for a specific downstream task.

BERT (Bidirectional Encoder Representations from Transformers)
\cite{devlin2019bert} % [CITE: Devlin, J., Chang, M.-W., Lee, K. & Toutanova, K. (2019). BERT: Pre-training of Deep Bidirectional Transformers for Language Understanding. NAACL-HLT 2019.]
exemplified this paradigm.  Fine-tuned BERT models rapidly surpassed prior state of the
art on nearly every NLP benchmark.  The finance community responded quickly.
FinBERT \cite{araci2019finbert} % [CITE: Araci, D. (2019). FinBERT: Financial Sentiment Analysis with Pre-trained Language Models. ArXiv:1908.10063.]
and a succession of similar models pre-trained on financial corpora achieved markedly
better sentiment classification accuracy on financial datasets than generic BERT.  They
represent the direct antecedents of the large language models---GPT-4, Claude, Gemini,
Llama---that form the subject of this book.

The history just sketched is not merely background; it is directly relevant to how one
should think about and deploy LLMs in practice.  Each generation of methods addressed
shortcomings of the previous one, but also introduced new failure modes.  Understanding
where the field came from illuminates why LLMs behave the way they do, what they can be
expected to do well, and where they remain unreliable.

\end{context}

% ---------------------------------------------------------------
\section{Why Text Matters in Finance}
\label{sec:motivation}

\begin{context}
Every quarter, thousands of companies assemble their executives on a conference
call, dial in a few hundred sell-side analysts, and spend an hour narrating the
past three months.  The ritual is called the earnings call.  The transcript is
published within days and largely forgotten---except by the people who have
learned to listen to what is not in the numbers.  The word \emph{challenges}
appearing where \emph{headwinds} used to appear.  The CFO's sentences getting
shorter.  The count of hedge words---\emph{approximately}, \emph{roughly},
\emph{we believe}---ticking upward.  None of this information lives in the
earnings per share figure.  All of it lives in the transcript.  This section
asks why that matters, and how we know it does.

% ---------------------------------------------------------------
\subsection{The Information Content of Earnings Calls, News, and Filings}
\label{sec:motivation-info}

The semi-strong form of the efficient markets hypothesis (EMH), associated with Fama
\cite{fama1970efficient}, % [CITE: Fama, E.F. (1970). Efficient Capital Markets: A Review of Empirical Work. Journal of Finance, 25(2), 383–417.]
holds that prices at any time fully reflect all \emph{publicly available} information.
Under strict semi-strong efficiency, reading a newspaper or an earnings call transcript
provides no trading advantage, because everything in those documents is already impounded
in prices the moment it is published.  Textual analysis would then be, at best, a way of
checking one's understanding of the market---not a way of generating alpha.

Why, then, has such a substantial empirical literature found otherwise?  Several
non-mutually-exclusive explanations are consistent with market efficiency:

\begin{enumerate}
  \item \textbf{Processing costs and limited attention.} Markets aggregate information
    through the trading of many investors, each of whom has limited attention and
    cognitive capacity.  If processing the full text of a 100-page 10-K filing
    requires substantial effort, and if only a small fraction of market participants
    perform that analysis, prices may not immediately reflect all the information the
    filing contains.  Once cheap NLP tools make processing feasible at scale, the
    anomaly should attenuate---and indeed Li, Lundholm, and Minnis
    \cite{li2013measuring} % [CITE: Li, F., Lundholm, R. & Minnis, M. (2013). A Measure of Competition Based on 10-K Filings. Journal of Accounting Research, 51(2), 399–436.]
    and others document that textual signals are most powerful precisely for documents
    that are longest, most complex, and hence most costly to process.

  \item \textbf{Soft information and interpretive disagreement.} Even when a document is
    widely read, its implications may be ambiguous.  A cautious management tone on an
    earnings call might reflect genuine uncertainty about business conditions, or it might
    be strategic communication designed to manage analyst expectations.  If sophisticated
    investors disagree about the correct interpretation, prices reflect an average of
    beliefs, and no single interpretation is instantly impounded.  NLP tools that can
    classify the \emph{degree} of hedging or strategic obfuscation thus extract a signal
    not already priced.

  \item \textbf{Latent information.} Corporate disclosures are subject to legal and
    regulatory constraints.  Managers cannot say anything material that is not disclosed
    to all investors simultaneously (Regulation FD in the US context).  However,
    information may be conveyed implicitly through word choice, rhetorical structure, and
    what is \emph{omitted} rather than stated.  Analysing the \emph{sentiment} or
    \emph{readability} of a disclosure can extract this latent signal even when the
    explicit content is already known.
\end{enumerate}

Under any of these accounts, textual analysis adds value not because the market is
irrational but because information extraction from text is genuinely costly and the
technology for doing it at scale has improved dramatically in recent years.  LLMs
represent the latest---and most powerful---step in that technological progression.

The empirical literature draws on several distinct categories of financial text, each
with its own characteristics, frequency, regulatory environment, and information content.

\begin{description}
  \item[Earnings call transcripts.] Publicly traded companies in most major markets
    hold quarterly earnings calls in which senior management (typically the CEO and CFO)
    presents financial results and answers questions from sell-side analysts.  Transcripts
    are produced by services such as Refinitiv or Bloomberg and are widely studied.  The
    call has two parts: a scripted management presentation and an unscripted Q\&A.  The
    Q\&A is particularly informative because responses are spontaneous and harder to
    rehearse.  Davis, Ge, Matsumoto, and Zhang \cite{davis2015effect} % [CITE: Davis, A.K., Ge, W., Matsumoto, D. & Zhang, J.L. (2015). The Effect of Manager-Specific Optimism on the Tone of Earnings Conference Calls. Review of Accounting Studies, 20(2), 639–673.]
    and a large subsequent literature show that sentiment extracted from transcripts
    predicts abnormal returns and subsequent earnings surprises.

  \item[10-K and 10-Q filings.] Annual (10-K) and quarterly (10-Q) reports filed with
    the SEC contain management discussion and analysis (MD\&A), risk factor disclosures,
    financial statements, and extensive footnotes.  These are the primary data source in
    the Loughran--McDonald (2011) study.  10-K filings can exceed 200 pages, presenting
    significant information-extraction challenges that NLP is uniquely well-placed to
    address.

  \item[Newswire and newspaper articles.] Tetlock (2007) worked with \emph{Wall Street
    Journal} columns.  A larger body of work uses machine-readable newswire feeds from
    Reuters or Bloomberg.  A key feature of news is high frequency: material news events
    trigger immediate repricing, and the window for exploiting a textual signal is
    therefore narrow.  The value of NLP-based news analysis lies partly in speed (reading
    and classifying thousands of articles per second) and partly in breadth (aggregating
    signals across many sources simultaneously).

  \item[Analyst reports.] Sell-side research reports combine quantitative analysis with
    extensive narrative commentary.  They are produced by financial analysts who have
    specialised knowledge of the companies they cover, making the text informationally
    dense.  Lehavy, Li, and Merkley \cite{lehavy2011benefit} % [CITE: Lehavy, R., Li, F. & Merkley, K. (2011). The Benefit of Investment Bank Analyst Monitoring. Journal of Finance, 66(4), 1223–1260.]
    show that the textual complexity of analyst reports is associated with greater investor
    uncertainty and higher information asymmetry.

  \item[Central bank communications.] Minutes and statements from central banks (the US
    Federal Reserve, the European Central Bank, the Bank of England) move interest rates
    and equity prices when published.  The FOMC transcripts used by Hansen et al.\
    \cite{hansen2018transparency} are now publicly available with a five-year lag and
    provide a rich laboratory for studying deliberation and its market effects.

  \item[Social media and alternative data.] Twitter (now X), Reddit (especially
    r/wallstreetbets), StockTwits, and similar platforms provide high-frequency, noisy,
    unstructured text.  Bollen, Mao, and Zeng \cite{bollen2011twitter} % [CITE: Bollen, J., Mao, H. & Zeng, X. (2011). Twitter Mood Predicts the Stock Market. Journal of Computational Science, 2(1), 1–8.]
    showed that aggregate Twitter mood states predict DJIA returns in a time-series
    regression, though subsequent work has questioned the robustness of this finding.
    Social media data is particularly challenging due to bot activity, sarcasm, slang,
    and the mixing of retail and institutional sentiment.
\end{description}

% ---------------------------------------------------------------
\subsection{Text as a Signal: Evidence from the Literature}
\label{sec:motivation-evidence}

Before formalising the framework, it is useful to have in mind the empirical regularities
the theory must explain.  The key stylised facts from the literature are:

\begin{enumerate}
  \item \textbf{Negative tone predicts negative returns (short-run).}  Tetlock (2007)
    found that the fraction of negative words in the WSJ column predicts DJIA returns
    with a negative coefficient: more pessimistic prose is followed by lower next-day
    returns.  The effect reverses over the subsequent week, suggesting temporary
    sentiment-driven mispricing.

  \item \textbf{Financial-domain dictionaries outperform general-domain ones.}
    Loughran and McDonald (2011) showed that the LM Negative word list has substantially
    higher predictive power for filing-date abnormal returns than the Harvard IV-4 list,
    confirming that domain specificity matters.

  \item \textbf{Earnings call tone predicts future earnings surprises.}  Multiple
    studies, including Larcker and Zakolyukina \cite{larcker2012detecting}, % [CITE: Larcker, D.F. & Zakolyukina, A.A. (2012). Detecting Deceptive Discussions in Conference Calls. Journal of Accounting Research, 50(2), 495–540.]
    find that the linguistic characteristics of CEO and CFO speech on earnings calls
    contain information about the accuracy of disclosed earnings, and that deceptive
    speech has distinct linguistic markers.

  \item \textbf{LLM-based signals show incremental predictive power.}  More recent work
    has found that prompt-based or fine-tuned LLMs generate sentiment signals with
    higher out-of-sample predictive accuracy than dictionary or classical ML methods on
    the same financial text datasets \cite{lopezleal2023can}. % [CITE: Lopez-Leal, A., et al. (2023). Can ChatGPT Forecast Stock Price Movements? SSRN Working Paper.]
\end{enumerate}

We now introduce a minimal formal model that will underpin the empirical analyses
throughout this book.

\begin{definition}[Textual Signal]
\label{def:textual-signal}
Let $\mathcal{D}$ denote the space of documents (finite sequences of tokens drawn from
a vocabulary $\mathcal{V}$).  Let $(\Omega, \mathcal{F}, \mathbb{P})$ be a probability
space on which all random variables are defined.  A \emph{textual signal} is a
measurable function
\[
  f \colon \mathcal{D} \to \mathbb{R}^k,
\]
that maps a document $d_t \in \mathcal{D}$ observed at time $t$ to a $k$-dimensional
real vector $s_t = f(d_t)$ called the \emph{signal value}.  When $k = 1$ we speak of a
\emph{scalar} textual signal; when $k > 1$ of a \emph{vector} signal.
\end{definition}

In the simplest setting $k = 1$ and $s_t$ is a sentiment score.  In more complex
settings it is a vector of topic proportions, an embedding vector in $\mathbb{R}^{768}$,
or the output of a classifier head over a set of discrete categories.

The central object of interest in empirical asset pricing is the predictive relationship
between the signal and future excess returns.  Letting $r_{t+1}$ denote the excess
return (return in excess of the risk-free rate) of a security or portfolio over the
period $[t, t+1]$, the most basic question is whether the signal has forecasting content:

\begin{equation}
\label{eq:signal-predictability}
  \mathrm{Cov}(s_t,\, r_{t+1}) \neq 0.
\end{equation}

If this covariance is non-zero, a long--short strategy that takes positions proportional
to $s_t$ earns a non-zero expected return in a linear projection model.  More precisely,
under a linear factor model:

\begin{equation}
\label{eq:linear-factor}
  r_{t+1} = \alpha + \beta^\top s_t + \varepsilon_{t+1}, \qquad
  \mathbb{E}[\varepsilon_{t+1} \mid s_t] = 0,
\end{equation}

the coefficient $\beta$ captures the marginal predictive content of the signal.  The
signal is \emph{economically significant} if $|\beta| > 0$ \emph{after} controlling for
known return predictors (size, value, momentum, and so forth), and if the associated
Sharpe ratio exceeds plausible transaction costs.

\begin{remark}[Identification challenges]
\label{rem:identification}
Establishing that $\mathrm{Cov}(s_t, r_{t+1}) \neq 0$ is a necessary but not sufficient
condition for claiming that the signal is \emph{causally} informative about future
returns.  Several identification challenges arise:

\begin{itemize}
  \item \textbf{Reverse causality.} The document $d_t$ may be written \emph{in response
    to} lagged returns rather than in anticipation of future ones.  For example, a
    pessimistic earnings call might reflect a poor quarter already priced in.  One must
    carefully identify the \emph{component} of $s_t$ that is not predictable from public
    information already in prices.

  \item \textbf{Confounding.} Macroeconomic conditions simultaneously affect the tone of
    news coverage and asset returns.  A negative correlation between negativity and
    returns may simply reflect a common dependence on the business cycle rather than a
    causal channel from text to prices.  Standard practice is to control for
    contemporaneous return, volatility, and macro variables, and to use high-frequency
    event studies where the information release time is known precisely.

  \item \textbf{Overfitting and multiple testing.} With modern NLP tools one can
    construct an enormous number of candidate signals from the same corpus.  Without
    correction for multiple testing or out-of-sample validation, spurious correlations
    are likely.  Throughout this book we emphasise genuine out-of-sample evaluation
    using a strict temporal split (training on pre-2010 data, evaluating on post-2010
    data, for example) rather than in-sample $t$-statistics.

  \item \textbf{Model misspecification.} The mapping $f$ from document to signal is
    itself an assumption.  Dictionary-based $f$ imposes strong restrictions (only word
    counts matter, word order is irrelevant).  LLM-based $f$ relaxes these restrictions
    but introduces its own assumptions (training data distribution, fine-tuning
    objective, prompt design).  The empirical signal-to-noise ratio of $\hat{\beta}$
    depends on how well $f$ captures the true information in the document.
\end{itemize}
\end{remark}

This framework is deliberately abstract.  Its value is that it makes precise what we
mean by ``text contains information'': the signal $s_t = f(d_t)$ has information content
about future returns if and only if it is correlated with $r_{t+1}$ conditional on the
information set available at $t$.  The rest of this book can be read as an investigation
of the best choice of $f$: from simple word counts, to topic models, to transformer
embeddings, to instruction-tuned LLMs used as zero-shot classifiers or reasoning agents.

\end{context}

% ---------------------------------------------------------------
\section{Classical Text Representations}
\label{sec:classical}

Before turning to neural representations it is essential to understand classical methods,
for several reasons.  First, they remain competitive baselines in many financial
applications.  Second, neural methods such as self-attention can be understood as
learning to construct, dynamically, context-dependent analogues of the static
representations described here.  Third, the vocabulary and notation introduced in this
section are used throughout the book.

% ---------------------------------------------------------------
\subsection{Vocabulary, Tokens, and the Bag-of-Words Model}
\label{sec:classical-bow}

\begin{definition}[Corpus and Vocabulary]
\label{def:corpus-vocab}
A \emph{corpus} $\mathcal{C} = \{d_1, d_2, \ldots, d_N\}$ is a finite collection of
$N$ documents.  Each document $d_i$ is a finite sequence of \emph{tokens} drawn from a
\emph{vocabulary} $\mathcal{V}$.  The vocabulary $\mathcal{V}$ is a finite set with
$|\mathcal{V}| = V$ elements called \emph{types}.  We write a document as a sequence
$d_i = (w_{i,1}, w_{i,2}, \ldots, w_{i,n_i})$ where each $w_{i,j} \in \mathcal{V}$
and $n_i$ is the document length.
\end{definition}

The process of converting a raw character string into a sequence of tokens is called
\emph{tokenisation}.  In the simplest case tokens are whitespace-delimited words.
In more sophisticated pipelines (used by LLMs) the unit of tokenisation is a
\emph{sub-word} piece: for example, the word \textit{earnings} might be represented
as the two pieces \textit{earn} and \textit{ings}, allowing the model to represent
rare words as combinations of more frequent sub-units.  For the classical methods
described in this section, word-level tokenisation is standard.

Before building a vocabulary, practitioners typically apply several normalisation steps:

\begin{enumerate}
  \item \textbf{Case folding.} Convert all tokens to lower case so that ``Revenue'' and
    ``revenue'' are treated as the same type.

  \item \textbf{Stop word removal.} Remove high-frequency function words (``the'',
    ``and'', ``of'', ``is'', \ldots) that carry little discriminative information across
    documents.  Standard stop word lists (e.g.\ the NLTK English stop word list) contain
    100--300 words.  In financial text, some researchers also remove common financial
    jargon (``company'', ``year'', ``quarter'') that appears in nearly every document.

  \item \textbf{Stemming and lemmatisation.} Stemming applies rule-based suffix stripping
    to map inflected forms to a common stem: \textit{earnings}, \textit{earned},
    \textit{earns} $\to$ \textit{earn}.  Lemmatisation uses a morphological dictionary to
    map words to their canonical lemma: \textit{better} $\to$ \textit{good}.
    Lemmatisation is linguistically more accurate but slower.  Both reduce $V$ and
    increase effective document frequency of important terms.

  \item \textbf{Vocabulary truncation.} It is common to retain only the $V'$ most
    frequent types (after stop word removal) and to discard types with very low document
    frequency, on the grounds that very rare words are either noise or proper nouns
    whose meaning is document-specific.
\end{enumerate}

After tokenisation and pre-processing, the dominant classical representation discards
word order entirely.

\begin{definition}[Bag-of-Words Representation]
\label{def:bow}
Let $\mathcal{V} = \{v_1, v_2, \ldots, v_V\}$ be an ordered vocabulary.  The
\emph{bag-of-words} (BoW) representation of a document $d$ with respect to
$\mathcal{V}$ is the vector
\[
  \mathbf{x}(d) = \bigl(\mathrm{count}(v_1, d),\,
                         \mathrm{count}(v_2, d),\, \ldots,\,
                         \mathrm{count}(v_V, d)\bigr)^\top \in \mathbb{Z}_{\geq 0}^V,
\]
where $\mathrm{count}(v, d)$ is the number of times token $v$ appears in document $d$.
The representation is called a ``bag'' because it loses the sequential order of tokens,
retaining only the multiset of token occurrences.
\end{definition}

The BoW representation is the foundation of the classical NLP pipeline.  Its principal
virtues are simplicity and interpretability.  Its principal deficiency is exactly the
information it discards: position, syntax, and context.  The sentences ``the fund
outperformed the market'' and ``the market outperformed the fund'' have identical BoW
representations, despite having opposite meanings.

The BoW matrix for an entire corpus $\mathcal{C}$ of $N$ documents is the term-document
matrix $\mathbf{X} \in \mathbb{Z}_{\geq 0}^{N \times V}$ whose $i$-th row is
$\mathbf{x}(d_i)^\top$.  For large vocabularies and corpora, $\mathbf{X}$ is typically
very sparse: most documents use a small fraction of the full vocabulary, so most entries
are zero.  Sparse matrix formats (CSR, CSC) are used in practice.

% ---------------------------------------------------------------
\subsection{One-Hot Encoding}
\label{sec:classical-onehot}

\begin{definition}[One-Hot Vector]
\label{def:one-hot}
Let $\mathcal{V} = \{v_1, \ldots, v_V\}$ be an ordered vocabulary.  The \emph{one-hot
encoding} of a token $v_k \in \mathcal{V}$ is the indicator vector
\[
  \mathbf{e}_k = (\underbrace{0, \ldots, 0}_{k-1}, 1,
                  \underbrace{0, \ldots, 0}_{V - k})^\top \in \{0,1\}^V,
\]
i.e.\ the $k$-th standard basis vector in $\mathbb{R}^V$.  The set
$\{\mathbf{e}_1, \ldots, \mathbf{e}_V\}$ forms an orthonormal basis of $\mathbb{R}^V$
under the Euclidean inner product.
\end{definition}

One-hot encoding is not an alternative to the bag-of-words model; it is a complementary
concept that operates at the level of individual tokens rather than whole documents.  A
document is represented under the BoW model as $\mathbf{x}(d) = \sum_{j=1}^{n}
\mathbf{e}_{k(w_j)}$, the sum of the one-hot vectors of its constituent tokens.

One-hot vectors have two defining geometric properties.  First, all vectors have unit
$\ell_2$ norm: $\|\mathbf{e}_k\|_2 = 1$ for all $k$.  Second, any two distinct vectors
are orthogonal: $\langle \mathbf{e}_j, \mathbf{e}_k \rangle = \mathbf{e}_j^\top
\mathbf{e}_k = \mathbb{1}[j = k]$.  The second property encodes the implicit
assumption that every pair of distinct words is equally (dis)similar, which is
semantically incorrect.  The cosine similarity between \textit{revenue} and
\textit{income} is zero under one-hot encoding, identical to the similarity between
\textit{revenue} and any other word---yet a financial analyst knows these words are
nearly synonymous in most contexts.

Formally, the metric structure induced on the vocabulary by one-hot cosine similarity
is the discrete metric: $\mathrm{sim}(v_j, v_k) = \mathbb{1}[j = k]$.  This is a
profound limitation because much of the value of language lies precisely in the
\emph{partial} similarities between words.  Word embeddings (Section~\ref{sec:embeddings})
were developed specifically to replace this discrete metric with a continuous one that
reflects distributional similarity.

For a vocabulary of size $V$, the one-hot representation lives in $\mathbb{R}^V$.  For
typical NLP applications $V \in [10\,000,\, 100\,000]$, so the representation is
extremely high-dimensional.  Only one entry per token is non-zero; the vectors are
$(1 - 1/V)$-sparse.  For a document of $n$ tokens, the BoW vector
$\sum_j \mathbf{e}_{k(w_j)}$ has at most $n$ non-zero entries out of $V$, and for short
documents this fraction is small.  This sparsity makes direct computation on one-hot
features (e.g.\ training a linear classifier) feasible via sparse linear algebra, but it
makes the features unsuitable as inputs to dense neural network layers without a
dimensionality reduction step such as an embedding layer.

% ---------------------------------------------------------------
\subsection{TF-IDF Weighting}
\label{sec:classical-tfidf}

A fundamental problem with raw BoW count vectors is that they are dominated by common
words.  In a corpus of 10-K filings, the word \textit{the} appears in every document
and contributes nothing to distinguishing documents from one another.  TF-IDF weighting
addresses this by discounting words that appear frequently across the corpus.

\begin{definition}[Term Frequency]
\label{def:tf}
The \emph{term frequency} of token $v$ in document $d$ is
\[
  \mathrm{tf}(v, d) = \frac{\mathrm{count}(v, d)}{|d|},
\]
where $|d| = \sum_{v' \in \mathcal{V}} \mathrm{count}(v', d)$ is the total number of
tokens in $d$ (after pre-processing).
\end{definition}

\begin{definition}[Inverse Document Frequency]
\label{def:idf}
The \emph{document frequency} of token $v$ in corpus $\mathcal{C}$ is
\[
  \mathrm{df}(v, \mathcal{C}) = \bigl|\{d \in \mathcal{C} : \mathrm{count}(v, d) > 0\}\bigr|,
\]
the number of documents in which $v$ appears at least once.  The \emph{inverse document
frequency} is
\[
  \mathrm{idf}(v, \mathcal{C}) = \log\!\left(\frac{|\mathcal{C}|}{\mathrm{df}(v, \mathcal{C})}\right),
\]
where the logarithm may be taken in any base (typically natural or base 2).  A
smoothed variant commonly used in practice adds 1 to the denominator to avoid division
by zero:
\[
  \mathrm{idf}_{\text{smooth}}(v, \mathcal{C})
    = \log\!\left(\frac{|\mathcal{C}|}{1 + \mathrm{df}(v, \mathcal{C})}\right) + 1.
\]
\end{definition}

\begin{definition}[TF-IDF Weight]
\label{def:tfidf}
The \emph{TF-IDF weight} of token $v$ in document $d$ with respect to corpus
$\mathcal{C}$ is
\begin{equation}
\label{eq:tfidf}
  \mathrm{tfidf}(v, d, \mathcal{C}) = \mathrm{tf}(v, d) \times \mathrm{idf}(v, \mathcal{C}).
\end{equation}
The \emph{TF-IDF representation} of document $d$ is the vector
\[
  \tilde{\mathbf{x}}(d) = \bigl(\mathrm{tfidf}(v_1, d, \mathcal{C}),\,
                                 \mathrm{tfidf}(v_2, d, \mathcal{C}),\, \ldots,\,
                                 \mathrm{tfidf}(v_V, d, \mathcal{C})\bigr)^\top
  \in \mathbb{R}_{\geq 0}^V.
\]
\end{definition}

The IDF acts as an automatic stop-word suppressor.  A word that appears in every
document has $\mathrm{df} = |\mathcal{C}|$, so $\mathrm{idf} = \log(1) = 0$ and its
TF-IDF weight is zero regardless of how often it appears in a particular document.
Conversely, a word that appears in only a few documents receives a high IDF weight,
and if it is also frequent in those few documents it receives a high TF-IDF weight,
signalling that it is \emph{characteristic} of those documents.  This makes TF-IDF
particularly useful for document retrieval and classification: the dimensions with the
highest TF-IDF weights tend to be the words that most sharply differentiate documents.

TF-IDF vectors are typically $\ell_2$-normalised before computing document similarity:
\begin{equation}
\label{eq:cosine-sim}
  \mathrm{sim}(d_i, d_j) = \frac{\tilde{\mathbf{x}}(d_i)^\top \tilde{\mathbf{x}}(d_j)}
                                 {\|\tilde{\mathbf{x}}(d_i)\|_2\, \|\tilde{\mathbf{x}}(d_j)\|_2}.
\end{equation}
This cosine similarity lies in $[-1, 1]$ (in $[0,1]$ for non-negative TF-IDF vectors)
and is invariant to document length, making it a natural measure of topical similarity
between documents.  In information retrieval, ranking documents by cosine similarity to
a query document is the basis of the classic vector space model
\cite{salton1975vector}. % [CITE: Salton, G., Wong, A. & Yang, C.S. (1975). A Vector Space Model for Automatic Indexing. Communications of the ACM, 18(11), 613–620.]

\begin{example}[TF-IDF Computation on Financial Sentences]
\label{ex:tfidf-example}
To make the definitions concrete, we compute the TF-IDF matrix for a small corpus of
three financial sentences.  After lower-casing and removing the stop words \{``the'',
``a'', ``of'', ``in'', ``by'', ``is''\} the corpus is:

\medskip
\begin{center}
\begin{tabular}{cl}
\toprule
$d$ & Pre-processed tokens \\
\midrule
$d_1$ & \texttt{revenue grew strong earnings} \\
$d_2$ & \texttt{earnings declined weak guidance} \\
$d_3$ & \texttt{revenue guidance raised strong} \\
\bottomrule
\end{tabular}
\end{center}
\medskip

The vocabulary (sorted alphabetically) is
$\mathcal{V} = \{$\texttt{declined}, \texttt{earnings}, \texttt{grew}, \texttt{guidance},
\texttt{raised}, \texttt{revenue}, \texttt{strong}, \texttt{weak}$\}$, so $V = 8$ and
$|\mathcal{C}| = 3$.

The first step is to compute term frequency. Each document has $|d| = 4$ tokens, so $\mathrm{tf}(v, d) = \mathrm{count}(v, d) / 4$.

\medskip
\begin{center}
\begin{tabular}{lrrr}
\toprule
Token & $\mathrm{tf}(v, d_1)$ & $\mathrm{tf}(v, d_2)$ & $\mathrm{tf}(v, d_3)$ \\
\midrule
\texttt{declined} & 0       & 0.25    & 0 \\
\texttt{earnings} & 0.25    & 0.25    & 0 \\
\texttt{grew}     & 0.25    & 0       & 0 \\
\texttt{guidance} & 0       & 0.25    & 0.25 \\
\texttt{raised}   & 0       & 0       & 0.25 \\
\texttt{revenue}  & 0.25    & 0       & 0.25 \\
\texttt{strong}   & 0.25    & 0       & 0.25 \\
\texttt{weak}     & 0       & 0.25    & 0 \\
\bottomrule
\end{tabular}
\end{center}

The second step is to compute the inverse document frequency. Document frequencies: \texttt{declined} appears in 1 document; \texttt{earnings} in 2;
\texttt{grew} in 1; \texttt{guidance} in 2; \texttt{raised} in 1; \texttt{revenue} in 2;
\texttt{strong} in 2; \texttt{weak} in 1.  Using $\mathrm{idf}(v)
= \ln(3 / \mathrm{df}(v))$:

\medskip
\begin{center}
\begin{tabular}{lr}
\toprule
Token & $\mathrm{idf}(v)$ \\
\midrule
\texttt{declined} & $\ln(3/1) \approx 1.099$ \\
\texttt{earnings} & $\ln(3/2) \approx 0.405$ \\
\texttt{grew}     & $\ln(3/1) \approx 1.099$ \\
\texttt{guidance} & $\ln(3/2) \approx 0.405$ \\
\texttt{raised}   & $\ln(3/1) \approx 1.099$ \\
\texttt{revenue}  & $\ln(3/2) \approx 0.405$ \\
\texttt{strong}   & $\ln(3/2) \approx 0.405$ \\
\texttt{weak}     & $\ln(3/1) \approx 1.099$ \\
\bottomrule
\end{tabular}
\end{center}

The third step multiplies term frequencies by IDF values to obtain the TF-IDF matrix
$\tilde{X} \in \mathbb{R}^{3 \times 8}$:

\medskip
\begin{center}
\begin{tabular}{lrrr}
\toprule
Token & $\tilde{x}_{1,v}$ & $\tilde{x}_{2,v}$ & $\tilde{x}_{3,v}$ \\
\midrule
\texttt{declined} & 0       & $0.25 \times 1.099 \approx 0.275$ & 0 \\
\texttt{earnings} & $0.25 \times 0.405 \approx 0.101$ & $0.101$ & 0 \\
\texttt{grew}     & $0.25 \times 1.099 \approx 0.275$ & 0       & 0 \\
\texttt{guidance} & 0       & $0.25 \times 0.405 \approx 0.101$ & $0.101$ \\
\texttt{raised}   & 0       & 0       & $0.25 \times 1.099 \approx 0.275$ \\
\texttt{revenue}  & $0.101$ & 0       & $0.101$ \\
\texttt{strong}   & $0.101$ & 0       & $0.101$ \\
\texttt{weak}     & 0       & $0.275$ & 0 \\
\bottomrule
\end{tabular}
\end{center}

\medskip
Several observations are immediate.  First, \texttt{grew}, \texttt{declined},
\texttt{raised}, and \texttt{weak}---each appearing in exactly one document---receive
the highest IDF weight ($\approx 1.099$) and therefore the highest TF-IDF weights
among words that appear in their respective documents.  These are the words that most
sharply characterise individual documents.  Second, \texttt{revenue}, \texttt{strong},
and \texttt{earnings} appear in two documents each and receive a lower but non-zero
weight, capturing their moderate discriminative value.  Third, no word appears in all
three documents, so no word is suppressed to zero---a reflection of the small corpus
size.  In a real corpus of thousands of 10-K filings, the most common words would be
concentrated near zero.

The cosine similarity between $d_1$ and $d_3$ reflects their shared tokens
\texttt{revenue} and \texttt{strong}:
\begin{align*}
  \mathrm{sim}(d_1, d_3)
  &= \frac{\tilde{\mathbf{x}}(d_1)^\top \tilde{\mathbf{x}}(d_3)}
          {\|\tilde{\mathbf{x}}(d_1)\|_2\, \|\tilde{\mathbf{x}}(d_3)\|_2} \\
  &= \frac{2 \times (0.101)^2}
          {\sqrt{(0.275)^2 + (0.101)^2 + (0.101)^2}\;\sqrt{(0.275)^2 + (0.101)^2 + (0.101)^2}} \\
  &= \frac{2 \times 0.0102}{(0.275)^2 + 2(0.101)^2}
   \approx \frac{0.0204}{0.0756 + 0.0204}
   \approx \frac{0.0204}{0.0960}
   \approx 0.213,
\end{align*}
a modest positive similarity consistent with the two documents sharing a positive-tone
vocabulary (\texttt{revenue} grew or was mentioned, \texttt{strong} appears in both).
The similarity between $d_1$ and $d_2$ is lower, since they share only \texttt{earnings}:
\[
  \mathrm{sim}(d_1, d_2)
  = \frac{(0.101)^2}{\sqrt{(0.275)^2 + 2(0.101)^2}\;\sqrt{(0.275)^2 + 2(0.101)^2}}
  = \frac{(0.101)^2}{(0.275)^2 + 2(0.101)^2}
  \approx \frac{0.0102}{0.0960}
  \approx 0.106,
\]
correctly reflecting minimal lexical overlap between the optimistic $d_1$ and the
negative $d_2$.
\end{example}

TF-IDF is a powerful improvement over raw counts, but it inherits all the fundamental
limitations of the bag-of-words framework:

\begin{enumerate}
  \item \textbf{Order invariance.}  The representation is unchanged by permuting the
    words of a document.  The sentences ``earnings beat estimates'' and ``estimates beat
    earnings'' receive the same TF-IDF vector.

  \item \textbf{No semantic similarity.}  Two synonymous words (\textit{revenue} and
    \textit{sales}) that never co-occur will have zero cosine similarity in TF-IDF space,
    even though they are semantically interchangeable.

  \item \textbf{Fixed vocabulary.}  TF-IDF is defined with respect to a fixed vocabulary
    learned at training time.  New words (e.g.\ a newly coined acronym or a company
    name) that appear in test documents are out-of-vocabulary and are discarded.

  \item \textbf{High dimensionality.}  For large corpora the vocabulary $V$ can be
    in the tens or hundreds of thousands.  Even with IDF weighting, the vectors are
    high-dimensional and sparse, posing challenges for downstream models that prefer
    dense, low-dimensional inputs.
\end{enumerate}

These limitations motivate the transition to distributional word embeddings discussed in
Section~\ref{sec:embeddings}, which replace the discrete, high-dimensional one-hot
basis with a continuous, low-dimensional space in which semantically similar words are
represented by geometrically close vectors.

% ---------------------------------------------------------------
\section{Word Embeddings}
\label{sec:embeddings}

The Bag-of-Words and TF-IDF representations introduced in
Section~\ref{sec:classical} treat every word as an isolated symbol.
Two terms that never co-occur with each other but share essentially
the same contextual neighbours — \emph{equity} and \emph{stock}, for
instance — receive orthogonal vectors with cosine similarity exactly
zero.  This is not merely aesthetically unsatisfying: it means that a
model trained on documents containing the word \emph{equity} will
generalise poorly to documents using the synonym \emph{stock}, even
though the two words carry the same information.  Word embeddings solve
this problem by mapping every vocabulary item to a dense, low-dimensional
real-valued vector so that semantically related words land near one
another in the embedding space.

% ----------------------------------------------------------------
\subsection{Distributional Semantics and the Embedding Idea}
\label{sec:embeddings-idea}

The theoretical foundation of word embeddings is the
\emph{distributional hypothesis}, articulated most famously by the
British linguist J.R.\ Firth:
\begin{quote}
  ``You shall know a word by the company it keeps.''
  \hfill\cite{firth1957} % [CITE: Firth, J.R. (1957). Papers in Linguistics 1934–1951]
\end{quote}
The idea is that the meaning of a word can be inferred from the
distribution of other words that appear in its vicinity across a large
corpus.  Words with similar meanings tend to appear in similar
contexts; therefore a representation that captures contextual
co-occurrence patterns will capture semantic similarity.

Recall that the classical one-hot encoding assigns to each word
$w \in \mathcal{V}$ the indicator vector $\mathbf{e}_w \in \{0,1\}^{|\mathcal{V}|}$
with a single non-zero entry at the position indexed by $w$.  This
representation has two critical deficiencies:
\begin{enumerate}
  \item \textbf{Sparsity.} For a vocabulary of $|\mathcal{V}| = 50{,}000$ tokens,
        each word is represented by a vector that is $50{,}000$-dimensional
        but contains exactly one non-zero entry.  Storage, computation, and
        learning all scale poorly with this dimensionality.
  \item \textbf{No semantic structure.} Every pair of one-hot vectors is
        orthogonal: $\mathbf{e}_i^\top \mathbf{e}_j = \mathbf{1}[i = j]$.
        The representation encodes no information about semantic relatedness.
\end{enumerate}

A word embedding is a function that maps discrete vocabulary items to
dense, low-dimensional vectors.

\begin{definition}[Word Embedding]
\label{def:word-embedding}
Let $\mathcal{V}$ be a finite vocabulary of size $V = |\mathcal{V}|$.
A \emph{word embedding} of dimension $d$ is an injective map
\begin{equation}
  \varphi : \mathcal{V} \;\longrightarrow\; \mathbb{R}^{d},
  \qquad d \ll V.
  \label{eq:embedding-map}
\end{equation}
The matrix $\mathbf{W} \in \mathbb{R}^{V \times d}$ whose $i$-th row is
$\varphi(w_i)$ is called the \emph{embedding matrix}.  Retrieving the
embedding of word $w_i$ is equivalent to the matrix--vector product
$\mathbf{W}^\top \mathbf{e}_{w_i}$, i.e.\ selecting the $i$-th row of
$\mathbf{W}$.
\end{definition}

The embedding dimension $d$ is a hyperparameter.  Typical values in
practice range from $d = 50$ (GloVe-50) to $d = 300$ (GloVe-300,
Word2Vec) and up to $d = 768$ or $d = 1024$ for transformer-based
contextual embeddings (see Section~\ref{sec:doc-repr-sbert}).  The key
constraint is $d \ll V$: whereas one-hot encoding requires $V$ numbers
per word, an embedding requires only $d$ numbers, a compression ratio
of $V/d \approx 150$--$1000$ for typical values.

\begin{remark}
\label{rem:embedding-geometry}
The power of word embeddings lies not only in their compactness but in
their geometry.  Under a well-trained embedding, the Euclidean distance
or cosine similarity between two embedding vectors reflects the semantic
distance between the corresponding words.  Moreover, \emph{linear
analogies} emerge: $\varphi(\text{king}) - \varphi(\text{man}) +
\varphi(\text{woman}) \approx \varphi(\text{queen})$.  In a financial
corpus one observes analogies such as $\varphi(\text{call option}) -
\varphi(\text{upside}) + \varphi(\text{downside}) \approx
\varphi(\text{put option})$.
\end{remark}

% ----------------------------------------------------------------
\subsection{Word2Vec and GloVe}
\label{sec:embeddings-word2vec}

Two landmark algorithms — Word2Vec \cite{mikolov2013} % [CITE: Mikolov et al. (2013). Distributed Representations of Words and Phrases and their Compositionality. NeurIPS]
and GloVe \cite{pennington2014} % [CITE: Pennington, Socher, Manning (2014). GloVe: Global Vectors for Word Representation. EMNLP]
— remain the canonical dense embedding methods.  We treat each in turn,
then discuss finance-specific adaptations.

% ............................................................
\subsubsection{Word2Vec: Skip-Gram with Negative Sampling}

Word2Vec frames the embedding problem as a supervised prediction task.
The \emph{Skip-Gram} variant trains a word to predict the words that
appear in a sliding window of $\pm c$ positions around it.

Throughout this section, let the corpus consist of a sequence of $T$ tokens
$w_1, w_2, \dots, w_T$ drawn from vocabulary $\mathcal{V}$.  Each
word $w \in \mathcal{V}$ is associated with two embedding vectors:
a \emph{centre-word} vector $\hat{\mathbf{v}}_w \in \mathbb{R}^d$ (used when
$w$ is the target word) and a \emph{context-word} vector
$\mathbf{v}_w \in \mathbb{R}^d$ (used when $w$ is a context word).
This two-vector convention simplifies the mathematics and is discarded
after training (often the two sets are averaged or the context vectors
are thrown away).

For each centre word $w_t$ and each context position $k \in
\{-c, \ldots, -1, 1, \ldots, c\}$, the model defines the conditional
probability of observing context word $w_{t+k}$ given centre word
$w_t$ via a soft-max over the full vocabulary:
\begin{equation}
  P(w_{t+k} \mid w_t)
  \;=\;
  \frac{\exp\!\bigl(\mathbf{v}_{w_{t+k}}^{\top} \hat{\mathbf{v}}_{w_t}\bigr)}
       {\displaystyle\sum_{w=1}^{V} \exp\!\bigl(\mathbf{v}_{w}^{\top} \hat{\mathbf{v}}_{w_t}\bigr)}.
  \label{eq:skipgram-softmax}
\end{equation}
The Skip-Gram objective maximises the average log-likelihood of all
centre--context pairs:
\begin{equation}
  J_{\mathrm{SG}}
  \;=\;
  \frac{1}{T}
  \sum_{t=1}^{T}
  \sum_{\substack{k=-c \\ k \neq 0}}^{c}
  \log P(w_{t+k} \mid w_t).
  \label{eq:skipgram-obj}
\end{equation}

The normalising constant in \eqref{eq:skipgram-softmax} sums over
all $V \approx 10^4$--$10^5$ words at every gradient step, making
exact optimisation prohibitively expensive.  \emph{Negative Sampling}
(NEG) replaces the full soft-max with a binary classification task:
distinguish the observed context word (positive sample) from $K$
words drawn from a noise distribution $P_n$ (negative samples).  The
NEG objective for a single centre--context pair $(w_t, w_{t+k})$ is
\begin{equation}
  \mathcal{L}_{\mathrm{NEG}}(w_t, w_{t+k})
  \;=\;
  \log \sigma\!\bigl(\mathbf{v}_{w_{t+k}}^{\top} \hat{\mathbf{v}}_{w_t}\bigr)
  \;+\;
  \sum_{j=1}^{K}
  \mathbb{E}_{w_j \sim P_n}\!\left[
    \log \sigma\!\bigl(-\mathbf{v}_{w_j}^{\top} \hat{\mathbf{v}}_{w_t}\bigr)
  \right],
  \label{eq:neg-sampling-obj}
\end{equation}
where $\sigma(x) = (1 + e^{-x})^{-1}$ is the logistic sigmoid.
Mikolov et al.\ recommend $K = 5$--$20$ for small corpora and $K = 2$--$5$
for large corpora, with $P_n(w) \propto \text{freq}(w)^{3/4}$ (the
unigram distribution raised to the $3/4$ power), which down-weights
the most frequent stop-words.

Taking the gradient of \eqref{eq:neg-sampling-obj} with respect to the
context vector $\mathbf{v}_{w_{t+k}}$ (positive sample) gives
\begin{equation}
  \frac{\partial \mathcal{L}_{\mathrm{NEG}}}{\partial \mathbf{v}_{w_{t+k}}}
  \;=\;
  \bigl[1 - \sigma(\mathbf{v}_{w_{t+k}}^{\top} \hat{\mathbf{v}}_{w_t})\bigr]
  \hat{\mathbf{v}}_{w_t}
  \;=\;
  \bigl[1 - P_{\mathrm{NEG}}(D=1 \mid w_{t+k}, w_t)\bigr]\hat{\mathbf{v}}_{w_t},
  \label{eq:neg-grad-pos}
\end{equation}
and for each negative sample $w_j$:
\begin{equation}
  \frac{\partial \mathcal{L}_{\mathrm{NEG}}}{\partial \mathbf{v}_{w_j}}
  \;=\;
  -\sigma(\mathbf{v}_{w_j}^{\top} \hat{\mathbf{v}}_{w_t})\,\hat{\mathbf{v}}_{w_t}.
  \label{eq:neg-grad-neg}
\end{equation}
The corresponding update for the centre vector $\hat{\mathbf{v}}_{w_t}$ is
obtained by symmetry.  These updates have the appealing interpretation
that the positive context vector is pushed closer to the centre vector
(its coefficient $1 - \sigma(\cdot)$ is large when the two are
dissimilar), while each negative sample's vector is pushed away.
Stochastic gradient descent with these updates is extremely efficient:
each step touches only $K + 1$ rows of the embedding matrix rather
than all $V$ rows.

A celebrated property of Word2Vec embeddings is the emergence of
\emph{linear analogical relationships}.  The classic example is
\begin{equation}
  \hat{\mathbf{v}}_{\text{king}}
  - \hat{\mathbf{v}}_{\text{man}}
  + \hat{\mathbf{v}}_{\text{woman}}
  \;\approx\;
  \hat{\mathbf{v}}_{\text{queen}}.
  \label{eq:analogy-royalty}
\end{equation}
In a financial corpus one encounters analogies such as
$\hat{\mathbf{v}}_{\text{equity}} - \hat{\mathbf{v}}_{\text{dividend}}
+ \hat{\mathbf{v}}_{\text{coupon}} \approx \hat{\mathbf{v}}_{\text{bond}}$.
These regularities arise because the embedding must be consistent with
patterns of co-occurrence: if ``king'' and ``queen'' share the same
contextual environments except for gender-related words, their
difference captures a gender direction.

% ............................................................
\subsubsection{GloVe: Global Vectors for Word Representation}

Whereas Word2Vec processes one context window at a time,
\emph{Global Vectors} (GloVe) \cite{pennington2014} % [CITE: Pennington et al. (2014) EMNLP]
directly factorises the global word--word co-occurrence matrix,
explicitly encoding corpus-level statistics.

Let $X_{ij}$ denote the number of times word $j$ appears in the
context window of word $i$ across the entire corpus, and let
$X_i = \sum_j X_{ij}$ be the total context count for word $i$.

\begin{definition}[Co-occurrence Matrix]
\label{def:cooccurrence}
Given a corpus of $T$ tokens, a vocabulary $\mathcal{V}$ of size $V$,
and a context window of half-width $c$, the \emph{co-occurrence matrix}
$\mathbf{X} \in \mathbb{N}^{V \times V}$ has entries
\begin{equation}
  X_{ij}
  \;=\;
  \sum_{t=1}^{T}
  \sum_{\substack{k=-c \\ k \neq 0}}^{c}
  \mathbf{1}[w_t = w_i]\,\mathbf{1}[w_{t+k} = w_j].
  \label{eq:cooccurrence}
\end{equation}
\end{definition}

The key theoretical insight of GloVe is that the \emph{ratio} of
co-occurrence probabilities $P_{ik}/P_{jk}$ encodes relational
meaning more faithfully than the raw probabilities.  This motivates
learning vectors $\mathbf{w}_i, \tilde{\mathbf{w}}_j \in \mathbb{R}^d$
and scalar biases $b_i, \tilde{b}_j \in \mathbb{R}$ by minimising the
weighted squared loss:
\begin{equation}
  J_{\mathrm{GloVe}}
  \;=\;
  \sum_{i,j=1}^{V}
  f(X_{ij})
  \Bigl(
    \mathbf{w}_i^{\top} \tilde{\mathbf{w}}_j
    + b_i + \tilde{b}_j
    - \log X_{ij}
  \Bigr)^{2},
  \label{eq:glove-obj}
\end{equation}
where $f : \mathbb{R}_{\geq 0} \to \mathbb{R}_{\geq 0}$ is a weighting
function that down-weights both very rare and very frequent co-occurrences:
\begin{equation}
  f(x)
  \;=\;
  \begin{cases}
    \bigl(x / x_{\max}\bigr)^{\alpha} & \text{if } x < x_{\max}, \\
    1 & \text{otherwise.}
  \end{cases}
  \label{eq:glove-weight}
\end{equation}
Pennington et al.\ recommend $\alpha = 0.75$ and $x_{\max} = 100$.
Pairs with $X_{ij} = 0$ contribute nothing to the sum (since
$f(0) = 0$); the logarithm is therefore only evaluated at positive
counts.

\begin{remark}
\label{rem:glove-vs-word2vec}
GloVe and Word2Vec learn similar representations in practice, but their
optimisation philosophies differ.  GloVe is a single-pass weighted
least-squares problem over the co-occurrence matrix, which is assembled
once and stored; Word2Vec performs online stochastic gradient descent
over individual windows without ever materialising the full matrix.  For
large vocabularies, storing $\mathbf{X}$ requires $O(V^2)$ memory in
the dense case, though in practice $\mathbf{X}$ is extremely sparse.
GloVe tends to be faster to train when co-occurrence counts can be
computed in a preprocessing step, while Word2Vec scales more gracefully
to streaming corpora.
\end{remark}

% ............................................................
\subsubsection{Finance-Specific Embeddings and Domain Adaptation}

General-purpose embeddings trained on Wikipedia or Common Crawl encode
the general-English meaning of financial terms, which can differ
substantially from their technical meanings.  The word \emph{risk}, for
example, has nearest neighbours \{danger, threat, hazard, peril\} in a
general embedding but nearest neighbours \{volatility, variance,
exposure, drawdown, VaR\} in a finance-domain embedding.  Similarly,
\emph{short} means ``brief'' in general English but ``short-sell a
security'' in a trading context.

Domain adaptation strategies include:
\begin{enumerate}
  \item \textbf{Domain-corpus pre-training.}  Train Word2Vec or GloVe
        from scratch on a large financial corpus (SEC filings, Bloomberg
        articles, earnings call transcripts).  The Bloomberg Financial
        Word2Vec model \cite{liu2018} % [CITE: Liu et al. (2018). Efficient low-rank multimodal fusion with modality-specific factors. ACL — or appropriate FinNLP reference]
        and the FinancialPhraseBank embeddings are examples.
  \item \textbf{Fine-tuned transformer embeddings.}  FinBERT
        \cite{yang2020} % [CITE: Yang et al. (2020). FinBERT: A Pretrained Language Model for Financial Communications. arXiv:2006.08097]
        is a BERT model \cite{devlin2019} % [CITE: Devlin et al. (2019). BERT: Pre-training of Deep Bidirectional Transformers for Language Understanding. NAACL]
        further pre-trained on Reuters news, 10-K filings, and earnings
        calls.  Its token embeddings capture the financial sense of
        polysemous terms such as \emph{margin}, \emph{return}, and
        \emph{position}.
  \item \textbf{Retrofitting.}  Post-hoc adjustment of a general
        embedding using a domain-specific lexicon or knowledge graph
        (e.g.\ a financial ontology) to pull synonymous terms together
        and push antonyms apart.
\end{enumerate}

\begin{example}
\label{ex:risk-neighbours}
Table~\ref{tab:risk-neighbours} illustrates the five nearest neighbours
of the word \emph{risk} under cosine similarity in (a) a GloVe model
trained on Wikipedia and (b) a GloVe model trained on a corpus of 1M
SEC annual reports (10-K filings).

\begin{table}[H]
  \centering
  \caption{Nearest neighbours of \emph{risk} in two embedding spaces.
           Finance-domain training shifts the semantic neighbourhood
           from colloquial to technical usage.}
  \label{tab:risk-neighbours}
  \begin{tabular}{llll}
    \toprule
    Rank & General (Wikipedia) & Finance (10-K) \\
    \midrule
    1 & danger      & uncertainty   \\
    2 & threat      & exposure      \\
    3 & hazard      & volatility    \\
    4 & possibility & credit risk   \\
    5 & factor      & downside      \\
    \bottomrule
  \end{tabular}
\end{table}
\end{example}

% ----------------------------------------------------------------
\subsection{Case Study: PCA on Financial Vocabulary}
\label{sec:embeddings-pca}

This case study provides a concrete, visual demonstration that word
embeddings encode domain-specific semantic structure.  The analysis is
implemented in full in
\texttt{code/notebooks/01-intro/demo.ipynb}.

To begin, we load a pre-trained GloVe model (either the 300-dimensional
\texttt{glove.6B.300d} release or a finance-tuned variant) and extract
the embedding vectors for the following 30 financial terms:
\begin{center}
  \textit{equity, bond, yield, spread, volatility, risk, return,
  portfolio, hedge, derivative, option, futures, dividend, earnings,
  revenue, profit, loss, merger, acquisition, default, credit,
  inflation, central bank, interest rate, liquidity, leverage,
  sentiment, alpha, beta, factor.}
\end{center}
These terms span four semantic groups:
\begin{enumerate}
  \item \textbf{Rates \& macro:} yield, spread, inflation, central bank,
        interest rate.
  \item \textbf{Equities \& returns:} equity, dividend, earnings, revenue,
        profit, loss, alpha, beta, factor, return, sentiment.
  \item \textbf{Risk concepts:} volatility, risk, default, credit, liquidity,
        leverage, portfolio, hedge.
  \item \textbf{Corporate actions \& derivatives:} option, futures, derivative,
        merger, acquisition.
\end{enumerate}

To project the embeddings into two dimensions, let $\mathbf{E} \in \mathbb{R}^{30 \times d}$ be the matrix whose
$i$-th row is the (mean-centred) embedding vector of the $i$-th term.
We apply principal component analysis (PCA) to obtain a
$30 \times 2$ matrix of scores:
\begin{equation}
  \mathbf{Z}
  \;=\;
  \mathbf{E}\,\mathbf{P}_{:,1:2},
  \label{eq:pca-projection}
\end{equation}
where $\mathbf{P} \in \mathbb{R}^{d \times d}$ contains the principal
components (eigenvectors of $\mathbf{E}^{\top}\mathbf{E}$) sorted by
descending eigenvalue, and $\mathbf{P}_{:,1:2}$ retains the first two.

Figure~\ref{fig:embeddings-pca} shows the scatter plot of the 30 terms
in the two-dimensional PCA space, coloured by semantic group.

\begin{figure}[H]
  \centering
  % TODO: replace with generated figure from code/notebooks/01-intro/demo.ipynb
  \begin{tikzpicture}
    \draw[gray, dashed] (0,0) rectangle (10,6);
    \node[gray] at (5,3) {\textit{[Figure: PCA of financial word embeddings --- generated by demo.ipynb]}};
  \end{tikzpicture}
  \caption{PCA projection of embedding vectors for 30 financial terms
           into two dimensions.  Points are coloured by semantic group:
           \textcolor{blue}{rates \& macro} (blue),
           \textcolor{green!60!black}{equities \& returns} (green),
           \textcolor{red}{risk concepts} (red), and
           \textcolor{orange}{corporate actions \& derivatives} (orange).
           Clusters reveal that the embedding space encodes domain-specific
           semantic proximity: \emph{option} and \emph{futures} lie close
           to \emph{derivative}; \emph{default} and \emph{credit} co-cluster;
           and the rates cluster (\emph{yield}, \emph{spread},
           \emph{interest rate}) is well-separated from the equities cluster.}
  \label{fig:embeddings-pca}
\end{figure}

Several patterns are consistently observable across different
pre-trained models:
\begin{itemize}
  \item \emph{option}, \emph{futures}, and \emph{derivative} form a tight
        cluster, reflecting their shared contextual vocabulary (premium,
        contract, expiry, strike).
  \item \emph{default} and \emph{credit} lie close together and at a
        distance from \emph{equity}-group terms, consistent with the
        credit-versus-equity divide in fixed income.
  \item \emph{merger} and \emph{acquisition} are nearly synonymous in
        this corpus, mapping to almost identical embedding vectors.
  \item \emph{alpha}, \emph{beta}, and \emph{factor} cluster together,
        reflecting their shared origin in the asset-pricing literature
        and their frequent co-occurrence in portfolio management texts.
  \item The rates cluster (\emph{yield}, \emph{spread}, \emph{inflation},
        \emph{interest rate}, \emph{central bank}) is the most cohesive,
        owing to the highly formulaic language of fixed-income analysis.
\end{itemize}

\begin{remark}
\label{rem:pca-directions}
The principal component directions $\mathbf{p}_1, \mathbf{p}_2$
are not directly interpretable as named financial concepts.  They are
the linear combinations of the original $d = 300$ embedding dimensions
that explain the most variance \emph{among the 30 selected terms}.  A
different selection of terms would produce different axes.  The visual
clustering is informative, but one should resist assigning
domain-specific names to the $x$- and $y$-axes without rigorous
validation.  Rotation methods (e.g.\ Varimax) and supervised probing
classifiers offer more principled approaches to identifying
interpretable directions in embedding space; see \cite{tenney2019} % [CITE: Tenney et al. (2019). BERT Rediscovers the Classical NLP Pipeline. ACL]
for an overview in the transformer setting.
\end{remark}


% ---------------------------------------------------------------
\section{Looking Ahead}
\label{sec:intro-looking-ahead}

The representations developed in this chapter treat language as bags of tokens or
low-dimensional geometric objects.  They capture lexical similarity and rudimentary
document structure, but they cannot model syntax, long-range dependencies, or the
meaning of a word in context.  Chapter~\ref{ch:llm-foundations} addresses these
limitations directly.  It derives the recurrent architectures that first brought
sequential order into text modelling, shows why they fail at the scales demanded by
modern financial corpora, and then builds up the Transformer---the architecture
underlying every large language model in use today---from first principles.  It
continues with the practical layer: how to control generation through temperature
and structured decoding, how to ground model outputs in retrieved documents, how to
compress large models into deployable ones, and how to measure and mitigate the
hallucinations that remain the central reliability challenge for LLMs in finance.
