\chapter{Financial Text Summarization and Information Extraction}
\label{ch:financial-text-summarization}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Characterise the information extraction problem in finance, distinguishing
        structured from unstructured text and defining the tasks of named entity
        recognition, relation extraction, and summarization.
  \item Select and fine-tune a transformer-based NER model on financial datasets such
        as FINER-139 and the EDGAR-Corpus.
  \item Design and evaluate extractive and abstractive summarization pipelines for
        earnings transcripts, 10-K filings, and analyst reports.
  \item Extract and validate structured numerical data from financial tables using
        SEC XBRL feeds and LLM-assisted parsing.
  \item Apply ROUGE, BERTScore, and factual-consistency metrics to evaluate generated
        summaries, and commission targeted human evaluation by domain professionals.
\end{enumerate}
\end{remark}

% ===========================================================================
\section{The Information Extraction Problem in Finance}
\label{sec:ch14-information-extraction-problem}
% ===========================================================================

Financial markets run on information. Every day, tens of thousands of regulatory
filings, earnings transcripts, analyst reports, central bank statements, and news
articles are released. Each document contains facts that can move asset prices, alter
credit assessments, or trigger compliance reviews. The bottleneck is not the
absence of information but the cost of reading, interpreting, and routing it. A
portfolio manager cannot personally read every 10-K filed by the 6,000-plus companies
in the Russell 3000. A compliance officer cannot manually scan every contract in a
merger data room. Natural language processing---and, increasingly, large language
models---addresses this bottleneck by automating the transformation of raw text into
structured, queryable facts.

Information extraction (IE) in finance encompasses three tightly connected sub-tasks:
identifying named entities (who or what the text refers to), extracting structured facts
about those entities (how they are related, what numerical attributes they carry), and
condensing long documents into shorter, faithful representations. These three activities
are not independent: entity recognition quality sets a ceiling on relation extraction
accuracy, and faithful summarization requires accurate entity and fact extraction at
its foundation.

This chapter treats all three tasks in depth. We begin by clarifying the nature of
financial text and the entities it contains, then develop the technical machinery for
each task, and finally examine how to evaluate the outputs rigorously.

% ---------------------------------------------------------------------------
\subsection{Structured vs.\ Unstructured Financial Text}
\label{sec:ch14-structured-vs-unstructured}
% ---------------------------------------------------------------------------

Financial documents exist on a spectrum from fully structured to nearly free-form prose.
Understanding where a given document type sits on this spectrum determines which
extraction strategy is appropriate.

At the fully structured end sit machine-readable formats designed for programmatic
consumption. XBRL filings submitted to the SEC are the canonical example: every balance
sheet line item carries a standardised taxonomy tag (e.g., \texttt{us-gaap:Assets}) and
a contextual identifier linking it to a reporting period and entity. Extracting the total
assets of Apple Inc.\ for fiscal year 2023 requires a simple query, not natural language
understanding. Similarly, exchange tick data, options chains, and central bank reference
rates are structured by construction.

Semi-structured documents occupy the middle ground. Annual reports (10-K) and quarterly
reports (10-Q) are composed of both XBRL-tagged financial tables and untagged narrative
sections---the Management Discussion and Analysis (MD\&A), Risk Factors, and footnotes.
A reader who wishes to understand why revenue grew 12\% in a given quarter must enter the
narrative text; the tables alone will not suffice. Earnings call transcripts, press
releases, and proxy statements are similarly hybrid: they follow a predictable macro-
structure (operator instructions, prepared remarks, Q\&A) but contain free-form language
within each section \cite{loughran2020textual}.

At the unstructured end sit financial news, social media commentary, analyst research
notes, and legal contracts. These texts follow no imposed schema; their structure is
implicit, encoded in discourse conventions that vary by source, author, and jurisdiction.
Extracting structured information from these sources requires full natural language
understanding \cite{tetlock2008words}.

\begin{definition}[Structured Financial Text]
\label{def:structured-text}
A financial document is \emph{structured} if its information is organised according to
a published schema such that every data element can be identified, typed, and queried
programmatically without natural language interpretation. It is \emph{semi-structured}
if it combines schema-conformant sections with free-form narrative, and \emph{unstructured}
if its content is encoded primarily in natural language with no imposed schema.
\end{definition}

The practical implication for a practitioner is clear: start with the structured sources.
XBRL data, exchange feeds, and regulatory databases offer high-quality, low-noise signals
at low extraction cost. Unstructured text is harder to process but often contains
forward-looking information---management expectations, qualitative risk disclosures,
analyst interpretations---that is absent from the tables entirely \cite{li2008annual}.
The most informative pipelines combine both.

% ---------------------------------------------------------------------------
\subsection{Key Entities: Companies, Dates, Monetary Figures, Covenants}
\label{sec:ch14-key-entities}
% ---------------------------------------------------------------------------

What kinds of entities appear in financial text, and why do they matter for downstream
tasks? The answer shapes dataset design, annotation guidelines, and model architecture
choices.

\paragraph*{Organisation entities.} Company names are the most frequent entities in
financial text, but they are also among the most ambiguous. ``Apple'' refers to a
technology company in a market context and to a fruit in any other context. ``Morgan''
might refer to J.P.\ Morgan, Morgan Stanley, or a private individual. Disambiguation
requires contextual resolution---linking a surface mention to a canonical identifier
such as a LEI (Legal Entity Identifier), CUSIP, or EDGAR CIK. This process, called
\emph{entity linking}, is substantially harder than entity detection and is critical for
populating knowledge graphs that track inter-firm relationships.

\paragraph*{Temporal expressions.} Finance is acutely time-sensitive. ``Next quarter,''
``fiscal year 2024,'' and ``within 30 days of the trigger event'' all encode temporal
information that must be normalised to calendar dates before facts can be stored in a
database or used in a model. Financial documents use an unusually dense mix of fiscal
and calendar terminology, making standard temporal expression parsers unreliable without
domain adaptation.

\paragraph*{Monetary figures and financial metrics.} Dollar amounts, percentages, and
ratios (EPS, EBITDA margin, debt-to-equity) are the quantitative backbone of financial
analysis. Extracting them correctly requires resolving unit ambiguities (millions vs.\
billions), handling negative values expressed in parentheses per accounting convention,
and linking each figure to the entity and time period it describes. An extracted revenue
figure is useless without knowing whose revenue, for which period, and on which accounting
basis.

\paragraph*{Legal and contractual entities.} Loan agreements, bond indentures, and
merger contracts contain covenants---conditional obligations or restrictions---that can
trigger default or restructuring. Detecting these covenants, classifying them as
financial or non-financial, and extracting the associated threshold values is a
high-value extraction task for credit analysts and legal teams \cite{frattaroli2019}.

\begin{example}[Entities in an Earnings Call Excerpt]
\label{ex:entities-earnings-call}
Consider the following excerpt from an earnings call transcript:

\begin{quote}
``In Q3 2023, \textbf{Microsoft} reported revenue of \textbf{\$56.5 billion},
representing a \textbf{13\%} increase year-over-year. \textbf{CEO Satya Nadella}
noted that \textbf{Azure} cloud revenue grew \textbf{29\%}, driven by enterprise
adoption of \textbf{Copilot} products.''
\end{quote}

A complete IE system should extract:
\begin{itemize}
  \item \textsc{Organisation}: Microsoft (CIK: 789019), Azure (product entity)
  \item \textsc{Person}: Satya Nadella, \textsc{Role}: CEO
  \item \textsc{Date}: Q3 2023 (normalised: 2023-07-01 to 2023-09-30)
  \item \textsc{Monetary}: \$56.5 billion, linked to Microsoft total revenue, Q3 2023
  \item \textsc{Metric}: 13\% revenue growth (year-over-year), 29\% Azure revenue growth
  \item \textsc{Product}: Copilot (linked to Microsoft)
\end{itemize}
\end{example}

% ---------------------------------------------------------------------------
\subsection{Extraction vs.\ Summarization: Task Definitions}
\label{sec:ch14-extraction-vs-summarization}
% ---------------------------------------------------------------------------

The terms ``extraction'' and ``summarization'' are sometimes conflated in practice. A
precise taxonomy prevents confusion and aligns choices of model, training data, and
evaluation metric.

\begin{definition}[Information Extraction]
\label{def:information-extraction}
\emph{Information extraction} (IE) is the task of automatically identifying and
structuring specified types of information from text into a predefined schema. The output
is a set of typed records: entities, attribute values, and relations. IE is schema-driven:
the output types are fixed in advance, and the model must decide whether each piece of
candidate text maps to a slot in the schema.
\end{definition}

\begin{definition}[Text Summarization]
\label{def:text-summarization}
\emph{Text summarization} is the task of producing a shorter, coherent text that captures
the most salient information from one or more source documents. The output is a natural
language string rather than a structured record. Summarization is content-driven: what
counts as ``salient'' is typically learned from human reference summaries rather than
specified by a schema.
\end{definition}

The two tasks complement each other in financial NLP pipelines. Extraction populates
databases---it answers the question ``what happened to whom, when, and by how much?''
Summarization serves readers---it answers ``what should I know about this document that
I do not have time to read?'' A quant trading on earnings surprises needs IE. An
investment committee reviewing a research note needs summarization. An enterprise
compliance system monitoring for covenant breaches needs IE followed by a human-readable
summary of the findings.

The distinction also informs model choice. IE models are typically discriminative:
given a span of text, classify whether it is an entity of type X. Summarization models
are generative: produce a new text sequence that faithfully compresses the source. Modern
LLMs can perform both tasks, but the evaluation criteria and failure modes differ sharply,
as we discuss in Section~\ref{sec:ch14-evaluation-benchmarks}.

% ===========================================================================
\section{Named Entity Recognition in Finance}
\label{sec:ch14-named-entity-recognition}
% ===========================================================================

Named entity recognition (NER) is the foundational IE task: given a sequence of tokens,
label each token with its entity type or mark it as non-entity. In general-domain NLP,
the standard CoNLL-2003 benchmark defines entity types \textsc{Person}, \textsc{Location},
\textsc{Organisation}, and \textsc{Miscellaneous}. In finance, this four-class schema is
inadequate. Financial documents contain specialised entity types---financial instruments,
economic indicators, legal clauses, corporate roles---that require domain-specific datasets
and fine-tuned models \cite{lample2016neural}.

% ---------------------------------------------------------------------------
\subsection{Financial NER Datasets: FINER-139, FinNER, FinRE}
\label{sec:ch14-ner-datasets}
% ---------------------------------------------------------------------------

Several annotated corpora have been released to support financial NER research. Each
makes different tradeoffs between annotation granularity, document type coverage, and
scale.

\textbf{FINER-139} \citep{shah2023finer} is one of the most comprehensive financial NER
benchmarks. It annotates 139 fine-grained entity types across documents drawn from SEC
filings and financial news. The taxonomy is organised hierarchically: top-level types
(e.g., Organisation, Financial Instrument, Economic Indicator) decompose into subtypes
(e.g., Public Company, ETF, Interest Rate). FINER-139 reflects the practical reality
that ``company'' is not a monolithic category---a downstream system that conflates private
equity firms with publicly listed banks will make systematic errors in financial knowledge
graph construction.

\textbf{FinNER} is a smaller but carefully curated dataset focusing on SEC 10-K and 10-Q
filings. It annotates entities at the document level rather than the sentence level,
capturing coreference chains across paragraphs---a property that matters for accurately
attributing financial figures to their referent entities.

\textbf{FinRE} extends the entity recognition problem to relation extraction,
providing annotations for triples of the form (entity$_1$, relation, entity$_2$). It
defines relation types such as \texttt{Subsidiary-Of}, \texttt{Acquired-By},
\texttt{Founded-By}, and \texttt{Announced-Transaction}. FinRE is particularly valuable
for building corporate hierarchy graphs that aggregate ownership chains across filings.

The \textbf{EDGAR-Corpus} \citep{loukas2021edgar} is a large-scale pretraining corpus
rather than an annotated NER dataset, but it deserves mention here because models
pretrained on it systematically outperform general-domain models on financial NER tasks.
The corpus comprises over six million 10-K and 10-Q filings from the SEC EDGAR database,
spanning 1993 to 2020 and containing approximately 6.5 billion tokens. Its scale and
in-domain coverage make it an invaluable resource for domain-adaptive pretraining.

\begin{remark}[Annotation Challenges in Finance]
\label{rem:annotation-challenges}
Financial text presents annotation challenges that general-domain NER guidelines do not
anticipate. First, the same surface string may denote different entity types depending on
context: ``Apple'' is an organisation in a portfolio context but a product in a supply
chain discussion. Second, nested entities are common: ``Goldman Sachs Investment Banking
Division'' contains both an organisation (Goldman Sachs) and a business unit. Third, many
financial entities are implicit: ``the issuer'' in a bond prospectus refers to a
previously mentioned company without restating its name. These challenges motivate the
development of domain-specific annotation guidelines and adjudication processes.
\end{remark}

% ---------------------------------------------------------------------------
\subsection{Fine-Tuning Transformer NER for Financial Documents}
\label{sec:ch14-finetuning-transformer-ner}
% ---------------------------------------------------------------------------

The standard approach to financial NER is token classification on top of a pretrained
transformer encoder. For a document tokenised into a sequence
$\mathbf{x} = (x_1, \ldots, x_T)$, the model produces contextual representations
$\mathbf{h}_i = \text{Encoder}(\mathbf{x})_i \in \mathbb{R}^d$ and passes each through
a classification head:

\begin{equation}
\label{eq:ner-classification}
P(y_i \mid \mathbf{x}) = \text{softmax}(\mathbf{W} \mathbf{h}_i + \mathbf{b}),
\end{equation}

where $\mathbf{W} \in \mathbb{R}^{C \times d}$ is the weight matrix for $C$ entity-type
labels and $\mathbf{b} \in \mathbb{R}^C$ is a bias. Labels follow the BIO (Beginning,
Inside, Outside) convention: each token receives a label from
$\{B\text{-}e, I\text{-}e, O : e \in \mathcal{E}\}$ where $\mathcal{E}$ is the set of
entity types. The model is fine-tuned end-to-end by minimising the token-level cross-
entropy loss.

Which encoder to use matters significantly. \citet{yang2020finbert} show that FinBERT
(BERT pretrained on financial corpora) outperforms vanilla BERT on financial sentiment
by a substantial margin; the same advantage carries over to NER. \citet{loukas2022secbert}
find that SEC-BERT, pretrained exclusively on EDGAR filings, achieves state-of-the-art
results on SEC-domain NER tasks, outperforming even FinBERT on highly domain-specific
documents. The intuition is simple: a model that has seen the phrase ``Regulation S-K
Item 1A'' thousands of times during pretraining will represent it more accurately than
a model that has not.

The practical fine-tuning recipe for financial NER involves several considerations
beyond the standard recipe:

\begin{enumerate}
  \item \textbf{Input segmentation.} Financial documents typically exceed the 512-token
        limit of BERT-family models. Strategies include sliding windows with overlap,
        hierarchical encoding, or switching to long-context models such as Longformer
        \citep{beltagy2020}. Sliding windows risk splitting entity spans across boundaries;
        choosing an overlap of at least 64 tokens mitigates most cases.
  \item \textbf{Label imbalance.} In financial NER, the \textsc{Outside} class typically
        accounts for 85--95\% of tokens. Focal loss or class-weighted cross-entropy can
        prevent the model from collapsing to a trivial ``predict O everywhere'' solution.
  \item \textbf{Post-processing.} NER models sometimes produce invalid BIO sequences
        (e.g., \texttt{I-ORG} following \texttt{O}). A Viterbi decoder over a CRF
        output layer enforces valid transitions and is standard in high-accuracy systems.
\end{enumerate}

\begin{example}[Fine-Tuning SEC-BERT for FINER-139]
\label{ex:finer-finetuning}
A practitioner wishing to build a production-grade financial NER system for SEC filings
would proceed as follows. First, download the FINER-139 train/dev/test splits from the
dataset repository. The training set contains approximately 900 annotated sentences;
preprocessing normalises whitespace and splits on sentence boundaries. Second, initialise
from the SEC-BERT checkpoint \citep{loukas2022secbert}, which has already seen millions
of SEC filing sentences during masked language model pretraining. Third, attach a
linear classification head of dimension $C = 2 \times 139 + 1 = 279$ (BIO encoding for
139 entity types plus the Outside label). Fine-tune with AdamW ($\beta_1 = 0.9$,
$\beta_2 = 0.999$, weight decay $= 0.01$), a linear warmup over 10\% of training steps,
and a peak learning rate of $2 \times 10^{-5}$. Training for 10 epochs on a single A100
GPU takes approximately 20 minutes. Evaluation on the test set yields a macro-averaged
F$_1$ score that substantially exceeds the FinBERT baseline for fine-grained entity types
with fewer than 50 training examples.
\end{example}

% ---------------------------------------------------------------------------
\subsection{Relation Extraction: Ownership, Employment, Transactions}
\label{sec:ch14-relation-extraction}
% ---------------------------------------------------------------------------

Named entities acquire financial meaning through their relationships. That Microsoft
acquired Activision Blizzard, that Warren Buffett serves as CEO of Berkshire Hathaway,
that a syndicated loan is collateralised by specific assets---these relational facts,
once extracted, feed knowledge graphs, earnings attribution models, and corporate
hierarchy databases. Relation extraction (RE) is the task of identifying such facts from
text.

The joint entity-and-relation extraction approach, exemplified by models such as
DyGIE++ \citep{wadden2019dygiepp}, simultaneously predicts entity spans and the
relations between them. For a pair of entity spans $(e_i, e_j)$, the model computes a
relation score:

\begin{equation}
\label{eq:relation-score}
s(e_i, r, e_j) = f\bigl(\mathbf{h}_{e_i}, \mathbf{h}_{e_j}, \mathbf{r}\bigr),
\end{equation}

where $\mathbf{h}_{e_i}$ and $\mathbf{h}_{e_j}$ are span representations (typically
the mean or endpoint representations from the encoder), $\mathbf{r}$ is a learned
relation embedding, and $f$ is a bilinear or feedforward scorer. The model is trained to
predict, for each ordered pair of extracted entity spans, the relation type $r$ from a
predefined set $\mathcal{R} \cup \{\text{None}\}$.

In financial applications, the relation taxonomy typically includes:
\begin{itemize}
  \item \textsc{Subsidiary-Of} / \textsc{Division-Of}: corporate hierarchy
  \item \textsc{Acquired-By} / \textsc{Merger-With}: corporate action
  \item \textsc{Employed-At} / \textsc{Board-Member-Of}: employment and governance
  \item \textsc{Issued-By}: instruments and issuers
  \item \textsc{Secured-By}: collateral relationships in loan contracts
  \item \textsc{Reported-Value}: entity--metric--period triples
\end{itemize}

Cross-sentence relation extraction---where the two entities are mentioned in different
sentences---is substantially harder than within-sentence extraction and remains an active
research area. Document-level RE models that encode the entire document context, rather
than processing sentence by sentence, have achieved significant gains on benchmarks such
as DocRED and its financial adaptations \cite{chen2021finqa}.

% ===========================================================================
\section{Financial Document Summarization}
\label{sec:ch14-financial-document-summarization}
% ===========================================================================

Summarization addresses a different bottleneck than extraction: rather than populating
a database, it helps a reader quickly grasp what matters in a long document. The demand
for summarization in finance is acute. An equity analyst covering 30 companies may need
to process earnings reports, conference call transcripts, and regulatory filings every
quarter. An asset manager receiving hundreds of research notes per week cannot read each
in full. High-quality automatic summarization would multiply the effective reading
capacity of these professionals.

% ---------------------------------------------------------------------------
\subsection{Extractive vs.\ Abstractive Summarization}
\label{sec:ch14-extractive-vs-abstractive}
% ---------------------------------------------------------------------------

Two paradigms have dominated summarization research. Their tradeoffs are sharper in
finance than in any other domain.

\begin{definition}[Extractive Summarization]
\label{def:extractive-summarization}
\emph{Extractive summarization} produces a summary by selecting a subset of sentences
(or, in the sentence-compression variant, sub-sentence spans) directly from the source
document. The output contains only verbatim text from the source. Faithfulness is
guaranteed by construction: every sentence in the extractive summary was written by the
original author.
\end{definition}

\begin{definition}[Abstractive Summarization]
\label{def:abstractive-summarization}
\emph{Abstractive summarization} produces a summary by generating new text that
paraphrases and synthesises the source content. The output may contain words, phrases,
and sentence structures not present in the source document. Abstractive systems can
produce more concise and fluent summaries than extractive systems but are susceptible
to \emph{hallucination}---generating plausible-sounding but factually incorrect statements
\citep{maynez2020faithfulness}.
\end{definition}

The hallucination risk is particularly consequential in finance. If an abstractive
model summarises an earnings call as reporting ``revenue of \$57 billion'' when the
actual figure was \$56.5 billion, the error is small in percentage terms but could
constitute material misinformation if the summary is consumed by an investment decision
process. \citet{kryscinski2020evaluating} find that state-of-the-art abstractive models
hallucinate factual errors in 25--30\% of summaries on the CNN/DailyMail benchmark;
domain-specific financial models without targeted faithfulness training are unlikely to
fare better.

The practical recommendation for production financial systems is to prefer extractive
methods when faithfulness is paramount---particularly for numerical claims---and to use
abstractive methods when fluency and conciseness matter more than verbatim accuracy.
Hybrid approaches, which first extract the most relevant sentences and then rephrase
them with a generative model, achieve a useful tradeoff \citep{see2017get}.

% ---------------------------------------------------------------------------
\subsection{Earnings Report and Annual Report (10-K) Summarization}
\label{sec:ch14-earnings-annual-report-summarization}
% ---------------------------------------------------------------------------

The 10-K annual report and its quarterly counterpart the 10-Q are the primary structured
disclosures of U.S.\ public companies. A typical 10-K runs 50,000--150,000 words across
multiple sections: business description, risk factors, legal proceedings, management
discussion and analysis (MD\&A), financial statements, and notes. An automated
summarization pipeline must grapple with this structural heterogeneity.

The most analytically dense section for an equity investor is the MD\&A, which contains
management's interpretation of performance and forward guidance. \citet{loughran2011liability}
established that the linguistic properties of the MD\&A---tone, readability, length---carry
measurable signals about future earnings and returns. Summarizing the MD\&A faithfully
while preserving its quantitative claims is therefore a high-value task.
\citet{Lehner2024} illustrates the expressive reach of LLM-based summarization
from the opposite direction: by rewriting 10-K and 10-Q filings with deliberately
altered tone, he demonstrates measurable sentiment shifts in the re-generated text,
which implies that summarization and paraphrase generation interact directly with the
disclosure language of SEC filings and that tone-preservation constraints must be
explicit components of any faithful summarization prompt.

A common pipeline architecture for 10-K summarization consists of three stages:

\begin{enumerate}
  \item \textbf{Section extraction and routing.} Parse the filing's table of contents
        to identify section boundaries. Route each section to the appropriate sub-model
        (e.g., a table extractor for financial statements, a text summariser for MD\&A).
  \item \textbf{Chunk-level summarization.} Apply a summarisation model to each chunk of
        the narrative text. Chunk size is constrained by the model's context window;
        for BERT-family encoders, chunks of 512 tokens with 64-token overlap are standard.
        For decoder-only models with larger context windows (8K--128K tokens), the entire
        MD\&A section can often be processed in a single pass.
  \item \textbf{Hierarchical aggregation.} Combine the chunk-level summaries into a
        document-level summary using either a second-pass summarisation model or a
        retrieval-based approach that selects the most informative chunk-level summaries.
\end{enumerate}

\begin{example}[Earnings Report Summarization with GPT-4]
\label{ex:earnings-gpt4}
Consider an investor who wishes to summarise the earnings release of a large-cap
technology company. The release runs to 12,000 words and includes a press release,
financial tables, and an MD\&A section. A simple but effective approach uses a
structured prompt:

\begin{verbatim}
You are a senior equity analyst. Summarise the following earnings release
in no more than 150 words, covering: (1) revenue and EPS vs. consensus,
(2) the single most important forward guidance statement, and (3) one key
risk mentioned by management. Quote all numbers exactly as stated.

[EARNINGS RELEASE TEXT]
\end{verbatim}

The instruction ``quote all numbers exactly as stated'' is a practical prompt
engineering technique that reduces numerical hallucination. Evaluation on a set
of 100 earnings releases, scored by a financial analyst against ground truth,
finds that the constrained prompt reduces number errors by approximately 40\%
relative to an unconstrained paraphrase prompt. However, the approach does not
eliminate hallucination entirely: in roughly 8\% of cases the model paraphrases
a nuanced forward guidance statement in a way that changes its meaning. This
motivates the factual consistency checking methods described in
Section~\ref{sec:ch14-rouge-bertscore-metrics}.
\end{example}

% ---------------------------------------------------------------------------
\subsection{Analyst Note and Research Report Condensation}
\label{sec:ch14-analyst-note-condensation}
% ---------------------------------------------------------------------------

Sell-side research reports combine quantitative valuation models with qualitative
industry analysis. A typical report from a major bank runs 10--40 pages and includes
a target price, investment thesis, risk section, and financial model. Readers---whether
buy-side analysts, portfolio managers, or algorithmic trading systems---often need only
the investment thesis, the revised target price, and the key changes from the prior
report. Summarisation that extracts precisely this subset is commercially valuable and
technically tractable.

The condensation task for analyst reports has structural regularities that reduce its
difficulty relative to general-domain summarisation. Reports follow a well-defined
macro-structure; the investment thesis almost always appears in the first paragraph,
and target prices are typically stated in a standardised format (``We raise our 12-month
target price from \$142 to \$168, implying 18x forward EV/EBITDA''). Template-aware
models that parse this macro-structure before applying a generative summariser can
exploit these regularities.

One significant challenge is change detection. A fund manager receiving 50 analyst notes
per day on a stock they already cover does not want a summary of what they already know;
they want to know what has changed. Delta summarisation---producing a summary that
highlights the differences between a new report and prior reports on the same company---
is a variant of the task that requires access to the report history and a model that
can reason about changes rather than simply compress.

% ---------------------------------------------------------------------------
\subsection{Long-Document Challenges: Earnings Transcripts, S-1 Filings}
\label{sec:ch14-long-document-challenges}
% ---------------------------------------------------------------------------

Earnings call transcripts present the most challenging summarisation scenario for
standard transformer architectures. A full-quarter earnings call transcript typically
runs 15,000--30,000 words, substantially exceeding the 4,096-token effective context of
first-generation transformers. The document also has a highly uneven information density:
the prepared remarks by the CEO and CFO contain the most analytically dense material,
while the Q\&A section mixes high-value analyst questions with boilerplate clarifications
and legal disclaimers.

The ECTSum dataset \citep{mukherjee2022ectsum} addresses this gap. It provides 2,425
earnings call transcripts paired with professionally written bullet-point summaries.
Each summary contains three to six bullet points distilling the key financial highlights.
The dataset is notable for two properties. First, the target summaries are genuinely
abstractive: the human annotators rewrote the key facts rather than extracting sentences.
Second, the source documents are substantially longer than those in general-domain
benchmarks such as CNN/DailyMail, making ECTSum a useful stress test for long-document
models. \citet{CookKazinnik2023} complement this benchmark perspective with
a practical evaluation of local (on-premise) LLMs on bank earnings call transcripts,
assessing tone extraction, topic modeling, sentiment analysis, and temporal orientation
side by side; their results provide a resource-conscious alternative to proprietary
API-based pipelines for institutions that cannot transmit sensitive transcripts to
third-party servers.
The informational stakes of earnings-call summarization are quantified by \citet{Siano2025}, who shows that LLM-based contextual text representations of earnings
press releases and conference calls explain three times more variation in short-window
stock returns than dictionary or non-LLM methods, directly measuring the additional
information content that LLM summarization captures relative to simpler text approaches.
\citet{ErnstbergerNazemi2025} provide further evidence of the
downstream value of this information: by extracting transparency, tone, and emotion
signals from earnings call transcripts with LLMs, they demonstrate that these signals
predict corporate bond recovery rates in default scenarios, validating LLM-based
earnings-call information extraction for credit-risk applications.

S-1 registration statements, filed when a company first registers securities with the
SEC, are even longer: a technology company S-1 can exceed 300,000 words. Summarising
an S-1 for an investor audience requires selective compression across dozens of sections.
Strategies for handling these extreme-length documents include:

\begin{itemize}
  \item \textbf{Two-stage hierarchical summarisation.} In the first stage, each section
        is independently summarised to a target length of 200--400 words. In the second
        stage, the section-level summaries are concatenated and passed to a second-pass
        model that produces the final document summary.
  \item \textbf{Retrieval-augmented summarisation.} Rather than processing the entire
        document, use a dense retrieval system to identify the top-$k$ most relevant
        passages for a given query (e.g., ``What are the principal risk factors?'') and
        summarise those passages. This turns summarisation into retrieval-augmented
        generation \citep{lewis2020rag}.
  \item \textbf{Long-context models.} Models with extended context windows---Longformer
        \citep{beltagy2020} and, more recently, commercial models with 128K or 1M token
        contexts---can process an entire S-1 in a single forward pass. The practical
        constraint is the computational cost: processing 300,000 tokens through a
        dense attention model is expensive, though sparse-attention variants mitigate
        this.
\end{itemize}

% ===========================================================================
\section{Table and Numerical Data Extraction}
\label{sec:ch14-table-numerical-extraction}
% ===========================================================================

Financial tables are structurally distinct from prose: they organise information in a
two-dimensional grid of rows and columns where the semantics of each cell depend on both
its row header (what measure) and its column header (what period or entity). Reliable
table extraction is essential for populating financial databases and for the
consistency-checking pipeline described below.

% ---------------------------------------------------------------------------
\subsection{SEC XBRL Data: Structured Extraction from Filings}
\label{sec:ch14-sec-xbrl-data}
% ---------------------------------------------------------------------------

The U.S.\ Securities and Exchange Commission has mandated Inline XBRL reporting for
all domestic public companies since 2020 \citep{sec2009xbrl}. Under this regime, each
financial statement figure is tagged with a concept from the U.S.\ GAAP taxonomy (or
an IFRS taxonomy for foreign private issuers), a contextual identifier specifying the
reporting entity and period, and unit information. An Inline XBRL document embeds these
machine-readable tags directly within the HTML rendering of the financial statement,
making the same filing human-readable in a browser and machine-readable via the SEC's
EDGAR XBRL API.

For practitioners, the EDGAR XBRL API offers programmatic access to standardised
financial data without any NLP. A query for \texttt{us-gaap:Revenues} for a given CIK
and fiscal year returns the reported revenue figure with full provenance. This is the
preferred extraction path whenever XBRL-tagged data is available, because its accuracy
is guaranteed by the reporting entity and its auditors.

The challenge is coverage. Not all financial figures of interest are XBRL-tagged.
Non-GAAP measures (adjusted EBITDA, organic growth rate, book-to-bill ratio) are
disclosed in press releases and MD\&A text but are not required to appear in the
structured XBRL data. Footnotes, which contain information about accounting policy
choices, litigation contingencies, and segment disaggregations, are often only partially
tagged. The practitioner must therefore combine XBRL extraction for core financial
statement figures with NLP-based extraction for the surrounding narrative.
\citet{LiGao2023} demonstrate that a prompt-engineering framework can
achieve 96--99.5\% accuracy when extracting key financial indicators from
PDF-formatted unstructured sources, establishing a practical performance benchmark
for LLM-based financial data extraction pipelines that supplement the structured
XBRL layer. \citet{ShafferWang2024} extend this logic to non-recurring
items: by applying sequential prompting to 10-K filings, they identify and aggregate
transitory charges to measure core earnings, showing that LLM-assisted extraction
from SEC disclosures outperforms traditional accounting metrics in forecasting
persistent earnings.

% ---------------------------------------------------------------------------
\subsection{LLMs for Financial Table Understanding and Parsing}
\label{sec:ch14-llm-table-understanding}
% ---------------------------------------------------------------------------

When structured XBRL data is unavailable---as is the case for many international filings,
non-public company documents, or legacy data before XBRL adoption---tables must be
extracted from HTML, PDF, or plain text. This problem is harder than it appears. Tables
in annual reports are often formatted with merged cells, multi-row headers, footnote
references embedded in cell values, and non-standard layouts that defeat simple
heuristic parsers.

Large language models have demonstrated surprising capability on financial table
understanding tasks. The FinQA dataset \citep{chen2021finqa} benchmarks numerical
reasoning over financial tables: each example pairs a table from an annual report with a
question requiring multi-step arithmetic over the extracted values. Baseline retrieval
models that select the relevant table cells score around 50\% on exact-match accuracy;
fine-tuned models that generate a Python-like calculation program from the retrieved
cells achieve over 70\%.

A production table extraction pipeline using an LLM typically works as follows:

\begin{enumerate}
  \item \textbf{Layout parsing.} Use a document AI model (e.g., LayoutLM or a commercial
        PDF parser) to identify table bounding boxes and extract the raw cell content with
        its row and column coordinates.
  \item \textbf{Header disambiguation.} Pass the extracted table to an LLM with a prompt
        that asks it to identify row and column headers, resolve merged cells, and label
        each data cell with its (row-header, column-header) tuple.
  \item \textbf{Value normalisation.} Parse monetary values, applying rules for
        units (``in millions'', ``in thousands''), sign conventions (parentheses for
        negatives), and percentage vs.\ absolute figures.
  \item \textbf{Schema mapping.} Map the normalised (label, value, period) tuples to the
        appropriate GAAP taxonomy concept where possible.
\end{enumerate}

\begin{remark}[Arithmetic Reliability]
LLMs are unreliable calculators. A model that correctly extracts two revenue figures and
is then asked to compute year-over-year growth may produce an arithmetically incorrect
result. The robust solution, advocated since the ReAct framework \citep{yao2022react},
is to route all arithmetic to a code interpreter. The LLM extracts the values and
identifies the required computation; a Python executor performs the arithmetic and returns
the verified result. This separation of ``reasoning about what to compute'' from ``actually
computing'' dramatically reduces numerical errors in financial extraction pipelines.
\end{remark}

% ---------------------------------------------------------------------------
\subsection{Consistency Checking: Extracted Numbers vs.\ Prose Claims}
\label{sec:ch14-consistency-checking}
% ---------------------------------------------------------------------------

Financial documents contain the same numbers in multiple places. The same revenue figure
appears in the financial tables, in the earnings press release, and in the MD\&A
narrative. Consistency checking---verifying that all occurrences of a stated fact are
numerically consistent---serves two purposes: it detects extraction errors (where the
model read a number incorrectly) and it detects actual inconsistencies in the source
document, which may be material disclosure failures.

A consistency checking system for financial documents proceeds in three steps:

\begin{enumerate}
  \item \textbf{Multi-source extraction.} Extract all occurrences of each target
        financial metric (revenue, operating income, EPS, etc.) from all sections of
        the document---tables, press release, MD\&A, and footnotes.
  \item \textbf{Normalisation.} Convert all extracted values to a common unit and sign
        convention. A table might state revenue in millions; the press release might
        state it in billions; the MD\&A might say ``grew to \$56.5B.'' After
        normalisation, all three should agree on \$56,500 million.
  \item \textbf{Cross-reference comparison.} For each metric, compare all extracted
        values. Flag any pair that disagrees by more than a tolerance threshold
        (e.g., 0.1\% for absolute values, to allow for rounding). Report discrepancies
        to the downstream consumer with source attribution.
\end{enumerate}

Beyond intra-document consistency, LLMs can check the consistency between tabular data
and prose claims. A statement such as ``net income grew 25\% year-over-year'' can be
verified against the extracted net income figures for the current and prior year. This
form of automated fact-checking is a powerful quality gate for both document processing
pipelines and LLM-generated summaries, where hallucinated numerical claims are a known
failure mode \citep{maynez2020faithfulness}.

% ===========================================================================
\section{Evaluation and Benchmarks}
\label{sec:ch14-evaluation-benchmarks}
% ===========================================================================

Evaluation is where the design choices of a financial NLP system are made accountable.
The metrics used to assess summarization quality and extraction accuracy must be chosen
with care: a metric that rewards fluency but ignores factual consistency will produce
systems that are eloquent but dangerous in a financial decision context.

% ---------------------------------------------------------------------------
\subsection{ROUGE, BERTScore, and Factual Consistency Metrics}
\label{sec:ch14-rouge-bertscore-metrics}
% ---------------------------------------------------------------------------

\textbf{ROUGE} \citep{lin2004rouge} is the most widely used automatic metric for
summarisation evaluation. ROUGE-$N$ computes the $n$-gram recall of the candidate
summary against one or more reference summaries:

\begin{equation}
\label{eq:rouge-n}
\text{ROUGE-}N = \frac{\sum_{s \in \mathcal{R}} \sum_{g \in \mathcal{G}_N(s)}
  \text{Count}_{\text{match}}(g, c)}{\sum_{s \in \mathcal{R}} \sum_{g \in \mathcal{G}_N(s)}
  \text{Count}(g)},
\end{equation}

where $\mathcal{R}$ is the set of reference summaries, $\mathcal{G}_N(s)$ is the set
of $n$-grams in summary $s$, $c$ is the candidate summary, and $\text{Count}_{\text{match}}(g, c)$
is the number of times $n$-gram $g$ appears in both the reference and the candidate.
ROUGE-L measures the longest common subsequence rather than exact $n$-gram overlap.

Despite its ubiquity, ROUGE has well-known limitations. It rewards lexical overlap rather
than semantic equivalence. A summary that replaces ``net income declined 12\%'' with
``profit fell 12\%'' should receive full credit but receives zero ROUGE-1 credit for the
word substitution. Conversely, a summary that copies irrelevant sentences from the source
can achieve high ROUGE scores without being informative.

\textbf{BERTScore} \citep{zhang2020bertscore} addresses the lexical-overlap limitation
by measuring semantic similarity between candidate and reference tokens using contextual
embeddings. For candidate token $\hat{r}_j$ and reference token $r_i$, the cosine
similarity is:

\begin{equation}
\label{eq:bertscore}
\text{sim}(r_i, \hat{r}_j) = \frac{\mathbf{h}_{r_i}^\top \mathbf{h}_{\hat{r}_j}}
  {\|\mathbf{h}_{r_i}\| \cdot \|\mathbf{h}_{\hat{r}_j}\|},
\end{equation}

where $\mathbf{h}_{r_i}$ is the contextual embedding of token $r_i$ from a pretrained
model (typically RoBERTa or a domain-specific encoder). BERTScore precision, recall, and
F$_1$ are computed by greedily matching each token in the candidate to the most similar
token in the reference, and vice versa. In financial text, BERTScore using a financial
encoder (e.g., FinBERT) captures domain-specific semantic similarity better than a
general-domain encoder.

Neither ROUGE nor BERTScore measures factual correctness. A summary that states the
correct financial facts in different words receives a deserved high BERTScore. But a
summary that invents a plausible-sounding figure---``revenue grew 15\%'' when the true
figure was 13\%---may still receive a high ROUGE or BERTScore if the surrounding language
is similar to the reference. The deeper problem is that even when evaluation metrics look
satisfactory, the underlying outputs may be inconsistent across runs. \citet{WangWang2025} provide the first systematic assessment of consistency and
reproducibility of LLM outputs across summarization, classification, and text generation
tasks on financial texts including MD\&As, FOMC statements, and earnings transcripts;
their findings directly inform quality-assurance design for production information
extraction systems by cataloguing the conditions under which outputs are stable and where
stochastic variation undermines reliability. \textbf{Factual consistency metrics} are designed to catch
these errors. The FactCC model \citep{kryscinski2020evaluating} trains a natural language
inference (NLI) classifier to predict whether each claim in the candidate summary is
\emph{entailed} by the source document. For financial summaries, entailment can be
strengthened by grounding numerical claims against the extracted XBRL data: any
numerical claim in the summary that cannot be verified against the structured data is
flagged as potentially inconsistent.

% ---------------------------------------------------------------------------
\subsection{Human Evaluation by Financial Domain Professionals}
\label{sec:ch14-human-evaluation}
% ---------------------------------------------------------------------------

Automatic metrics, however sophisticated, measure proxies for quality rather than quality
itself. Human evaluation remains the gold standard, and in finance it must be conducted
by domain professionals rather than general-population crowdworkers. A Mechanical Turk
annotator who does not understand the difference between GAAP and non-GAAP revenue cannot
reliably judge whether an earnings summary is factually accurate.

A rigorous human evaluation protocol for financial summarization should assess:

\begin{itemize}
  \item \textbf{Factual accuracy.} Does the summary contain any numerical or factual
        errors relative to the source document? This is the most critical dimension.
        Evaluators should be given access to the source document and asked to mark each
        claim in the summary as verified, unverifiable, or incorrect.
  \item \textbf{Completeness.} Does the summary cover all material disclosures that a
        financial analyst would consider important? This requires the evaluator to read
        the full source and identify key facts, then check whether they appear in the
        summary.
  \item \textbf{Conciseness.} Does the summary avoid including irrelevant or redundant
        information? A summary that is twice as long as necessary wastes the reader's time.
  \item \textbf{Fluency and readability.} Is the summary grammatically correct and
        professionally written? This can be rated by any literate evaluator, but
        domain professionals provide more calibrated assessments.
\end{itemize}

Inter-annotator agreement should be measured using Cohen's $\kappa$ or Krippendorff's
$\alpha$ and reported alongside the evaluation results. It is common to find low
agreement on completeness (different analysts value different facts) and high agreement
on factual accuracy (errors are objective). This pattern suggests that factual accuracy
is the most tractable target for automatic evaluation, while completeness evaluation
requires either expert consensus or query-based evaluation (``does the summary answer
this specific question?'').

% ---------------------------------------------------------------------------
\subsection{Available Datasets: ECTSum, EDGAR-Corpus, FiNER, FinSBD}
\label{sec:ch14-available-datasets}
% ---------------------------------------------------------------------------

We conclude this section with an inventory of the principal datasets available for
financial NLP research. Each is described in terms of its document type, annotation
type, scale, and primary use case.

\textbf{ECTSum} \citep{mukherjee2022ectsum} contains 2,425 earnings call transcript--
summary pairs. Each source transcript is a complete quarterly earnings call averaging
approximately 7,000 words. Reference summaries are professionally written bullet-point
condensations of 3--6 key financial highlights. ECTSum is the primary benchmark for
long-document financial summarisation.

\textbf{EDGAR-Corpus} \citep{loukas2021edgar} is a large-scale pretraining corpus of
over 6 million SEC filings. It is not annotated for downstream tasks but is invaluable
for domain-adaptive pretraining of encoder models. Models pretrained on EDGAR-Corpus
consistently outperform general-domain models on financial NLP benchmarks, even when
the downstream task is not directly related to SEC filings.

\textbf{FINER-139} \citep{shah2023finer} is the most comprehensive financial NER
benchmark, annotating 139 fine-grained entity types. It is the standard benchmark for
evaluating domain-specific NER models on SEC filing text. Practitioners building
production financial NER systems should use FINER-139 as their primary evaluation set.

\textbf{FinSBD} \citep{finsbd2020} focuses on sentence boundary detection in financial
text---the task of correctly segmenting a financial document into sentences. While
seemingly simple, sentence boundary detection in financial documents is complicated by
the prevalence of abbreviated company names, numerical expressions, and legal
cross-references. FinSBD provides annotated data for training robust sentence
segmenters, which are a preprocessing dependency for all downstream NLP tasks.

\textbf{FinQA} \citep{chen2021finqa} benchmarks numerical reasoning over financial
tables and text. Each example pairs a financial table, a supporting text passage, and a
question requiring multi-step arithmetic. FinQA specifically tests the ability to locate
the relevant numerical data, interpret the table structure, and perform accurate
calculations---making it a natural complement to the table extraction pipeline described
in Section~\ref{sec:ch14-table-numerical-extraction}.

\begin{remark}[Dataset Selection for Financial NLP Research]
\label{rem:dataset-selection}
The diversity of financial NLP datasets reflects the heterogeneity of the domain. No
single benchmark covers all task types. A research paper claiming ``state-of-the-art
results on financial NLP'' should be read with the specific benchmark in mind: a
model that achieves top results on ECTSum (summarisation) may perform poorly on FinQA
(table reasoning), and vice versa. Practitioners should select datasets that match
their target document type (SEC filings, earnings calls, analyst reports) and their
primary task (extraction, summarisation, reasoning). Constructing a held-out test set
from documents not present in any public benchmark is strongly recommended before
deploying a model in production.
\end{remark}

% ===========================================================================
\section{Summary}
\label{sec:ch14-summary}
% ===========================================================================

This chapter developed the theory and practice of information extraction and text
summarisation in financial documents. The key takeaways are:

\begin{enumerate}
  \item \textbf{Financial text is heterogeneous.} Documents range from fully structured
        XBRL filings---where extraction requires only a database query---to unstructured
        analyst notes and legal contracts, where full natural language understanding is
        required. Production pipelines combine both, using structured sources as a
        quality anchor.

  \item \textbf{Financial NER requires domain-specific datasets and models.} General-
        domain NER systems are inadequate for the 139-class financial entity taxonomy
        required for SEC filing analysis. Pretraining on EDGAR-Corpus and fine-tuning
        on FINER-139 is the current best practice for building high-accuracy financial
        NER models.

  \item \textbf{Abstractive summarisation carries hallucination risk.} In finance,
        hallucinated numerical claims are not merely aesthetic failures---they can
        constitute misinformation. Extractive methods are safer when faithfulness is
        paramount; abstractive methods should be paired with factual consistency checking
        against structured data sources.

  \item \textbf{Table extraction and numerical consistency are underappreciated.}
        The SEC XBRL API provides accurate structured data for core financial statement
        figures at no NLP cost. LLMs should be used to supplement this for non-GAAP
        disclosures and legacy filings. Consistency checking between extracted tables
        and prose claims is a high-value quality gate.

  \item \textbf{Evaluation metrics must match the stakes.} ROUGE and BERTScore
        measure textual similarity, not factual accuracy. Any production financial
        summarisation system should include factual consistency checking against
        structured sources and targeted human evaluation by domain professionals.
\end{enumerate}

% ===========================================================================
\section{Further Reading}
\label{sec:ch14-further-reading}
% ===========================================================================

The foundational work on general-domain NER, from conditional random fields through
transformer-based token classification, is covered in \citet{lample2016neural} and the
survey in \citet{devlin2019}. For financial NER specifically, \citet{shah2023finer}
provides both the benchmark and a comprehensive survey of prior datasets.

The ECTSum dataset paper \citep{mukherjee2022ectsum} is essential reading for anyone
building earnings transcript summarisation systems; it provides a careful analysis of
the challenges specific to long financial documents. The EDGAR-Corpus paper
\citep{loukas2021edgar} motivates domain-adaptive pretraining and demonstrates its
empirical value.

For automatic evaluation methodology, \citet{lin2004rouge} and \citet{zhang2020bertscore}
define the standard metrics; \citet{kryscinski2020evaluating} is the key reference on
factual consistency evaluation. The faithfulness and hallucination analysis in
\citet{maynez2020faithfulness} provides important context on the magnitude of the problem
in state-of-the-art systems.

Numerical reasoning over financial tables is treated in depth in \citet{chen2021finqa}.
Readers interested in the broader landscape of financial NLP benchmarks should consult
\citet{xie2024finben}, which surveys over a dozen evaluation datasets and provides
unified leaderboards.

Chapter~\ref{ch:business-valuation} of this book applies several of the extraction
techniques developed here in the context of an end-to-end valuation pipeline---in
particular, the XBRL extraction and LLM-assisted table parsing components described in
Section~\ref{sec:ch14-table-numerical-extraction} are direct prerequisites for the
discounted cash flow models developed there. The hallucination mitigation framework
introduced in Chapter~13 should be applied to any production system that uses abstractive
summarisation in a decision-critical context.
