\documentclass{beamer}
\input{../_style/finance-beamer.tex}

\title{Lecture 5: LLMs for Business Valuation}
\subtitle{Large Language Models in Finance}
\author{Juan F.\ Imbet}
\institute{Paris Dauphine -- PSL University}
\date{}

\begin{document}

% ============================================================
% FRAME 1: Title
% ============================================================
\begin{frame}
\frametitle{}
\titlepage
\end{frame}

% ============================================================
% FRAME 2: Outline
% ============================================================
\begin{frame}
\frametitle{Outline}
\tableofcontents
\end{frame}

% ============================================================
\section{Business Valuation: Frameworks and LLM Opportunity}
% ============================================================

% ============================================================
% FRAME 3: What is business valuation?
% ============================================================
\begin{frame}
\frametitle{What Is Business Valuation?}

\begin{block}{Definition}
  Business valuation maps observable and projected financial variables --- cash flows,
  discount rates, growth rates --- into a scalar enterprise value $V \in \mathbb{R}_+$
  that a rational, fully informed market participant would assign to the claim.
\end{block}

\smallskip
\textbf{Professional contexts:}
{\small
\begin{itemize}\itemsep0pt
  \item \textbf{M\&A}: spread between buyer and seller estimates defines zone of possible agreement
  \item \textbf{IPOs}: investment banks produce valuation ranges to price new shares
  \item \textbf{Portfolio management}: model value vs.\ market price generates trade signals
  \item \textbf{Financial reporting}: IFRS 13 / ASC 820 Level 3 assets require documented models
\end{itemize}}
\end{frame}

% ============================================================
% FRAME 3b: Why valuation is forward-looking
% ============================================================
\begin{frame}
\frametitle{What Is Business Valuation? (cont.)}

\begin{bigpicture}
  Valuation is forward-looking. A balance sheet records historical cost;
  the market prices a claim on \emph{future} earnings.
  Translating current information into a credible forward estimate is where LLMs add the most value.
\end{bigpicture}
\end{frame}

% ============================================================
% FRAME 4: Three valuation frameworks
% ============================================================
\begin{frame}
\frametitle{Three Valuation Frameworks}

\textbf{Income approach} — DCF:
\[
  V = \sum_{t=1}^{n} \frac{\text{FCF}_t}{(1+\text{WACC})^t}
      + \underbrace{\frac{\text{FCF}_{n+1}/(\text{WACC}-g)}{(1+\text{WACC})^n}}_{\text{Terminal Value}}
\]
A 100-bps change in WACC or $g$ alters terminal value by \textbf{20--40\%}. Scenario analysis is indispensable.

\smallskip
\textbf{Market approach} — trading multiples (EV/EBITDA, P/E, P/Book) from comparable firms.
Challenge: finding genuinely comparable peers. \textit{LLMs and embeddings help here.}

\smallskip
\textbf{Asset approach} — net asset value:
\[
  V_{\text{NAV}} = \sum_i FV(\text{Asset}_i) - \sum_j FV(\text{Liability}_j)
\]
Best for asset-intensive firms or as a floor valuation cross-check.

\smallskip
\begin{block}{In practice}
Triangulate across all three; present a valuation \emph{range}, not a point estimate.
\end{block}
\end{frame}

% ============================================================
% FRAME 5: Where LLMs add value
% ============================================================
\begin{frame}
\frametitle{How LLMs Transform Each Valuation Step}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Step} & \textbf{Traditional approach} & \textbf{LLM contribution} \\
\midrule
Data collection & Manual EDGAR parsing & Automated extraction pipeline \\
FCF computation & Spreadsheet formulas & Code interpreter (error-free) \\
Forecasting & Analyst judgment & CoT prompting + evaluation \\
Scenario analysis & Manual scenarios & Narrative coherence checking \\
Comparable selection & SIC codes + manual & Embedding + LLM hybrid \\
Report generation & Analyst writing & LLM first-draft synthesis \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\begin{bigpicture}
\textbf{Key principle}: LLMs are most valuable where the task is
\emph{qualitative reasoning over text} and least reliable where the task
is \emph{numerical computation}. Route arithmetic to a code interpreter.
\end{bigpicture}
\end{frame}

% ============================================================
\section{Extracting Financial Data from SEC EDGAR}
% ============================================================

% ============================================================
% FRAME 6: EDGAR ecosystem
% ============================================================
\begin{frame}
\frametitle{The SEC EDGAR Ecosystem}

\textbf{Key filing types for valuation:}
\begin{itemize}
  \item \textbf{10-K}: annual audited financials + MD\&A. Primary source.
  \item \textbf{10-Q}: quarterly unaudited interim statements.
  \item \textbf{8-K}: material events within 4 business days (earnings, M\&A, CEO changes).
  \item \textbf{DEF 14A}: proxy statement --- executive comp, board composition.
\end{itemize}

\medskip
\textbf{Programmatic access:}
\begin{itemize}
  \item Submissions API: \texttt{data.sec.gov/submissions/CIK\{cik\}.json}
  \item Full-text search: \texttt{efts.sec.gov/LATEST/search-index?q=...}
  \item XBRL companyfacts: \texttt{data.sec.gov/api/xbrl/companyfacts/CIK\{cik\}.json}
\end{itemize}

\medskip
\begin{block}{XBRL: the first-pass extractor}
Since 2009, the SEC requires XBRL tagging. XBRL covers the majority of standard line items reliably.
LLM parsing fills the gap for non-standard layouts, merged cells, and extension elements.
\end{block}
\end{frame}

% ============================================================
% FRAME 7: LLM table extraction
% ============================================================
\begin{frame}
\frametitle{Parsing Financial Tables with LLMs}

\textbf{Why rule-based approaches fail:} merged cells, multi-level headers, footnote references
inside cells, non-standard layouts. Classical parsers break on the long tail.

\medskip
\begin{underhood}
\textbf{LLM extraction pattern:}
\begin{enumerate}
  \item Extract raw HTML of candidate table with BeautifulSoup
  \item Convert to compact plain-text (preserve headers, strip HTML tags)
  \item Send to LLM with structured extraction prompt requesting JSON schema
  \item Validate output against schema; retry on parse failure
\end{enumerate}
\end{underhood}
\end{frame}

% ============================================================
% FRAME 7b: LLM table extraction (continued)
% ============================================================
\begin{frame}
\frametitle{Parsing Financial Tables with LLMs (cont.)}

\textbf{Design principles:}
\begin{itemize}
  \item Request JSON directly --- reduces post-processing; enables automatic validation
  \item Add ``Respond only with valid JSON. No markdown code fences.'' if model lacks JSON mode
  \item Use XBRL as ground truth; flag conflicts between XBRL tags and LLM extractions
\end{itemize}

\medskip
\begin{alertblock}{Avoid look-ahead bias}
Always record the \emph{filing date}, not the fiscal year end date.
Using fiscal year end in a backtest can include information not yet available.
\end{alertblock}
\end{frame}

% ============================================================
\section{Computing Free Cash Flows}
% ============================================================

% ============================================================
% FRAME 8: FCFF and FCFE definitions
% ============================================================
\begin{frame}
\frametitle{Free Cash Flow: Two Canonical Definitions}

\begin{underhood}
\textbf{Free Cash Flow to Firm (FCFF):}
\[
  \text{FCFF}_t = \text{EBIT}_t(1-\tau) + \text{DA}_t - \text{CapEx}_t - \Delta\text{NWC}_t
\]

\textbf{Free Cash Flow to Equity (FCFE):}
\[
  \text{FCFE}_t = \text{NI}_t + \text{DA}_t - \text{CapEx}_t - \Delta\text{NWC}_t + \text{NB}_t
\]

\textbf{Relationship} (Modigliani--Miller with taxes):
\[
  \text{FCFE}_t = \text{FCFF}_t - \text{InterestExp}_t(1-\tau) + \text{NB}_t
\]
\end{underhood}

\smallskip
\textbf{Which to use?} When leverage is changing materially, FCFF is more robust:
WACC can be held constant while cost of equity must be re-levered per scenario.
(See appendix for the discount-rate/output mapping.)
\end{frame}

% ============================================================
% FRAME 9: Worked FCFF example
% ============================================================
\begin{frame}
\frametitle{Worked Example: FCFF Computation}

\textbf{Hypothetical software company (USD millions):}
\begin{center}
\begin{tabular}{lr}
\toprule
Revenue & 4{,}200 \\
EBIT & 840 \\
Depreciation \& Amortisation & 310 \\
Capital Expenditure & 180 \\
NWC (current year) & 620 \\
NWC (prior year) & 540 \\
Tax rate $\tau$ & 21\% \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\begin{align*}
  \Delta\text{NWC} &= 620 - 540 = 80 \\
  \text{FCFF} &= 840(1-0.21) + 310 - 180 - 80 \\
              &= 663.6 + 310 - 180 - 80 = \mathbf{713.6\text{ M}}
\end{align*}

With WACC = 9\%, 5-year explicit growth = 8\%, terminal growth = 3\%:
Enterprise value $\approx$ \textbf{\$15.2B} $\approx$ 3.6$\times$ trailing revenue.
\end{frame}

% ============================================================
% FRAME 10: Handling non-recurring items
% ============================================================
\begin{frame}
\frametitle{Normalising EBIT: LLM-Assisted Non-Recurring Items}

Three categories that distort FCFF if treated naively:
\begin{enumerate}
  \item \textbf{One-time items}: restructuring charges, litigation settlements, impairments.
        Remove from base EBIT used for forecasting.
  \item \textbf{Capitalised operating leases} (IFRS 16, ASC 842): mismatch between reported
        EBIT, depreciation, and cash flows; must be reconciled.
  \item \textbf{Acquired intangible amortisation}: from purchase accounting step-up;
        often excluded in ``cash EBIT'' calculations.
\end{enumerate}

\smallskip
\textbf{LLM-assisted normalisation:} prompt the model on the MD\&A to identify non-recurring items;
return JSON with description, amount, and confidence (High / Medium / Low).

\[
  \text{EBIT}^* = \text{EBIT} + \sum_{k \in \mathcal{K}} a_k
\]

\smallskip
\begin{block}{Human-in-the-loop design}
High confidence $\to$ auto-strip. Medium $\to$ flag for analyst review. Low $\to$ pass through.
The LLM accelerates identification; the analyst makes the final adjustment.
\end{block}
\end{frame}

% ============================================================
\section{Forecasting Cash Flows with LLMs}
% ============================================================

% ============================================================
% FRAME 11: Classical baselines
% ============================================================
\begin{frame}
\frametitle{Classical Forecasting Baselines}

\textbf{Random walk with drift:}
\[
  \text{FCF}_{t+1} = \mu + \text{FCF}_t + \varepsilon_{t+1}, \quad \varepsilon \sim \mathcal{N}(0,\sigma^2)
\]

\textbf{Mean-reverting model:}
\[
  \text{FCF}_{t+1} - \bar{\text{FCF}} = \phi(\text{FCF}_t - \bar{\text{FCF}}) + \varepsilon_{t+1}
\]

Empirical estimates: $\phi \approx 0.5$--$0.7$ for earnings --- substantial persistence, non-trivial mean reversion.

\medskip
\textbf{Key limitation of pure time-series models:}
They rely exclusively on historical values.
They ignore qualitative information in management guidance, industry trends, and macroeconomic conditions.

\medskip
\begin{block}{LLM opportunity}
LLMs are trained on financial disclosures, analyst reports, and economic news.
They can incorporate qualitative dimensions that time-series models miss.
The empirical question is whether this yields a measurable improvement in forecast accuracy.
\end{block}
\end{frame}

% ============================================================
% FRAME 12: Chain-of-thought forecasting
% ============================================================
\begin{frame}
\frametitle{Chain-of-Thought Cash Flow Forecasting}

Wei et al.\ (2022): CoT prompting elicits intermediate reasoning steps before the final answer.
For forecasting, CoT forces the model to articulate causal mechanisms --- making output auditable.

\medskip
\textbf{Four-stage CoT forecasting prompt:}
\begin{enumerate}
  \item \textbf{Macro context}: how do interest rates, GDP, and inflation affect revenue and margins?
  \item \textbf{Industry dynamics}: drivers and headwinds; competitive dynamics; regulatory changes.
  \item \textbf{Firm-specific}: competitive position; margin trajectory; CapEx plans from MD\&A.
  \item \textbf{Synthesis}: point estimate + upside (P90) + downside (P10) per year; 2-sentence justification.
\end{enumerate}

\medskip
\begin{alertblock}{Why separating reasoning from numbers matters}
If the reasoning says ``highly competitive market with stable pricing'' but the number says 20\% revenue growth,
an automated consistency-checker catches it before the number enters the DCF model.
\end{alertblock}
\end{frame}

% ============================================================
% FRAME 13: Forecast evaluation metrics
% ============================================================
\begin{frame}
\frametitle{Evaluating Forecast Quality}

\textbf{Mean Absolute Percentage Error (MAPE):}
\[
  \text{MAPE} = \frac{1}{n}\sum_{t=1}^{n} \frac{|y_t - \hat{y}_t|}{|y_t|} \times 100
\]

\textbf{Root Mean Squared Error (RMSE):}
\[
  \text{RMSE} = \sqrt{\frac{1}{n}\sum_{t=1}^{n}(y_t - \hat{y}_t)^2}
\]

\textbf{Directional Accuracy (DA):}
\[
  \text{DA} = \frac{1}{n}\sum_{t=1}^{n}
  \mathbf{1}\!\left[\mathrm{sign}(y_t - y_{t-1}) = \mathrm{sign}(\hat{y}_t - y_{t-1})\right]
\]

\smallskip
\begin{center}
{\footnotesize
\begin{tabular}{lll}
\toprule
Metric & Problem & When to use \\
\midrule
MAPE & Undefined when $y_t \approx 0$ & Stable, positive cash flows \\
RMSE & Scale-dependent & Within-company comparison \\
DA & Scale-invariant & Investment decision relevance \\
\bottomrule
\end{tabular}}
\end{center}
Always compare against random-walk baseline and analyst consensus.
\end{frame}

% ============================================================
\section{Tool-Augmented Agents for Financial Arithmetic}
% ============================================================

% ============================================================
% FRAME 14: Why LLMs can't do arithmetic
% ============================================================
\begin{frame}
\frametitle{Why LLMs Cannot Be Trusted for Arithmetic}

\begin{block}{Architectural mismatch}
  LLMs are next-token predictors. Arithmetic is a deterministic algorithm.
  Statistical pattern-matching approximates arithmetic for simple cases;
  reliability degrades rapidly with complexity.
\end{block}

\medskip
\textbf{For business valuation this is serious:}
\begin{itemize}
  \item A 5-year DCF involves $\approx 15$--20 chained arithmetic operations
  \item An incorrectly applied tax rate propagates through every calculation
  \item Compounding errors: a 5\% FCFF extraction error + 1\% terminal growth bias
        can together cause a $>10\%$ error in enterprise value
\end{itemize}

\medskip
\textbf{Solution: architectural, not prompting.}
Remove arithmetic from the LLM entirely. Route all computation to a deterministic executor.

\medskip
\begin{alertblock}{Rule}
No financial calculation --- however simple --- should be performed by in-context LLM reasoning.
\end{alertblock}
\end{frame}

% ============================================================
% FRAME 15: Code interpreter and ReAct for DCF
% ============================================================
\begin{frame}[fragile]
\frametitle{Code Interpreter Pattern for DCF (1/2)}

\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
# ReAct agent trace --- DCF computation

Thought 1: I need to compute FCFF. EBIT=840, DA=310, CapEx=180,
           delta_NWC=80, tau=0.21. Use python_executor.

Action 1:  python_executor(code="""
               ebit=840; da=310; capex=180; delta_nwc=80; tau=0.21
               fcff = ebit*(1-tau) + da - capex - delta_nwc
               print(f"FCFF = {fcff:.1f} M")
           """)

Observation 1: FCFF = 713.6 M
\end{lstlisting}

The LLM emits structured Thought/Action steps; arithmetic is delegated to
\texttt{python\_executor} rather than performed in-context.
\end{frame}

% ============================================================
% FRAME 15b: Code interpreter and ReAct for DCF (continued)
% ============================================================
\begin{frame}[fragile]
\frametitle{Code Interpreter Pattern for DCF (2/2)}

\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
Thought 2: Now compute the 5-year DCF with g=8%, g_term=3%, WACC=9%.

Action 2:  python_executor(code="""
               import numpy as np
               fcff0=713.6; g=0.08; g_t=0.03; wacc=0.09; n=5
               fcfs=[fcff0*(1+g)**t for t in range(1,n+1)]
               tv=fcfs[-1]*(1+g_t)/(wacc-g_t)
               ev=sum(f/(1+wacc)**t for t,f in enumerate(fcfs,1))+tv/(1+wacc)**n
               print(f"EV = {ev:.1f} M")
           """)

Observation 2: EV = 15224.0 M
\end{lstlisting}

Generated code is an auditable artefact: human reviewers inspect it to verify formula, inputs, units.
\end{frame}

% ============================================================
% FRAME 16: LLM failure modes in valuation
% ============================================================
\begin{frame}
\frametitle{Four LLM Failure Modes in Valuation}

\begin{enumerate}
  \item \textbf{Phantom tickers}: the LLM generates realistic-sounding but non-existent tickers.
        \textit{Mitigation}: validate every ticker against a securities master before use.

  \item \textbf{Invented ratios}: confabulated plausible-looking numbers not in any filing.
        \textit{Mitigation}: trace every figure to a specific EDGAR accession number.
        Replace ``trust but verify'' with \textbf{``do not trust until verified.''}

  \item \textbf{Misapplied formulas}: syntactically correct code implementing the wrong formula
        (FCFE when FCFF was requested; missing tax shield on interest).
        \textit{Mitigation}: automated formula-level testing against known benchmark cases.

  \item \textbf{Stale knowledge}: LLMs have a knowledge cutoff. Financial data, tax rates,
        and accounting standards change.
        \textit{Mitigation}: always use EDGAR-retrieved information in preference to parametric
        knowledge for specific company facts.
\end{enumerate}
\end{frame}

% ============================================================
\section{Scenario Analysis and Monte Carlo}
% ============================================================

% ============================================================
% FRAME 17: Bull/base/bear cases
% ============================================================
\begin{frame}
\frametitle{Scenario Construction: Bull, Base, and Bear}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Scenario} & \textbf{Percentile} & \textbf{Description} \\
\midrule
Bull & P90 & Favourable macro, strong execution, above-consensus growth \\
Base & P50 & Median expectation; most likely trajectory \\
Bear & P10 & Adverse macro, margin compression, or competitive deterioration \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\textbf{Internal consistency requirement:} every scenario must specify a complete,
internally consistent set: revenue growth, EBIT margins, CapEx intensity, NWC ratios,
and terminal growth rate. A scenario combining 25\% revenue CAGR and 1\% terminal growth
is internally contradictory.

\medskip
\begin{block}{LLM role in scenarios}
LLMs excel at \emph{narrative coherence checking}: given a proposed set of scenario inputs,
does the narrative justify the numbers? Do the numbers contradict the stated macro environment?
\textbf{Not} at computing the resulting valuations --- that goes to the code interpreter.
\end{block}
\end{frame}

% ============================================================
% FRAME 18: Monte Carlo DCF
% ============================================================
\begin{frame}
\frametitle{Monte Carlo DCF: Full Distribution of Enterprise Value}

Let $\text{WACC} \sim \mathcal{N}(\mu_W, \sigma_W^2)$ and $g \sim \mathcal{N}(\mu_g, \sigma_g^2)$
(truncated so $g < \text{WACC}$). For $M$ simulation paths:

\[
  V^{(m)} = \sum_{t=1}^{n}
  \frac{\text{FCF}_t^{(m)}}{(1+\text{WACC}^{(m)})^t}
  + \frac{\text{FCF}_{n+1}^{(m)}}{(\text{WACC}^{(m)}-g^{(m)})(1+\text{WACC}^{(m)})^n}
\]

\medskip
\textbf{Typical parameterisation:}
$\mu_W = 0.09$, $\sigma_W = 0.015$, $\mu_g = 0.03$, $\sigma_g = 0.01$.
With $M = 10{,}000$: runtime $< 1$ second.

\medskip
\textbf{Output}: mean, median, standard deviation, P10, P25, P75, P90 of EV distribution.

\medskip
\begin{block}{Why Monte Carlo is not optional}
A 100-bps error in terminal growth $\to$ \textbf{15--18\% error in enterprise value}
(see error propagation analysis). Point estimates are not just imprecise --- they are misleading.
\end{block}
\end{frame}

% ============================================================
\section{Comparable Selection}
% ============================================================

% ============================================================
% FRAME 19: Traditional pitfalls and LLM approaches
% ============================================================
\begin{frame}
\frametitle{Comparable Selection: Pitfalls and LLM Solutions}

\textbf{Traditional pitfalls:}
\begin{enumerate}
  \item \textbf{Code obsolescence}: SIC codes last revised in 1987.
        A cloud computing company and a 1990s software distributor share the same SIC.
  \item \textbf{Diversified conglomerates}: manufacturing SIC company deriving revenue from financial services.
  \item \textbf{Cyclical timing}: temporarily distressed firms depress the median multiple.
\end{enumerate}

\medskip
\textbf{Two LLM-based approaches:}

\medskip
\textbf{Prompt-based}: describe the target in natural language; request 8--12 comparable firms as JSON.
\textit{Problem}: hallucinated tickers occur frequently.
\textit{Mandatory}: validate every ticker against a securities master.
\end{frame}

% ============================================================
% FRAME 19b: Comparable selection (continued)
% ============================================================
\begin{frame}
\frametitle{Comparable Selection: Embedding-Based Approach}

\begin{underhood}
\textbf{Embedding-based}: embed the MD\&A of all companies in the universe; retrieve top-$k$ by cosine similarity.
\[
  \text{sim}(\mathbf{a}, \mathbf{b}) = \frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\|\,\|\mathbf{b}\|}
\]
Empirical result: embedding-based peers have lower IQR of EV/EBITDA and lower deviation from transaction prices vs.\ SIC-based peers.
\end{underhood}

\medskip
\textbf{Trade-off:} embeddings are grounded (every retrieved peer is a real firm) but may
surface textual lookalikes whose business model differs from the target. The next slide
combines both approaches to capture their complementary strengths.
\end{frame}

% ============================================================
% FRAME 20: Hybrid pipeline
% ============================================================
\begin{frame}
\frametitle{Hybrid Comparable Pipeline}

\textbf{Neither approach alone is sufficient:}
\begin{itemize}
  \item Prompt-based: broad world knowledge but hallucinates tickers
  \item Embedding-based: grounded but may surface textual lookalikes with different businesses
\end{itemize}

\medskip
\textbf{Two-stage hybrid:}
\begin{enumerate}
  \item \textbf{Candidate generation}: embedding index returns top 30 by cosine similarity.
  \item \textbf{LLM filtering}: model receives candidates + brief company descriptions;
        removes firms that are geographically incomparable, in different competitive positions,
        or subject to idiosyncratic events (pending acquisition, bankruptcy).
        Filtered set of 8--12 comparables proceeds to multiple analysis.
\end{enumerate}

\medskip
\begin{block}{Principle: confine the LLM to its strengths}
Stage 1: grounded retrieval (no hallucination risk --- candidates come from real company vectors).
Stage 2: qualitative business-model reasoning (LLM excels).
Stage 2 \emph{never} needs to recall whether a specific ticker exists --- that is already guaranteed by Stage 1.
\end{block}
\end{frame}

% ============================================================
\section{End-to-End Pipeline and Best Practices}
% ============================================================

% ============================================================
% FRAME 21: Error propagation
% ============================================================
\begin{frame}
\frametitle{Error Propagation in Multi-Step Pipelines}

Suppose extraction produces FCF with error $\varepsilon_1$ and forecasting introduces a terminal growth bias $\varepsilon_2$. First-order approximation:
\[
  \Delta V \approx \frac{\partial V}{\partial \text{FCF}} \varepsilon_1
                 + \frac{\partial V}{\partial g} \varepsilon_2
\]

For the Gordon Growth perpetuity:
\[
  \frac{\partial V}{\partial \text{FCF}_n} = \frac{1+g}{\text{WACC}-g}, \qquad
  \frac{\partial V}{\partial g} = \frac{\text{FCF}_n(1+\text{WACC})}{(\text{WACC}-g)^2}
\]

\medskip
\textbf{At WACC = 9\%, $g$ = 3\%:}
\begin{itemize}
  \item 1\% relative error in FCF $\to$ $\approx$ 1\% error in enterprise value
  \item \textbf{100-bps error in terminal growth $\to$ $\approx$ 15--18\% error in enterprise value}
\end{itemize}

\medskip
\begin{alertblock}{The dominant uncertainty source}
Terminal growth rate uncertainty \emph{dominates} enterprise value uncertainty.
Always run Monte Carlo simulation; always present percentile ranges.
\end{alertblock}
\end{frame}

% ============================================================
% FRAME 24: Best practices
% ============================================================
\begin{frame}
\frametitle{Best Practices for Production Deployment}

\begin{enumerate}
  \item \textbf{XBRL first}: verify extracted numbers against XBRL tags.
        Flag any conflict for human review. LLM extraction is a complement, not a sole source.

  \item \textbf{Code interpreter for all arithmetic}: no financial calculation should be
        performed by in-context LLM reasoning. Generated code is the deliverable.

  \item \textbf{Run Monte Carlo}: a point-estimate DCF is misleading.
        Communicate the P10--P90 range alongside the central estimate.

  \item \textbf{Mandatory hallucination guards}: validate every ticker against a securities register;
        trace every financial figure to a specific EDGAR accession number;
        suppress unverifiable claims from final output.

  \item \textbf{Human review before publication}: LLMs are a powerful analytical assistant,
        not an autonomous decision-maker.
        A qualified analyst must take responsibility for the figures.
\end{enumerate}

\medskip
\textit{Consistent with FSB (2023) guidance on AI in finance: human accountability,
model transparency, and robust back-testing are prerequisites.}
\end{frame}

% ============================================================
% FRAME 25: Summary
% ============================================================
\begin{frame}
\frametitle{Summary: The LLM-Assisted Valuation Stack}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Component} & \textbf{Key mechanism} \\
\midrule
Data extraction & EDGAR API + LLM table parsing + XBRL validation \\
FCF computation & Pydantic schema + code interpreter \\
Forecasting & CoT prompting; evaluate with MAPE / RMSE / DA \\
Scenarios & Bull/base/bear + Monte Carlo over WACC and $g$ \\
Comparable selection & Embedding index + LLM qualitative filter \\
Orchestration & Multi-agent with parallel specialist execution \\
Governance & HITL review; hallucination guards; audit trail \\
\bottomrule
\end{tabular}
\end{center}

\medskip
\begin{block}{Connection to Chapter 6}
Chapter 6 applies the same data extraction and forecasting tools to credit risk ---
where target variables are probability of default and loss given default, and regulatory
constraints on model transparency are stricter.
\end{block}
\end{frame}

% ============================================================
% APPENDIX
% ============================================================
\appendixstart{Worked Case Study and Production Orchestration}

% FRAME (appendix): Case study --- relocated from core deck
\begin{frame}[noframenumbering]{Case Study: Mid-Cap SaaS Company}

\textbf{Profile:} Revenue \$4.2B (+12\% YoY), EBIT margin 20\%, D\&A \$310M,
CapEx \$180M, net debt \$800M, 280M shares outstanding.

\medskip
\textbf{After normalisation:} EBIT = \$915M, FCFF = \$789M
(two non-recurring items: \$45M restructuring + \$30M acquired intangible amortisation).

\medskip
\begin{center}
\begin{tabular}{lrr}
\toprule
\textbf{Method} & \textbf{Enterprise Value} & \textbf{Equity/share} \\
\midrule
DCF (WACC 9\%, $g$ 3\%) & \$17.0B & \$57.9 \\
Comparables (EV/EBITDA 24.5$\times$) & \$29.9B & --- \\
\bottomrule
\end{tabular}
\end{center}

\medskip
DCF vs.\ comparables gap: sector trading at elevated growth multiples.
Both are informative; neither is definitive.

\medskip
\textbf{Monte Carlo} (10{,}000 draws): mean \$17.2B, P10--P90 range \$12.5B--\$24.1B.
The range quantifies the uncertainty a point estimate conceals.
\end{frame}

% FRAME (appendix): Agent orchestration --- relocated from core deck
\begin{frame}[noframenumbering]{Multi-Agent Orchestration}

\textbf{Why not a monolithic script?}
A failure in step 6 (terminal value) requires rerunning all prior steps.
A monolithic pipeline is also untestable in parts.

\medskip
\textbf{Orchestrator + specialists:}
\begin{itemize}
  \item \texttt{edgar\_agent}: retrieval and extraction
  \item \texttt{fcf\_agent}: computation and normalisation
  \item \texttt{forecast\_agent}: CoT projections
  \item \texttt{comps\_agent}: comparable selection
  \item \texttt{dcf\_agent}: final valuation and scenario synthesis
\end{itemize}

\medskip
\textbf{Parallelism}: comparable selection has no dependency on the FCF forecast ---
both can run concurrently, reducing end-to-end latency.

\medskip
\begin{block}{Production latency}
6 LLM calls $\times$ 1--5 seconds each + parallelisation $\approx$
\textbf{15--30 seconds end-to-end} per company valuation.
Cost: \$0.50--\$3.00 per company at 2024 frontier-model pricing.
\end{block}
\end{frame}

% ============================================================
% APPENDIX: FCFF vs FCFE discount-rate mapping
% ============================================================
\begin{frame}[noframenumbering]{FCFF vs.\ FCFE: Discount-Rate Mapping}

For reference, the two free-cash-flow definitions are discounted at different
rates and yield different value measures:

\medskip
\begin{center}
\begin{tabular}{lll}
\toprule
Metric & Discounted at & Yields \\
\midrule
FCFF & WACC & Enterprise value \\
FCFE & Cost of equity & Equity value directly \\
\bottomrule
\end{tabular}
\end{center}

\medskip
Discounting FCFF at WACC gives enterprise value; subtract net debt to recover
equity value. Discounting FCFE at the cost of equity yields equity value directly.
\end{frame}

\end{document}
