\documentclass{beamer}
\input{../_style/finance-beamer.tex}

\title{Domain-Specific Financial LLMs}
\subtitle{Lecture 08 --- When Finance Needs Its Own Models}
\author{Juan F. Imbet}
\institute{Paris Dauphine -- PSL University}
\date{\today}

\begin{document}

\begin{frame}\titlepage\end{frame}

\begin{frame}{Outline}\tableofcontents\end{frame}

% ============================================================
\section{Why Domain-Specific LLMs?}
% ============================================================

\begin{frame}{Learning Objectives}
By the end of this lecture you should be able to:
\begin{enumerate}
  \item Explain \textit{why} general-purpose LLMs underperform on financial text,
        and name the linguistic features that create the gap.
  \item Place any financial model in a taxonomy: encoder-only vs.\ decoder-only,
        from-scratch vs.\ domain-adaptive vs.\ fine-tune-only.
  \item Apply \textbf{domain-adaptive pre-training (DAPT)} and design a balanced
        corpus.
  \item Read FinQA / FLUE / FinBen results in terms of the task taxonomy.
  \item Weigh licensing, compute, governance, and accuracy when choosing a model
        for production.
\end{enumerate}
\end{frame}

\begin{frame}{The Central Tension}
\begin{bigpicture}
General models have \emph{breadth} --- hundreds of billions of web tokens give
them world knowledge about options, CDS, and Basel III.
Finance demands \emph{specificity} --- a model that confuses ``basis risk'' with
``basis points'', misreads an income-statement table, or reads ``liability'' with
its everyday connotation produces outputs that are not merely imprecise but
\textbf{dangerous} downstream.
\end{bigpicture}

\medskip
The resolution is not breadth \emph{vs.}\ depth but \emph{how to combine them}.
Domain-specific financial LLMs orient pre-training, fine-tuning, or both toward
the linguistic distribution of financial text.

\medskip
\textbf{First question:} what actually makes financial language different?
\end{frame}

\begin{frame}{Financial Language: A Distinct Register}
\begin{block}{Three structural features underrepresented in general corpora}
\begin{enumerate}
  \item \textbf{Lexical specificity.} Common words carry precise, sometimes
        \emph{reversed} meanings: ``liability'', ``short'', ``haircut''. The
        Loughran--McDonald dictionary catalogues $\sim$350 words whose polarity
        is neutral or inverted in SEC filings vs.\ general English.
  \item \textbf{Numeric density.} ``Revenue grew 12.3\% YoY to \$4.7B'' is fluent
        English \emph{and} an arithmetic claim. FinQA requires multi-step
        reasoning over tables from 10-Ks.
  \item \textbf{Tabular embedding.} Income statements, balance sheets, loan tapes
        encode meaning in 2-D grids; standard tokenisation flattens them and loses
        the row/column alignment.
\end{enumerate}
\end{block}
A model is \emph{financially literate} to the extent it resolves token meaning on
all three dimensions.
\end{frame}

\begin{frame}{Vocabulary Mismatch is a Tokenisation Problem}
\begin{underhood}
The mismatch operates at the \textbf{sub-word} level. A BPE tokeniser trained on
general text may split \texttt{EBITDA} into 4--5 sub-tokens, none of which carries
financial meaning --- the burden shifts onto the attention layers to reassemble it.
\end{underhood}

\medskip
\begin{itemize}
  \item Tokenisers trained on financial corpora represent abbreviations
        (\texttt{EBITDA}, \texttt{CUSIP}, \texttt{LIBOR}) as single tokens.
  \item BloombergGPT trained a \emph{custom} BPE tokeniser on its combined
        general+financial corpus precisely so domain tokens stay compact.
\end{itemize}

\medskip
\begin{exampleblock}{Why it matters}
Fewer tokens per financial term $\Rightarrow$ shorter sequences, lower compute,
and crucially a single embedding the model can specialise --- instead of a
fragile composition over fragments.
\end{exampleblock}
\end{frame}

\begin{frame}{How General Models Fail --- Four Concrete Modes}
\small
\begin{tabular}{p{0.32\textwidth}p{0.6\textwidth}}
\toprule
\textbf{Failure mode} & \textbf{Evidence} \\
\midrule
Sentiment polarity inversion &
  ``strong'', ``leverage'', ``unprecedented'' flip valence; general model fine-tuned
  on Financial PhraseBank trails FinBERT despite \emph{more} parameters. \\[3pt]
Numerical hallucination &
  GPT-class models fabricate EPS / figures that diverge from the source 10-K
  (Kang \& Liu, 2023). \\[3pt]
Table-parsing failure &
  On FinQA, GPT-3 (175B) zero-shot $<$60\%; fine-tuned specialists $>$80\%.
  Prompting alone does not close the gap. \\[3pt]
Regulatory ambiguity &
  ISDA / Basel / SEC comment letters are a legal sub-dialect; SEC-BERT beats
  general BERT on NER and clause classification. \\
\bottomrule
\end{tabular}

\medskip
\begin{alertblock}{Common root}
All four reflect the gap between the financial distribution and the web-text
distribution that dominates pre-training. The remedy: shift the model's prior
toward finance.
\end{alertblock}
\end{frame}

% ============================================================
\section{A Taxonomy of Financial LLMs}
% ============================================================

\begin{frame}{Two Axes of Classification}
\begin{bigpicture}
Every financial model is a point in a 2-D design space. Where you land dictates
compute cost, capability profile, and deployment constraints.
\end{bigpicture}

\medskip
\begin{columns}[T]
\column{0.5\textwidth}
\textbf{Axis 1 --- Architecture}
\begin{itemize}
  \item Encoder-only (bidirectional): classification, extraction
  \item Decoder-only (autoregressive): generation, multi-step reasoning
  \item Encoder--decoder (seq2seq): summarisation, table-to-text
\end{itemize}
\column{0.5\textwidth}
\textbf{Axis 2 --- Training strategy}
\begin{itemize}
  \item From scratch (BloombergGPT)
  \item Domain-adaptive continuation / DAPT (FinBERT)
  \item Fine-tune / instruction-tune only (FinMA, InvestLM)
\end{itemize}
\end{columns}

\medskip
\begin{block}{Definition (Financial LLM)}
A model is \emph{financial} if its pre-training corpus has a significant financial
fraction, \emph{or} it is fine-tuned on financial NLP tasks, \emph{or} its
tokeniser is extended for financial vocabulary.
\end{block}
\end{frame}

\begin{frame}{Encoder--Decoder: the T5 Family in Finance}
\begin{underhood}
A bidirectional \textbf{encoder} reads the full source; an autoregressive
\textbf{decoder} emits output token by token. The separate encoding pass lets the
model attend to a long source \emph{before} writing a single output token ---
reducing copy errors and improving factual consistency.
\end{underhood}

\medskip
Canonical model: \textbf{T5} / instruction-tuned \textbf{FLAN-T5}. Financial uses:
\begin{itemize}
  \item Earnings-call summarisation (hours of commentary $\to$ structured abstract)
  \item Abstractive 10-K summarisation (100 pages $\to$ one-page narrative)
  \item Financial table-to-text generation
\end{itemize}

\medskip
\textbf{When still preferred:} constrained-output, extraction-heavy summarisation,
even though large decoder-only models (FinMA) now also handle these at scale.
\end{frame}

\begin{frame}{Encoder Family: FinBERT, SEC-BERT, FinBERT-tone}
\small
\begin{tabular}{p{0.20\textwidth}p{0.34\textwidth}p{0.38\textwidth}}
\toprule
\textbf{Model} & \textbf{Base / corpus} & \textbf{Specialisation} \\
\midrule
FinBERT (Araci 2019) & BERT-base, $\sim$1.8B words: Reuters TRC2, Bloomberg news, PhraseBank & 3-class sentiment; large F1 gain over BERT-base on PhraseBank \\[3pt]
FinBERT-tone (Yang et al.\ 2020) & BERT-large, earnings calls + analyst reports + filings & LM tone: optimistic / pessimistic / litigious / uncertain / constraining \\[3pt]
SEC-BERT (Loukas 2022) & BERT-base, $\sim$3B tokens of EDGAR only (10-K/Q, 8-K, proxies) & NER + clause classification on regulatory text \\
\bottomrule
\end{tabular}

\medskip
\begin{block}{Why default to encoders for classification?}
110--340M params, open weights, well-understood fine-tuning. EDGAR is a coherent
legal sub-dialect --- a news-trained model does \emph{not} fully transfer to it.
\end{block}
\end{frame}

\begin{frame}{Worked Example: FinBERT-tone on an Earnings Call}
\begin{exampleblock}{Transcript excerpt}
``While we are \textbf{cautiously optimistic} about the trajectory of our core
business, we want to flag meaningful \textbf{headwinds} in the European segment,
driven by continued \textbf{softness} in consumer demand and heightened
\textbf{competitive pressure}.''
\end{exampleblock}

\medskip
\begin{columns}[T]
\column{0.48\textwidth}
\textbf{General sentiment model}\\
Labels \textcolor{forestgreen}{\textbf{positive}} --- ``optimistic'' is a strong
positive signal in product-review / social-media text.
\column{0.48\textwidth}
\textbf{FinBERT-tone}\\
Labels \textcolor{crimson}{\textbf{uncertain / pessimistic}} --- ``cautiously
optimistic'' is a hedge; ``headwinds'', ``softness'' are near-universally negative
in this register.
\end{columns}

\medskip
\begin{alertblock}{Downstream stakes}
This tone feeds analyst sentiment indices that drive systematic trading strategies
(Tetlock 2007). A label error here is a position error.
\end{alertblock}
\end{frame}

\begin{frame}{Decoder Family: BloombergGPT, FinGPT, FinMA, InvestLM}
\small
\begin{tabular}{p{0.18\textwidth}p{0.18\textwidth}p{0.56\textwidth}}
\toprule
\textbf{Model} & \textbf{Scale} & \textbf{Approach} \\
\midrule
BloombergGPT & 50B params, 708B tokens & From scratch; $\sim$50/50 finance (363B FinPile) / general (345B Pile, Wiki, books). Beats peers on financial NLP, matches GPT-NeoX-20B on general tasks. \\[3pt]
FinGPT & LoRA on LLaMA/ChatGLM/Falcon & Open-source, \emph{data-centric}; fine-tune on a single A100 with live news APIs. Democratisation over peak accuracy. \\[3pt]
FinMA (PIXIU) & 7B LLaMA & Instruction-tuned on 136k financial instruction--response pairs across 5 task types. Strong in-distribution; weaker on unseen formats. \\[3pt]
InvestLM & 65B LLaMA & Instruction-tuned on a \emph{small, curated} set (CFA Qs, research reports). Large base beats 7B models with fewer examples. \\
\bottomrule
\end{tabular}
\end{frame}

\begin{frame}{Decoder Family --- the Autoregressive Objective}
\begin{underhood}
Decoder LLMs are autoregressive: $\mathcal{L}_{\text{CLM}} = -\sum_t \log
p_\theta(w_t \mid w_{<t})$. The roughly 50/50 corpus split in BloombergGPT is
deliberate --- a finance-only model loses the world knowledge that makes an LLM
useful on finance tasks requiring general reasoning.
\end{underhood}

\medskip
\begin{block}{Reading the four models as a design spectrum}
\begin{itemize}
  \item \textbf{BloombergGPT} buys peak accuracy with from-scratch compute few can match.
  \item \textbf{FinGPT} trades peak accuracy for reproducibility on a single A100.
  \item \textbf{FinMA} is strong in-distribution but brittle on unseen formats.
  \item \textbf{InvestLM} shows a large base + small curated set can beat 7B models.
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{Encoder vs.\ Decoder --- a Practical Heuristic}
\begin{block}{Decision rule}
\begin{itemize}
  \item \textbf{Fixed label space} (pos/neg/neutral, default/no-default)?
        $\Rightarrow$ start with an \textbf{encoder} (FinBERT, SEC-BERT): compact,
        fast, calibrated.
  \item \textbf{Open-ended output} (draft a risk section, answer from a 10-K)?
        $\Rightarrow$ a \textbf{decoder} is necessary.
\end{itemize}
\end{block}

\medskip
\begin{itemize}
  \item Encoders: 110--340M, fast to fine-tune, well-understood.
  \item Decoders: an order of magnitude larger, more inference memory, harder to
        calibrate for structured prediction.
  \item \textbf{Hybrid:} a small decoder + a well-indexed document store (RAG) can
        substitute for a much larger model (Lecture on LLM agents).
\end{itemize}
\end{frame}

% ============================================================
\section{Pre-training Strategies for Finance}
% ============================================================

\begin{frame}{From Scratch or Continue? The Build Decision}
\begin{bigpicture}
When Bloomberg built BloombergGPT it faced the choice every team confronts: train
a new model on a mixed corpus, or start from a general model and \emph{continue}
pre-training on financial text?
\end{bigpicture}

\medskip
\begin{columns}[T]
\column{0.5\textwidth}
\textbf{From scratch}
\begin{itemize}
  \item Clean control of tokeniser, positional encoding, corpus weights
  \item Requires compute + data that \emph{very few} organisations possess
\end{itemize}
\column{0.5\textwidth}
\textbf{Continue (DAPT)}
\begin{itemize}
  \item Reuses learned representations of a general foundation model
  \item Extends toward the domain at a fraction of the cost
  \item The dominant strategy in practice
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}{Domain-Adaptive Pre-training (DAPT)}
\begin{underhood}
\textbf{Definition.} Given pre-trained parameters $\theta_0$ of model $M_0$ and an
unlabelled domain corpus $\mathcal{D}_{\text{domain}}$,
\[
  \theta^\star = \arg\min_\theta \, \mathcal{L}_{\text{PT}}(\theta;\,
  \mathcal{D}_{\text{domain}}),
\]
where $\mathcal{L}_{\text{PT}}$ is MLM (encoders) or CLM (decoders), the optimisation
is \emph{initialised at} $\theta_0$, and the learning rate is reduced ($\sim 10\times$)
to mitigate catastrophic forgetting. Then fine-tune $M^\star$ on labelled task data.
\end{underhood}

\medskip
\begin{itemize}
  \item Gururangan et al.\ (2020): consistent downstream gains across 4 domains;
        \emph{largest} when the domain is most dissimilar from pre-training.
  \item Finance is strongly out-of-distribution for web-text models $\Rightarrow$
        DAPT gains should be large.
  \item Budget: typically \textbf{1--10B} domain tokens vs.\ hundreds of billions
        for original pre-training.
\end{itemize}
\end{frame}

\begin{frame}{DAPT: How Much Data, and Two Refinements}
\textbf{Diminishing returns.} The largest gain is in the first $\sim$1B domain
tokens; smaller increments thereafter.
\begin{itemize}
  \item Encoders (110--340M): \textbf{1--5B} domain tokens usually suffice.
  \item Decoders (7--70B): need proportionally larger corpora.
\end{itemize}

\medskip
\begin{columns}[T]
\column{0.5\textwidth}
\begin{block}{TAPT (task-adaptive)}
Further pre-train on \emph{unlabelled} examples from the task distribution before
supervised fine-tuning. Consistent gains, but needs task-specific unlabelled data.
\end{block}
\column{0.5\textwidth}
\begin{alertblock}{Catastrophic forgetting}
Narrow-domain continuation can degrade general ability. Mitigate with a reduced LR,
a large enough corpus, and \textbf{replay} --- mixing a small fraction of general
data back in.
\end{alertblock}
\end{columns}

\medskip
\footnotesize Hirano (2024): Japanese financial LLMs confirm DAPT gains generalise
across languages and sub-domains.
\end{frame}

\begin{frame}{Corpus Composition: Sources for Financial Pre-training}
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lllll}
\toprule
\textbf{Source} & \textbf{Scale} & \textbf{Quality} & \textbf{Access} & \textbf{Coverage} \\
\midrule
Bloomberg News      & $\sim$300B & Very high & Proprietary & News, markets \\
SEC EDGAR           & $\sim$50B  & High      & Open        & Regulatory, corporate \\
Reuters TRC2        & $\sim$2B   & High      & Licensed    & Financial news \\
Financial PhraseBank & $\sim$5M  & High      & Open        & Sentiment-labelled \\
Seeking Alpha       & $\sim$10B  & Medium    & Open/API    & Retail analysis \\
Earnings transcripts & $\sim$5B  & High      & Commercial  & Corporate comms \\
SSRN / arXiv finance & $\sim$2B  & High      & Open        & Academic research \\
General web (Pile)  & $\sim$800B & Mixed     & Open        & Broad coverage \\
\bottomrule
\end{tabular}

\medskip
\begin{block}{The defining fact}
General web text ($\sim$800B) dwarfs even the largest \emph{financial} source.
This asymmetry is exactly \emph{why DAPT beats from-scratch}: cheaply shift a model
that already has the breadth, rather than rebuild it.
\end{block}
\end{frame}

\begin{frame}{Example: A Central-Bank Communication Model}
\textbf{Goal:} classify FOMC paragraphs as hawkish vs.\ dovish --- needs the precise
vocabulary of monetary policy (``accommodation'', ``balance-sheet normalisation'',
``neutral rate'').

\medskip
\begin{block}{A deliberately narrow corpus}
\begin{enumerate}
  \item FOMC statements + minutes 1994--present ($\sim$50M tokens) --- target
        distribution.
  \item Other central-bank publications: ECB, BoE, Beige Books ($\sim$200M).
  \item Monetary-policy academic papers, SSRN/NBER ($\sim$500M).
  \item Financial news mentioning policy keywords (1--2B).
\end{enumerate}
\end{block}

\medskip
This model would lose on general financial tasks --- but on monetary-policy
communication it \emph{beats} a general model fine-tuned on the same labels,
because pre-training aligns with both the vocabulary and the pragmatics of
central-bank language.
\end{frame}

% ============================================================
\section{Benchmark Comparisons}
% ============================================================

\begin{frame}{The Standard Financial NLP Benchmarks}
\small
\begin{tabular}{p{0.13\textwidth}p{0.46\textwidth}p{0.32\textwidth}}
\toprule
\textbf{Bench} & \textbf{What it tests} & \textbf{Headline numbers} \\
\midrule
FinQA & 8{,}281 QA pairs over 10-K/Q tables + prose; multi-step arithmetic. The hardest standard benchmark. & Fine-tuned specialists $\sim$68--72\% exec.\ acc.; zero-shot GPT-4 class $\sim$55--60\%. \\[4pt]
FLUE & 5-task NLU suite (GLUE analogue): sentiment, headline class., NER, FiQA QA, STS. & FinBERT $>$ general BERT/RoBERTa; biggest margins on sentiment + NER. \\[4pt]
FinBen & 36 datasets, 24 tasks, 5 capability categories incl.\ generation + risk mgmt. & Designed to test \emph{decoder} open-ended generation, not just classification. \\
\bottomrule
\end{tabular}

\medskip
\begin{block}{How to read the headline numbers}
Each benchmark stresses a different capability. FinQA isolates table + arithmetic
reasoning; FLUE isolates lexical NLU; FinBen pushes open-ended generation. A model
strong on one is \emph{not} automatically strong on the others.
\end{block}
\end{frame}

\begin{frame}{FinQA Scoring: Execution Accuracy}
\begin{underhood}
\textbf{Execution accuracy} (FinQA):
\[
  \mathrm{EA} = \frac{1}{|\mathcal{Q}|}\sum_{q\in\mathcal{Q}}
  \mathbf{1}\!\left[\,|\hat a_q - a_q| \le \epsilon\,\right],
\]
$\epsilon = 0$ for exact match, or a small relative threshold ($\sim$1\%) for rounding.
\end{underhood}

\medskip
\begin{block}{Why a tolerance at all?}
A model can execute the right reasoning program yet differ from the gold answer by a
rounding step (e.g.\ reporting a ratio to two decimals). A small relative $\epsilon$
credits correct \emph{reasoning} without rewarding spurious over-precision.
\end{block}
\end{frame}

\begin{frame}{When Do FinLLMs Win? When Do They Lose?}
\begin{columns}[T]
\column{0.5\textwidth}
\begin{block}{FinLLMs win}
Tasks tied to domain \emph{lexical semantics}:
\begin{itemize}
  \item PhraseBank sentiment --- FinBERT $\gg$ BERT/RoBERTa, even after equal
        fine-tuning.
  \item EDGAR NER --- SEC-BERT $\gg$ general BERT (entity surface forms differ).
\end{itemize}
\end{block}
\column{0.5\textwidth}
\begin{alertblock}{General models win / tie}
Tasks needing broad knowledge or deep reasoning:
\begin{itemize}
  \item Macro / regulatory-history / geopolitical questions.
  \item FinQA with chain-of-thought on a large model often matches a smaller
        specialist --- reasoning is a function of \emph{scale}.
\end{itemize}
\end{alertblock}
\end{columns}
\end{frame}

\begin{frame}{The Scaling Crossover}
\begin{block}{Domain advantage vs.\ model scale}
The domain advantage is roughly \emph{inversely proportional to model scale} at fixed
difficulty.
\begin{itemize}
  \item $\le$7B: DAPT gives large, consistent gains.
  \item $\ge$70B: gains persist only for lexically sensitive tasks (sentiment, tone,
        NER); they fade for long-document reasoning.
\end{itemize}
\end{block}

\medskip
\begin{exampleblock}{A countervailing result}
Rahimikia \& Drinkall: year-specific FinLLMs can still beat far larger general models
on return prediction --- once one controls for look-ahead bias.
\end{exampleblock}
\end{frame}

\begin{frame}{Two Caveats Every Benchmark Number Carries}
\begin{alertblock}{Data contamination}
A benchmark may already sit in a model's pre-training corpus. Several large models
are \emph{suspected} to have seen PhraseBank / FinQA train splits, inflating apparent
performance. Rigorous evaluation needs held-out test sets with \textbf{temporal
cutoffs} after the model's knowledge cutoff --- which most published benchmarks do
not enforce.
\end{alertblock}

\medskip
\begin{alertblock}{Hallucination is not solved by domain pre-training}
It can be \emph{worsened}: a model trained on financial news learns to generate
plausible-but-unverified statistics. Kang \& Liu (2023) find finance-trained models
\emph{more} likely to fabricate specific figures. Retrieval augmentation, constrained
generation, and factual verification matter just as much for FinLLMs as for general
models.
\end{alertblock}
\end{frame}

% ============================================================
\section{Practical Deployment Considerations}
% ============================================================

\begin{frame}{Deployment is Orthogonal to Benchmark Accuracy}
\begin{bigpicture}
A model that tops FinQA but cannot be \emph{licensed} commercially, needs a compute
budget you cannot sustain, or produces outputs you cannot \emph{trace} to source for
audit --- is not a deployable model.
\end{bigpicture}

\medskip
Three consequential dimensions:
\begin{enumerate}
  \item Licensing \& data governance
  \item Hosting strategy (cloud API vs.\ on-prem)
  \item Cost-vs-performance $\Rightarrow$ hybrid architectures
\end{enumerate}
\end{frame}

\begin{frame}{Licensing, Hosting, Governance}
\small
\begin{tabular}{ll}
\toprule
\textbf{Model} & \textbf{License} \\
\midrule
BloombergGPT & Proprietary, not released (terminal subscribers only) \\
FinBERT (Araci); FinBERT (Yang) & Apache 2.0 --- commercial use OK \\
FinGPT & MIT \\
InvestLM / FinMA & Research-only (LLaMA-1 / LLaMA-2 license structures) \\
\bottomrule
\end{tabular}

\medskip
\begin{block}{Data governance}
GDPR / GLBA / CCPA constrain training on customer data. Customer balances,
transactions, credit scores generally \emph{cannot} train a model for \emph{other}
customers without consent. \textbf{Synthetic data} is a common workaround --- at the
cost of distribution shift.
\end{block}

\medskip
\textbf{Hosting.} FinBERT (110--340M): single A10G, sub-50ms, fits in memory.
7--50B decoders: multi-GPU; \textbf{quantisation} (GPTQ, AWQ, QLoRA) cuts memory
2--4$\times$ to fit 13--30B on one server. 50B+ : cluster-scale, few institutions.
\end{frame}

\begin{frame}{Cost vs.\ Performance: the Effective-Cost Model}
\begin{underhood}
\textbf{Effective cost per correct output.} With per-query model cost $c_m$, human
review cost $c_h$, model correctness $p_m$, and routed fraction $\rho$:
\[
  C_{\text{eff}}(m,\rho) = \frac{c_m + \rho\,c_h}{p_m + \rho(1-p_m)},
\]
assuming human review recovers all errors. Minimise subject to
$p_m + \rho(1-p_m) \ge p^\star$.
\end{underhood}

\medskip
\begin{block}{Hybrid routing pattern}
A small, fast FinLLM handles the easy majority; low-confidence queries route to a
larger model or a human. Effective when difficulty is \textbf{skewed} --- a large
``easy'' head + a small ``hard'' tail.
\end{block}
\end{frame}

\begin{frame}{Cost vs.\ Performance: RAG as a Complement}
\begin{block}{Grounding instead of growing}
A compact 7--13B decoder grounded in a well-indexed EDGAR store can match a larger
ungrounded model on document tasks --- at lower cost and with \textbf{full
traceability} (an SR 11-7 / regulatory requirement).
\end{block}

\medskip
\begin{itemize}
  \item The retrieved passage pins generated figures to source text, cutting
        numerical hallucination.
  \item Traceability is not a nicety: model-risk regimes \emph{require} an auditable
        link from output back to evidence.
  \item This makes a small grounded model a regulatory-grade alternative to a large
        ungrounded one.
\end{itemize}
\end{frame}

\begin{frame}{Example: Cost-Optimal Regulatory-Filing Classifier}
\textbf{Task:} classify ``risk factor'' sections of 10-Ks into 15 categories for
2{,}000 companies/year; 3--10 paragraphs each ($\approx$20{,}000--60{,}000 paragraphs).

\medskip
\begin{tabular}{lll}
\toprule
\textbf{Option} & \textbf{Accuracy} & \textbf{Annual cost} \\
\midrule
70B general model via API (all paragraphs) & 93\% & \$200--\$600 \\
Fine-tuned FinBERT-base on a T4 GPU         & 88\% & $\sim$\$400 server \\
\textbf{Hybrid} (FinBERT + route 12\% low-conf.\ to 70B) & \textbf{92\%+} & $\sim$\textbf{\$250} \\
\bottomrule
\end{tabular}

\medskip
\begin{alertblock}{Lesson}
The hybrid dominates either model \emph{alone} on the accuracy--cost frontier. The
break-even point is highly sensitive to the \emph{confidence distribution}, which must
be measured on a held-out validation set --- not assumed.
\end{alertblock}
\end{frame}

% ============================================================
\section{Summary}
% ============================================================

\begin{frame}{Lecture 08 --- Key Takeaways}
\begin{enumerate}
  \item \textbf{Register.} Financial language = specialised vocabulary + numeric
        density + tabular structure $\Rightarrow$ a systematic general-model gap.
        The LM dictionary quantifies the inversion starkly.
  \item \textbf{Two families.} Encoders (FinBERT, SEC-BERT) for fixed-label
        classification; decoders (BloombergGPT, FinGPT, FinMA, InvestLM) for
        open-ended generation and reasoning.
  \item \textbf{DAPT} is the broadly applicable strategy: continue the PT objective
        on a financial corpus blended with enough general text.
  \item \textbf{Benchmarks.} FinLLMs win on lexically sensitive tasks; lose the edge
        at large scale on reasoning. Watch contamination and hallucination.
  \item \textbf{Deployment.} License audit, governance, hosting, and the
        accuracy--cost trade-off. \textbf{Hybrid routing} dominates single-model
        deployments.
\end{enumerate}
\end{frame}

% ============================================================
% APPENDIX
% ============================================================
\appendixstart{Extended Material: Architectures, Sources, Mitigations}

\begin{frame}[noframenumbering]{Appendix A --- Encoder--Decoder Internals}
Full encoder--decoder (T5) processing of a financial document:
\begin{enumerate}
  \item \textbf{Bidirectional encoder} attends over the entire 10-K source
        $\Rightarrow$ each source token sees full context.
  \item \textbf{Cross-attention} lets each decoder step query the encoded source.
  \item \textbf{Autoregressive decoder} emits the summary token by token.
\end{enumerate}

\medskip
\begin{block}{Why factual consistency improves}
Because the encoder consumes the \emph{whole} source before generation starts, the
decoder can copy figures and entities against a complete representation --- fewer
fabricated numbers in extraction-heavy summarisation than a pure decoder reading
left-to-right.
\end{block}
\end{frame}

\begin{frame}[noframenumbering]{Appendix B --- Replay \& Forgetting Detail}
Catastrophic forgetting: continued pre-training on a narrow corpus degrades general
language ability.

\medskip
\textbf{Three mitigations, increasing cost:}
\begin{enumerate}
  \item \textbf{Reduced learning rate} ($\sim 10\times$ below original PT): cheapest;
        limits how far weights drift.
  \item \textbf{Sufficiently large domain corpus} relative to total tokens processed:
        keeps the update well-conditioned.
  \item \textbf{Replay}: interleave a small fraction of original general-corpus data
        with the domain corpus --- modest extra cost, materially less forgetting.
\end{enumerate}

\medskip
TAPT can be stacked on top: DAPT (broad domain) $\to$ TAPT (task-distribution,
unlabelled) $\to$ supervised fine-tuning.
\end{frame}

\begin{frame}[noframenumbering]{Appendix C --- Secondary Corpus Notes}
\begin{itemize}
  \item \textbf{Bloomberg archive:} decades of professionally edited news ---
        consistent terminology, calibrated tone, high factual density. The gold
        standard, but proprietary (barrier for academia / open source).
  \item \textbf{SEC EDGAR:} free since 1996; full US public-company universe;
        continuously updated. Caveat: uneven quality (small-firm boilerplate) and
        XML structure needs careful parsing to split prose from tables/metadata.
  \item \textbf{News aggregators} (Reuters, AP, Seeking Alpha, subreddits): broader
        topical coverage, lower editorial quality. Reuters TRC2 is a standard source.
  \item \textbf{Analyst reports:} among the richest, most precise texts --- almost
        entirely proprietary; licensing at pre-training scale is cost-prohibitive.
\end{itemize}
\end{frame}

\begin{frame}[noframenumbering]{Appendix D --- Quantisation for On-Prem Decoders}
To fit a 13--30B decoder on a single high-memory GPU server:
\small
\begin{tabular}{lll}
\toprule
\textbf{Method} & \textbf{Idea} & \textbf{Effect} \\
\midrule
GPTQ  & Post-training weight quantisation (4-bit) & 2--4$\times$ memory reduction \\
AWQ   & Activation-aware weight quantisation & Preserves salient channels \\
QLoRA & 4-bit base + LoRA adapters in fine-tuning & Train + serve cheaply \\
\bottomrule
\end{tabular}

\medskip
All trade a modest accuracy cost for a large memory saving --- the enabling
technology for on-prem deployment of mid-scale decoders where data residency
forbids a cloud API.
\end{frame}

\begin{frame}[noframenumbering]{Appendix E --- Hallucination Mitigation Stack}
Domain pre-training does \emph{not} remove hallucination. Production stack:
\begin{enumerate}
  \item \textbf{RAG}: ground generation in retrieved, verified passages.
  \item \textbf{Constrained / structured generation}: schema-enforced outputs.
  \item \textbf{Self-consistency}: $K$ samples, majority vote (numerical reasoning).
  \item \textbf{Claim decomposition} + automated factual verification against source.
  \item \textbf{Confidence elicitation + abstention}: route low-confidence to humans.
\end{enumerate}

\medskip
\begin{block}{Governance hook}
SR 11-7 model-risk management and (forthcoming lecture) the EU AI Act / MiFID II
make \emph{traceability} a requirement --- which is itself an argument for
RAG-grounded architectures over ungrounded large models.
\end{block}
\end{frame}

\end{document}
