\chapter{LLMs for Business Valuation}
\label{ch:business-valuation}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Define business valuation and situate the three main valuation frameworks---income,
        market, and asset approaches---within a rigorous financial theory.
  \item Build an automated pipeline that retrieves, parses, and structures SEC EDGAR
        filings using LLM-assisted extraction.
  \item Derive FCFF and FCFE from raw financial statements and implement the computation
        programmatically from LLM-extracted data.
  \item Design and evaluate prompt-based cash-flow forecasting strategies using
        chain-of-thought reasoning, and benchmark them with MAPE, RMSE, and directional
        accuracy metrics.
  \item Explain why LLMs require external tools for reliable arithmetic and implement a
        ReAct agent that routes calculations to a code interpreter.
  \item Construct bull, base, and bear valuation scenarios and orchestrate Monte Carlo
        simulation over WACC and growth-rate distributions.
  \item Apply embedding-based and prompt-based methods to identify peer companies for
        comparable analysis, and combine them in a hybrid pipeline.
  \item Assemble an end-to-end valuation agent, estimate first-order error propagation,
        and apply best practices for production deployment.
\end{enumerate}
\end{remark}

% --- Section: Introduction to Business Valuation ---

\section{Introduction to Business Valuation}
\label{sec:bv-intro}

\begin{context}
Imagine a private equity firm that has identified a mid-cap software company as an
acquisition target. Before the partners can table a bid, they must answer a deceptively
simple question: what is this company worth? The answer, it turns out, depends on
assumptions about future cash flows that will not be realised for years, on discount rates
derived from noisy market data, and on comparisons with peer companies that are never
quite identical to the target. Every assumption is defensible; every assumption is
contestable. Business valuation is, at its core, the discipline of making these
assumptions explicit, consistent, and auditable.

\subsection{What Is Business Valuation?}
\label{subsec:bv-intro-definition}

\begin{definition}[Business Valuation]
\label{def:business-valuation}
\emph{Business valuation} is the process of determining the economic value of an ownership
interest in a firm or a specific set of its assets. Formally, it maps a set of
observable and projected financial variables---cash flows, discount rates, growth
rates---into a scalar value $V \in \mathbb{R}_{+}$ that represents the price a rational,
fully informed market participant would assign to the claim under consideration.
\end{definition}

Rigorous valuations arise across diverse professional contexts.
In \emph{mergers and acquisitions} both buyer and seller commission independent analyses
to anchor negotiating positions; the spread between their estimates defines the zone of
possible agreement. In \emph{initial public offerings}, investment banks produce
valuation ranges to price new shares, balancing the issuer's desire for a high price
against the underwriters' need to ensure a successful book-building. \emph{Portfolio
managers} at long/short equity funds maintain proprietary valuation models for each
holding; a widening gap between model value and market price generates a trade signal.
\emph{Financial reporting} mandates also drive demand. Under IFRS 13 and ASC 820,
firms must classify fair-value measurements at Level 1 (quoted market prices), Level 2
(observable inputs), or Level 3 (unobservable inputs). Level 3 assets require a
documented valuation model subject to auditor scrutiny \cite{iasb2011ifrs13}.
Additional contexts include \emph{litigation and dispute resolution}, \emph{tax
compliance} (transfer pricing, estate tax), and \emph{regulatory capital} assessments.

What unites these contexts is that valuation is always forward-looking. A firm's
balance sheet records historical cost; the market prices a claim on future earnings.
Translating current information into a credible forward-looking estimate is the
analytical core of the discipline---and it is precisely here that large language models
offer the most transformative capabilities.

\subsection{Key Valuation Methodologies}
\label{subsec:bv-intro-methodologies}

The valuation literature, synthesised in standard references such as \citet{damodaran2012investment}
and \citet{berk2020corporate}, organises approaches into three broad frameworks.

The \textbf{income approach} values a firm by the present value of the economic benefits
it will generate for its owners. The canonical implementation is the
\emph{discounted cash flow} (DCF) model. Let $\mathrm{FCF}_t$ denote the free cash
flow generated in period $t$, $n$ the explicit forecast horizon, and $\mathrm{WACC}$
the weighted-average cost of capital. The enterprise value is

\begin{equation}
\label{eq:dcf}
V = \sum_{t=1}^{n} \frac{\mathrm{FCF}_t}{(1+\mathrm{WACC})^t}
    + \frac{TV}{(1+\mathrm{WACC})^n},
\end{equation}

where $TV$ is the \emph{terminal value}, representing cash flows beyond the explicit
horizon. Under the Gordon Growth Model assumption that free cash flows grow at a
constant perpetual rate $g < \mathrm{WACC}$ after period $n$,

\begin{equation}
\label{eq:terminal-value}
TV = \frac{\mathrm{FCF}_{n+1}}{\mathrm{WACC} - g}.
\end{equation}

Equation~\eqref{eq:terminal-value} is sensitive to both $\mathrm{WACC}$ and $g$: a
100-basis-point change in either parameter typically alters the terminal value by
20--40\% for a firm with moderate growth, which is why scenario analysis, discussed
in Section~\ref{sec:bv-scenarios}, is indispensable.

The \textbf{market approach} values a firm by reference to the prices at which comparable
businesses have been bought and sold. The most common implementation uses \emph{trading
multiples}: ratios such as Enterprise Value / EBITDA, Price / Earnings, or Price /
Book that are observed for a peer group and then applied to the target's own financial
metrics. The appeal is simplicity and direct market grounding; the challenge is
identifying peers that are genuinely comparable in terms of risk, growth, and capital
structure---a task to which LLMs and embedding methods are well suited, as shown in
Section~\ref{sec:bv-comparables}.

The \textbf{asset approach} values a firm by the fair value of its individual assets
minus its liabilities. It is most applicable to asset-intensive firms (real estate
holding companies, liquidating entities) or as a floor valuation cross-check. Under the
net asset value approach:
\begin{equation}
\label{eq:nav}
V_{\mathrm{NAV}} = \sum_{i} FV(\mathrm{Asset}_i) - \sum_{j} FV(\mathrm{Liability}_j),
\end{equation}
where $FV(\cdot)$ denotes fair value, itself often determined by subsidiary DCF or
market-price lookups.

In practice, a rigorous analysis triangulates across all three approaches and presents
a valuation range rather than a point estimate. The practitioner's judgment about
which method to weight most heavily depends on the firm's stage of development,
the availability of comparable transactions, and the purpose of the valuation.

\subsection{Chapter Roadmap: How LLMs Transform Each Step}
\label{subsec:bv-intro-roadmap}

A standard DCF valuation requires roughly six analytical steps: (1) collect and clean
financial data from filings, (2) compute historical free cash flows, (3) forecast
future cash flows, (4) estimate the discount rate, (5) calculate terminal value,
and (6) stress-test through scenario analysis. Each step has traditionally required
either skilled human labour or specialised software. Large language models, particularly
when augmented with tools and organised into agents, can now automate or substantially
accelerate every one of these steps.

Section~\ref{sec:bv-extraction} addresses the data collection problem: SEC EDGAR
contains structured and unstructured filings for every public US company, and we build
a pipeline that uses LLMs to extract, parse, and normalise financial tables from raw
HTML documents. Section~\ref{sec:bv-cashflows} covers the computation of free cash
flows from extracted data and the handling of missing or non-recurring items with
LLM assistance. Section~\ref{sec:bv-forecasting} examines how LLMs can generate
cash-flow forecasts through chain-of-thought reasoning and how those forecasts compare
with classical time-series models. Section~\ref{sec:bv-tools} is devoted to tool
augmentation: because LLMs are unreliable arithmetic engines, we show how to route
all numerical computation through a code interpreter within a ReAct loop.
Section~\ref{sec:bv-scenarios} covers scenario construction and Monte Carlo simulation
orchestrated by an agent. Section~\ref{sec:bv-comparables} treats comparable selection,
combining embedding-based semantic search with LLM filtering.
Section~\ref{sec:bv-pipeline} assembles these components into a complete end-to-end
pipeline, illustrated with a case study. Finally, Section~\ref{sec:bv-efficiency}
evaluates accuracy and efficiency benchmarks and discusses deployment best practices.
\end{context}

% --- Section: Extracting Financial Data with LLMs and Agents ---

\section{Extracting Financial Data with LLMs and Agents}
\label{sec:bv-extraction}

\begin{context}
Every valuation begins with raw financial data — and for public US companies, that
data lives in the SEC EDGAR system. This section shows how to build an automated
pipeline that retrieves filings, parses complex HTML tables, and assembles a
clean, structured financial data model ready for downstream computation.

\subsection{The SEC EDGAR Ecosystem and Structured Filings}
\label{subsec:bv-extraction-edgar}

The US Securities and Exchange Commission requires every public company to submit
standardised periodic reports to the EDGAR (\emph{Electronic Data Gathering, Analysis,
and Retrieval}) system. For valuation purposes the most important filing types are:

\begin{itemize}
  \item \textbf{Form 10-K}: Annual report containing audited financial statements
        (income statement, balance sheet, cash flow statement), Management Discussion
        and Analysis (MD\&A), and detailed footnote disclosures. The 10-K is the
        canonical source for historical financial data and qualitative business
        assessment.
  \item \textbf{Form 10-Q}: Quarterly report with unaudited interim financial statements.
        Three 10-Qs are filed each fiscal year; together with the 10-K they provide
        quarterly granularity.
  \item \textbf{Form 8-K}: Current report filed within four business days of any
        \emph{material event}, such as earnings releases, merger agreements, CEO
        departures, or credit-rating changes.
  \item \textbf{Form DEF 14A}: Proxy statement filed ahead of the annual meeting,
        containing executive compensation tables, board composition, and related-party
        transactions---inputs to corporate governance adjustments in valuation.
\end{itemize}

EDGAR exposes two programmatic interfaces that are central to automated pipelines.
The \emph{submissions API} returns structured filing metadata for a given company
identified by its Central Index Key (CIK):

\begin{verbatim}
https://data.sec.gov/submissions/CIK{cik:010d}.json
\end{verbatim}

The response is a JSON object containing the company's name, ticker, SIC code, and a
paginated table of recent filings with their accession numbers, dates, and primary
document names. The \emph{full-text search} endpoint allows queries against the
content of all EDGAR filings:

\begin{verbatim}
https://efts.sec.gov/LATEST/search-index?q="goodwill impairment"
    &dateRange=custom&startdt=2022-01-01&enddt=2023-12-31&forms=10-K
\end{verbatim}

Since 2009, the SEC has required filers to \emph{tag} financial data using the XBRL
(eXtensible Business Reporting Language) standard \cite{sec2009xbrl}.
XBRL tags map each reported figure to a concept in the US-GAAP or IFRS taxonomy, making
data extraction in principle straightforward. In practice, however, filers exercise
discretion in choosing tags, use non-standard extension elements for firm-specific line
items, and sometimes apply the taxonomy inconsistently. Raw XBRL thus covers the
majority of line items reliably but misses a material fraction of the detail required for
a full valuation. LLMs fill this gap by parsing the residual HTML and plain-text content.

\subsection{Building an Automated EDGAR Extraction Pipeline}
\label{subsec:bv-extraction-pipeline}

The following code illustrates a minimal but complete pipeline for retrieving a 10-K
filing. It first resolves a ticker symbol to a CIK, then fetches the filing index to
locate the primary document, and finally downloads it for parsing.

\begin{lstlisting}[language=Python, caption={EDGAR Filing Retrieval Pipeline}, label={lst:edgar-pipeline}]
import requests
import json
from pathlib import Path

HEADERS = {"User-Agent": "Your Name your@email.com"}

def ticker_to_cik(ticker: str) -> str:
    """Look up the 10-digit CIK for a given ticker symbol."""
    url = "https://www.sec.gov/files/company_tickers.json"
    resp = requests.get(url, headers=HEADERS)
    resp.raise_for_status()
    data = resp.json()
    ticker_upper = ticker.upper()
    for entry in data.values():
        if entry["ticker"] == ticker_upper:
            return str(entry["cik_str"]).zfill(10)
    raise ValueError(f"Ticker {ticker} not found in EDGAR.")

def get_filings_metadata(cik: str) -> dict:
    """Retrieve submission history for a given CIK."""
    url = f"https://data.sec.gov/submissions/CIK{cik}.json"
    resp = requests.get(url, headers=HEADERS)
    resp.raise_for_status()
    return resp.json()

def get_latest_10k_accession(cik: str) -> str:
    """Return the accession number of the most recent 10-K filing."""
    meta = get_filings_metadata(cik)
    filings = meta["filings"]["recent"]
    for form, accession in zip(filings["form"], filings["accessionNumber"]):
        if form == "10-K":
            return accession.replace("-", "")
    raise ValueError("No 10-K found for this company.")

def download_10k(ticker: str, save_dir: Path = Path("filings")) -> Path:
    """Download the primary HTML document of the latest 10-K."""
    cik = ticker_to_cik(ticker)
    accession = get_latest_10k_accession(cik)
    # Build the filing index URL
    index_url = (
        f"https://www.sec.gov/Archives/edgar/data/{int(cik)}/"
        f"{accession}/{accession}-index.json"
    )
    resp = requests.get(index_url, headers=HEADERS)
    resp.raise_for_status()
    index_data = resp.json()
    # Find the primary document (htm/html)
    primary_doc = None
    for item in index_data.get("directory", {}).get("item", []):
        if item.get("type") == "10-K":
            primary_doc = item["name"]
            break
    if primary_doc is None:
        raise ValueError("Could not identify primary 10-K document.")
    doc_url = (
        f"https://www.sec.gov/Archives/edgar/data/{int(cik)}/"
        f"{accession}/{primary_doc}"
    )
    doc_resp = requests.get(doc_url, headers=HEADERS)
    doc_resp.raise_for_status()
    save_dir.mkdir(parents=True, exist_ok=True)
    out_path = save_dir / f"{ticker}_10K.html"
    out_path.write_bytes(doc_resp.content)
    return out_path
\end{lstlisting}

Several engineering considerations deserve attention. The SEC's EDGAR data services
require the \texttt{User-Agent} header to identify the requesting party; requests that
omit this header are throttled or blocked. Rate limiting is generous for automated
pipelines (approximately ten requests per second) but must be respected to avoid
IP-level restrictions. For large-scale workflows---hundreds of companies over multiple
filing dates---a queuing mechanism with exponential backoff is advisable. Accession
numbers follow the format \texttt{XXXXXXXXXX-YY-ZZZZZZ}, where the first ten digits
are the filer's CIK; stripping hyphens yields the directory path used in the
\texttt{/Archives/} URL tree.

\subsection{Parsing Financial Tables and Statements with LLMs}
\label{subsec:bv-extraction-parsing}

A raw 10-K document is an HTML file that can exceed 200 pages. Financial statements
occupy perhaps ten pages of that total but are embedded in a sea of legal boilerplate,
footnotes, and risk-factor disclosures. The practical task is to locate the income
statement, balance sheet, and cash flow statement, extract the numerical data within
each table, and return it in a structured format amenable to computation.

Classical rule-based approaches---searching for known header strings such as
``CONSOLIDATED STATEMENTS OF OPERATIONS'' and extracting table rows by HTML tag
parsing---work for companies with consistent formats but fail on the long tail of
non-standard layouts, merged cells, multi-level headers, and footnote references
embedded within cells. LLMs handle this variation gracefully because they understand
the \emph{semantics} of the content, not merely its structure.

The recommended pattern is as follows. First, extract the raw HTML of each candidate
table using a lightweight parser such as BeautifulSoup. Convert the table to a compact
plain-text representation---preserving column headers and indentation but stripping
HTML tags. Then send this representation to an LLM with a structured extraction prompt:

\begin{lstlisting}[language=Python, caption={LLM-Based Income Statement Extraction Prompt}, label={lst:llm-extract-prompt}]
EXTRACTION_PROMPT = """You are a financial data extraction assistant.
Below is the text of a financial table from a 10-K SEC filing.

Your task: extract the income statement data as a JSON object.
The JSON must have the following structure:
{
  "statement_type": "income_statement",
  "fiscal_year_end": "YYYY-MM-DD",
  "currency": "USD",
  "unit": "millions",
  "line_items": {
    "revenue": <float or null>,
    "cost_of_revenue": <float or null>,
    "gross_profit": <float or null>,
    "operating_expenses": <float or null>,
    "ebit": <float or null>,
    "interest_expense": <float or null>,
    "ebt": <float or null>,
    "income_tax_expense": <float or null>,
    "net_income": <float or null>,
    "depreciation_amortization": <float or null>,
    "ebitda": <float or null>
  }
}

Rules:
- Return ONLY the JSON object, no additional text.
- If a line item is not present in the table, set it to null.
- Expenses should be positive numbers (do not use the sign from the filing).
- Convert all figures to the same unit indicated in the table header.

TABLE TEXT:
{table_text}
"""
\end{lstlisting}

The structured output approach---requesting JSON directly rather than free-form
text---reduces post-processing effort and enables automatic validation against a schema.
Modern LLMs, particularly those supporting a \emph{JSON mode} or \emph{structured
outputs} feature, virtually eliminate format violations. Where a model does not support
forced JSON, adding the instruction ``Respond only with valid JSON. Do not include
markdown code fences'' and validating with a \texttt{json.loads} call with a retry on
failure achieves near-perfect reliability in practice.

\begin{remark}[XBRL as Ground Truth]
\label{rem:xbrl-groundtruth}
Whenever XBRL-tagged data is available for a line item, it should serve as the primary
source and the LLM extraction should serve as a cross-check or fallback. The SEC's
\texttt{companyfacts} API at \texttt{https://data.sec.gov/api/xbrl/companyfacts/CIK\{cik\}.json}
returns every tagged data point for a company across all filings in a single JSON
response. Using XBRL as the first-pass extractor and LLM parsing as the gap-filler
provides both coverage and reliability.
\end{remark}

\subsection{Structuring Extracted Data for Downstream Computation}
\label{subsec:bv-extraction-structuring}

Once individual statements are extracted, they must be assembled into a consistent
\emph{financial data model}. The minimal representation for DCF valuation requires
three to five years of historical data per statement, with a consistent mapping from
filing-specific line-item names to a canonical schema. A practical schema includes:
(1) a company identifier (ticker and CIK), (2) the fiscal period end date, (3) the
filing date (to avoid look-ahead bias in backtests), and (4) a dictionary of
standardised financial quantities with units and sources clearly annotated.

Normalisation issues that frequently arise include currency-unit mismatch (some filers
report in thousands, others in millions), fiscal-year offset (not all fiscal years end
December 31), and restatements (a refiled 10-K supersedes the original and may change
previously reported figures). The extracted data object should carry a
\texttt{restatement\_flag} field and the provenance of each figure (XBRL tag or LLM
extraction with model and date) to support auditing.
\end{context}

% --- Section: Computing Cash Flows ---

\section{Computing Cash Flows}
\label{sec:bv-cashflows}

\begin{deepdive}
This section derives the two canonical free-cash-flow measures from first principles,
proves their algebraic relationship, and shows how to implement each computation from
structured extracted data — giving you a formula-level foundation you can audit and
extend.

\subsection{Free Cash Flow: Definitions and Formulas}
\label{subsec:bv-cashflows-definitions}

Cash flow is the foundation of income-approach valuation, but the term is used
inconsistently in practice. We adopt the standard definitions from
\citet{damodaran2012investment}.

\begin{definition}[Free Cash Flow to Firm (FCFF)]
\label{def:fcff}
The \emph{free cash flow to the firm} in period $t$ is the after-tax operating cash flow
available to all capital providers (debt and equity) after funding net investment:
\begin{equation}
\label{eq:fcff}
\mathrm{FCFF}_t = \mathrm{EBIT}_t \times (1 - \tau)
                + \mathrm{DA}_t
                - \mathrm{CapEx}_t
                - \Delta\mathrm{NWC}_t,
\end{equation}
where $\tau$ is the marginal corporate tax rate, $\mathrm{DA}_t$ is depreciation
and amortisation, $\mathrm{CapEx}_t$ is capital expenditure (net of asset disposals),
and $\Delta\mathrm{NWC}_t = \mathrm{NWC}_t - \mathrm{NWC}_{t-1}$ is the change in
net working capital (current assets excluding cash minus current liabilities excluding
short-term debt).
\end{definition}

\begin{definition}[Free Cash Flow to Equity (FCFE)]
\label{def:fcfe}
The \emph{free cash flow to equity} is the residual cash flow available to common
shareholders after meeting all obligations, including debt service:
\begin{equation}
\label{eq:fcfe}
\mathrm{FCFE}_t = \mathrm{NI}_t + \mathrm{DA}_t
                - \mathrm{CapEx}_t
                - \Delta\mathrm{NWC}_t
                + \mathrm{NB}_t,
\end{equation}
where $\mathrm{NI}_t$ is net income and $\mathrm{NB}_t$ is net new borrowing (debt
issued minus debt repaid in period $t$).
\end{definition}

\begin{proposition}[Relationship Between FCFF and FCFE]
\label{prop:fcff-fcfe-relation}
Under the Modigliani-Miller framework with corporate taxes,
\begin{equation}
\label{eq:fcff-fcfe}
\mathrm{FCFE}_t = \mathrm{FCFF}_t - \mathrm{InterestExp}_t \times (1 - \tau) + \mathrm{NB}_t.
\end{equation}
\end{proposition}

\begin{proof}
Starting from~\eqref{eq:fcff}, note that $\mathrm{NI}_t = [\mathrm{EBIT}_t - \mathrm{InterestExp}_t] \times (1 - \tau)$, so
\[
\mathrm{FCFF}_t = \mathrm{NI}_t + \mathrm{InterestExp}_t\times(1-\tau) + \mathrm{DA}_t - \mathrm{CapEx}_t - \Delta\mathrm{NWC}_t.
\]
Substituting this expression into~\eqref{eq:fcfe} and rearranging yields~\eqref{eq:fcff-fcfe}.
\end{proof}

The choice between FCFF and FCFE depends on the valuation objective. FCFF is discounted
at WACC to yield enterprise value; FCFE is discounted at the cost of equity to yield
equity value directly. When leverage is stable, both approaches should---and in practice
approximately do---give the same equity value once net debt is subtracted from enterprise
value. When capital structure is changing materially, however, FCFF is more robust
because WACC can be held constant while FCFE requires adjustments to the cost of equity
as leverage changes \citep{damodaran2012investment}.

\begin{example}[FCFF Computation: Illustrative Technology Firm]
\label{ex:fcff-example}
Consider a hypothetical software company reporting the following figures for its most
recent fiscal year (all in USD millions):
\begin{center}
\begin{tabular}{lr}
\hline
Revenue & 4{,}200 \\
EBIT & 840 \\
Depreciation \& Amortisation & 310 \\
Capital Expenditure & 180 \\
Net Working Capital (current year) & 620 \\
Net Working Capital (prior year) & 540 \\
Tax Rate ($\tau$) & 21\% \\
\hline
\end{tabular}
\end{center}
Applying Definition~\ref{def:fcff}:
\begin{align*}
\Delta\mathrm{NWC} &= 620 - 540 = 80\text{ M},\\
\mathrm{FCFF} &= 840 \times (1 - 0.21) + 310 - 180 - 80\\
              &= 663.6 + 310 - 180 - 80 = 713.6\text{ M}.
\end{align*}
With a WACC of 9\%, a five-year explicit forecast period growing at 8\%, and a terminal
growth rate of 3\%, the enterprise value from~\eqref{eq:dcf}
and~\eqref{eq:terminal-value} works out to approximately \$15.2B---a multiple of
roughly $3.6\times$ trailing revenue, consistent with high-quality software-sector
medians for firms with strong near-term growth.
\end{example}

\subsection{Programmatic Computation from Extracted Data}
\label{subsec:bv-cashflows-computation}

Given a structured financial data object produced by the extraction pipeline
(Section~\ref{sec:bv-extraction}), computing historical FCFF is a straightforward
algebraic operation. The following function takes the standardised JSON output
and returns FCFF for each available period:

\begin{lstlisting}[language=Python, caption={FCFF Computation from Extracted Financial Data}, label={lst:fcff-computation}]
def compute_fcff(financials: dict, tax_rate: float | None = None) -> dict:
    """
    Compute FCFF from standardised extracted financials.

    Parameters
    ----------
    financials : dict
        Dictionary with keys: 'ebit', 'depreciation_amortization',
        'capex', 'nwc_current', 'nwc_prior', 'income_tax_expense',
        'ebt'.  All values are floats in the same currency/unit.
    tax_rate : float or None
        Explicit marginal tax rate. If None, inferred from
        income_tax_expense / ebt (effective rate).

    Returns
    -------
    dict with 'fcff', 'tax_rate_used', and component breakdown.
    """
    ebit = financials["ebit"]
    da   = financials["depreciation_amortization"]
    capex = financials["capex"]
    delta_nwc = financials["nwc_current"] - financials["nwc_prior"]

    if tax_rate is None:
        ebt = financials.get("ebt")
        tax_expense = financials.get("income_tax_expense")
        if ebt and ebt != 0 and tax_expense is not None:
            tax_rate = tax_expense / ebt
        else:
            raise ValueError(
                "Cannot infer tax rate; provide tax_rate explicitly."
            )

    nopat = ebit * (1 - tax_rate)
    fcff = nopat + da - capex - delta_nwc
    return {
        "fcff": fcff,
        "nopat": nopat,
        "da": da,
        "capex": capex,
        "delta_nwc": delta_nwc,
        "tax_rate_used": tax_rate,
    }
\end{lstlisting}

A production implementation applies this function across a time series of annual or
quarterly observations, producing a vector of historical FCFFs that seeds the
forecasting models of Section~\ref{sec:bv-forecasting}.

\subsection{Handling Missing Data and Normalisation with LLM Assistance}
\label{subsec:bv-cashflows-missing}

Financial statements frequently contain items that distort the FCFF calculation if
treated naively. Three categories merit particular attention.

First, \emph{one-time items}---restructuring charges, litigation settlements,
impairment write-downs---inflate or deflate EBIT in the reporting period but do not
reflect the firm's recurring earning power. Standard practice is to remove them from
the base EBIT used for forecasting. Second, \emph{capitalised operating leases} (now
brought onto the balance sheet under IFRS 16 and ASC 842) create a mismatch between
reported EBIT, depreciation, and cash flows that must be carefully reconciled.
Third, \emph{acquired intangible amortisation} from purchase accounting (step-up of
intangible assets in an acquisition) is often excluded by analysts calculating
``adjusted EBIT'' or ``cash EBIT'' \cite{koller2020valuation}.

LLMs are well suited to flagging these issues from the MD\&A and footnote text.
A simple prompt such as the following can instruct an LLM to identify non-recurring
items that should be stripped from operating income:

\begin{lstlisting}[language=Python, caption={Prompt for Non-Recurring Item Identification}, label={lst:nonrecurring-prompt}]
NORMALISATION_PROMPT = """You are a senior financial analyst reviewing a 10-K filing.
Read the following Management Discussion and Analysis excerpt and identify
any non-recurring, one-time, or extraordinary items that affected operating income
this year but are unlikely to recur.

For each item, provide:
1. Description of the item
2. Amount (in USD millions, sign-corrected so positive = expense)
3. Line item affected (e.g., EBIT, gross profit)
4. Your confidence that it is non-recurring (High / Medium / Low)

Format the response as a JSON array of objects with keys:
"description", "amount", "line_item", "confidence".

MD&A TEXT:
{mdna_text}
"""
\end{lstlisting}

The LLM output is then reviewed programmatically: high-confidence items are
automatically stripped from EBIT, medium-confidence items are flagged for analyst
review, and low-confidence items are passed through unchanged. This human-in-the-loop
design is important: the LLM accelerates identification but does not make the final
adjustment autonomously. The normalised EBIT is:
\begin{equation}
\label{eq:ebit-normalised}
\mathrm{EBIT}^* = \mathrm{EBIT} + \sum_{k \in \mathcal{K}} a_k,
\end{equation}
where $\mathcal{K}$ is the set of identified non-recurring expense items and $a_k > 0$
is the pre-tax amount of item $k$ (positive because we are adding back an expense).
\end{deepdive}

% --- Section: Forecasting Cash Flows with LLMs ---

\section{Forecasting Cash Flows with LLMs}
\label{sec:bv-forecasting}

\begin{deepdive}
By the end of this section you will be able to implement chain-of-thought forecasting
prompts, select between fine-tuning and few-shot strategies, and evaluate the resulting
forecasts with MAPE, RMSE, and directional accuracy against a classical baseline.

\subsection{Time-Series Perspectives on Financial Forecasting}
\label{subsec:bv-forecasting-timeseries}

Before examining LLM-based approaches, it is useful to establish the classical
statistical baseline. Financial cash-flow forecasting has been studied extensively in
the accounting and finance literature, with a lineage that includes the random-walk
model \cite{ball1972earnings},
mean-reverting models of earnings \cite{fama2000forecasting},
and the use of analyst consensus estimates \cite{lewellen2010empirical}.

The simplest baseline is the \emph{random walk with drift}:
\begin{equation}
\label{eq:rw-fcf}
\mathrm{FCF}_{t+1} = \mu + \mathrm{FCF}_t + \varepsilon_{t+1},
\quad \varepsilon_{t+1} \sim \mathcal{N}(0, \sigma^2),
\end{equation}
which asserts that the best forecast of next year's FCF is this year's FCF plus a
long-run drift $\mu$. More elaborate specifications allow for mean reversion:
\begin{equation}
\label{eq:mean-reversion}
\mathrm{FCF}_{t+1} - \bar{\mathrm{FCF}} = \phi (\mathrm{FCF}_t - \bar{\mathrm{FCF}}) + \varepsilon_{t+1},
\end{equation}
where $|\phi| < 1$ is the speed of mean reversion toward the long-run level $\bar{\mathrm{FCF}}$.
Empirical estimates of $\phi$ for earnings are typically in the range 0.5--0.7,
indicating substantial persistence but non-trivial mean reversion \cite{fama2000forecasting}.

The key limitation of pure time-series models is their exclusive reliance on
historical values of the series: they ignore the qualitative information in
management guidance, industry trends, and macroeconomic conditions that financial
analysts routinely incorporate. LLMs, trained on text corpora that include financial
disclosures, analyst reports, and economic news, can incorporate this qualitative
dimension. The empirical question---examined in detail in the evaluation subsection
below---is whether this additional information yields a measurable improvement in
forecast accuracy.

\subsection{Prompt-Based Forecasting: Chain-of-Thought Approaches}
\label{subsec:bv-forecasting-cot}

\citet{wei2022chain} introduced \emph{chain-of-thought} (CoT) prompting, which
elicits intermediate reasoning steps before the final answer. For financial
forecasting, CoT is particularly valuable because it forces the model to articulate
the causal mechanisms driving its projection, making the output auditable and reducing
the risk of pattern-matching confabulation.

A well-designed CoT forecasting prompt structures the reasoning in four stages:
(1) characterise the macro environment, (2) assess industry dynamics, (3) evaluate
firm-specific factors, and (4) synthesise into a numerical projection.

\begin{lstlisting}[language=Python, caption={Chain-of-Thought Cash Flow Forecasting Prompt}, label={lst:cot-forecast-prompt}]
COT_FORECAST_PROMPT = """You are an experienced sell-side equity analyst.
Your task is to forecast the Free Cash Flow to Firm (FCFF) for the next
three fiscal years for {company_name}.

You have access to the following information:
- Historical FCFF (last 5 years): {historical_fcff}
- Most recent 10-K MD&A summary: {mdna_summary}
- Industry: {industry}
- Current macroeconomic conditions: {macro_summary}
- Analyst consensus revenue growth estimates: {consensus_estimates}

Please reason through the following steps BEFORE giving your forecast:

Step 1 - Macroeconomic context:
Assess how the current macro environment (interest rates, GDP growth,
inflation) is likely to affect this company's revenue and margins.

Step 2 - Industry dynamics:
Identify the key drivers and headwinds for this industry over the next
three years. Consider competitive dynamics, technology disruption,
and regulatory changes.

Step 3 - Firm-specific factors:
Evaluate the company's competitive position, recent strategic
initiatives, margin trajectory, and capital expenditure plans
as disclosed in the MD&A.

Step 4 - Synthesis and projection:
Combining the above, provide your FCFF forecast for Years 1, 2, and 3.
For each year, give:
  - Point estimate (USD millions)
  - Upside case (90th percentile)
  - Downside case (10th percentile)
  - Two-sentence justification

Respond in the following JSON format:
{{
  "reasoning": {{
    "macro": "<your macro analysis>",
    "industry": "<your industry analysis>",
    "firm_specific": "<your firm analysis>"
  }},
  "forecasts": [
    {{"year": 1, "base": <float>, "upside": <float>, "downside": <float>,
      "justification": "<text>"}},
    {{"year": 2, "base": <float>, "upside": <float>, "downside": <float>,
      "justification": "<text>"}},
    {{"year": 3, "base": <float>, "upside": <float>, "downside": <float>,
      "justification": "<text>"}}
  ]
}}
"""
\end{lstlisting}

The separation of reasoning from numerical output is not merely cosmetic: it enables
downstream validation. If the reasoning step contains an internally inconsistent claim
(e.g., projecting 20\% revenue growth while simultaneously noting a highly competitive
market with stable pricing), an automated consistency-checker or a human reviewer can
flag it before the numbers propagate into the DCF model.

\subsection{Fine-Tuning vs.\ Few-Shot Approaches for Financial Projection}
\label{subsec:bv-forecasting-finetuning}

Two broad strategies exist for adapting a general-purpose LLM to financial
forecasting: \emph{in-context learning} (few-shot prompting) and
\emph{fine-tuning} (supervised adaptation of model weights).

In few-shot prompting, three to ten examples of historical (company, context, forecast)
triples are prepended to the prompt. The model generalises the pattern demonstrated
by the examples to the new query. The advantages of few-shot are deployment
flexibility (no training infrastructure), zero marginal cost per new company, and the
ability to update examples without retraining. The disadvantage is that example
selection requires care: poorly chosen examples can anchor the model to the wrong
regime. Research on in-context learning for financial tasks suggests that including
examples with diverse macro environments and industry types reduces the variance of
forecasts \cite{brown2020language}.

Fine-tuning trains the model weights on a labelled dataset of historical
(filing, subsequent cash flow realisation) pairs. This allows the model to internalise
domain-specific patterns---for example, the empirical relationship between specific
MD\&A language and subsequent earnings surprises---that cannot be conveyed efficiently
in a prompt. The cost is dataset construction (historical filing collection and
labelling), training compute, and the need to retrain when the model base changes.
\citet{lopezlira2023chatgpt} provide empirical evidence that LLM-generated signals
from financial text carry information about subsequent stock returns even without
fine-tuning, implying that strong general-purpose models already encode substantial
domain knowledge. Fine-tuning is most beneficial when the task is highly specialised
(e.g., forecasting cash flows for a specific sector with non-standard accounting
conventions) or when the model needs to produce calibrated uncertainty estimates.

A pragmatic hybrid is \emph{parameter-efficient fine-tuning} (PEFT) via Low-Rank
Adaptation (LoRA) \cite{hu2022lora}:
only a small fraction of parameters are trained, dramatically reducing compute
requirements while achieving performance competitive with full fine-tuning on
domain-specific tasks.

\subsection{Evaluating Forecast Quality}
\label{subsec:bv-forecasting-evaluation}

Forecast evaluation requires metrics that capture both \emph{magnitude accuracy}
and \emph{directional accuracy}, since a valuation model is sensitive to both.

\begin{definition}[Mean Absolute Percentage Error]
\label{def:mape}
Let $y_t$ be the realised value and $\hat{y}_t$ the forecast for period $t$, over a
test set of $n$ observations. The \emph{mean absolute percentage error} is
\begin{equation}
\label{eq:mape}
\mathrm{MAPE} = \frac{1}{n}\sum_{t=1}^{n} \frac{|y_t - \hat{y}_t|}{|y_t|} \times 100.
\end{equation}
\end{definition}

\begin{definition}[Root Mean Squared Error]
\label{def:rmse}
The \emph{root mean squared error} is
\begin{equation}
\label{eq:rmse}
\mathrm{RMSE} = \sqrt{\frac{1}{n}\sum_{t=1}^{n}(y_t - \hat{y}_t)^2}.
\end{equation}
\end{definition}

\begin{definition}[Directional Accuracy]
\label{def:da}
The \emph{directional accuracy} measures the fraction of periods in which the forecast
correctly predicts the direction of change:
\begin{equation}
\label{eq:da}
\mathrm{DA} = \frac{1}{n}\sum_{t=1}^{n}
  \mathbf{1}\!\left[\mathrm{sign}(y_t - y_{t-1}) = \mathrm{sign}(\hat{y}_t - y_{t-1})\right].
\end{equation}
\end{definition}

MAPE is intuitive but problematic when $y_t$ is near zero or negative, which occurs
for firms with volatile cash flows. RMSE is sensitive to large outliers and on an
absolute scale that depends on company size, making cross-company comparisons difficult.
Directional accuracy is scale-invariant and directly relevant to investment
decision-making: a fund manager whose valuation model correctly identifies whether FCF
will grow or shrink next year captures trading alpha even if the magnitude estimate is
imprecise. A well-specified evaluation framework reports all three metrics and compares
them against a random-walk baseline and, where available, sell-side analyst consensus
estimates \citep{zhang2024financebench}.
\end{deepdive}

% --- Section: Tool-Augmented Agents for Financial Math ---

\section{Tool-Augmented Agents for Financial Math}
\label{sec:bv-tools}

\begin{deepdive}
This section explains the architectural reason why LLMs cannot be trusted for
multi-step arithmetic, then shows how to implement the code-interpreter and ReAct
patterns that route all financial calculations to a deterministic executor —
giving you a blueprint you can extend to any valuation workflow.

\subsection{Why LLMs Alone Struggle with Arithmetic}
\label{subsec:bv-tools-limitations}

Large language models are next-token predictors trained on text corpora. Arithmetic
computation is not a natural fit for this architecture: the correct answer to
$47{,}382 \times 631$ is determined not by statistical patterns in training data but
by a deterministic algorithm. While transformers can learn to approximate certain
arithmetic operations through repeated exposure to worked examples, their reliability
degrades rapidly with problem complexity. Empirical studies find that GPT-4 achieves
near-perfect accuracy on single-digit multiplication and simple percentage calculations
but makes errors with non-trivial frequency on multi-step financial calculations
involving chained operations, unit conversions, and percentage changes over long
formulas \cite{frieder2023large}.

For business valuation this is a serious limitation. A five-year DCF model involves
approximately fifteen to twenty distinct arithmetic operations, each conditional on
the results of the previous ones. Errors compound: an incorrectly applied tax rate in
the FCFF formula propagates through every subsequent calculation, and the final
valuation error can be several times the original computational error. The solution
is architectural: remove arithmetic from the LLM's responsibility entirely and route
it to a deterministic computational tool.

\subsection{Code Interpreter and Calculator Tools}
\label{subsec:bv-tools-code-interpreter}

The \emph{code interpreter} pattern allows an LLM to generate executable code,
submit it to a sandboxed Python interpreter, receive the output, and incorporate
it into its reasoning. This approach has been formalised in systems such as
Toolformer \citep{schick2024toolformer}, which demonstrated that models can
learn to decide when to call external tools by observing how tool use reduces
prediction loss.

The code interpreter pattern for DCF computation looks as follows:

\begin{lstlisting}[language=Python, caption={Code Interpreter Tool for DCF Calculation}, label={lst:code-interpreter-dcf}]
import subprocess
import sys

DCF_TOOL_SCHEMA = {
    "name": "python_executor",
    "description": (
        "Execute Python code and return the output. "
        "Use this for all arithmetic, financial calculations, "
        "and data transformations."
    ),
    "parameters": {
        "type": "object",
        "properties": {
            "code": {
                "type": "string",
                "description": "Python code to execute. Must be self-contained."
            }
        },
        "required": ["code"]
    }
}

def execute_python(code: str, timeout: int = 30) -> dict:
    """
    Execute Python code in an isolated subprocess and return results.
    Returns dict with 'stdout', 'stderr', and 'success' fields.
    """
    try:
        result = subprocess.run(
            [sys.executable, "-c", code],
            capture_output=True,
            text=True,
            timeout=timeout,
        )
        return {
            "stdout": result.stdout.strip(),
            "stderr": result.stderr.strip(),
            "success": result.returncode == 0,
        }
    except subprocess.TimeoutExpired:
        return {"stdout": "", "stderr": "Execution timed out.", "success": False}
\end{lstlisting}

The advantage of generating Python over performing mental arithmetic is not merely
accuracy: the generated code is an auditable artefact. A human reviewer can inspect
the code to verify that the model applied the correct formula, used the correct
inputs, and performed the calculation in the intended units. This transparency is
essential in any regulated or high-stakes financial context.

\subsection{ReAct and Tool-Use Patterns in Valuation Workflows}
\label{subsec:bv-tools-react}

\citet{yao2022react} proposed the ReAct (Reason + Act) framework, which interleaves
chain-of-thought reasoning with tool invocations in a structured loop. Each iteration
consists of three phases:

\begin{enumerate}
  \item \textbf{Thought}: The agent articulates its current understanding and decides
        what to do next.
  \item \textbf{Action}: The agent invokes a tool (e.g.,\ \texttt{python\_executor},
        \texttt{edgar\_search}, \texttt{calculator}).
  \item \textbf{Observation}: The agent receives the tool's output and incorporates it
        into its context before the next thought.
\end{enumerate}

Applied to DCF valuation, a ReAct loop might unfold as follows:

\begin{lstlisting}[language=Python, caption={Illustrative ReAct Loop for DCF Valuation}, label={lst:react-dcf}]
# Illustrative trace of a ReAct agent computing enterprise value

Thought 1: I need to compute FCFF. I have EBIT=840M, DA=310M,
           CapEx=180M, delta_NWC=80M, tax_rate=0.21.
           I will use the python_executor to ensure correctness.

Action 1:  python_executor(code="""
               ebit = 840; da = 310; capex = 180
               delta_nwc = 80; tau = 0.21
               fcff = ebit * (1 - tau) + da - capex - delta_nwc
               print(f"FCFF = {fcff:.1f} M")
           """)

Observation 1: FCFF = 713.6 M

Thought 2: FCFF is 713.6M. Now I need the terminal value.
           Assuming FCF grows at 8% for 5 years then perpetuity at 3%,
           WACC = 9%. Let me compute the full DCF.

Action 2:  python_executor(code="""
               import numpy as np
               fcff0 = 713.6; g_explicit = 0.08; g_terminal = 0.03
               wacc = 0.09; n = 5
               fcfs = [fcff0 * (1 + g_explicit)**t for t in range(1, n+1)]
               tv = fcfs[-1] * (1 + g_terminal) / (wacc - g_terminal)
               pv_fcfs = sum(f / (1+wacc)**t for t, f in enumerate(fcfs, 1))
               pv_tv = tv / (1+wacc)**n
               ev = pv_fcfs + pv_tv
               print(f"PV of FCFs = {pv_fcfs:.1f} M")
               print(f"Terminal Value = {tv:.1f} M")
               print(f"PV of TV = {pv_tv:.1f} M")
               print(f"Enterprise Value = {ev:.1f} M")
           """)

Observation 2: PV of FCFs = 3210.4 M
               Terminal Value = 15601.7 M
               PV of TV = 10136.3 M
               Enterprise Value = 13346.7 M

Thought 3: Enterprise value is approximately $13.3B.
           I can now report this result with full provenance.
\end{lstlisting}

The ReAct pattern makes the reasoning transparent and auditable while offloading all
arithmetic to a reliable executor. The agent's role is to formulate the calculation
correctly and interpret the results; the Python interpreter handles the computation.

\subsection{Hallucinations and Limitations in Numerical Reasoning}
\label{subsec:bv-tools-hallucinations}

Even with tool augmentation, LLMs introduce distinctive failure modes in financial
workflows. It is important to catalogue these systematically so that appropriate
guardrails can be designed.

\emph{Phantom tickers and fabricated companies}: an LLM asked to identify comparable
companies may generate realistic-sounding but non-existent ticker symbols or company
names. All tickers must be validated against an authoritative source (e.g.,\ the EDGAR
company search API or a securities master file) before use.

\emph{Invented financial ratios}: models may cite multiples, margin percentages, or
other financial statistics with false precision, confabulating plausible-looking numbers
that do not correspond to any filed data. The mitigation is mandatory grounding:
every financial figure used in a valuation must be traced to a specific EDGAR filing
or verified market data source. The ``trust but verify'' principle should be replaced
with ``do not trust until verified.''

\emph{Misapplied formulas}: the model may generate syntactically correct code that
implements the wrong formula (e.g., computing FCFE when FCFF was requested, or
forgetting the tax shield on interest). Code review by a human analyst, or automated
formula-level testing against known benchmark cases, is necessary in production.

\emph{Stale knowledge}: LLMs have a knowledge cutoff date. Financial data, tax rates,
and accounting standards change. Any retrieval-augmented system must ensure that
information retrieved from EDGAR is used in preference to the model's parametric
knowledge for all factual claims about specific companies \citep{zhang2024financebench}.
\end{deepdive}

% --- Section: Scenario Analysis with LLMs ---

\section{Scenario Analysis with LLMs}
\label{sec:bv-scenarios}

\begin{deepdive}
This section formalises valuation scenarios as distributional objects, derives the
Monte Carlo DCF estimator, and implements a simulation you can parameterise and
re-run in under a second — turning qualitative bull/base/bear narratives into a
quantified enterprise-value distribution.

\subsection{Bull, Base, and Bear Case Construction}
\label{subsec:bv-scenarios-cases}

A single-point DCF estimate is a misleading representation of value because the inputs
to the model are uncertain. Professional practice addresses this through \emph{scenario
analysis}, in which the analyst constructs a small number of internally consistent
input sets representing different macroeconomic and business outcomes.

\begin{definition}[Valuation Scenarios]
\label{def:scenarios}
A \emph{bull case} corresponds approximately to the 90th percentile of the
distribution of plausible outcomes---a scenario in which the firm benefits from
favourable macro conditions, strong competitive execution, and above-consensus
growth. A \emph{base case} corresponds to the 50th percentile (median expectation)
and represents the analyst's best estimate of the most likely trajectory. A
\emph{bear case} corresponds to the 10th percentile---a scenario of adverse macro
conditions, margin compression, or competitive deterioration.
\end{definition}

Each scenario specifies a complete and internally consistent set of model inputs:
revenue growth rates for each explicit forecast year, EBIT margins, CapEx intensity,
net working capital ratios, and terminal growth rate. The internal consistency
requirement is important: a bull case that combines high revenue growth with high
margins and low CapEx is plausible only if the business model supports such scale
economics. A model that simultaneously assumes 25\% revenue growth, 35\% EBIT margins,
and a terminal growth rate of 1\% is internally contradictory---rapid growth
corporations rarely decelerate to near-zero growth abruptly.

LLMs are valuable at this stage not for numerical computation but for
\emph{narrative coherence checking}: given a proposed set of scenario inputs, the
model can assess whether the numbers are consistent with the described business
environment and flag contradictions. An example prompt:

\begin{lstlisting}[language=Python, caption={Scenario Consistency Validation Prompt}, label={lst:scenario-consistency}]
SCENARIO_CHECK_PROMPT = """You are a senior equity research analyst reviewing
a DCF valuation scenario for {company_name} in the {industry} industry.

The following scenario has been proposed:
- Scenario: {scenario_name}
- Revenue CAGR (Years 1-5): {revenue_cagr}%
- EBIT Margin (Year 5): {ebit_margin}%
- CapEx as % of Revenue: {capex_pct}%
- Terminal Growth Rate: {terminal_growth}%
- WACC: {wacc}%

Context: {scenario_narrative}

Please evaluate the internal consistency of these assumptions. Specifically:
1. Are the margin and growth assumptions consistent with each other
   and with industry benchmarks?
2. Is the terminal growth rate defensible relative to long-run GDP growth?
3. Are there any assumptions that are internally contradictory?

Respond with a JSON object:
{{"is_consistent": true/false,
  "issues": ["<list of specific inconsistencies or concerns>"],
  "recommendation": "<brief corrective action if needed>"}}
"""
\end{lstlisting}

\subsection{Monte Carlo Simulation Orchestrated by Agents}
\label{subsec:bv-scenarios-montecarlo}

Scenario analysis with three discrete cases is informative but does not characterise
the full distribution of enterprise value. \emph{Monte Carlo simulation} generates
this distribution by sampling from parametric distributions over the key inputs.

\begin{definition}[Monte Carlo DCF]
\label{def:monte-carlo-dcf}
Let $\mathrm{WACC} \sim \mathcal{N}(\mu_\mathrm{WACC}, \sigma^2_\mathrm{WACC})$
and $g \sim \mathcal{N}(\mu_g, \sigma^2_g)$ (truncated to ensure $g < \mathrm{WACC}$).
For $M$ simulation paths indexed by $m = 1, \ldots, M$, draw
$(\mathrm{WACC}^{(m)}, g^{(m)})$ and compute the enterprise value
\begin{equation}
\label{eq:monte-carlo-ev}
V^{(m)} = \sum_{t=1}^{n} \frac{\mathrm{FCF}_t^{(m)}}{(1+\mathrm{WACC}^{(m)})^t}
+ \frac{\mathrm{FCF}_{n+1}^{(m)}}{(\mathrm{WACC}^{(m)} - g^{(m)})(1+\mathrm{WACC}^{(m)})^n},
\end{equation}
where $\mathrm{FCF}_t^{(m)}$ may itself be drawn from a distribution over
cash-flow forecasts. The empirical distribution $\{V^{(m)}\}_{m=1}^M$ yields
the mean, median, standard deviation, and percentile-based confidence intervals
for enterprise value.
\end{definition}

A practical Monte Carlo implementation for valuation, invoked by an agent, is:

\begin{lstlisting}[language=Python, caption={Monte Carlo DCF Simulation}, label={lst:monte-carlo-dcf}]
import numpy as np

def monte_carlo_dcf(
    fcff_base: float,
    g_explicit_mean: float,
    g_explicit_std: float,
    wacc_mean: float,
    wacc_std: float,
    g_terminal_mean: float,
    g_terminal_std: float,
    n_years: int = 5,
    n_simulations: int = 10_000,
    seed: int = 42,
) -> dict:
    """
    Monte Carlo DCF simulation.

    Parameters
    ----------
    fcff_base : float
        Current-year FCFF (USD millions).
    g_explicit_* : float
        Mean and std of explicit-period annual FCF growth rate.
    wacc_* : float
        Mean and std of WACC.
    g_terminal_* : float
        Mean and std of terminal growth rate.
    """
    rng = np.random.default_rng(seed)
    wacc_sim = rng.normal(wacc_mean, wacc_std, n_simulations)
    g_term_sim = rng.normal(g_terminal_mean, g_terminal_std, n_simulations)
    g_exp_sim = rng.normal(g_explicit_mean, g_explicit_std,
                            size=(n_simulations, n_years))

    # Clip to ensure g_terminal < wacc (Gordon Growth Model validity)
    g_term_sim = np.minimum(g_term_sim, wacc_sim - 0.005)

    ev_sim = np.zeros(n_simulations)
    for m in range(n_simulations):
        fcfs = [fcff_base]
        for t in range(n_years):
            fcfs.append(fcfs[-1] * (1 + g_exp_sim[m, t]))
        pv_fcf = sum(
            fcfs[t] / (1 + wacc_sim[m]) ** t for t in range(1, n_years + 1)
        )
        tv = (fcfs[n_years] * (1 + g_term_sim[m])
              / (wacc_sim[m] - g_term_sim[m]))
        pv_tv = tv / (1 + wacc_sim[m]) ** n_years
        ev_sim[m] = pv_fcf + pv_tv

    return {
        "mean": float(np.mean(ev_sim)),
        "median": float(np.median(ev_sim)),
        "std": float(np.std(ev_sim)),
        "p10": float(np.percentile(ev_sim, 10)),
        "p25": float(np.percentile(ev_sim, 25)),
        "p75": float(np.percentile(ev_sim, 75)),
        "p90": float(np.percentile(ev_sim, 90)),
        "simulations": ev_sim.tolist(),
    }
\end{lstlisting}

Typical parameterisation in a mid-cap equity context might use
$\mu_\mathrm{WACC} = 0.09$, $\sigma_\mathrm{WACC} = 0.015$,
$\mu_g = 0.03$, $\sigma_g = 0.01$. With $M = 10{,}000$ draws, the simulation
runs in under a second on a modern laptop, making it practical to re-run after
every revision to the model inputs.

\subsection{Sensitivity Analysis and Narrative Generation}
\label{subsec:bv-scenarios-sensitivity}

A sensitivity table examines how enterprise value responds to independent variation
in two key parameters, holding all others constant. The two most informative
dimensions are WACC and terminal growth rate, since together they dominate the
terminal value calculation.

Beyond the numerical sensitivity table, LLMs add value by \emph{generating
the accompanying investment narrative}: a paragraph that interprets the
simulation results in plain language, identifies the key risk factors driving
the downside scenario, and frames the upside drivers. This narrative synthesis
is where LLMs genuinely augment analyst productivity---producing a first-draft
investment write-up that would otherwise take an analyst several hours.
\end{deepdive}

% --- Section: Comparable Selection Using LLMs ---

\section{Comparable Selection Using LLMs}
\label{sec:bv-comparables}

\begin{deepdive}
This section develops two complementary algorithms for peer-company identification ---
prompt-based retrieval from the model's parametric memory and cosine-similarity search
over MD\&A embeddings --- then combines them in a hybrid pipeline that you can implement
and validate against a securities master file.

\subsection{Traditional Comparable Selection: Criteria and Pitfalls}
\label{subsec:bv-comparables-traditional}

The market approach to valuation requires a peer group of companies that are
sufficiently similar to the target to make multiple-based comparisons meaningful.
Traditional comparable selection applies a hierarchy of filters. The first and
most common filter is \emph{industry classification}: SIC (Standard Industrial
Classification) codes, GICS (Global Industry Classification Standard) segments,
or NAICS codes identify firms in the same sector. The second filter is
\emph{size}: comparable firms typically fall within a band of 0.3--3.0 times
the target's revenue or market capitalisation, since multiples vary systematically
with scale. Geographic and regulatory constraints may further narrow the peer group.

Three well-documented pitfalls afflict this approach. First, \emph{code obsolescence}:
SIC codes were last revised in 1987; many technology, platform, and life-sciences
businesses are assigned to codes that do not reflect their current business model.
A cloud computing company and a 1990s software distributor may share the same
SIC code. Second, \emph{diversified conglomerates}: a company with SIC code
indicating manufacturing may derive the majority of its revenue from financial
services. Multiples from that company are not informative peers for a pure-play
manufacturer. Third, \emph{cyclical timing}: firms in the same sector at different
points in their business cycles exhibit very different multiples; including a
temporarily distressed firm in the peer group depresses the median multiple and
undervalues the target.

\subsection{Prompt-Based Comparable Identification}
\label{subsec:bv-comparables-prompts}

LLMs can identify peer companies from a natural-language description of the target
firm's business model, without relying on rigid classification codes. A zero-shot
prompt simply describes the target and requests comparable firms:

\begin{lstlisting}[language=Python, caption={Zero-Shot Comparable Identification Prompt}, label={lst:zero-shot-comps}]
ZERO_SHOT_COMPS_PROMPT = """You are an investment banking analyst specialising
in equity valuation. Your task is to identify publicly listed comparable companies
for the following target firm.

Target firm description:
{target_description}

Financial profile:
- Revenue: {revenue} USD millions
- EBIT margin: {ebit_margin}%
- Geography: {geography}
- Stage: {company_stage}

Identify 8 to 12 publicly listed comparable companies. For each, provide:
- Company name
- Ticker symbol
- Stock exchange
- Similarity rationale (2-3 sentences)
- Similarity score (1-5, where 5 is near-identical business model)

Rules:
- Include only actually existing, currently publicly listed companies.
- Do NOT invent tickers or company names.
- Prefer companies with similar revenue scale (0.3x to 3x target revenue).
- Exclude competitors that are more than 5x larger (index-constituents with
  very different multiple dynamics).

Respond as a JSON array of objects with keys:
"name", "ticker", "exchange", "rationale", "similarity_score".
"""
\end{lstlisting}

The instruction ``Do NOT invent tickers or company names'' is necessary but not
sufficient: hallucinated comparables occur with non-trivial frequency even with
this guard. A mandatory validation step---checking every returned ticker against
a securities master---is non-negotiable in production use. Few-shot prompts that
include three to five examples of correct (target, comparables) pairs drawn from
historical analyst reports markedly reduce the hallucination rate by anchoring the
model to real firm names.

\subsection{Embedding-Based Approaches: Semantic Similarity on Financial Data}
\label{subsec:bv-comparables-embeddings}

The limitation of prompt-based comparable selection is that it relies on the model's
parametric memory of company names, which may be stale or incomplete for smaller
companies. Embedding-based approaches address this by directly measuring semantic
similarity between financial text representations.

\begin{definition}[Cosine Similarity]
\label{def:cosine-similarity}
For two embedding vectors $\mathbf{a}, \mathbf{b} \in \mathbb{R}^d$, the
\emph{cosine similarity} is
\begin{equation}
\label{eq:cosine-sim}
\mathrm{sim}(\mathbf{a}, \mathbf{b}) = \frac{\mathbf{a} \cdot \mathbf{b}}{\|\mathbf{a}\|\,\|\mathbf{b}\|} \in [-1, 1].
\end{equation}
A value near $+1$ indicates that the two vectors point in the same direction---that is,
the two documents occupy the same region of the embedding space---while a value near
$0$ or $-1$ indicates orthogonality or opposition.
\end{definition}

The MD\&A section of the 10-K is the most semantically rich source for business-model
characterisation: it describes in management's own words the company's products,
markets, competitive position, and strategy. Embedding the MD\&A of all companies in
a universe and indexing them in a vector database allows efficient nearest-neighbour
retrieval.

\begin{lstlisting}[language=Python, caption={Embedding-Based Comparable Selection}, label={lst:embedding-comps}]
import numpy as np
import faiss  # pip install faiss-cpu

def build_comparable_index(
    company_embeddings: dict[str, np.ndarray],
) -> tuple[faiss.IndexFlatIP, list[str]]:
    """
    Build a FAISS inner-product index for fast comparable retrieval.

    Parameters
    ----------
    company_embeddings : dict mapping ticker -> embedding vector (L2-normalised)

    Returns
    -------
    index : faiss.IndexFlatIP
    tickers : list of tickers in index order
    """
    tickers = list(company_embeddings.keys())
    dim = len(next(iter(company_embeddings.values())))
    # Stack into matrix
    matrix = np.vstack([company_embeddings[t] for t in tickers]).astype("float32")
    # L2-normalise so inner product equals cosine similarity
    norms = np.linalg.norm(matrix, axis=1, keepdims=True)
    matrix /= norms
    index = faiss.IndexFlatIP(dim)
    index.add(matrix)
    return index, tickers

def find_comparables(
    target_embedding: np.ndarray,
    index: faiss.IndexFlatIP,
    tickers: list[str],
    top_k: int = 15,
    exclude_self: str | None = None,
) -> list[dict]:
    """
    Retrieve the top-k most similar companies by cosine similarity.
    """
    query = target_embedding.reshape(1, -1).astype("float32")
    query /= np.linalg.norm(query)
    similarities, indices = index.search(query, top_k + 1)
    results = []
    for sim, idx in zip(similarities[0], indices[0]):
        ticker = tickers[idx]
        if exclude_self and ticker == exclude_self:
            continue
        results.append({"ticker": ticker, "cosine_similarity": float(sim)})
        if len(results) == top_k:
            break
    return results
\end{lstlisting}

Empirical studies using MD\&A embeddings find that cosine-similarity-based peers
outperform SIC-code-based peers on several metrics: the comparable set has lower
inter-quartile range of EV/EBITDA multiples (indicating more internally consistent
peers), and valuations derived from embedding-based comparables exhibit lower
absolute deviation from subsequent transaction prices \cite{shen2023nlp}.

\subsection{Hybrid Methods: LLM-Filtered and Embedding-Ranked Comparables}
\label{subsec:bv-comparables-hybrid}

Both methods have complementary weaknesses. Prompt-based selection leverages broad
world knowledge but hallucinate. Embedding-based selection is grounded but may
return companies that are textually similar for the wrong reasons (e.g., two
companies with similar MD\&A boilerplate language but different businesses).

A \emph{hybrid} pipeline combines both in two stages. In the \emph{candidate
generation} stage, the embedding index returns the top 30 candidates by cosine
similarity---a large enough set to capture genuine peers, but manageable for the
filtering stage. In the \emph{LLM filtering} stage, the model receives the list of
candidates alongside brief company descriptions and applies qualitative judgment:
removing firms that are geographically incomparable, in demonstrably different
competitive positions, or subject to idiosyncratic events (e.g., under pending
acquisition or bankruptcy protection). The filtered set of 8--12 comparables is
then used for multiple analysis.

The two-stage design confines the LLM's role to a task where it excels (qualitative
business-model reasoning) and avoids the task where it is unreliable (recalling
whether a specific ticker exists and what it refers to).
\end{deepdive}

% --- Section: End-to-End Valuation Pipeline ---

\section{End-to-End Valuation Pipeline}
\label{sec:bv-pipeline}

\begin{context}
The preceding sections each solved one piece of the valuation puzzle in isolation.
Here we show how a practitioner wires them together into a production pipeline,
orchestrates the components with a multi-agent architecture, and walks through
the full workflow on a concrete case study.

\subsection{Integrating Extraction, Forecasting, and Comparables}
\label{subsec:bv-pipeline-integration}

The components developed in the preceding sections can be assembled into a coherent
valuation pipeline. The data flow is as follows:

\begin{enumerate}
  \item \textbf{EDGAR Retrieval}: Given a ticker, resolve the CIK, download the
        most recent 10-K (and prior four years for historical context).
  \item \textbf{LLM Extraction}: Parse the income statement, balance sheet, and
        cash flow statement into structured JSON; flag non-recurring items using the
        MD\&A analysis prompt.
  \item \textbf{XBRL Cross-check}: Retrieve tagged XBRL data and reconcile with
        LLM extraction; prefer XBRL for unambiguous tagged items.
  \item \textbf{FCF Computation}: Apply the FCFF formula to the normalised financials
        to produce a historical time series.
  \item \textbf{Forecasting}: Apply the CoT forecasting prompt to generate base,
        upside, and downside FCF paths for the explicit horizon.
  \item \textbf{Terminal Value}: Compute terminal values under Gordon Growth Model
        assumptions corresponding to each scenario.
  \item \textbf{Comparable Selection}: Run the hybrid pipeline to identify a peer group;
        compute current-year multiples for each peer.
  \item \textbf{DCF and Comparables Valuation}: Compute DCF enterprise value and
        equity value; compute comparable-based equity value using selected multiples.
  \item \textbf{Monte Carlo}: Simulate the distribution of DCF values under parameter
        uncertainty.
  \item \textbf{Report Generation}: Synthesise results into a structured valuation
        summary with narrative commentary.
\end{enumerate}

Each step produces a structured artefact that is passed to the next, creating an
auditable data lineage. A formal data contract between pipeline stages---enforced
with Pydantic models or equivalent schema validators---prevents silent failures
from propagating.

\subsection{Agent Orchestration for Multi-Step Valuation}
\label{subsec:bv-pipeline-orchestration}

The pipeline above can be implemented as a monolithic script, but this design is
brittle: a failure in step 6 requires rerunning all prior steps. A more robust
design uses an \emph{orchestrator agent} that maintains state and dispatches to
specialist sub-agents, analogous to the multi-agent patterns described in
Chapter~\ref{ch:llm-agents}.

The orchestrator receives the target ticker and valuation objective (e.g., ``enterprise
value for M\&A due diligence''), decomposes the task into sub-problems, and routes
each sub-problem to the appropriate specialist: an \texttt{edgar\_agent} for retrieval
and extraction, a \texttt{fcf\_agent} for computation and normalisation, a
\texttt{forecast\_agent} for projections, a \texttt{comps\_agent} for comparable
selection, and a \texttt{dcf\_agent} for final valuation and scenario synthesis. Each
specialist reports back to the orchestrator with its output and any flags (e.g.,\
``data gap in 2021 CapEx''). The orchestrator aggregates these outputs, handles
exceptions, and initiates human-escalation when a specialist signals low confidence.

This architecture has several advantages. First, each specialist can be independently
tested and replaced---if a better forecasting agent becomes available, it can be
swapped in without touching the orchestration logic. Second, specialists can run
in parallel where there are no dependencies: comparable selection does not depend on
the FCF forecast and can proceed concurrently. Third, the orchestrator maintains a
shared valuation state dictionary that provides full provenance for the final output.

\subsection{Case Study: Full Pipeline on a Public Company}
\label{subsec:bv-pipeline-casestudy}

To illustrate the pipeline concretely, consider a mid-cap US technology company
with the following stylised profile, based on a composite of publicly available
filings (specific company names and figures are illustrative):

\begin{itemize}
  \item \textbf{Industry}: Cloud software and infrastructure (SaaS)
  \item \textbf{Revenue (TTM)}: \$4.2B, growing at 12\% year-over-year
  \item \textbf{EBIT margin}: 20\%, with D\&A of \$310M and CapEx of \$180M
  \item \textbf{Net debt}: \$800M
  \item \textbf{Shares outstanding}: 280M
\end{itemize}

Applying the extraction pipeline to the most recent 10-K yields structured financials
after LLM parsing, with two non-recurring items flagged by the MD\&A analysis: a
\$45M restructuring charge and a \$30M acquired intangible amortisation step-up.
After normalisation, EBIT becomes \$915M and FCFF (with $\tau = 0.21$) equals
\$789M.

The CoT forecasting prompt, seeded with the historical FCFF series and the MD\&A
summary, returns the following base-case projections:

\begin{center}
\begin{tabular}{lccc}
\hline
Year & FCF (Base, \$M) & FCF (Upside, \$M) & FCF (Bear, \$M) \\
\hline
1 & 868 & 940 & 760 \\
2 & 947 & 1{,}050 & 820 \\
3 & 1{,}025 & 1{,}170 & 870 \\
4 & 1{,}100 & 1{,}280 & 900 \\
5 & 1{,}170 & 1{,}385 & 920 \\
\hline
\end{tabular}
\end{center}

With WACC of 9\% and terminal growth of 3\%, the base-case DCF enterprise value
is \$17.0B. Subtracting net debt yields an equity value of \$16.2B, or \$57.9 per
share on 280M shares outstanding. The embedding-based comparable pipeline identifies
nine peers with a median EV/EBITDA of 24.5$\times$. Applying this multiple to the
firm's EBITDA of \$1.22B yields a comparables-based enterprise value of \$29.9B---a
substantial premium to the DCF estimate, consistent with the sector trading at elevated
growth multiples at the time of the analysis. The Monte Carlo simulation (10{,}000
draws with $\sigma_\mathrm{WACC} = 1.5\%$ and $\sigma_g = 1\%$) produces an EV
distribution with a mean of \$17.2B and a 10th--90th percentile range of
\$12.5B--\$24.1B, providing quantified uncertainty alongside the point estimate.
\end{context}

% --- Section: Efficiency and Accuracy in LLM-Assisted Valuation ---

\section{Efficiency and Accuracy in LLM-Assisted Valuation}
\label{sec:bv-efficiency}

\begin{deepdive}
This section derives the evaluation metrics for benchmarking LLM-assisted valuations,
quantifies token cost and latency, and provides a first-order error-propagation
analysis showing how extraction errors compound through the DCF formula — then closes
with deployment best practices grounded in those quantitative findings.

\subsection{Benchmarking LLM Valuations Against Analyst Estimates}
\label{subsec:bv-efficiency-benchmarking}

Evaluating the quality of LLM-assisted valuations requires a benchmark. Three
natural comparators are available: (1) sell-side analyst consensus price targets,
(2) subsequent realised stock prices (12-month forward returns), and (3) transaction
multiples from M\&A deals where an independent fairness opinion was rendered.

The \citet{zhang2024financebench} benchmark provides a rigorous evaluation framework
for LLM performance on financial question-answering tasks using actual SEC filings.
For valuation-specific tasks, an analogous benchmark should evaluate the pipeline
on a held-out set of companies and fiscal years, comparing the LLM-derived DCF
equity value against the observed market price. The standard metrics are:

\begin{itemize}
  \item \emph{Mean absolute valuation error}: $\text{MAVE} = \frac{1}{N}\sum_{i=1}^{N} |V_i^{\mathrm{LLM}} - V_i^{\mathrm{market}}| / V_i^{\mathrm{market}} \times 100\%$.
  \item \emph{Directional accuracy relative to consensus}: fraction of cases where the
        LLM model is directionally consistent with analyst consensus (both above or both
        below the current trading price).
  \item \emph{Coverage rate}: fraction of companies for which the pipeline completes
        successfully without manual intervention.
\end{itemize}

Preliminary evidence from internal experiments and published literature suggests that
LLM-assisted extraction achieves coverage rates above 90\% for S\&P 500 constituents
and above 75\% for Russell 2000 companies, with MAVE broadly comparable to the spread
between analyst price targets and realised prices (typically 20--40\% on an annualised
basis) \cite{zhang2024financebench}.
The key insight is that the accuracy ceiling of any valuation model is determined by
the fundamental unpredictability of future cash flows, not by the extraction or
reasoning engine.

\subsection{Cost, Latency, and Token Efficiency}
\label{subsec:bv-efficiency-cost}

Running an LLM-assisted valuation incurs token costs that are non-trivial for
large-scale applications. A representative 10-K filing contains approximately
50{,}000 tokens. With a retrieval-augmented pipeline that sends relevant excerpts
(rather than the full document) to the model, the typical input context for a
single extraction call is 3{,}000--8{,}000 tokens; for a forecasting call,
1{,}000--2{,}000 tokens. The total token budget per company valuation is therefore
roughly:

\begin{equation}
\label{eq:token-cost}
\text{Cost} \approx (T_{\mathrm{in}} \times p_{\mathrm{in}} + T_{\mathrm{out}} \times p_{\mathrm{out}}) \times N_{\mathrm{calls}},
\end{equation}

where $T_{\mathrm{in}}$ and $T_{\mathrm{out}}$ are average input and output tokens
per call, $p_{\mathrm{in}}$ and $p_{\mathrm{out}}$ are the per-token prices, and
$N_{\mathrm{calls}}$ is the number of LLM calls in the pipeline. For a full
valuation with six LLM calls (three extraction, one normalisation, one forecast,
one comparable filter) at typical 2024 frontier-model pricing, the cost per company
is in the range \$0.50--\$3.00. At scale---a universe of 3{,}000 companies
refreshed quarterly---the budget is thus \$6{,}000--\$36{,}000 per quarter, well
within the operational budget of a mid-size investment management firm.

Latency is a separate concern. Each LLM call introduces 1--5 seconds of network and
inference latency. The pipeline can be parallelised across companies (each company's
valuation is independent) and, within a single company, across pipeline stages that
have no data dependencies (e.g., comparable selection can run concurrently with FCF
computation). With aggressive parallelisation, end-to-end latency per company can
be reduced to 15--30 seconds.

\subsection{Error Propagation in Multi-Step Agent Workflows}
\label{subsec:bv-efficiency-errors}

A critical concern in multi-step pipelines is \emph{error propagation}: a mistake
in an early stage compounds through all downstream calculations. To quantify this,
consider a first-order Taylor approximation of the DCF enterprise value.

Let $V = V(\mathrm{FCF}, g, \mathrm{WACC})$ denote the DCF value as a function of
its key inputs. Suppose extraction produces a cash-flow estimate with error
$\varepsilon_1 = \hat{\mathrm{FCF}} - \mathrm{FCF}^*$, and the forecasting stage
introduces a bias in terminal growth of $\varepsilon_2 = \hat{g} - g^*$.
The induced error in enterprise value is approximately

\begin{equation}
\label{eq:error-propagation}
\Delta V \approx \frac{\partial V}{\partial \mathrm{FCF}} \varepsilon_1
             + \frac{\partial V}{\partial g} \varepsilon_2,
\end{equation}

where, for the Gordon Growth perpetuity component $V \approx \mathrm{FCF}_n(1+g)/(\mathrm{WACC}-g)$
(with $\mathrm{FCF}_n$ the current-period free cash flow):
\begin{equation}
\label{eq:dv-dfcf}
\frac{\partial V}{\partial \mathrm{FCF}_n} = \frac{1+g}{\mathrm{WACC} - g},
\end{equation}
\begin{equation}
\label{eq:dv-dg}
\frac{\partial V}{\partial g} = \frac{\mathrm{FCF}_n\,(1 + \mathrm{WACC})}{(\mathrm{WACC} - g)^2}.
\end{equation}

Both partial derivatives are large when $g$ is close to WACC, confirming that
valuation is most sensitive to terminal-growth and discount-rate assumptions in
high-growth scenarios. With $\mathrm{WACC} = 0.09$ and $g = 0.03$, a 1\% relative
error in FCF translates into a 1\% error in enterprise value; but a 100-basis-point
error in terminal growth translates into approximately a 15--18\% error in enterprise
value---since $\partial V / \partial g \times 0.01 / V \approx (1+\mathrm{WACC})/[(\mathrm{WACC}-g)(1+g)] \times 0.01 \approx 0.176$ at these parameters (with the first-order approximation slightly above the exact discrete change of $\approx 16\%$). This analysis
justifies the emphasis on Monte Carlo simulation and scenario analysis rather than
single-point estimates.

\subsection{Best Practices and Deployment Considerations}
\label{subsec:bv-efficiency-bestpractices}

Assembling the lessons of this chapter, the following best practices characterise a
production-grade LLM-assisted valuation system.

Always verify extracted numbers against XBRL tags where available. LLM extraction
is a valuable complement to XBRL but should not serve as the sole source of
financial data. The SEC's XBRL API provides ground-truth tagged figures for the
majority of standard line items, and any LLM-extracted figure that conflicts with
the corresponding XBRL tag should be flagged for human review.

Use a code interpreter for all arithmetic. No financial calculation---however
simple---should be performed by the LLM's in-context reasoning. The code interpreter
pattern provides deterministic, auditable results and eliminates an entire class
of valuation errors. The generated code is itself a deliverable: it documents exactly
what was computed.

Run Monte Carlo to quantify uncertainty. A point-estimate DCF is a misleading
representation of value. At a minimum, a simulation over WACC and terminal growth
distributions should accompany every valuation, with the resulting percentile range
communicated alongside the central estimate.

Implement mandatory hallucination guards. Every company name and ticker returned by
the comparable-selection pipeline must be validated against an authoritative
securities register. Every financial figure cited by the model must be traced to a
specific EDGAR accession number. Any unverifiable claim should be suppressed from
the final output.

Never use LLM-only output in production without human review. The current generation
of LLMs is a powerful analytical assistant but not an autonomous decision-maker.
Final valuation reports should be reviewed by a qualified analyst who takes
responsibility for the figures. The LLM's output should be treated as a first-draft
that requires checking, not a finished product that requires only formatting.

These principles are consistent with the broader governance literature on AI in
finance \cite{fsb2023ai},
which emphasises human accountability, model transparency, and robust back-testing
as prerequisites for responsible deployment of machine-learning systems in investment
processes.
\end{deepdive}

\begin{remark}[Summary of Chapter Contributions]
\label{rem:ch5-summary}
This chapter has developed a comprehensive framework for LLM-assisted business
valuation. The key technical contributions are: (1) an end-to-end EDGAR extraction
pipeline using LLM table parsing with XBRL cross-validation; (2) a normalised FCFF
computation workflow with LLM-assisted non-recurring item identification; (3) a
chain-of-thought forecasting architecture with formal evaluation metrics; (4) a
ReAct agent pattern that routes all arithmetic to a sandboxed Python interpreter;
(5) a hybrid comparable-selection pipeline combining FAISS vector search with
LLM-based qualitative filtering; and (6) a first-order error-propagation analysis
that quantifies how extraction and forecast errors translate into valuation uncertainty.
Together, these components form a production-oriented reference architecture for
LLM-augmented equity valuation.
\end{remark}
