\chapter{Explainability and Interpretability of LLMs in Finance}
\label{ch:xai-explainability}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Articulate the legal and regulatory requirements for model explainability under
        the EU AI Act, SR 11-7, ECOA, and MiFID II, and explain why LLMs create novel
        compliance challenges.
  \item Apply SHAP and LIME to language model predictions, derive the Shapley value
        formula from cooperative game theory, and interpret feature attribution maps in a
        financial credit context.
  \item Critically evaluate the role of attention weights as explanations and distinguish
        between interpretable and faithful explanations.
  \item Design chain-of-thought prompts that function as auditable reasoning traces and
        generate counterfactual explanations for automated financial decisions.
  \item Draft ECOA-compliant adverse action notices and MiFID II suitability disclosures
        using LLM-assisted generation.
  \item Measure explanation quality using faithfulness, completeness, and plausibility
        metrics, and design human-subject evaluation studies with financial domain experts.
\end{enumerate}
\end{remark}

% ---------------------------------------------------------------
\section{Why Explainability Matters in Regulated Finance}
\label{sec:ch12-why-explainability}

\begin{context}
Consider a small business owner who applies for a working-capital loan at a mid-sized
regional bank. The bank's credit decision system---a transformer-based model trained on
millions of historical applications---returns a rejection within seconds. The loan officer
is asked to convey the decision. Under the Equal Credit Opportunity Act, the officer
must provide the applicant with specific reasons for the denial. The model, however, has
no explanations to offer: it produces a probability, not a rationale. The loan officer is
left to reverse-engineer a justification from features she may not even know the model
used. This everyday scenario, replicated thousands of times daily across the financial
system, is the practical heart of the explainability problem.
\end{context}

Explainability in machine learning refers to the degree to which a human can understand
the cause of a model's decision \cite{miller2019explanation}. In most industries, a lack
of explainability is an inconvenience. In finance, it is a legal, fiduciary, and
reputational problem of the first order. Financial institutions are not free to deploy
opaque models simply because those models are accurate; they must demonstrate that model
outputs are appropriate, non-discriminatory, and auditable \cite{sr117}.

The challenge has grown sharper as the industry has adopted large language models. Earlier
generations of credit and fraud models---logistic regression, gradient-boosted trees---
could at least be interrogated with classical XAI techniques. LLMs add new layers of
opacity: billions of parameters, uninterpretable intermediate activations, and outputs
that depend on the precise formulation of a prompt in ways that are difficult to
characterise \cite{arrieta2020explainable}. \citet{RaneChoudharyRane2023} survey how XAI approaches have been developed to address
the distinct transparency and accountability requirements of the financial domain,
providing conceptual grounding for why regulators and practitioners demand interpretable
models even when accuracy gains from opaque architectures are substantial. This chapter
develops the conceptual and technical toolkit for addressing that challenge.

\subsection{Legal Requirements: EU AI Act, SR 11-7, ECOA}
\label{sec:ch12-legal-requirements}

Three regulatory frameworks are particularly consequential for practitioners deploying
LLMs in financial decision-making.

The \emph{EU AI Act} \cite{euaiact2024}, which entered into force in August 2024,
categorises AI systems according to risk level. Credit scoring, insurance pricing, and
financial eligibility assessments for individuals are classified as \emph{high-risk}
applications (Annex III). For systems in this category the Act mandates transparency
measures that allow users to interpret and override outputs, logging of inputs and outputs
for post-hoc audit, and human oversight provisions that prevent the system from being
deployed without a qualified human in the loop for consequential decisions. The Act
defines ``transparency'' operationally: it is not sufficient to state that a model exists;
the deployer must be able to explain what factors drove any specific decision in terms
meaningful to the affected person.

\emph{SR 11-7}, the Federal Reserve's supervisory guidance on model risk management
\cite{sr117}, predates modern machine learning but remains the principal framework used
by U.S.\ bank examiners when assessing AI models. It requires that model developers
document the conceptual soundness of a model---why does this model work?---and that
validators conduct adversarial testing. Critically, it requires that model outputs be
explainable to risk managers and executives who must take accountability for them. The
guidance has been interpreted to require that complex models, including neural networks,
be accompanied by sensitivity analyses and post-hoc attribution methods that can satisfy
a non-technical examiner.

The \emph{Equal Credit Opportunity Act} (ECOA) and its implementing regulation, Regulation
B \cite{cfpb2013ecoa}, require creditors to notify applicants of the specific reasons for
adverse credit actions. The regulatory text specifies that reasons must be the
``principal reasons'' and must be stated in plain language that the applicant can
understand. A generic statement such as ``our model determined you were not creditworthy''
does not satisfy this requirement. The creditor must identify the four to five factors
that most negatively affected the decision. When the underlying model is a gradient-
boosted tree this is tractable; when it is a fine-tuned language model processing free-
text business descriptions and financial narrative, producing these reasons becomes
technically non-trivial.

Similar requirements exist under the Fair Credit Reporting Act (FCRA) for decisions that
rely on consumer reports, and in the European context under the General Data Protection
Regulation (GDPR) Article 22, which confers on data subjects a right not to be subject
to solely automated decisions that significantly affect them, along with a right to
obtain a ``meaningful explanation'' of the logic involved.

\begin{definition}[Explanation vs.\ Interpretation]
\label{def:explanation-vs-interpretation}
Following the framework of \citet{doshivelez2017rigorous}, we distinguish two related
concepts. An \emph{interpretation} is a mapping from an abstract model representation
to a domain intelligible to humans---it answers the question ``how does this model
work in general?'' An \emph{explanation} is a local object that justifies a specific
prediction---it answers the question ``why did this model produce this output for this
input?'' Regulatory requirements are primarily explanatory: they require that specific
decisions be justified to specific persons.
\end{definition}

\subsection{Stakeholder Expectations: Clients, Regulators, Auditors}
\label{sec:ch12-stakeholder-expectations}

Explainability in finance serves at least three distinct audiences, each with different
requirements and tolerances for technical complexity.

\emph{Clients} and applicants---the natural persons whose financial lives are affected
by model decisions---require explanations that are actionable and comprehensible without
specialist knowledge. An explanation that says ``your debt-to-income ratio of 52\% was
10 percentage points above our threshold'' is more useful than one that says ``feature
$x_{17}$ had a SHAP value of $-0.43$.'' Research in human-computer interaction
consistently finds that users prefer causal, contrastive explanations---``you were
rejected because your income was insufficient; a \$15,000 increase in annual income
would likely have resulted in approval''---over purely statistical ones
\cite{miller2019explanation}.

\emph{Regulators and examiners} require explanations at a higher level of technical
rigour. A Federal Reserve examiner reviewing a mortgage origination model will ask not
only whether individual decisions can be explained but whether the model, taken as a whole,
exhibits systematic disparate impact along protected-class lines. This requires
interpretations of global model behaviour---which features does the model rely on most
heavily?---as well as the ability to audit specific cases when complaints are filed. The
According to the CFA Institute \cite{cfainstitute2025xai}, institutional investors increasingly require explainability assurances before
deploying third-party AI systems for portfolio construction.

\emph{Internal auditors and model validators} occupy an intermediate position. They need
both global interpretations, to assess whether the model makes conceptual sense given the
problem domain, and local explanations, to perform targeted testing. Crucially, they must
be able to reproduce explanations---an XAI method that produces different explanations on
repeated runs from the same input is not suitable for audit purposes.

The mismatch between these audiences poses a genuine design challenge. Methods that
satisfy regulators may overwhelm retail clients; methods that are accessible to clients
may be too simplified to satisfy auditors. A comprehensive explainability strategy
typically requires multiple explanation types deployed in parallel.

\subsection{The Black-Box Problem for LLMs}
\label{sec:ch12-black-box-problem}

Classical machine learning models such as logistic regression and decision trees are
\emph{intrinsically interpretable}: their structure directly reveals the prediction
function. A logistic regression score is a weighted sum of inputs; the weights are the
interpretation. A decision tree produces a sequence of binary splits that can be read
as a set of human-intelligible rules. These models have a long track record in regulated
finance precisely because their interpretability is straightforward to document.

LLMs are not interpretable in this sense. A transformer with billions of parameters
encodes knowledge in distributed representations spread across all its layers. No single
parameter or small set of parameters corresponds to a human concept. The attention
mechanism, often invoked as a window into the model's reasoning, does not straightforwardly
indicate which input tokens caused an output---a point we examine in detail in
Section~\ref{sec:ch12-attention-visualization}.

The challenge is compounded by the nature of LLM inputs. Classical credit models consume
a fixed-length vector of structured numerical features. LLMs can consume free text: a
business plan, a financial statement narrative, a news article about a company's
counterparty. The high dimensionality of the input space---the set of possible sentences
is effectively infinite---makes exhaustive sensitivity analysis impossible.

\begin{remark}
\citet{rudin2019stop} argues forcefully that for high-stakes decisions, the correct
response to the black-box problem is not better post-hoc explanations but the deployment
of intrinsically interpretable models in the first place. This is a legitimate position
where it is technically feasible. The rejoinder in the financial LLM context is that
structured and unstructured inputs often cannot be disentangled: a language model that
simultaneously processes a borrower's financial ratios and the qualitative text of their
business description may extract signal from the text that a purely numerical model
cannot recover. The question of whether that extra signal justifies the cost in
interpretability is ultimately a business and regulatory judgement, not a technical one.
\end{remark}

% ---------------------------------------------------------------
\section{Classical XAI Methods Applied to LLMs}
\label{sec:ch12-classical-xai}

Before examining methods designed specifically for language models, it is instructive to
understand how classical post-hoc XAI methods can be adapted to the LLM setting. Three
methods have been most widely applied: SHAP, LIME, and attention visualisation. Each has
clear strengths and non-trivial limitations when applied to text-based models. \citet{MohsinNasim2025} provide bibliometric evidence for this taxonomy: their
systematic review of XAI in finance confirms that SHAP, attention mechanisms, and
feature importance are the three dominant techniques across financial applications
ranging from credit scoring to fraud detection, establishing these methods as the
effective standard against which any new approach must be benchmarked.

\subsection{SHAP Values: From Tree Models to Language Models}
\label{sec:ch12-shap-values}

Shapley values originate in cooperative game theory \cite{shapley1953value}. In that
context a game involves a set of players $N = \{1, \ldots, n\}$ who cooperate to produce
a value $v(S)$ for every coalition $S \subseteq N$. The Shapley value $\phi_i$ assigns
each player a fair share of the total value, defined as the average marginal contribution
of player $i$ across all possible orderings of players joining the coalition.

\begin{definition}[Shapley Value]
\label{def:shapley-value}
For a cooperative game $(N, v)$ where $N$ is the set of players and $v: 2^N \to \mathbb{R}$
is the characteristic function satisfying $v(\emptyset) = 0$, the Shapley value of
player $i$ is
\begin{equation}
\label{eq:shapley-value}
\phi_i(v) \;=\; \sum_{S \subseteq N \setminus \{i\}} \frac{|S|!\,(n - |S| - 1)!}{n!}
\bigl[v(S \cup \{i\}) - v(S)\bigr],
\end{equation}
where the sum is over all coalitions $S$ that do not contain $i$, the factor
$|S|!\,(n - |S| - 1)!/n!$ is the probability that $i$ joins the coalition at this
position in a uniform random ordering, and $v(S \cup \{i\}) - v(S)$ is $i$'s marginal
contribution.
\end{definition}

The SHAP framework of \citet{lundberg2017unified} reinterprets equation~\eqref{eq:shapley-value}
as a feature attribution problem. The players are the features $x_1, \ldots, x_n$ of an
input instance; the value function is the model's expected prediction when only the
features in coalition $S$ are ``known'' and the remainder are marginalised out. The SHAP
value $\phi_i$ then measures how much feature $i$ pushes the model's prediction above
or below the baseline (the expected prediction over the data distribution).

Three properties make Shapley values particularly appealing for financial explainability:

\begin{enumerate}
  \item \emph{Efficiency}: $\sum_{i=1}^n \phi_i = f(x) - \mathbb{E}[f(X)]$. The SHAP
        values sum to the prediction gap from the baseline, so the explanation accounts
        for the entire prediction.
  \item \emph{Symmetry}: if two features contribute identically in all coalitions, they
        receive equal SHAP values.
  \item \emph{Dummy}: features that never affect the prediction receive zero SHAP value.
\end{enumerate}

For tree-based models, TreeSHAP \cite{lundberg2020local} computes exact Shapley values
in polynomial time. For neural networks---including LLMs---the computation requires
approximation because exact Shapley values have exponential complexity in the number of
features.

Applying SHAP to LLMs requires defining what a ``feature'' is. The natural choice is the
token: each position in the input sequence is treated as a player, and its SHAP value
measures its contribution to the output logit or predicted probability. The challenge is
that marginalising out a token---replacing it with its expected value---is not
straightforward for discrete text: the expectation must be taken over the marginal
distribution of tokens given the context, which requires a language model itself. In
practice, the token is either replaced by a mask token, by a randomly sampled token from
the vocabulary, or by sampling from a smaller reference language model.

\begin{example}[SHAP Attribution for a Credit Decision]
\label{ex:shap-credit}
Consider a fine-tuned BERT model that classifies short business descriptions as
``creditworthy'' or ``not creditworthy.'' The input is:
\begin{quote}
``Our restaurant operates in a highly competitive market and has struggled with declining
footfall since the pandemic. Revenue is down 30\% year-on-year.''
\end{quote}
The model assigns probability 0.14 to the creditworthy class (baseline: 0.50 from the
training data balance). Running a KernelSHAP approximation with 500 samples identifies
the following top attributions:

\begin{center}
\begin{tabular}{lr}
\hline
Token & SHAP value \\
\hline
``struggled'' & $-0.089$ \\
``declining'' & $-0.073$ \\
``down 30\%'' & $-0.061$ \\
``competitive'' & $-0.047$ \\
``restaurant'' & $-0.038$ \\
``operates'' & $+0.012$ \\
``pandemic'' & $-0.029$ \\
\hline
\end{tabular}
\end{center}

The negative tokens account for most of the $0.50 - 0.14 = 0.36$ prediction gap. This
attribution can be translated into a partial ECOA adverse action statement: ``Your
application was primarily affected by language indicating declining revenue, operational
struggle, and market difficulties.'' An auditor can verify that the identified tokens
are conceptually consistent with creditworthiness concerns. The numerical attribution
values provide evidence that the model is reasoning from economically relevant signals
rather than from protected-class proxies.
\end{example}

One important limitation deserves mention. SHAP values, even when computed exactly,
measure correlational feature importance within the model's computation---they do not
identify causal relationships between features and the real-world outcome. If the model
has learned a spurious correlation (e.g., certain industry-specific jargon is correlated
with protected-class membership in the training data), SHAP will faithfully report that
correlation as a driver of predictions, without flagging it as problematic. Fairness
auditing requires a separate step beyond attribution \cite{mehrabi2021survey}.

\subsection{LIME for Local, Model-Agnostic Explanations}
\label{sec:ch12-lime}

LIME (Local Interpretable Model-Agnostic Explanations) takes a different approach to the
same problem \cite{ribeiro2016lime}. Rather than decomposing the model's prediction into
additive feature contributions, LIME fits a simple interpretable surrogate model---
typically a sparse linear model---in the neighbourhood of the instance being explained.

The key intuition is that while a complex model $f$ may be globally non-linear, it is
locally approximately linear in a small region around any given input $x$. LIME exploits
this by:
\begin{enumerate}
  \item Generating a set of perturbed inputs $z_1, \ldots, z_K$ in the neighbourhood of
        $x$ by randomly removing or masking features.
  \item Evaluating the model $f(z_k)$ on each perturbation.
  \item Fitting a sparse linear model $g$ on the perturbations $(z_k, f(z_k))$, weighted
        by proximity to $x$ according to a kernel function $\pi_x(z)$.
\end{enumerate}

\begin{definition}[LIME Objective]
\label{def:lime-objective}
The LIME explanation for input $x$ is
\begin{equation}
\label{eq:lime-objective}
\xi(x) \;=\; \arg\min_{g \in \mathcal{G}} \;
\mathcal{L}(f, g, \pi_x) + \Omega(g),
\end{equation}
where $\mathcal{G}$ is a class of interpretable models (e.g., sparse linear classifiers),
$\mathcal{L}(f, g, \pi_x) = \sum_k \pi_x(z_k)\bigl(f(z_k) - g(z_k)\bigr)^2$ is a
locality-weighted fidelity loss, and $\Omega(g)$ is a regularisation term penalising
model complexity (typically the $\ell_0$ norm of the coefficient vector, controlling
sparsity).
\end{definition}

For text inputs, LIME treats contiguous segments (words or short phrases) as the
interpretable features. A segment is either ``present'' (the original text) or ``absent''
(replaced by the empty string or a padding token). The perturbation space is the set of
binary vectors indicating which segments are present.

LIME has practical advantages over SHAP in the LLM setting. It is truly model-agnostic:
it requires only the ability to query $f$ on new inputs, with no access to gradients or
internal representations. This makes it applicable to closed-source API-based models
such as GPT-4 or Claude, where internal activations are inaccessible. The main
disadvantage is instability: the stochastic sampling of perturbations means that running
LIME twice on the same input may produce meaningfully different explanations
\cite{arrieta2020explainable}. For audit purposes, this instability is problematic; a
deterministic seed should be fixed for reproducibility.

\subsection{Attention Visualization: Promise and Limitations}
\label{sec:ch12-attention-visualization}

The attention mechanism \cite{vaswani2017attention} assigns, for each query position $i$
and key position $j$, a weight $\alpha_{ij}$ that determines how much representation $j$
contributes to the output at position $i$. These weights are normalised to sum to one
and are therefore tempting to read as a kind of explanation: position $j$ is ``attended
to'' by position $i$ with weight $\alpha_{ij}$. Early work on BERT interpretability
explored this idea systematically \cite{clark2019bert}, identifying heads that
appeared to attend to syntactic relations such as subject-verb agreement.

The financial practitioner community has also adopted attention visualisation as a quick
explainability tool, often colouring input tokens by their attention weight to indicate
which parts of a document ``drove'' a classification. Unfortunately, the theoretical
and empirical underpinnings for this interpretation are weak.

\citet{jain2019attention} showed that, across a range of NLP classification tasks,
alternative attention distributions that produce identical predictions can be constructed
by adversarial perturbation. This means the attention weights are not a unique
explanation: different attention patterns may yield the same output, so knowing the
pattern does not tell you what caused the output. \citet{wiegreffe2019attention}
countered that attention can function as explanation in a weaker, diagnostic sense---it
does not causally explain model behaviour, but it correlates with explanations obtained
by other means and can be a useful diagnostic.

The resolution, formalised by \citet{jacovi2020towards}, draws on a distinction between
two desiderata for explanations:

\begin{definition}[Faithfulness and Plausibility]
\label{def:faithfulness-plausibility}
An explanation is \emph{faithful} if it accurately reflects the causal process by which
the model produced its output---that is, if highlighting a feature as important correctly
identifies that feature as having caused (in a manipulationist sense) the output.
An explanation is \emph{plausible} if it looks convincing to human evaluators as a
reasonable justification for the prediction, regardless of whether it reflects the actual
computation.
\end{definition}

Attention visualisations can be plausible---they often highlight tokens that a human
expert would agree are relevant---while failing faithfulness. The consequence for finance
is severe: a plausible but unfaithful explanation can mislead a human auditor into
certifying a model decision as well-reasoned when the actual drivers of the prediction
are different from those highlighted.

Gradient-based attribution methods, such as Integrated Gradients \cite{sundararajan2020shapley},
provide stronger faithfulness guarantees by measuring how much the output changes as a
function of infinitesimal changes to each input. While they require access to model
gradients (ruling out closed-source APIs), they are generally preferred over attention
weights for audit-grade explanation.

% ---------------------------------------------------------------
\section{LLM-Native Explainability Mechanisms}
\label{sec:ch12-llm-native}

The methods discussed in Section~\ref{sec:ch12-classical-xai} treat the LLM as a black
box and apply post-hoc explanations. A complementary approach exploits the generative
capabilities of LLMs themselves to produce natural language explanations. These
\emph{LLM-native} mechanisms are arguably better suited to the regulatory requirement
for plain-language justifications, though they carry their own faithfulness risks.

\subsection{Chain-of-Thought Reasoning as an Explanation Trace}
\label{sec:ch12-chain-of-thought}

Chain-of-thought (CoT) prompting \cite{wei2022chain} elicits an explicit reasoning trace
from the model before it produces a final answer. Instead of asking the model to classify
a credit application directly, one appends to the prompt instructions such as ``think
step by step'' or a few-shot example that demonstrates a desirable reasoning pattern. The
model then generates a series of intermediate statements---premises, inferences,
conclusions---before outputting the decision.

In a financial explainability context, CoT serves a dual function. First, it can improve
decision quality: models that reason through intermediate steps tend to produce more
accurate outputs on tasks requiring multi-step inference, such as evaluating whether a
debt restructuring covenant has been breached. Second, the reasoning trace itself
constitutes a natural language explanation that can be delivered to the applicant,
reviewed by the loan officer, and archived for regulatory audit.

The key limitation is faithfulness. A language model generating a chain of thought may
produce a sequence of plausible-sounding reasoning steps that do not reflect the actual
computation---recall that the model's underlying inference is not sequential reasoning
but parallel matrix computation over learned representations. Empirical studies
\cite{wang2022selfconsistency} have found that CoT explanations become more reliable when
self-consistency is enforced: one samples multiple independent CoT traces and identifies
the reasoning pattern that leads to the majority answer, which has been shown to produce
more accurate outputs and more consistent reasoning patterns.

\begin{definition}[Explanation Trace]
\label{def:explanation-trace}
An \emph{explanation trace} is a sequence of natural language statements
$r_1, r_2, \ldots, r_T$ generated by the model prior to its final output $\hat{y}$,
such that each $r_t$ is a logical step that follows from previous steps and the input
$x$, and together $r_1, \ldots, r_T$ entail $\hat{y}$. A trace is \emph{faithful} if
replacing any $r_t$ with its negation causes the model to change $\hat{y}$ accordingly.
\end{definition}

Checking the faithfulness of explanation traces in practice is non-trivial. One approach,
proposed for the self-consistency method \cite{wang2022selfconsistency}, is to generate
multiple traces independently and check whether the traces that lead to different final
answers are detectable by their structure, thus identifying cases where the explanation
and decision are decoupled.

\subsection{Counterfactual Explanations: What Would Change the Decision?}
\label{sec:ch12-counterfactual}

Counterfactual explanations answer a fundamentally different question from feature
attributions: not ``what drove the current decision?'' but ``what would have to change
for the decision to be different?'' \cite{wachter2017counterfactual}. In a credit context
the question becomes: what is the closest possible state of affairs, in terms of
applicant characteristics, in which the application would have been approved?

Counterfactual explanations are actionable by construction. Telling an applicant that
their debt-to-income ratio of 54\% was a key negative factor is informative, but telling
them that a ratio below 45\% would likely lead to approval under otherwise identical
conditions gives them a concrete path to an improved outcome. This actionability aligns
closely with what regulators and consumer advocates consider a genuinely useful adverse
action notification.

\begin{definition}[Counterfactual Explanation]
\label{def:counterfactual-explanation}
Given a model $f: \mathcal{X} \to \mathcal{Y}$, an input $x \in \mathcal{X}$, and a
desired output class $c \ne f(x)$, a counterfactual explanation is an input
$x' \in \mathcal{X}$ such that:
\begin{align}
\label{eq:counterfactual}
x' &= \arg\min_{z \in \mathcal{X}} d(x, z) \quad \text{subject to} \quad f(z) = c,
\end{align}
where $d$ is a distance metric on the input space that reflects the practical cost of
the change. For structured numerical inputs, $d$ is typically a weighted $\ell_1$ or
$\ell_2$ norm; for text, it may be the minimum number of token substitutions required.
\end{definition}

For LLMs processing text inputs, generating counterfactual explanations is challenging
because the input space is discrete. Gradient-based optimisation cannot directly minimise
\eqref{eq:counterfactual}; instead, one typically uses a two-stage approach: first
identify the feature dimensions (token positions, or semantic dimensions in embedding
space) along which small perturbations change the prediction; then articulate those
perturbations in natural language. LLMs themselves can be used as generators of
counterfactuals---prompted to describe a minimally different version of the applicant's
situation that would receive the opposite decision.

\begin{remark}
The proximity constraint in Definition~\ref{def:counterfactual-explanation} requires
careful calibration. A counterfactual that is technically closest in input space may be
practically impossible (e.g., ``your application would have been approved if your age
were different'') or legally impermissible (since age is a protected characteristic
under ECOA). Practitioners must augment the optimisation with feasibility and
non-discrimination constraints.
\end{remark}

\subsection{Natural Language Justifications for Automated Decisions}
\label{sec:ch12-nl-justifications}

Beyond reasoning traces and counterfactuals, LLMs can be used as a post-processing layer
that synthesises attributions and counterfactuals into a polished natural language
justification. The pipeline is as follows: (1) a primary model---possibly an LLM,
possibly a traditional model---makes the decision; (2) an attribution method such as SHAP
identifies the top contributing features; (3) an LLM is prompted to translate those
attributions into a coherent, plain-language explanation suitable for the target audience.

This separation of concerns is important. The primary model is evaluated for predictive
accuracy; the attribution method is evaluated for faithfulness to that model; and the
LLM generator is evaluated for clarity, completeness, and regulatory compliance. Each
component can be independently tested and replaced. \citet{TatsatShater2025} extend this picture by demonstrating how mechanistic
interpretability techniques---specifically the analysis of attention heads and circuits
within transformer architectures---can be applied directly to financial LLMs, making
it possible to trace model behaviour in tasks ranging from sentiment analysis to
regulatory compliance without relying solely on post-hoc attribution wrappers. Building
on this, \citet{CetintavEtAl2024} demonstrate a complementary
interface layer in which GPT-4o is used to translate complex SHAP outputs and XAI
visualisations into natural-language explanations accessible to non-expert users,
illustrating how the chain-of-thought generation step can simultaneously improve
explanation fidelity and audience accessibility.

A critical consideration is that the generating LLM must not be allowed to hallucinate
attributions. If the primary model indicated that ``revenue decline'' was the key factor,
the generating LLM must not spontaneously add ``poor personal credit history'' to the
explanation, even if such a statement would sound plausible. Architecturally, this
requires that the generator receive the attribution scores as a structured input and be
constrained---through prompt engineering or constrained decoding---to refer only to the
attributions provided.

% ---------------------------------------------------------------
\section{Case Studies in Financial Explainability}
\label{sec:ch12-case-studies}

The preceding sections developed the theoretical foundations. We now examine three applied
case studies that illustrate how explainability requirements manifest in concrete financial
workflows and how LLM-based pipelines can be designed to satisfy them.

\subsection{Credit Scoring: Explaining Rejection to Applicants}
\label{sec:ch12-credit-scoring}

Credit scoring is the canonical application domain for financial XAI. The prediction
problem is binary: will this borrower default within a specified horizon? The input is
typically a mix of structured financial variables---income, debt levels, payment history,
credit utilisation---and, increasingly, unstructured text such as the stated purpose of
the loan, supporting documentation, and social media signals (where legally permissible).

When an LLM contributes to this decision---for instance, by encoding a free-text business
description to supplement a traditional scorecard---the explainability challenge involves
explaining the contribution of the unstructured component. The practitioner must
communicate, in four to five phrases, the key reasons for rejection, drawing on evidence
from both the structured features and the text.

\begin{example}[Hybrid Credit Scoring Explanation Pipeline]
\label{ex:hybrid-credit-pipeline}
A fintech lender uses the following three-stage pipeline for SME loan decisions:

\textbf{Stage 1 (Structured Features):} A gradient-boosted tree scores the application
on 42 structured financial variables. TreeSHAP produces exact Shapley values for each
variable.

\textbf{Stage 2 (Unstructured Text):} A fine-tuned DistilBERT model encodes the
applicant's business description. KernelSHAP with 1,000 perturbation samples produces
token-level attributions. The top five negative tokens are extracted.

\textbf{Stage 3 (Adverse Action Generation):} A GPT-4o model is prompted with the
following structured input:

\begin{verbatim}
You are a credit officer drafting a regulatory-compliant adverse
action notice. The application was declined. The primary reasons,
in order of importance, are:

1. Debt-to-income ratio: 54% (threshold: 45%) — SHAP: -0.18
2. Revenue trend: -28% year-on-year — SHAP: -0.12
3. Business description mentions "struggling", "declining",
   "competitive pressure" — text SHAP: -0.09
4. Cash reserve: 0.8 months operating expenses (threshold: 3) — SHAP: -0.07
5. Industry: Hospitality (elevated default rate sector) — SHAP: -0.05

Draft an adverse action notice that explains these reasons in
plain English. Do not introduce any factors not listed above.
Do not mention the applicant's protected characteristics.
Limit to 150 words.
\end{verbatim}

The output is a compliant adverse action notice suitable for delivery to the applicant.
The pipeline is auditable at every stage: the tree's SHAP values are deterministic,
the text SHAP is reproducible given the random seed, and the LLM output is prompted in
a way that constrains it to the provided attributions.
\end{example}

This pipeline illustrates a general principle: LLMs are most safely deployed in the
explanation layer when their inputs are tightly constrained by upstream attribution
methods. The LLM adds value by transforming structured attributions into natural
language; it should not be permitted to exercise substantive credit judgement at this
stage. \citet{Schmitt2024} provides empirical support for the complementarity of
automated machine learning and SHAP in credit scoring, demonstrating that combining
AutoML with post-hoc Shapley attribution improves both predictive accuracy and model
transparency, and that the resulting explanations satisfy interpretability requirements
under regulatory review without sacrificing performance relative to uninterpreted black-box
baselines.

\subsection{Loan Denial Letters Compliant with ECOA/FCRA}
\label{sec:ch12-loan-denial-letters}

Regulation B, implementing ECOA, specifies that adverse action notices must be provided
to applicants within 30 days of a credit decision and must include: (1) the specific
reasons for the adverse action, (2) the applicant's right to a statement of reasons, and
(3) notice of the applicant's right to a free copy of their consumer report if the
decision was based on it \cite{cfpb2013ecoa}. The reasons must be ``the principal reasons
for the adverse action''---typically interpreted as four to five reasons, stated
specifically rather than generally.

There are two distinct failure modes that LLM-generated denial letters must avoid.

The first is \emph{vagueness}. Regulation~B's specificity requirement means that denial letters stating reasons in overly general terms---such as \emph{`your credit profile did not meet our criteria'}---may be insufficient~\citep{cfpb2013ecoa}. An LLM generation step must produce concrete,
quantified reasons: ``your debt-to-income ratio of 54\% exceeded our maximum of 45\%''
is compliant; ``your financial situation presented concerns'' is not.

The second is \emph{impermissible factor exposure}. ECOA prohibits credit decisions based
on race, colour, religion, national origin, sex, marital status, or age. An LLM generation
step that has access to demographic signals in the text may inadvertently construct
explanations that reference, or could be construed as referencing, these protected
characteristics. Robust pipelines should include a post-generation filtering step that
scans the output for protected-characteristic language and flags any output that contains
such language for human review before dispatch.

A practical implementation requires:
\begin{enumerate}
  \item A \emph{structured attribution feed} containing only non-protected variables and
        their associated attribution scores.
  \item A \emph{system prompt} that explicitly prohibits the LLM from introducing any
        factors not in the feed and from referencing protected characteristics.
  \item A \emph{post-generation compliance check} using a classifier or keyword filter
        trained on adverse-action compliance requirements.
  \item A \emph{human review queue} for any letter flagged by the compliance check.
\end{enumerate}

\citet{Desai2024} offers a formal grounding for this architecture by developing a
mathematical framework that quantifies explainability via Shapley values and
information-theoretic measures as an optimisation constraint, applying it across five
regulatory audit categories including BSA/AML and Consumer Compliance, and demonstrating
that institutions can satisfy CAMELS-style examination requirements by embedding
explainability as a design objective rather than a post-hoc annotation. \citet{Lakarasu2024} operationalises this further by proposing an XAI-prompted compliance
platform for financial transactions that automatically generates data-analytic compliance
documentation with feature attributions, directly addressing the EU AI Act's Article 13
transparency obligations and AML reporting requirements.

\subsection{Investment Recommendation Disclosures under MiFID II}
\label{sec:ch12-mifid-disclosures}

MiFID II (Markets in Financial Instruments Directive II) and its delegated regulation
require that investment firms provide retail clients with a suitability assessment
explaining why a recommended product is appropriate for that client's investment
objectives, risk tolerance, financial situation, and knowledge and experience
\cite{esma2018mifid2}. This suitability report must be provided before the transaction
is executed and must be sufficiently detailed that the client can understand the
rationale.

LLMs are being explored as engines for automated suitability report generation. The
challenge is that a MiFID II suitability report must reflect the specific
characteristics of both the client and the product---it cannot be a generic template.
The relevant explainability question is: given a client profile and a recommended product,
can an LLM generate a personalised suitability narrative that is accurate,
non-misleading, and grounded in the disclosed client characteristics?

The CFA Institute \cite{cfainstitute2025xai} identified three key requirements for compliant AI-generated suitability disclosures:
(1) the reasoning must be traceable to the underlying client data on file; (2) the
language must not overstate expected returns or understate risks; and (3) the document
must be auditable in the sense that, given the client data and the model configuration,
a third party could reproduce a materially equivalent document.

\begin{example}[MiFID II Suitability Disclosure Generation]
\label{ex:mifid-disclosure}
A robo-adviser platform needs to generate personalised suitability disclosures for
$n = 12{,}000$ clients at the point of annual portfolio review. Each client has a
structured profile including risk questionnaire responses, investment horizon, liquidity
needs, and existing portfolio composition. The recommended portfolio is a risk-adjusted
mix of equity and fixed-income ETFs.

The generation pipeline:
\begin{itemize}
  \item Structured client features are passed to a scoring model that produces a risk
        tier and a suitability score for the recommended portfolio.
  \item Integrated Gradients are computed to identify which client-profile features most
        strongly support the recommendation.
  \item An LLM is prompted with the client tier, the product's risk profile, and the top
        three attribution drivers. The prompt explicitly instructs the model to: use the
        client's stated investment horizon and risk tolerance as the primary justification;
        acknowledge relevant risks including capital loss; and not mention any expected
        return figure not contained in the product's Key Information Document.
  \item The output is passed through a regulatory compliance filter that checks for
        prohibited language (guaranteed returns, excessive return projections, missing risk
        disclosures).
\end{itemize}

A compliance officer reviews a random 2\% sample of generated disclosures each quarter.
For illustration, suppose the review identified a rate of outputs requiring revision of around 0.3\%,
all attributable to edge cases in the prompt where the client's stated
objective and the recommended product's objectives were in mild tension.
\end{example}

% ---------------------------------------------------------------
\section{Evaluating the Quality of Explanations}
\label{sec:ch12-evaluating-quality}

Having constructed explanations, practitioners must evaluate whether those explanations
are good. This is non-trivial: unlike predictions, where one can simply compare the
model's output to a ground truth label, there is no ground-truth explanation against
which to benchmark. The evaluation literature has developed three complementary
evaluation paradigms: faithfulness metrics, completeness and plausibility metrics, and
human-subject studies \cite{nauta2023anecdotal}.

\subsection{Faithfulness: Does the Explanation Reflect the Model?}
\label{sec:ch12-faithfulness}

Faithfulness measures whether the explanation accurately describes the model's actual
computation. The most common operationalisation is the \emph{sufficiency test}: if the
explanation claims that features $S$ are responsible for the prediction, then a model
given only features $S$ should produce approximately the same prediction.

Formally, let $e(x)$ be the set of features identified as important by the explanation
for input $x$, with complement $\bar{e}(x)$. The sufficiency of the explanation is
\begin{equation}
\label{eq:sufficiency}
\text{Suff}(e, f, x) \;=\; 1 - |f(x) - f(x_{e(x)})|,
\end{equation}
where $x_{e(x)}$ denotes the input $x$ with all features not in $e(x)$ marginalised out.
A sufficiency of 1 indicates perfect faithfulness; lower values indicate that the
explanation omits features that are actually important to the prediction.

A complementary measure is \emph{comprehensiveness}: if the explanation claims that
features $S$ are important, then a model given all features \emph{except} $S$ should
produce a substantially different prediction.
\begin{equation}
\label{eq:comprehensiveness}
\text{Comp}(e, f, x) \;=\; |f(x) - f(x_{\bar{e}(x)})|.
\end{equation}
High comprehensiveness means that removing the explained features significantly degrades
the prediction, confirming that those features genuinely drove the output.

SHAP values, by construction, satisfy the efficiency property (equation~\eqref{eq:shapley-value}),
which implies a form of sufficiency for the full feature set. LIME explanations are
trained to maximise local fidelity but may fail sufficiency because the surrogate model
is deliberately sparse---it may omit features that matter but are correlated with the
included features. Attention-based explanations typically fail comprehensiveness: one can
often ablate the attended tokens without changing the prediction, precisely because the
model can route information through alternative paths.

\subsection{Completeness and Plausibility Metrics}
\label{sec:ch12-completeness-plausibility}

Beyond faithfulness, two further dimensions matter for financial explanations.

\emph{Completeness} refers to whether the explanation accounts for all relevant drivers
of the prediction, not only the single most important one. An ECOA adverse action notice
that identifies only one reason when five significant reasons exist is technically
incomplete, and a regulator might question whether the lender is obscuring relevant
factors. Completeness can be measured as the fraction of the total prediction gap (from
baseline to final prediction) attributable to the explanations provided; for SHAP, this
is directly measurable as $\sum_{i \in e(x)} \phi_i / (f(x) - \mathbb{E}[f(X)])$.

\emph{Plausibility} is the degree to which a domain expert would agree that the
explanation makes sense given their knowledge of the domain. A credit model explanation
that attributes a rejection to a high debt-to-income ratio is plausible; one that
attributes it to the word ``the'' appearing in the application is implausible, even if
technically accurate. Plausibility correlates with but is not identical to faithfulness:
a plausible explanation may be faithful or unfaithful; an implausible explanation is
likely, but not necessarily, unfaithful.

Quantitative plausibility metrics include agreement with a ``ground truth'' explanation
set annotated by domain experts, or correlation with the outputs of a reference XAI
method known to be faithful on the domain. For text inputs, readability metrics can
capture whether the explanation is sufficiently comprehensible for the target audience,
which is relevant for ECOA compliance.

\subsection{Human-Subject Studies with Financial Domain Experts}
\label{sec:ch12-human-subject-studies}

The ultimate test of an explanation's quality, at least for regulatory purposes, is
whether a qualified human expert can use it to make appropriate decisions. Human-subject
evaluation has three standard designs for the explanation literature \cite{doshivelez2017rigorous}:

\emph{Forward simulation}: participants are given a model explanation and asked to predict
what the model would output for a held-out input. If the explanation is faithful and
complete, participants should be able to simulate the model accurately.

\emph{Decision support}: participants perform a decision task---for example, approving or
rejecting credit applications---first without model assistance, then with model
predictions and explanations. The quality of the explanation is measured by the degree
to which it helps participants catch model errors and override incorrect predictions.

\emph{Trust calibration}: participants are asked to rate their trust in model predictions
before and after seeing explanations. A good explanation should increase trust on correct
predictions and decrease trust on incorrect ones, calibrating the human's reliance on
the model.

In the financial domain, human-subject studies face practical constraints. Qualified
credit officers and compliance specialists are expensive participants; their time cannot
easily be allocated to academic experiments. A practical alternative is the use of
\emph{vignette studies} in which realistic but synthetic decision scenarios are constructed
and submitted to practitioners recruited via professional networks. Research consistent with CFA Institute guidance~\cite{cfainstitute2025xai} suggests that explanations may reduce overreliance on AI recommendations, particularly in high-uncertainty conditions.

A persistent challenge is that human experts often find plausible-but-unfaithful
explanations just as satisfying as faithful ones. This implies that human evaluation
alone is not a reliable quality gate: it must be complemented by the quantitative
faithfulness metrics discussed in Section~\ref{sec:ch12-faithfulness}.

% ---------------------------------------------------------------
\section{Summary}
\label{sec:ch12-summary}

This chapter has examined the explainability challenge that arises when large language
models are deployed in regulated financial contexts. Five takeaways are worth
highlighting.

First, the demand for explainability in finance is not principally an ethical aspiration
but a legal requirement encoded in multiple overlapping regulatory frameworks. The EU
AI Act classifies financial applications as high-risk and mandates human oversight and
interpretability; SR 11-7 requires that model outputs be explainable to non-technical
risk managers; ECOA requires specific plain-language reasons for adverse credit actions;
and MiFID II requires personalised suitability disclosures grounded in disclosed client
characteristics.

Second, classical XAI methods---SHAP and LIME in particular---can be adapted to LLMs,
but adaptation is not trivial. SHAP requires that features be defined in the input space
(tokens, phrases, or semantic segments) and that the marginalisation distribution be
specified. LIME requires that the neighbourhood and perturbation scheme be calibrated for
the text domain. Both methods are subject to instability, and their outputs should be
tested for faithfulness before use in regulatory contexts.

Third, attention visualisation, while intuitively appealing, does not in general provide
faithful explanations. Alternative attention distributions can produce identical outputs,
and ablating attended tokens often does not change predictions. Gradient-based attribution
methods provide stronger guarantees at the cost of requiring access to model internals.

Fourth, LLMs themselves can generate chain-of-thought reasoning traces, counterfactual
explanations, and natural language justifications. These LLM-native mechanisms are
well-suited to producing the plain-language output that regulators require, but their
faithfulness cannot be taken for granted. Robust pipelines constrain the generating LLM
to draw only from upstream attribution outputs and include compliance filters that screen
for prohibited language and protected-characteristic references.

Fifth, explanation quality should be evaluated on multiple dimensions: faithfulness,
completeness, plausibility, and human utility. No single metric captures all dimensions,
and no evaluation is fully satisfactory without complementary quantitative and
human-subject components. The field is young; practitioners should expect both the
metrics and the methods to continue evolving rapidly.

% ---------------------------------------------------------------
\section{Further Reading}
\label{sec:ch12-further-reading}

The theoretical foundations of feature attribution originate in cooperative game theory.
\citet{shapley1953value} established the axiomatic uniqueness of the Shapley value, and
\citet{lundberg2017unified} showed how this framework unifies existing feature importance
methods under a single additive attribution formalism. For readers wishing to understand
the approximation methods required to apply SHAP to neural networks, the survey in
\citet{lundberg2020local} is the natural starting point.

For LIME, the original paper by \citet{ribeiro2016lime} remains highly readable and
includes worked examples across text, image, and tabular domains. The evaluation
framework developed in \citet{doshivelez2017rigorous} provides a principled vocabulary
for comparing XAI methods and is essential reading before designing any human-subject
evaluation.

The counterfactual explanation literature was founded by \citet{wachter2017counterfactual},
which also addresses the legal context of GDPR Article 22 in depth. \citet{miller2019explanation}
places the technical XAI literature in the context of the social science literature on
explanation and is particularly valuable for understanding what human users actually want
from explanations.

For the attention-as-explanation debate, both \citet{jain2019attention} and
\citet{wiegreffe2019attention} should be read together. The resolution by
\citet{jacovi2020towards} formalising the faithfulness-plausibility distinction provides
the conceptual clarity that the debate was lacking.

On the regulatory side, the EU AI Act \cite{euaiact2024} should be read alongside the
ESMA guidelines on MiFID II suitability \cite{esma2018mifid2} and the Federal Reserve's
SR 11-7 guidance \cite{sr117} for a complete picture of the compliance landscape. The
CFA Institute report \cite{cfainstitute2025xai} is the most practically oriented survey
of industry explainability practices in investment management. The comprehensive survey
by \citet{arrieta2020explainable} covers the full taxonomy of XAI methods across all
input modalities.

Within this book, Chapter~\ref{ch:credit-risk} covers the credit scoring context in more
depth, including probability calibration and model validation methods that are
prerequisites for the SHAP pipelines described here. Chapter~\ref{ch:llm-agents} discusses
chain-of-thought reasoning as an agent planning mechanism, which complements the
explanation trace perspective developed in this chapter. Chapter~\ref{ch:responsible-llms}
addresses fairness and bias issues that are closely related to the non-discrimination
requirements of ECOA.

% ---------------------------------------------------------------
\section{Data Lab}\label{sec:ch12-data-lab}

The following exercises use live data from the SEC EDGAR REST API
(\url{https://data.sec.gov}). Starter code is in
\texttt{code/notebooks/12-xai-explainability/exercises.ipynb}, Data Lab section.

\begin{exercise}[Logistic Regression Coefficients]\label{ex:ch12-edgar-b}
\textbf{[B]} Build a logistic regression model (implemented in NumPy) to
predict whether a company's next-year stock return beats the S\&P~500.
Features: leverage, ROE, and revenue growth (computed from EDGAR XBRL).
(\textit{a})~Train on ten companies over three years of data.
(\textit{b})~Print the coefficient vector and identify which feature has the
largest absolute weight.
See \texttt{code/notebooks/12-xai-explainability/exercises.ipynb}, Data Lab section.
\end{exercise}

\begin{exercise}[Manual SHAP-Style Feature Attribution]\label{ex:ch12-edgar-i}
\textbf{[I]} Implement leave-one-feature-out attribution (a simplified SHAP
method): for each test company, compute the predicted probability with all
features, then with each feature set to the training mean.
Attribution for feature $j$ $= P(\text{all features}) - P(\text{without }j)$.
(\textit{a})~Compute attributions for three companies.
(\textit{b})~Display a waterfall chart for one company.
See \texttt{code/notebooks/12-xai-explainability/exercises.ipynb}, Data Lab section.
\end{exercise}

\begin{exercise}[Model Card Design]\label{ex:ch12-edgar-a}
\textbf{[A]} Write a structured model card (as a Python dictionary printed as
formatted text) for the logistic regression trained in
Exercises~\ref{ex:ch12-edgar-b}--\ref{ex:ch12-edgar-i}.
Include the fields: \texttt{model\_type}, \texttt{training\_data},
\texttt{features} (with XBRL concept names), \texttt{performance}
(LOO-CV accuracy), \texttt{known\_limitations} (at least three),
\texttt{intended\_use}, \texttt{out\_of\_scope\_use}.
Discuss what additional validation a financial regulator (e.g.\ under
SR~11-7 guidance) would require before deployment.
See \texttt{code/notebooks/12-xai-explainability/exercises.ipynb}, Data Lab section.
\end{exercise}
