\chapter{Portfolio Optimization and Quantitative Trading with LLMs}
\label{ch:portfolio-quant-trading}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Derive the mean-variance frontier and explain why estimation error
        limits its practical usefulness.
  \item Describe how the Black-Litterman model combines equilibrium returns with
        investor views, and identify where LLM-generated views can enter that framework.
  \item Construct LLM-based factors for cross-sectional asset pricing and assess their
        marginal explanatory power beyond classical factor models.
  \item Design and critically evaluate an LLM sentiment signal pipeline for
        algorithmic trading, including realistic transaction-cost adjustments.
  \item Implement a walk-forward backtest and identify common pitfalls such as
        look-ahead bias, survivorship bias, and multiple-testing inflation.
  \item Apply CVaR-based risk management and connect textual risk signals to
        tail-loss estimation.
\end{enumerate}
\end{remark}

% ============================================================
\section{Classical Portfolio Theory and Its Limits}
\label{sec:ch10-classical-portfolio-theory}
% ============================================================

Portfolio theory attempts to answer a deceptively simple question: given a universe of
risky assets, how should a rational investor allocate wealth across them? The modern
mathematical answer, attributed to \citet{markowitz1952portfolio}, reduces the problem to
a quadratic program whose solution traces the \emph{efficient frontier}---a locus of
portfolios that offer the highest expected return for a given level of volatility.
Decades of refinement have layered more sophisticated machinery on top: Black-Litterman
priors, robust covariance estimators, risk-factor decompositions. Yet a stubborn
empirical finding persists: across many realistic settings, a naive equal-weight portfolio
outperforms the theoretically optimal one \citep{demiguel2009optimal}. This paradox---
that more sophisticated optimization frequently underperforms---is the central motivation
for revisiting portfolio construction with LLM-augmented signals.

% -----------------------------------------------------------
\subsection{Markowitz Mean-Variance Optimization: The Efficient Frontier}
\label{sec:ch10-markowitz-mean-variance}
% -----------------------------------------------------------

Let $\mathbf{r} \in \mathbb{R}^N$ be the random vector of excess returns on $N$ risky
assets over a holding period. Denote by $\boldsymbol{\mu} = \mathbb{E}[\mathbf{r}]$ the
vector of expected excess returns and by $\boldsymbol{\Sigma} = \mathrm{Cov}[\mathbf{r}]$
the $N \times N$ covariance matrix, assumed positive definite. A long-only portfolio is
described by a weight vector $\mathbf{w} \in \mathbb{R}^N$ with $\mathbf{w} \geq \mathbf{0}$
and $\mathbf{1}^\top \mathbf{w} = 1$. The portfolio's expected excess return and variance are
\begin{align}
  \mu_p &= \boldsymbol{\mu}^\top \mathbf{w}, \label{eq:ch10-port-return} \\
  \sigma_p^2 &= \mathbf{w}^\top \boldsymbol{\Sigma}\, \mathbf{w}. \label{eq:ch10-port-var}
\end{align}

\begin{definition}[Mean-Variance Efficient Portfolio]
\label{def:mv-efficient}
A portfolio $\mathbf{w}^*$ is \emph{mean-variance efficient} at target return
$\bar{\mu}$ if it solves
\begin{equation}
  \min_{\mathbf{w}} \; \mathbf{w}^\top \boldsymbol{\Sigma}\, \mathbf{w}
  \quad \text{subject to} \quad
  \boldsymbol{\mu}^\top \mathbf{w} = \bar{\mu}, \quad
  \mathbf{1}^\top \mathbf{w} = 1, \quad \mathbf{w} \geq \mathbf{0}.
  \label{eq:ch10-mv-program}
\end{equation}
The set of all such portfolios as $\bar{\mu}$ varies traces the \emph{efficient frontier}.
\end{definition}

The maximum Sharpe ratio portfolio occupies a special position on the frontier: it
maximises the ratio of expected excess return to volatility,
\begin{equation}
  \mathbf{w}^{\text{SR}} = \argmax_{\mathbf{w}\geq 0,\,\mathbf{1}^\top\mathbf{w}=1}
  \frac{\boldsymbol{\mu}^\top \mathbf{w}}{\sqrt{\mathbf{w}^\top \boldsymbol{\Sigma}\, \mathbf{w}}}.
  \label{eq:ch10-sharpe-max}
\end{equation}
Without the non-negativity constraint the unconstrained maximum Sharpe ratio portfolio has
the closed form $\mathbf{w}^{\text{SR}} \propto \boldsymbol{\Sigma}^{-1}\boldsymbol{\mu}$,
revealing that the solution scales inversely with risk and linearly with the return signal.

The practical problem is that both $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$ must be
estimated from data. Sample estimates of $\boldsymbol{\mu}$ are notoriously noisy: even
with 60 months of returns---a common institutional window---the standard error of a
monthly mean is roughly $\hat{\sigma}/\sqrt{60}$, which for a typical stock implies an
uncertainty on the order of the expected return itself. The sample covariance matrix
presents a separate challenge when $N$ is large relative to $T$: it becomes singular or
nearly singular, and its inverse amplifies estimation errors. A portfolio optimizer that
receives noisy inputs will allocate heavily to recently lucky assets, producing
concentrated, unstable positions that perform poorly out of sample.

Several remedies exist. \emph{Shrinkage estimators} \citep{demiguel2009optimal} pull the
sample covariance toward a structured target---often the identity or a factor model---
reducing variance at the cost of some bias. \emph{Robust optimization} replaces point
estimates with uncertainty sets, seeking portfolios that perform acceptably across the
set. \emph{Factor models} impose a low-rank structure:
$\boldsymbol{\Sigma} = \mathbf{B}\mathbf{F}\mathbf{B}^\top + \mathbf{D}$, where
$\mathbf{B}$ is the factor loading matrix, $\mathbf{F}$ is the factor covariance, and
$\mathbf{D}$ is a diagonal idiosyncratic variance matrix. Reducing estimation to $K$
factors rather than $N(N+1)/2$ entries dramatically improves stability. The celebrated
Fama-French three-factor model \citep{fama1993common} and its five-factor extension
\citep{fama2015five} operationalise this structure for equity portfolios.

Despite these refinements, \citet{demiguel2009optimal} showed in an influential comparison
across 14 empirical datasets that the $1/N$ equal-weight benchmark outperforms sophisticated
optimisers on out-of-sample Sharpe ratios. The intuition is that estimation error in
$\boldsymbol{\mu}$ dominates the gains from covariance-aware weighting unless the signal
is very precise. This finding directly motivates the research agenda of using LLMs:
if a model can extract more precise estimates of $\boldsymbol{\mu}$ from text---earnings
calls, news, analyst reports---the optimization gains that have been theoretically
available but practically elusive may finally materialise.

% -----------------------------------------------------------
\subsection{Black-Litterman Model: Incorporating Views and Priors}
\label{sec:ch10-black-litterman}
% -----------------------------------------------------------

The \emph{Black-Litterman model} \citep{blacklitterman1992} resolves the estimation
problem by Bayesian reasoning: rather than using sample means directly, it starts from a
prior that reflects market equilibrium and updates that prior with investor views.

The equilibrium prior is derived by reverse-engineering implied returns from the market
portfolio. Let $\mathbf{w}^{\text{mkt}}$ denote the market-capitalisation weights and
$\delta$ the risk-aversion coefficient. The implied equilibrium excess-return vector is
\begin{equation}
  \boldsymbol{\Pi} = \delta\, \boldsymbol{\Sigma}\, \mathbf{w}^{\text{mkt}}.
  \label{eq:ch10-bl-pi}
\end{equation}
These equilibrium returns serve as the prior mean: $\boldsymbol{\mu} \sim
\mathcal{N}(\boldsymbol{\Pi}, \tau\boldsymbol{\Sigma})$, where $\tau$ is a small scalar
reflecting uncertainty in the prior.

An investor holds $K$ views on linear combinations of asset returns, expressed as
\begin{equation}
  \mathbf{P}\boldsymbol{\mu} = \mathbf{q} + \boldsymbol{\varepsilon},
  \quad \boldsymbol{\varepsilon} \sim \mathcal{N}(\mathbf{0}, \boldsymbol{\Omega}),
  \label{eq:ch10-bl-views}
\end{equation}
where $\mathbf{P} \in \mathbb{R}^{K \times N}$ picks out assets, $\mathbf{q} \in
\mathbb{R}^K$ states the view values, and $\boldsymbol{\Omega}$ captures view uncertainty.
Applying Bayes' rule yields a posterior mean
\begin{equation}
  \hat{\boldsymbol{\mu}} =
  \bigl[(\tau\boldsymbol{\Sigma})^{-1} + \mathbf{P}^\top\boldsymbol{\Omega}^{-1}\mathbf{P}\bigr]^{-1}
  \bigl[(\tau\boldsymbol{\Sigma})^{-1}\boldsymbol{\Pi} + \mathbf{P}^\top\boldsymbol{\Omega}^{-1}\mathbf{q}\bigr].
  \label{eq:ch10-bl-posterior}
\end{equation}
The posterior mean $\hat{\boldsymbol{\mu}}$ then replaces $\boldsymbol{\mu}$ in the
mean-variance programme~\eqref{eq:ch10-mv-program}. When no views are held,
$\hat{\boldsymbol{\mu}} \to \boldsymbol{\Pi}$ and the optimal portfolio reduces to the
market portfolio---a useful sanity check.

The key virtue of Black-Litterman is that it decouples \emph{where} views are held from
\emph{what} the views are. An investor who has a view on three sectors out of thirty need
only specify those three rows of $\mathbf{P}$; the optimizer handles the rest
consistently. This modularity maps naturally onto LLM-generated signals: an LLM that
processes earnings calls can produce a view $q_k$ for each company it covers, together
with a confidence score that informs $\Omega_{kk}$. Sections~\ref{sec:ch10-analyst-views-extraction}
and~\ref{sec:ch10-llm-factors-fama-french} detail how to operationalise this pipeline.
\citet{MantshimuliMwamba2025} provide direct empirical validation of this architecture:
they aggregate sentiment views from multiple fine-tuned LLMs via an LSTM and feed the
resulting view vector into the Black-Litterman framework, finding that the multi-LLM
approach significantly improves both the Sharpe ratio and annualised return over
single-LLM and market-capitalisation benchmarks. More broadly, \citet{Yuksel2025alpha}
show that an LLM-guided evolutionary tournament can select portfolios from large asset
universes by generating and testing candidate allocation strategies in a closed loop,
offering an empirical baseline for treating LLMs as portfolio optimizers rather than
merely view providers.

\begin{example}[A Single Absolute View in Black-Litterman]
\label{ex:ch10-bl-single-view}
Suppose an LLM processes Apple's quarterly earnings call and outputs a probability of
$0.72$ that Apple will beat consensus EPS estimates next quarter. We translate this into
an \emph{absolute view}: $\mathbf{P} = \mathbf{e}_{\text{AAPL}}^\top$ (a row vector
selecting Apple), $q = 1\%$ (an expected monthly excess return), and $\Omega_{11} =
(0.5\%)^2$ (reflecting moderate confidence).

Before this view, the equilibrium implied return for Apple might be $\Pi_{\text{AAPL}} =
0.3\%$ per month. After applying equation~\eqref{eq:ch10-bl-posterior}, the posterior
return shifts upward, with the magnitude of the shift proportional to the distance
$q - \mathbf{P}\boldsymbol{\Pi}$ relative to the total uncertainty
$\tau \mathbf{P}\boldsymbol{\Sigma}\mathbf{P}^\top + \boldsymbol{\Omega}$. If the view
confidence is high ($\Omega_{11}$ small), the posterior closely matches $q = 1\%$; if
the view confidence is low, the posterior barely deviates from $\Pi_{\text{AAPL}}$.
Intuitively, the BL framework prevents any single noisy signal from dominating the
portfolio.
\end{example}

% -----------------------------------------------------------
\subsection{Where LLMs Can and Cannot Add Value}
\label{sec:ch10-llm-limits}
% -----------------------------------------------------------

The case for using LLMs in portfolio construction rests on a specific hypothesis: that
financial text contains information about future returns that is not yet fully impounded
in prices, and that LLMs can extract that information more accurately than legacy NLP
methods. Both conjuncts deserve scrutiny.

The \emph{information hypothesis} is grounded in the Grossman-Stiglitz paradox
\citep{grossman1980impossibility}: perfectly efficient markets cannot exist because no one would
then have an incentive to collect information. In equilibrium, some amount of
informationally-sensitive trading must earn a return that compensates for research costs.
The empirical question is whether LLM-extracted signals constitute such information. Early
evidence is mixed. \citet{LopezLiraTang2023} found that GPT-based sentiment scores for
firm-specific news predict short-horizon returns with statistically significant
coefficients, but the magnitudes are small and the evidence of economic significance---
after transaction costs---is limited. \citet{tetlock2007giving} showed more than a decade
ago that simple dictionary sentiment from newspaper columns predicts excess returns, so
the incremental value of LLMs over dictionaries is the relevant comparison.

The \emph{extraction hypothesis} is more unambiguous: LLMs understand financial language
better than term-frequency or rule-based methods. They handle negation, hedging, sarcasm,
and domain-specific idiom. A line such as ``while we remain cautiously optimistic about
the macro backdrop, we are managing expenses prudently to protect margins'' contains
nuanced signals that a bag-of-words classifier would misread, whereas a fine-tuned LLM
would correctly identify the defensive tone \citep{araci2019finbert}.

Where LLMs add clear value:
\begin{itemize}
  \item \textbf{Alternative data structuring.} Converting unstructured earnings calls,
    analyst reports, and regulatory filings into structured factors or views at scale.
  \item \textbf{Risk monitoring.} Flagging emerging risks---supply-chain disruptions,
    geopolitical events, management credibility concerns---before they appear in financial
    statements.
  \item \textbf{Analyst view aggregation.} Synthesising heterogeneous analyst
    reports into a single posterior view with calibrated uncertainty.
\end{itemize}

Where LLMs do \emph{not} add clear value:
\begin{itemize}
  \item \textbf{Return prediction at horizons beyond a few days.} At weekly or monthly
    horizons, macroeconomic and factor risks dominate idiosyncratic news signals.
  \item \textbf{Replacing quantitative factor models.} A Fama-French factor model
    captures covariance structure efficiently; LLMs are not designed to estimate
    covariance matrices.
  \item \textbf{Timing decisions in isolation.} LLMs have no privileged access to
    order-flow, price-impact, or microstructure data---all of which are essential for
    execution timing.
\end{itemize}

The most productive framing is that LLMs are \emph{information extraction engines}, not
portfolio optimizers. They produce signals; classical theory dictates how those signals
should be combined into weights. \citet{Mann2024} surveys this landscape in practitioner
terms, documenting how LLMs are transforming quantitative workflows from signal generation
through execution and arguing that the most durable value lies precisely in the
information-extraction layer rather than in the decision layer.

% ============================================================
\section{LLMs as Alternative Data Processors}
\label{sec:ch10-llms-alternative-data}
% ============================================================

Alternative data---any data source beyond traditional price and accounting series---has
become a defining feature of quantitative investing. Satellite imagery, credit-card
transaction flows, web scraping, and social media all provide informational edges that
erode quickly once widely adopted. Text remains the most abundant and least saturated
category: earnings call transcripts, analyst reports, SEC filings, news articles, and
central-bank communications collectively generate terabytes of structured natural language
each year. LLMs can process this at scale, converting prose into numerical signals suitable
for factor models and portfolio construction.

% -----------------------------------------------------------
\subsection{Extracting Analyst Views and Market Expectations from Text}
\label{sec:ch10-analyst-views-extraction}
% -----------------------------------------------------------

Analyst reports occupy an unusual informational position: they aggregate expert
judgment, proprietary channel checks, and management guidance into a single narrative.
Research has long documented that the \emph{tone} of an analyst report---independent of
the explicit rating upgrade or target-price revision---contains incremental information
about future returns \citep{lehavy2011benefit}. LLMs can extract this tone more reliably
than rule-based systems.

A practical extraction pipeline proceeds in four stages. First, the raw document is
retrieved (from a proprietary feed or from SEC EDGAR for public filings) and split into
semantically coherent chunks. Second, an LLM---typically a fine-tuned variant of a general
model or a specialist such as FinBERT \citep{araci2019finbert}---assigns a sentiment score
to each chunk. Third, the chunk-level scores are aggregated (e.g., weighted by sentence
position, since forward-looking paragraphs near the summary carry more weight) into a
document-level signal $s_i \in [-1, 1]$. Fourth, the signal is normalised within
industry and time period to remove sector-level trends.

For the Black-Litterman framework, the mapping from signal to view is
\begin{equation}
  q_i = \alpha \cdot s_i + \Pi_i,
  \quad \Omega_{ii} = \sigma_q^2 \cdot (1 - |s_i|),
  \label{eq:ch10-signal-to-view}
\end{equation}
where $\alpha$ is a calibrated scaling constant (estimated by regressing past signals
against realized excess returns), and the uncertainty $\Omega_{ii}$ is inversely
proportional to signal magnitude: a strong, unambiguous signal ($|s_i| \approx 1$)
warrants low view uncertainty; a neutral signal ($|s_i| \approx 0$) warrants high
uncertainty, effectively allowing the equilibrium prior to dominate.

Beyond earnings calls, LLMs can be applied to longer-horizon strategic documents.
\citet{MeskovskisKenyon2024} introduce a SWOT-based portfolio construction methodology
in which an LLM analyses company strategic filings to generate views for investment
horizons where statistical backtesting power is too limited to validate purely
price-based signals---demonstrating a fundamentally different pathway for generating
LLM views that complements the sentiment pipeline described above.

A practical concern is \emph{information staleness}: an analyst report may be published
after the market has already reacted to management guidance on an earnings call.
Timestamp management is therefore critical. The signal should use only text available
before the close of the trading day on which the portfolio is rebalanced. Systems that
accidentally include after-hours summaries in next-day decisions introduce look-ahead
bias---a topic we return to in Section~\ref{sec:ch10-backtesting-pitfalls}.

\begin{example}[Extracting BL Views from Earnings Call Transcripts]
\label{ex:ch10-earnings-call-bl}
Consider a universe of 100 S\&P 500 companies. Each quarter, earnings calls are
transcribed and chunked by speaker turn. An LLM is prompted as follows:

\begin{quote}
\textit{You are a sell-side equity analyst. Given the following excerpt from an earnings
call, assign a return sentiment score from $-1$ (strongly bearish) to $+1$ (strongly
bullish) for the stock over the next month. Output only the numeric score.}
\end{quote}

For a company whose CFO states ``we are raising our full-year revenue guidance by 5\%
and see no signs of demand softening'', the LLM reliably returns scores in $[0.6, 0.9]$.
For a company whose CEO states ``we are cautious about the macro environment and are
taking steps to right-size our cost structure'', scores cluster in $[-0.6, -0.3]$.

After calibrating $\alpha$ on two years of held-out data, the correlation between
the extracted signal and one-month-ahead excess returns is found to be approximately
$0.08$---statistically significant in a $T = 800$ observation sample, but implying a
modest information ratio of about 0.25. When inserted into a BL optimizer with
$\delta = 3$ and $\tau = 0.05$, the resulting portfolio achieves a Sharpe ratio 15\%
above the equal-weight benchmark before transaction costs.
\end{example}

% -----------------------------------------------------------
\subsection{Constructing LLM-Based Factors for Fama-French Models}
\label{sec:ch10-llm-factors-fama-french}
% -----------------------------------------------------------

Factor models decompose the return of any asset into a systematic component explained by
a small number of common factors and an idiosyncratic residual:
\begin{equation}
  r_{i,t} = \alpha_i + \sum_{k=1}^{K} \beta_{ik} f_{k,t} + \varepsilon_{i,t},
  \label{eq:ch10-factor-model}
\end{equation}
where $f_{k,t}$ are the factor returns and $\beta_{ik}$ are the loadings. Fama and
French \citep{fama1993common} identified three economically motivated factors---market
($\text{MKT}$), size ($\text{SMB}$), and value ($\text{HML}$)---later extended to
five by adding profitability ($\text{RMW}$) and investment ($\text{CMA}$)
\citep{fama2015five}.

LLMs enable the construction of \emph{text-based factors}---long-short portfolios
formed on signals derived from unstructured text. The construction mirrors classical
accounting factors. Define the LLM sentiment score for firm $i$ at time $t$ as $s_{i,t}
\in \mathbb{R}$. A \emph{language factor} (LF) is constructed as follows:

\begin{definition}[Language Factor]
\label{def:language-factor}
Let $s_{i,t}$ denote the cross-sectionally demeaned LLM sentiment score for firm $i$
at rebalancing date $t$. Sort firms into quintiles based on $s_{i,t}$.
The \emph{language factor} is
\begin{equation}
  \text{LF}_t = R_t^{\text{top quintile}} - R_t^{\text{bottom quintile}},
  \label{eq:ch10-language-factor}
\end{equation}
where $R_t^{\text{quintile}}$ denotes the equal-weighted portfolio return in that quintile
over the subsequent holding period.
\end{definition}

The key empirical question is whether LF earns a significant alpha after controlling
for classical factors. If the regression
\begin{equation}
  \text{LF}_t = a + \beta_{\text{MKT}} \text{MKT}_t + \beta_{\text{SMB}} \text{SMB}_t
  + \beta_{\text{HML}} \text{HML}_t + \beta_{\text{RMW}} \text{RMW}_t
  + \beta_{\text{CMA}} \text{CMA}_t + \varepsilon_t
  \label{eq:ch10-lf-alpha-regression}
\end{equation}
yields a statistically significant $\hat{a} > 0$, then LF captures information
orthogonal to the classical factors. Early evidence from earnings-call sentiment suggests
this alpha exists in the short run---up to five trading days after the call---but decays
rapidly as market participants process the same text \citep{tetlock2007giving,
LopezLiraTang2023}.

The process of constructing new quantitative factors can itself be automated by LLMs.
\citet{YukselAlphaQuant2025} present AlphaQuant, an LLM-evolutionary loop that
iteratively proposes, evaluates, and refines alpha features for systematic strategies,
replacing manual feature engineering with a guided search over the space of potential
factor definitions and offering a concrete methodology for factor discovery beyond the
earnings-call sentiment pathway described above.

An important methodological caveat: \citet{harvey2016cross} documented that with hundreds
of proposed factors tested against the same dataset, conventional significance thresholds
($t > 2$) produce many false discoveries. A language factor that survives a Bonferroni
correction or a Sharpe ratio test accounting for multiple comparisons provides much stronger
evidence than a raw $p$-value below 0.05. This concern interacts directly with the backtest
pitfalls discussed in Section~\ref{sec:ch10-backtesting-pitfalls}.

% ============================================================
\section{LLM-Guided Algorithmic Trading}
\label{sec:ch10-llm-algorithmic-trading}
% ============================================================

Algorithmic trading applies computational methods to generate, transmit, and execute
orders with minimal human intervention. The space ranges from \emph{high-frequency
trading} (holding periods of microseconds to seconds, driven by microstructure signals)
to \emph{systematic macro} (holding periods of weeks to months, driven by fundamental
signals). LLMs are relevant primarily in the latter regime, where information latency is
measured in hours rather than milliseconds and where the quality of the signal---rather
than the speed of its delivery---determines performance.

% -----------------------------------------------------------
\subsection{LLMs as Sentiment-Driven Signal Generators}
\label{sec:ch10-sentiment-signals}
% -----------------------------------------------------------

A trading signal $s_t \in \mathbb{R}$ drives a position $w_t = f(s_t)$, where $f$
is a mapping from signal to portfolio weight. The signal-to-weight mapping must satisfy
several practical constraints: the position should scale with signal strength, be bounded
to prevent excessive concentration, and decay as the signal ages.

A standard approach uses a \emph{z-score normalisation} of the raw LLM output followed
by a position-sizing rule. If the raw sentiment score for asset $i$ at time $t$ is
$\tilde{s}_{i,t}$, the normalised signal is
\begin{equation}
  z_{i,t} = \frac{\tilde{s}_{i,t} - \hat{\mu}_s}{\hat{\sigma}_s},
  \label{eq:ch10-zscore}
\end{equation}
where $\hat{\mu}_s$ and $\hat{\sigma}_s$ are computed over a rolling backward window
(e.g., 252 trading days). The position in asset $i$ is then
\begin{equation}
  w_{i,t} = \frac{z_{i,t}}{\sum_j |z_{j,t}|} \cdot W_{\max},
  \label{eq:ch10-position-sizing}
\end{equation}
with $W_{\max}$ a maximum gross exposure parameter. Equation~\eqref{eq:ch10-position-sizing}
produces a dollar-neutral portfolio when positive and negative signals are balanced across
the universe---desirable for a long-short strategy that seeks to be insulated from
broad market moves.

The scale of predictive power available from LLM sentiment is illustrated by
\citet{KirtacGermano2024}, who score 965,375 news articles with a GPT-3-class model
and find that the resulting sentiment scores predict next-day stock returns with 74.4\%
directional accuracy; a long-short strategy exploiting these scores earns a Sharpe ratio
of 3.05 after transaction costs on their evaluation sample, providing a large-scale
empirical baseline for the magnitude of LLM sentiment alpha before crowding erodes it.

The dynamics of news sentiment decay must be modelled explicitly. A sentiment signal
derived from a news article at time $t_0$ carries diminishing predictive content as the
market digests the news. A common functional form is exponential decay:
\begin{equation}
  s_{i,t} = \sum_{\tau \leq t} \tilde{s}_{i,\tau} \cdot e^{-\lambda(t - \tau)},
  \label{eq:ch10-ewma-sentiment}
\end{equation}
where $\lambda > 0$ controls the half-life of the signal. Calibrating $\lambda$ is an
empirical matter: short-horizon price impact suggests $\lambda$ around $0.1$--$0.5$
day$^{-1}$ (half-lives of one to seven trading days) for most equity news events
\citep{grinold2000active}.

% -----------------------------------------------------------
\subsection{LLM-Guided Reinforcement Learning for Order Execution}
\label{sec:ch10-rl-order-execution}
% -----------------------------------------------------------

Order execution is the problem of slicing a large parent order into child orders that are
dispatched to the exchange over a specified horizon, minimising total market impact and
shortfall relative to a benchmark price (typically the arrival price or VWAP). Classical
approaches, such as the Almgren-Chriss model \citep{almgren2001optimal}, derive optimal
execution schedules by minimising the expected cost of trading plus a penalty for
execution risk. In the linear-impact version the objective is:
\begin{equation}
  C(\mathbf{x}) = \gamma \sum_{k=1}^{N} x_k v_k
                + \eta  \sum_{k=1}^{N} \tau_k v_k^2
                + \lambda \sigma^2 \sum_{k=1}^{N} \tau_k x_k^2,
  \label{eq:ch10-ac-cost}
\end{equation}
where $v_k = \dot{x}_k$ is the trading rate in period $k$, $x_k$ is the remaining
inventory, $\tau_k$ is the length of trading interval $k$, $\gamma$ is the permanent
impact coefficient (capturing the lasting price displacement from each trade),
$\eta$ is the temporary impact coefficient (capturing per-trade slippage), $\lambda$
is the trader's risk aversion, and $\sigma$ is return volatility.  The first sum is
the permanent-impact cost, the second is the temporary-impact (slippage) cost, and the
third penalises the variance of the execution shortfall.
The Almgren-Chriss solution is a closed-form time-varying schedule that front-loads
execution when risk-aversion is high and spreads it uniformly when it is low.

Reinforcement learning (RL) extends this framework by learning state-dependent execution
policies directly from market data \citep{mnih2015human, liu2022finrl}. The RL agent
observes a state $s_t$ comprising the residual inventory $X_t$, time remaining $T - t$,
current bid-ask spread, and recent price momentum. It selects an action $a_t \in [0, X_t]$
(the volume to execute in the current period) and receives a reward $r_t = -\text{cost}(a_t)$.

The role of LLMs in this pipeline is \emph{state augmentation}: the raw market-microstructure
state is enriched with a textual signal summarising breaking news about the stock or the
broader market. When a negative earnings surprise appears in the news feed, an LLM can
rapidly classify the event and trigger an accelerated execution schedule, reducing
adverse selection risk by completing the order before the full price impact is realised.
The LLM acts as a real-time alert system that modifies the RL agent's state representation
rather than replacing the execution logic.

Concretely, the augmented state is $\tilde{s}_t = [s_t^{\text{micro}},
\text{LLM-alert}_t]$, where $\text{LLM-alert}_t \in \{-1, 0, 1\}$ represents a negative,
neutral, or positive news event in the previous five minutes. In simulations based on the
FinRL framework \citep{liu2022finrl}, adding news-alert features reduces implementation
shortfall by approximately 8--12\% relative to a baseline RL agent without text
augmentation, though results are highly sensitive to the quality of the event-detection
classifier and the latency of the LLM inference pipeline. A more ambitious integration
is proposed by \citet{CoriatBenhamou2025}, who develop HARLF, a three-tier hierarchical
RL architecture in which lightweight LLMs inject news sentiment into base RL agents at
each level of the hierarchy; tested on 2018--2024 data, the system achieves a 26\%
annualised return with a Sharpe ratio of 1.2, illustrating the gains available when LLM
sentiment signals are woven into the reward structure rather than merely appended to
the state vector.

% -----------------------------------------------------------
\subsection{Why LLMs Are Not Stock Pickers: Market Efficiency Arguments}
\label{sec:ch10-llm-not-stock-pickers}
% -----------------------------------------------------------

A recurring claim in the popular press and some academic papers is that LLMs can
``predict the stock market''. This claim deserves careful scrutiny, because it conflates
statistically significant predictability with economically profitable trading---a
distinction that the market-efficiency literature has drawn for decades.

The weak form of the efficient market hypothesis states that all past price information
is reflected in current prices \citep{fama1970efficient}. The semi-strong form further
incorporates all publicly available information. An LLM that reads public news articles
to generate trading signals is, by definition, attempting to exploit publicly available
information. For large-cap equities with extensive analyst coverage, prices adjust to
news within minutes of publication. By the time an LLM has processed a news article---
even with a low-latency API---the price impact of that news has already been absorbed by
algorithmic traders with faster pipelines.

The Grossman-Stiglitz equilibrium \citep{grossman1980impossibility} provides a more nuanced lens.
Some positive return must accrue to informed trading, otherwise no one would bear the
cost of information collection. LLMs reduce the cost of processing text, potentially
expanding the set of investors who can act on textual signals. But if every large
investor subscribes to the same LLM API, the signal becomes crowded: many traders act
on the same information simultaneously, accelerating price discovery and compressing the
available alpha. This dynamic is already observable in the equity sentiment literature:
signals based on broad newspaper coverage have largely decayed over time as the market
has become more sophisticated at processing news \citep{tetlock2007giving}.

There is also a \emph{calibration problem}. LLMs generate probabilities that are not
inherently calibrated to financial outcomes. A model that assigns a 70\% probability to
a ``positive earnings outcome'' is not making a claim that the stock will rise 70\% of
the time---the mapping from textual prediction to price movement depends on the current
market expectation, the volatility of the stock, and countless other factors. Proper
calibration requires out-of-sample evaluation against realized returns, not in-sample
training accuracy.

The appropriate conclusion is not that LLMs are useless for trading, but that their
value lies in specific niches: processing \emph{heterogeneous, low-coverage, or novel}
information sources (small-cap filings, emerging-market news, niche regulatory
disclosures) where human analyst coverage is thin and price discovery is slower. In
these markets, the Grossman-Stiglitz equilibrium implies that the LLM-derived signal
can still earn a return commensurate with the cost of collecting and processing the
information.

% ============================================================
\section{Backtesting with LLM Signals}
\label{sec:ch10-backtesting}
% ============================================================

Backtesting is the empirical backbone of quantitative finance: it evaluates whether a
proposed strategy would have been profitable on historical data. A well-designed backtest
is a necessary (though not sufficient) condition for deploying a strategy with real
capital. The central challenge is producing a \emph{realistic} simulation that accounts
for market frictions, data limitations, and the multiple-comparison problem.

% -----------------------------------------------------------
\subsection{Walk-Forward Evaluation and Out-of-Sample Tests}
\label{sec:ch10-walk-forward}
% -----------------------------------------------------------

The standard backtesting protocol exposes a strategy to a single contiguous historical
period, uses the full-period data to estimate parameters, then reports in-sample
performance. This protocol is nearly useless for evaluating predictive strategies: the
model has implicitly seen the future when its parameters are calibrated, and the
in-sample Sharpe ratio substantially overstates live performance.

\emph{Walk-forward validation} is the time-series analogue of cross-validation. The
history is partitioned into an initial training window of length $T_{\text{train}}$, a
validation window of length $T_{\text{val}}$, and a succession of out-of-sample test
windows of length $T_{\text{test}}$:

\begin{definition}[Walk-Forward Evaluation Protocol]
\label{def:walk-forward}
Fix a training length $T_{\text{train}}$, a validation length $T_{\text{val}}$, and a
step size $\Delta$. For fold $m = 0, 1, 2, \ldots$, define:
\begin{itemize}
  \item \emph{Training set}: observations $[m\Delta,\; m\Delta + T_{\text{train}})$.
  \item \emph{Validation set}: observations $[m\Delta + T_{\text{train}},\;
        m\Delta + T_{\text{train}} + T_{\text{val}})$.
  \item \emph{Test window}: observations $[m\Delta + T_{\text{train}} + T_{\text{val}},\;
        m\Delta + T_{\text{train}} + T_{\text{val}} + \Delta)$.
\end{itemize}
Model parameters are estimated on the training set, hyperparameters are selected on
the validation set, and performance is reported only on the concatenation of all
test windows.
\end{definition}

For LLM-based signals, the ``parameters'' being estimated include the calibration
constant $\alpha$ in equation~\eqref{eq:ch10-signal-to-view}, the decay parameter
$\lambda$ in equation~\eqref{eq:ch10-ewma-sentiment}, and the position-sizing
scale $W_{\max}$. These should never be optimised using knowledge of the test period.

The \emph{out-of-sample Sharpe ratio} is the primary performance metric:
\begin{equation}
  \text{SR}^{\text{OOS}} = \frac{\bar{r}_p^{\text{OOS}}}{\hat{\sigma}_p^{\text{OOS}}}
  \cdot \sqrt{252},
  \label{eq:ch10-oos-sharpe}
\end{equation}
where $\bar{r}_p^{\text{OOS}}$ and $\hat{\sigma}_p^{\text{OOS}}$ are the mean and
standard deviation of daily portfolio returns over the concatenated out-of-sample windows.
A rule of thumb used in practice: a backtest Sharpe ratio above $1.5$ on at least three
years of out-of-sample data is a reasonable threshold for proceeding to paper trading.

% -----------------------------------------------------------
\subsection{Transaction Costs, Slippage, and Capacity Analysis}
\label{sec:ch10-transaction-costs}
% -----------------------------------------------------------

A strategy that looks profitable before costs may be unprofitable after them. Transaction
costs for a systematic equity strategy include:
\begin{itemize}
  \item \textbf{Bid-ask spread:} Half the spread is paid on each trade. For US large-cap
    equities, this is on the order of 1--3 basis points per trade; for small-caps, 10--50
    basis points is common.
  \item \textbf{Market impact:} Large orders move prices against the trader. The
    Almgren-Chriss model \citep{almgren2001optimal} parameterises this as a temporary
    impact $\eta x^2$ and a permanent impact $\gamma x$, where $x$ is the fraction of
    daily volume traded.
  \item \textbf{Financing costs:} Short positions require borrowing securities, at a cost
    that varies with stock supply (``borrow rate''). For hard-to-borrow stocks, this can
    exceed 5\% per annum.
\end{itemize}

Let $\Delta w_{i,t} = w_{i,t} - w_{i,t-1}$ be the weight change in asset $i$ at
rebalancing date $t$. A simple linear cost model charges a constant $c$ per unit of
turnover:
\begin{equation}
  \text{TC}_t = c \sum_{i} |\Delta w_{i,t}|,
  \label{eq:ch10-transaction-cost}
\end{equation}
with $c$ expressed in percentage points per unit of portfolio value rebalanced. For a
daily-rebalanced LLM sentiment strategy with average one-way turnover of 15\%, a cost
of $c = 0.05\%$ per trade implies annualised costs of approximately $0.15 \times 252
\times 0.05\% \approx 1.89\%$---substantial enough to erase the alpha of a marginal
signal.

\emph{Capacity analysis} asks: how much capital can the strategy absorb before market
impact erodes profitability? A common rule of thumb is that a strategy should not trade
more than 5--10\% of average daily volume in any single stock. If the strategy requires
position sizes of $\$10$M in a stock with $\$50$M daily volume, the 20\% participation
rate will meaningfully move the price. Strategies based on LLM signals derived from
earnings calls face a particularly acute capacity constraint: many funds process the same
text simultaneously, creating crowded trades in the first hours after a call.

% -----------------------------------------------------------
\subsection{Common Backtesting Pitfalls: Look-Ahead Bias, Survivorship}
\label{sec:ch10-backtesting-pitfalls}
% -----------------------------------------------------------

\emph{Look-ahead bias} occurs when a backtest uses data that was not available at the
time of the simulated trade. For LLM-based strategies, look-ahead bias can arise in
several subtle forms:

\begin{enumerate}
  \item \textbf{Embedding leakage.} If the LLM was fine-tuned on data that includes
    the evaluation period, its representations encode future information. A model
    fine-tuned on all available earnings-call transcripts as of the backtest date is
    contaminated if those transcripts post-date the simulated trading date.
  \item \textbf{Model selection leakage.} Choosing the best-performing LLM variant
    by looking at full-period performance, then backtesting on the same period, overfits
    the model to that period. Hyperparameter selection must use only data predating the
    evaluation window.
  \item \textbf{Timestamp errors.} Earnings calls are typically released after market
    close. If the backtest data assigns a release date of $t$ but the strategy trades
    at the open on day $t$ rather than day $t+1$, the strategy is trading on information
    not yet available at the open.
\end{enumerate}

\emph{Survivorship bias} affects any study that uses a historical universe of assets
without accounting for delistings, bankruptcies, and mergers. A database of S\&P 500
constituents as of today excludes all companies that were removed from the index due to
financial distress---precisely the companies on which a risk-signal strategy would have
suffered its worst drawdowns. Using a point-in-time constituent database (available from
vendors such as Compustat or FactSet) is mandatory for a credible backtest.

\citet{bailey2014probability} formalize the multiple-comparison problem through the
\emph{Probability of Backtest Overfitting} (PBO), which estimates the probability that
the strategy selected as best in-sample will underperform the median out-of-sample. When
many parameter combinations are tried, PBO approaches one even for strategies with no
genuine predictive power. \citet{lopezdeprado2018advances} recommends reporting the \emph{minimum
track record length} needed to confirm a given Sharpe ratio at a specified confidence
level:
\begin{equation}
  T^* = 1 + \hat{\sigma}^2\left(1 - \gamma_1 \text{SR}^* + \frac{\gamma_2 - 1}{4}
    (\text{SR}^*)^2 \right) \cdot \frac{Z_{1-\alpha}^2}{(\text{SR}^*)^2},
  \label{eq:ch10-min-track}
\end{equation}
where $\gamma_1$ and $\gamma_2$ are the skewness and kurtosis of the return distribution
and $Z_{1-\alpha}$ is the standard normal quantile. For non-normal strategy returns
(typical of LLM sentiment strategies, which exhibit fat tails), $T^*$ can be several
years, underlining why short backtests are insufficient.

% ============================================================
\section{Risk Management Applications}
\label{sec:ch10-risk-management}
% ============================================================

Risk management in quantitative finance concerns the identification, measurement, and
control of potential losses. Classical tools---Value at Risk (VaR), Expected Shortfall
(CVaR), stress testing---rely primarily on historical return distributions or
parametric models. LLMs extend this toolkit by processing the enormous volume of
qualitative information that precedes financial stress: management warnings, regulatory
disclosures, macroeconomic commentary, and geopolitical developments that shift the
tail of the return distribution before they appear in historical prices.

% -----------------------------------------------------------
\subsection{Tail Risk Estimation from Textual Signals}
\label{sec:ch10-tail-risk}
% -----------------------------------------------------------

\emph{Conditional Value at Risk} (CVaR, also called Expected Shortfall) at confidence
level $\alpha$ is defined as the expected loss in the worst $\alpha$-fraction of outcomes:
\begin{equation}
  \text{CVaR}_\alpha(R) = -\mathbb{E}\bigl[R \mid R \leq q_\alpha(R)\bigr],
  \label{eq:ch10-cvar}
\end{equation}
where $q_\alpha(R)$ is the $\alpha$-quantile of the return distribution $R$.
\citet{rockafellar2000optimization} showed that CVaR is a coherent risk measure and can be
minimised via a linear program, making it attractive for portfolio construction. The
textual risk management question is: can LLM-based signals improve the estimation of
$\text{CVaR}_\alpha$ in real time?

The mechanism is as follows. Let $v_t$ denote a \emph{textual volatility signal}---a
scalar derived from LLM processing of news at time $t$ that captures the intensity of
risk-relevant language (references to uncertainty, litigation, credit deterioration,
sovereign risk, etc.). A conditionally heteroskedastic return model augmented with the
textual signal is:
\begin{equation}
  r_t = \mu + \varepsilon_t, \quad \varepsilon_t = \sigma_t z_t, \quad z_t \sim F(0, 1),
  \label{eq:ch10-risk-model-1}
\end{equation}
\begin{equation}
  \sigma_t^2 = \omega + \alpha_1 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2
               + \kappa v_{t-1},
  \label{eq:ch10-risk-model-2}
\end{equation}
where equations~\eqref{eq:ch10-risk-model-1}--\eqref{eq:ch10-risk-model-2} define a
GARCH(1,1) model augmented by the textual signal through the coefficient $\kappa$
(here $\alpha_1$ is the ARCH coefficient, not to be confused with the CVaR confidence
level $\alpha$).
Positive $\kappa$ means that elevated textual risk language forecasts higher return
variance---and thus fatter tails---in the next period.

Empirically, this specification connects to the News Implied Volatility (NVIX) of
\citet{manela2017news}, who constructed a text-based measure of disaster risk from
newspaper archives and showed it predicted stock-market variance and crash risk.
LLMs can update this estimate in near real time at the individual-firm level rather
than the aggregate market level, enabling position-specific tail-risk monitoring.

For portfolio construction, CVaR-constrained optimization becomes:
\begin{equation}
  \min_{\mathbf{w}} \; \mathbf{w}^\top \boldsymbol{\Sigma}_t \mathbf{w}
  \quad \text{subject to} \quad \text{CVaR}_\alpha(\mathbf{w}^\top \mathbf{r}_t)
  \leq \bar{C},
  \label{eq:ch10-cvar-constrained}
\end{equation}
where $\boldsymbol{\Sigma}_t$ is the time-varying covariance matrix (estimated via DCC
\citep{engle2002dynamic} or a similar model) and $\bar{C}$ is the tail-risk budget.
When $v_{t-1}$ is high for a particular sector, $\sigma_{i,t}^2$ rises for firms in
that sector, reducing their contribution to $\text{CVaR}$ tolerance and triggering
automatic de-risking.

% -----------------------------------------------------------
\subsection{Real-Time Portfolio Monitoring and Alert Systems}
\label{sec:ch10-portfolio-monitoring}
% -----------------------------------------------------------

Portfolio monitoring transforms the one-shot portfolio construction problem into a
continuous surveillance task. The goal is to detect material risks before they trigger
large losses, not after. LLMs are well suited to this role because they can process
streams of unstructured data---regulatory filings, corporate announcements, earnings
pre-announcements, news wires---and flag events that warrant human review or automated
rebalancing.

An alert system built on LLMs operates as a pipeline with three layers. The
\emph{ingestion layer} polls data feeds (EDGAR, Reuters/Bloomberg news APIs, social
media) at a configurable frequency and collects new text documents. The
\emph{classification layer} applies an LLM to each document with a prompt designed to
elicit a structured assessment of material risk:

\begin{quote}
\textit{You are a portfolio risk analyst. Read the following SEC filing excerpt and output
a JSON object with fields: \texttt{risk\_level} (low/medium/high/critical), \texttt{risk\_type}
(legal/financial/operational/macro/ESG), and \texttt{affected\_positions} (a list of
ticker symbols mentioned in the context that may be directly affected). Be concise and
specific.}
\end{quote}

The \emph{action layer} routes high-severity alerts to a risk officer for review, or
directly to the execution system if the firm operates under a pre-approved automated
response protocol. For example, a \texttt{critical} legal-risk alert for a position
exceeding 2\% of portfolio value might automatically reduce the position to 1\% pending
human review.

Several calibration challenges arise in practice. First, the LLM must be calibrated to
the firm's specific portfolio and risk tolerance---a ``high'' risk event for a sovereign
bond portfolio is not the same as for an equity long-short fund. Fine-tuning on
historical alert data labelled by the risk team significantly improves precision.
Second, alert fatigue is real: a system that generates 200 alerts per day will be
ignored. Precision-recall trade-offs should be tuned to produce a manageable number of
actionable alerts (typically fewer than 20 per day for a 100-position portfolio). Third,
the LLM inference pipeline must be fast enough to be useful: for end-of-day monitoring,
minutes of latency are acceptable; for intraday risk, seconds matter.

\begin{example}[An LLM Alert Triggered by an 8-K Filing]
\label{ex:ch10-8k-alert}
At 4:17 PM Eastern time on a Tuesday, a company in the portfolio files a Form 8-K with
the SEC announcing an SEC investigation into its revenue recognition practices. Within
30 seconds, the ingestion layer retrieves the filing. The classification layer invokes
an LLM with the excerpt and receives the response:

\begin{verbatim}
{
  "risk_level": "critical",
  "risk_type": "legal",
  "affected_positions": ["TICKER_X"],
  "summary": "SEC formal order of investigation into revenue recognition.
              Historical precedent suggests 20--40% downside risk."
}
\end{verbatim}

The action layer flags the alert and, under the firm's protocol, executes a partial
position reduction from 2.1\% to 1.0\% of the portfolio at the next opening. The next
morning, the stock opens down 28\% on the news. The automated response, triggered within
seconds of the filing, protected the portfolio from roughly half of the maximum loss
that a slow-response system would have sustained.
\end{example}

The example illustrates both the promise and the limitation of LLM-based risk monitoring.
The system acted faster than a human analyst could have. But it could not foresee the
exact magnitude of the market reaction, nor could it distinguish between an investigation
that will be closed with no action and one that will lead to a restatement. The system
reduces tail exposure but cannot eliminate it; human judgment remains essential for
calibrating the automated response thresholds.

% ============================================================
\section{Summary}
\label{sec:ch10-summary}
% ============================================================

This chapter developed the intersection of large language models with classical portfolio
theory, algorithmic trading, and risk management. Five key takeaways deserve emphasis.

The efficient frontier \citep{markowitz1952portfolio} and Black-Litterman framework
\citep{blacklitterman1992} remain the correct tools for translating return signals into
portfolio weights. LLMs contribute by generating better estimates of
$\boldsymbol{\mu}$---the expected return vector---from unstructured text, not by
replacing the optimisation machinery; they are information extractors, not optimisers.

Factor construction is the most defensible application channel. Building long-short
portfolios based on LLM-derived sentiment scores and testing their alpha against
Fama-French factors \citep{fama1993common, fama2015five} provides a principled
evaluation framework. Statistical rigour demands adjustment for multiple comparisons
\citep{harvey2016cross} and out-of-sample validation.

Market efficiency limits but does not eliminate LLM alpha. The Grossman-Stiglitz
paradox \citep{grossman1980impossibility} implies that some informed return must exist,
but also that LLM signals will be most valuable in low-coverage markets where price
discovery is slowest. Large-cap, heavily covered equities are the hardest target.

Backtesting discipline is non-negotiable. Walk-forward evaluation, realistic
transaction cost modelling, point-in-time data, and correction for backtest overfitting
\citep{bailey2014probability, lopezdeprado2018advances} are not optional refinements---they
are the difference between a research artefact and a deployable strategy.

Real-time risk monitoring represents a high-value, lower-competition application.
Extracting structured risk signals from regulatory filings and news wires, augmenting
GARCH-family volatility models \citep{engle2002dynamic}, and maintaining continuous
portfolio surveillance deliver tangible, auditable value with less exposure to the
market-efficiency critique.

% ============================================================
\section*{Exercises}
% ============================================================

\begin{exercise}[Mean-Variance Portfolio with an LLM Signal]
\label{ex:ch10-mv-llm}
\textbf{[B]} Suppose you have five assets with expected excess returns
$\boldsymbol{\mu} = (0.08,\, 0.06,\, 0.04,\, 0.03,\, 0.02)^{\top}$ and a diagonal
covariance matrix $\boldsymbol{\Sigma} = \mathrm{diag}(0.04, 0.09, 0.06, 0.03, 0.05)$.
(\textit{a})~Compute the minimum-variance portfolio weights (long-only, weights summing
to one) analytically or numerically. (\textit{b})~An LLM processes recent earnings calls
and produces a positive sentiment update for asset~1, raising its expected return to
$0.12$. Recompute the weights. (\textit{c})~Explain intuitively why the shift in
$\boldsymbol{\mu}$ for one asset affects the weights of the others.
\end{exercise}

\begin{exercise}[Almgren-Chriss Optimal Liquidation]
\label{ex:ch10-ac-liquidation}
\textbf{[I]} A trader holds $X_0 = 10{,}000$ shares and must liquidate over $N = 5$
equally-spaced periods. The permanent impact coefficient is $\gamma = 2.5 \times
10^{-6}$, the temporary impact coefficient is $\eta = 6.25 \times 10^{-7}$, the
per-period time step is $\tau = 0.2$ days, return volatility is $\sigma = 0.02$
(per day), and risk aversion is $\lambda = 10^{-6}$.
(\textit{a})~Compute the Almgren-Chriss optimal trading trajectory using the standard
three-term cost function from Section~\ref{sec:ch10-execution}. (\textit{b})~Plot
remaining inventory $x_k$ against period $k$. (\textit{c})~An LLM monitoring news
wires detects a negative sentiment spike at $k=2$, suggesting the permanent impact
coefficient will double for the remaining periods. Recompute the optimal trajectory
from $k=2$ onward and compare total expected cost with and without the LLM signal.
\end{exercise}

\begin{exercise}[Walk-Forward Backtest of an LLM Sentiment Factor]
\label{ex:ch10-backtest}
\textbf{[A]} Design and implement a walk-forward backtest of a long-short factor
constructed from LLM-derived sentiment scores on earnings call transcripts.
(\textit{a})~Obtain or simulate a panel of 50 stocks over 24 months with monthly
sentiment scores $S_{i,t}$ and returns $r_{i,t+1}$. (\textit{b})~Each month, rank
stocks by $S_{i,t}$ and form a value-weighted long portfolio (top decile) minus a
value-weighted short portfolio (bottom decile). Compute annualised Sharpe ratio and
information ratio. (\textit{c})~Apply realistic transaction costs using the cost model
$c_{i,t} = 0.001 \cdot |w_{i,t} - w_{i,t-1}|$ for each stock. Report the net-of-cost
Sharpe ratio. (\textit{d})~Run the Deflated Sharpe Ratio test \citep{bailey2014probability}
and interpret whether the observed Sharpe is statistically significant after adjusting
for the number of strategy variants you tested.
\end{exercise}

% ============================================================
\section{Further Reading}
\label{sec:ch10-further-reading}
% ============================================================

Readers seeking deeper treatment of classical portfolio theory should consult
\citet{markowitz1952portfolio} for the original mean-variance framework and
\citet{blacklitterman1992} for the Bayesian extension. The empirical limits of portfolio
optimization are surveyed comprehensively by \citet{demiguel2009optimal}.

For the factor-model foundation of portfolio construction, \citet{fama1993common} and
\citet{fama2015five} are essential primary sources. \citet{carhart1997persistence}
documents the momentum factor that the original three-factor model omits.
\citet{harvey2016cross} provides a sobering assessment of multiple-comparison problems
in the factor-zoo literature.

On LLM applications to return prediction and sentiment, \citet{LopezLiraTang2023}
and \citet{tetlock2007giving} bracket the evidence from primitive to modern methods.
\citet{kim2023llm} extends this to corporate risk disclosure.

For reinforcement learning in trading, \citet{liu2022finrl} provides the most complete
open-source framework, while \citet{almgren2001optimal} remains the canonical reference
for optimal execution. \citet{mnih2015human} and \citet{schulman2017proximal} are the
foundational RL papers.

Backtesting methodology is treated at book length by \citet{lopezdeprado2018advances}, which
provides Python implementations of the minimum track record length, deflated Sharpe
ratio, and combinatorially symmetric cross-validation. \citet{bailey2014probability}
develops the theoretical foundations for backtest overfitting. For risk measurement,
\citet{rockafellar2000optimization} derives the CVaR linear programme, and
\citet{engle2002dynamic} introduces DCC-GARCH.

The emerging literature on autonomous LLM agents for portfolio management is surveyed
by \citet{SahaLyuEtAl2025}, who provide a structured taxonomy of agent architectures---
including multi-agent, reflection-based, and tool-augmented designs---and identify open
benchmarks and research frontiers in agentic trading system design. For the LLM-as-optimizer
strand of this literature, \citet{Yuksel2025alpha} and \citet{YukselAlphaQuant2025}
offer complementary entry points into portfolio selection via evolutionary search and
automated feature engineering, respectively. \citet{MantshimuliMwamba2025} and
\citet{CoriatBenhamou2025} provide the most recent empirical results on multi-LLM
Black-Litterman integration and hierarchical RL sentiment fusion.

Readers interested in the regulatory and institutional context of LLM deployment in
risk management should consult Chapter~7 on regulatory compliance and the Federal
Reserve's model risk management guidance \citep{sr117}, which requires
that all model outputs be explainable, validated, and auditable---challenges that apply
directly to LLM-based risk signals.
