\chapter{Domain-Specific Financial Large Language Models}
\label{ch:domain-specific-llms}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Explain why general-purpose language models underperform on financial tasks
        and identify the specific linguistic features of financial text that create
        this gap.
  \item Describe the principal families of financial LLMs---encoder-only (BERT-based)
        and decoder-only---and select the appropriate architecture for a given task.
  \item Apply domain-adaptive pre-training (DAPT) to a pre-trained model using a
        financial corpus, and design a corpus composition strategy that balances
        coverage and quality.
  \item Evaluate a financial LLM on the standard benchmarks (FinQA, FLUE, FinBen)
        and interpret the results in terms of the task taxonomy developed in this chapter.
  \item Weigh the practical trade-offs---licensing, compute cost, data governance,
        and performance---when choosing between a domain-specific model, a fine-tuned
        general model, or a hybrid retrieval-augmented approach for a production
        finance application.
\end{enumerate}
\end{remark}

% ---------------------------------------------------------------
\section{Why Domain-Specific LLMs?}
\label{sec:ch08-why-domain-specific}

\begin{context}
The promise of large language models in finance rests on a tension that every
practitioner eventually encounters.  On one side stands the sheer breadth of
general-purpose models: systems trained on hundreds of billions of tokens drawn
from the web, books, and scientific literature have absorbed enough world knowledge
to answer questions about option pricing, credit default swaps, and Basel III
capital requirements without any finance-specific training.  On the other side
stands the specificity that serious financial work requires: a model that confuses
``basis risk'' with ``basis points'', misreads a tabular income statement, or
applies general-English connotations to the word ``liability'' will produce
outputs that are not merely imprecise but potentially dangerous to any workflow
that depends on them.

The resolution to this tension is not a choice between breadth and depth but a
question of how best to combine them.  Domain-specific financial LLMs are one
answer: models whose pre-training, fine-tuning, or both have been oriented
toward the linguistic distribution of financial text.  Understanding when those
models add value---and when they do not---requires first characterising what
makes financial language genuinely different.
\end{context}

% ---------------------------------------------------------------
\subsection{Financial Language: Jargon, Numerics, Tables}
\label{sec:ch08-financial-language}

Financial text is not simply formal English with occasional technical terms.
It is a distinct register with three structural features that are systematically
underrepresented in general pre-training corpora.

The first feature is specialised vocabulary.  Finance has accumulated, over
centuries of practice and decades of regulation, a lexicon in which common
English words carry precise and sometimes counterintuitive technical meanings.
A ``liability'' in accounting is a specific obligation on a balance sheet, not
a source of risk or responsibility in the everyday sense.  A ``short'' position
is an investment posture, not a description of duration.  ``Haircut'' refers to
a collateral discount, not a grooming service.  The Loughran--McDonald (LM)
dictionary \cite{loughran2011liability} catalogues roughly 350 words that are
positive or negative in general English but have neutral or reversed valence in
SEC filings.  A model trained on web text that assigns the standard sentiment
polarity to ``liability'' will systematically misclassify financial documents.

The second feature is the prevalence and centrality of numeric reasoning.
Financial documents are saturated with numbers: percentage changes, ratio
computations, multi-period comparisons across fiscal years.  A general-purpose
model may parse a sentence like ``revenue grew 12.3\% year-over-year to
\$4.7~billion'' as a fluent English sentence without internalising the arithmetic
relationships that make the claim meaningful.  Tasks like FinQA \cite{chen2021finqa}
require multi-step numerical reasoning over tables extracted from 10-K filings---a
capability that demands more than linguistic fluency.

The third feature is tabular structure.  The most information-dense sections of
financial documents---income statements, balance sheets, cash flow statements,
loan-level data tapes---are presented in tables whose semantics depend on row and
column headings that are spatially distant from the numbers they label.  Standard
transformer tokenisation treats a table as a flat sequence of tokens, losing the
two-dimensional alignment that gives the numbers their meaning.  Pre-training on
a corpus that contains millions of financial tables can partially compensate for
this structural blindness, but the problem is not fully solved by any existing
model.

\begin{definition}[Financial Language Register]
\label{def:ch08-register}
A \emph{language register} is a variety of a language used for a specific
purpose or in a specific social situation, characterised by a distinctive
vocabulary, syntax, and set of pragmatic conventions.  Financial language
constitutes a distinct register along three dimensions:
\begin{enumerate}
  \item \textit{Lexical specificity}: a large vocabulary of technical terms,
        many of which are homonyms with common English words but carry domain-specific
        meanings.
  \item \textit{Numeric density}: a high ratio of numerals, equations, and
        percentage expressions relative to prose text.
  \item \textit{Tabular embedding}: structured information encoded in two-dimensional
        grids whose interpretation requires spatial reasoning over non-adjacent tokens.
\end{enumerate}
A language model is \emph{financially literate} to the extent that it correctly
resolves the meaning of tokens in each of these three dimensions.
\end{definition}

\begin{remark}[Vocabulary coverage and tokenisation]
The vocabulary mismatch problem operates at the sub-word level.  A Byte-Pair
Encoding (BPE) tokeniser trained on general text may split the term
``EBITDA'' into four or five subword tokens, each of which individually carries
no financial meaning.  Tokenisers trained on financial corpora are more likely
to represent such abbreviations as single tokens, reducing the representational
burden on the model's attention layers.  BloombergGPT \cite{wu2023bloomberggpt}
trained a custom BPE tokeniser on its combined general and financial corpus
precisely to ensure that domain-specific tokens received compact representations.
\end{remark}

% ---------------------------------------------------------------
\subsection{Limitations of General-Purpose Models on Financial Tasks}
\label{sec:ch08-limitations-general-purpose}

The empirical case for domain specialisation is best made through specific failure
modes, because the aggregate benchmark numbers---discussed at length in
Section~\ref{sec:ch08-benchmark-comparisons}---can be misleading.

\textit{Sentiment polarity inversion.}  As documented by \citet{loughran2011liability},
words like ``unprecedented,'' ``strong,'' and ``leverage'' carry very different
connotations in financial versus general text.  A general-purpose model fine-tuned
on the Financial PhraseBank dataset \cite{malo2014phrasebank} achieves lower
accuracy than FinBERT \cite{araci2019finbert} on the same task, even when the
general model has more parameters, because its pre-trained representations encode
the wrong prior over these lexical items.

\textit{Numerical hallucination.}  Large general-purpose models frequently
hallucinate specific financial figures.  When asked to reproduce earnings per
share from a given 10-K passage, GPT-class models produce plausible-looking
numbers that diverge from the source text at rates that make them unreliable for
any workflow requiring numerical precision \cite{kang2023hallucination}.

\textit{Table parsing failures.}  On the FinQA benchmark \cite{chen2021finqa},
which requires chaining arithmetic operations over financial tables, general-purpose
models in the GPT-3 class (175B parameters) achieve accuracy below 60\%, while
fine-tuned specialist systems exceed 80\%.  The gap cannot be closed by prompting
alone; the model must have seen enough financial table structures during training
to internalise the mapping from table position to semantic role.

\textit{Regulatory language ambiguity.}  Documents like SEC comment letters, ISDA
master agreements, and Basel III regulatory technical standards use a legally
precise sub-dialect that differs from both standard English and casual financial
commentary.  Models trained on SEC EDGAR filings---as in SEC-BERT
\cite{loukas2022secbert}---systematically outperform general models on entity
recognition and clause classification tasks over regulatory text.

These failure modes share a common root: the gap between the distribution of
financial text and the distribution of text that dominates general pre-training
corpora.  The remedy is to shift the model's prior toward the financial distribution,
either by pre-training from scratch on financial data, by continued pre-training on
financial corpora, or by targeted fine-tuning.

% ---------------------------------------------------------------
\section{A Taxonomy of Financial LLMs}
\label{sec:ch08-taxonomy}

The landscape of financial language models can be organised along two axes:
architecture (encoder-only versus decoder-only) and training strategy
(from-scratch versus domain-adaptive continuation versus fine-tuning only).
Each combination of choices implies different computational cost, different
capability profile, and different deployment constraints.

\begin{definition}[Financial LLM Taxonomy]
\label{def:ch08-taxonomy}
Let $\mathcal{M}$ denote the space of language models.  A \emph{financial LLM}
is any $m \in \mathcal{M}$ for which at least one of the following holds:
\begin{enumerate}
  \item The pre-training corpus includes a significant fraction of financial text
        (news, filings, analyst reports, or financial reference data).
  \item The model has been fine-tuned or instruction-tuned on financial NLP tasks
        (sentiment, NER, summarisation, numerical QA).
  \item The model's tokeniser has been trained on or extended for financial vocabulary.
\end{enumerate}
Financial LLMs are further classified by architecture as \emph{encoder-only}
(bidirectional, suited to classification and extraction) or \emph{decoder-only}
(autoregressive, suited to generation and multi-step reasoning).
\end{definition}

% ---------------------------------------------------------------
\subsection{Encoder--Decoder Models: T5 Family in Finance}
\label{sec:ch08-encoder-decoder}

A third architectural family that the taxonomy above implicitly omits is the
\emph{encoder--decoder} (sequence-to-sequence) model, whose canonical
representative is T5 \cite{raffel2020t5} and its instruction-tuned descendant
FLAN-T5.  Encoder--decoder models process an input sequence through a
bidirectional encoder---capturing full context on the source side---and then
generate output token by token through an autoregressive decoder, making them
architecturally well suited to sequence-to-sequence tasks: summarisation,
translation, question answering, and structured text generation.  In financial
NLP, this architecture has been applied to earnings call summarisation (condensing
hours of executive commentary into a structured abstract), abstractive summarisation
of SEC filings (reducing a 100-page 10-K to a one-page narrative summary), and
financial table-to-text generation (converting a tabular income statement into
a plain-language paragraph).  Although recent instruction-tuned decoder-only
models such as FinMA have demonstrated that the decoder-only paradigm can handle
these tasks with sufficient scale and instruction data, encoder--decoder models
retain practical advantages in constrained-output settings: the separate encoding
pass allows the model to attend fully to a long source document before generating
a single output token, which reduces copy errors and improves factual consistency
in extraction-heavy summarisation tasks.

% ---------------------------------------------------------------
\subsection{BERT-Family: FinBERT, SEC-BERT, FinBERT-tone}
\label{sec:ch08-bert-family}

The earliest and most widely used financial LLMs are BERT-based
\cite{devlin2019} encoder models adapted to the financial domain.  Their small
size (110--340M parameters), open weights, and well-understood fine-tuning
behaviour make them the default choice for classification and extraction tasks
in research and production environments.

\textit{FinBERT (Araci 2019).}  The first model to carry the FinBERT name was
introduced by \citet{araci2019finbert}.  The model is BERT-base continued on a
corpus of approximately 1.8~billion words drawn from Reuters TRC2 news,
Bloomberg news, and Financial PhraseBank.  Its primary task is financial
sentiment classification---positive, negative, or neutral---over financial news
headlines and short articles.  On the Financial PhraseBank benchmark
\cite{malo2014phrasebank}, FinBERT improved F1 by roughly 15 percentage points
over the baseline BERT-base model, establishing that domain-specific pre-training
adds measurable value even for a relatively straightforward classification task.

\textit{FinBERT (Yang, Uy, and Huang 2020).}  A second model also called
FinBERT \cite{yang2020finbert} was published independently by
\citet{yang2020finbert}.  This version continues BERT-large on a corpus of
financial communication text---earnings call transcripts, analyst reports, and
SEC filings---and introduces a \emph{tone} classification variant
(FinBERT-tone) fine-tuned to distinguish the rhetorical tone of corporate
communications: optimistic, pessimistic, litigious, uncertain, and constraining,
following the Loughran--McDonald taxonomy \cite{loughran2011liability}.  The
tone classification task is particularly demanding because it requires the model
to integrate a sentence's lexical content with discourse-level context: the phrase
``subject to regulatory approval'' carries very different tone depending on whether
it appears in a merger announcement or a quarterly guidance withdrawal.

\textit{SEC-BERT.}  \citet{loukas2022secbert} pre-train a BERT-base model
exclusively on SEC EDGAR filings: 10-K annual reports, 10-Q quarterly reports,
8-K current reports, and proxy statements, totalling approximately 3~billion
tokens.  The resulting model outperforms general BERT and domain-adapted FinBERT
on named entity recognition over financial entities (companies, instruments,
persons, dates, monetary amounts) and on clause-level classification tasks over
regulatory language.  The key insight is that EDGAR text constitutes a coherent
and legally precise sub-dialect; a model pre-trained on news and analyst reports
does not fully transfer to it.

\begin{example}[FinBERT for earnings call sentiment]
\label{ex:ch08-finbert-earnings}
Consider the following excerpt from an earnings call transcript:

\begin{quote}
``While we are cautiously optimistic about the trajectory of our core business,
we want to flag meaningful headwinds in the European segment, driven by
continued softness in consumer demand and heightened competitive pressure.''
\end{quote}

A general-purpose sentiment classifier---trained on product reviews or social
media---will likely label this passage \textit{positive} because ``optimistic''
and ``optimistic about'' are strongly positive signals in general text.
FinBERT-tone, by contrast, correctly assigns \textit{uncertain/pessimistic}
because it has learned that ``cautiously optimistic'' is a hedging qualifier
in corporate earnings language, and that ``headwinds,'' ``softness,'' and
``competitive pressure'' are near-universally negative in this register.
The tone assigned to this passage has real downstream consequences: it feeds
directly into analyst sentiment indices that drive systematic trading strategies
\cite{tetlock2007giving}.
\end{example}

% ---------------------------------------------------------------
\subsection{Decoder-Family: BloombergGPT, FinGPT, FinMA, InvestLM}
\label{sec:ch08-decoder-family}

The decoder-only (autoregressive) family of financial LLMs emerged from 2022
onward, driven by the success of GPT-style models at generation and multi-step
reasoning tasks.  These models are substantially larger than the encoder-only
family and are capable of open-ended financial question answering, report
generation, and numerical reasoning.

\textit{BloombergGPT.}  The landmark paper in this family is
\citet{wu2023bloomberggpt}, which describes a 50-billion-parameter autoregressive
decoder model trained on a corpus of approximately 708 billion tokens, comprising
a 363-billion-token Bloomberg-proprietary slice (news, filings, earnings
transcripts, web text, and Bloomberg-specific data) and a 345-billion-token
public component drawn from The Pile \cite{gao2020pile}, Wikipedia, and books.
The roughly 50/50 split between financial and general text is deliberate: the authors argue that
a model trained \emph{exclusively} on financial text would sacrifice the general
reasoning and commonsense capabilities that make an LLM useful for financial
tasks requiring world knowledge.  On financial NLP benchmarks (FinQA, NER,
headline classification, and others), BloombergGPT outperformed all existing
models of comparable size, and also matched or exceeded GPT-NeoX-20B on several
general language benchmarks, demonstrating that domain-specific pre-training need
not degrade general capability.  \citet{Keshri2025} survey the full scope of
BloombergGPT's architecture, training corpus, and downstream applications---including
news summarisation, fraud detection, and trading strategy optimisation---providing
a useful secondary account of the model's capabilities for practitioners who lack
access to the full technical report.

\textit{FinGPT.}  In contrast to the closed, proprietary BloombergGPT,
\citet{Yang2023} introduced FinGPT as an open-source framework for
building financial LLMs.  Rather than training from scratch, FinGPT applies
low-rank adaptation (LoRA \cite{hu2022lora}) to open foundation models
(LLaMA, ChatGLM, Falcon) using continuously updated financial news data fetched
via APIs.  The design philosophy prioritises \emph{democratisation}: a practitioner
with access to a single A100 GPU can fine-tune a capable financial model using
FinGPT's pipeline, whereas replicating BloombergGPT would require a cluster-scale
compute budget.  \citet{Yang2023} formalise this data-centric philosophy,
demonstrating that the LoRA-based fine-tuning pipeline keeps the cost of
producing a financially capable model accessible to teams without proprietary
infrastructure while still achieving competitive performance on standard financial
NLP tasks.  Benchmark performance lags BloombergGPT on tasks requiring deep
financial knowledge encoded in the pre-training corpus, but FinGPT's modular,
data-efficient approach makes it the natural choice for teams without the resources
for pre-training at scale.

\textit{FinMA (via PIXIU).}  \citet{xie2023pixiu} introduce FinMA as part of
the PIXIU suite: a 7-billion-parameter LLaMA model instruction-tuned on a
curated dataset of 136,000 financial instruction--response pairs spanning
sentiment analysis, named entity recognition, news headline classification,
financial question answering, and stock movement prediction.  The value of
reframing even small amounts of supervised data into instruction-tuning format
is underscored by \citet{Zhang2023}, who show that transforming a limited set
of financial sentiment examples into instructions yields strong improvements in
sentiment classification, validating the approach for practitioners with
constrained labelled datasets.  FinMA achieves strong performance on tasks
covered by its instruction dataset but, as the authors document, generalises
less reliably to tasks and data formats not seen during instruction tuning---a
common limitation of instruction-tuned models when deployed outside their
training distribution.  \citet{Hirano2024b} propose an alternative that sidesteps
this data requirement entirely: by combining continual domain pre-training with
model merging, they build instruction-following financial LLMs without any
instruction data, offering a practical path for teams in low-resource settings
who cannot curate large instruction corpora.

\textit{InvestLM.}  \citet{yang2023investlm} take a complementary approach:
they instruction-tune LLaMA-65B on a smaller but highly curated set of financial
instructions derived from CFA exam questions, investment research reports, and
earnings call analyses.  The resulting model is evaluated on practical investment
tasks including portfolio commentary generation, risk factor identification, and
earnings analysis, where its larger base model capacity allows it to outperform
7B instruction-tuned models despite seeing fewer fine-tuning examples.

\begin{remark}[The encoder--decoder choice in practice]
The selection between encoder-only and decoder-only models is not merely a
theoretical design decision; it has immediate operational implications.
Encoder-only models (FinBERT, SEC-BERT) are compact, fast to fine-tune, and
well-understood for classification tasks.  Decoder-only models (BloombergGPT,
FinGPT) are an order of magnitude larger, require more memory at inference time,
and are harder to calibrate for structured prediction.  A useful heuristic: if
the task has a fixed label space (positive/negative/neutral, default/no-default),
start with an encoder model.  If the task requires open-ended output
(draft a risk section, answer a question from a 10-K), a decoder model is necessary.
Hybrid retrieval-augmented generation architectures---discussed in
Chapter~\ref{ch:llm-agents}---can sometimes substitute for a larger model by
connecting a smaller decoder to a well-indexed financial document store.
\end{remark}

% ---------------------------------------------------------------
\section{Pre-training Strategies for Finance}
\label{sec:ch08-pretraining-strategies}

\begin{context}
When Bloomberg decided to build BloombergGPT, they faced a choice every
team contemplating a domain-specific model must confront: train a new model
from scratch on a mixed general-and-financial corpus, or start from an existing
general-purpose model and continue pre-training on financial text?  The from-scratch
option offers the cleanest architectural control---tokeniser, positional encoding,
and corpus weights can all be optimised jointly for the financial domain---but it
requires compute and data resources that very few organisations possess.  The
continued pre-training option, known as domain-adaptive pre-training (DAPT),
leverages the already-learned representations of a general foundation model and
extends them toward the target domain at a fraction of the computational cost.
Understanding both strategies, and the conditions under which each is preferable,
is essential for any team designing a financial NLP system.
\end{context}

% ---------------------------------------------------------------
\subsection{Domain-Adaptive Pre-training (DAPT)}
\label{sec:ch08-dapt}

The systematic study of domain-adaptive pre-training was formalised by
\citet{gururangan2020dont}, who trained RoBERTa on four domains---biomedicine,
computer science, news, and reviews---and demonstrated consistent downstream
improvement when the pre-training corpus was restricted to the target domain,
even when the base model had already seen some domain text during its initial
pre-training.  The effect was largest when the target domain was most dissimilar
from the original training distribution.  Finance, with its specialised lexicon
and numeric density, constitutes a domain that is substantially out-of-distribution
for models trained on general web text, suggesting that DAPT should yield
particularly large gains in this context.

DAPT operates by continuing the original pre-training objective---masked language
modelling for encoder models, causal language modelling for decoders---on the
domain corpus, using the pre-trained weights as initialisation.  The training
budget is typically a small fraction of the original pre-training budget: it is
common to process 1--10~billion tokens of domain text, compared to hundreds of
billions for the initial pre-training.  The learning rate is usually set lower
than the original pre-training rate (a factor of 10 is typical) to prevent
catastrophic forgetting of the general representations.  Empirical support for
this methodology extends beyond English-language models: \citet{Hirano2024a}
demonstrate through a Japanese financial LLM that continual pre-training on a
domain-specific corpus consistently improves performance on financial benchmarks,
confirming that the gains documented for English in \citet{gururangan2020dont}
generalise across languages and financial sub-domains.

\begin{definition}[Domain-Adaptive Pre-training]
\label{def:ch08-dapt}
Let $\theta_0$ denote the parameters of a pre-trained language model $M_0$, and
let $\mathcal{D}_\mathrm{domain}$ denote an unlabelled corpus from the target domain.
\emph{Domain-adaptive pre-training} (DAPT) produces updated parameters
\begin{equation}
  \theta^\star = \arg\min_\theta \, \mathcal{L}_\mathrm{PT}(\theta;\, \mathcal{D}_\mathrm{domain}),
  \label{eq:ch08-dapt}
\end{equation}
where $\mathcal{L}_\mathrm{PT}$ is the pre-training loss (MLM or CLM) and the
optimisation is initialised at $\theta_0$ with a reduced learning rate to
mitigate catastrophic forgetting of the original model's representations.
The resulting model $M^\star(\theta^\star)$ is then fine-tuned on labelled task
data for downstream evaluation.
\end{definition}

The practical question in DAPT is how much domain data is enough.  Empirically,
the gain from DAPT exhibits diminishing returns: the largest improvements occur
in the first billion tokens of domain text, with smaller incremental gains
thereafter.  For encoder models (BERT-scale, 110--340M parameters),
1--5~billion domain tokens is generally sufficient to capture the domain's lexical
distribution.  For decoder models at the 7--70B parameter scale, larger domain
corpora are needed to update a proportionally larger parameter set.

\textit{Task-adaptive pre-training (TAPT).}  A variant introduced in the same
paper \cite{gururangan2020dont} further fine-tunes the DAPT model on unlabelled
examples drawn from the specific task distribution (e.g., the unlabelled portion
of the sentiment dataset) before the final supervised fine-tuning step.  TAPT
consistently improves performance but requires task-specific unlabelled data,
which may not be available in all financial applications.

\textit{Catastrophic forgetting.}  A recurring concern with DAPT is that
continued pre-training on a narrow domain corpus may degrade performance on
general language tasks.  In practice, the degradation is modest when the
learning rate is appropriately reduced and the domain corpus is sufficiently
large relative to the total tokens processed.  Replay strategies---mixing a
small fraction of general pre-training data into the domain corpus---can further
reduce forgetting at modest additional cost \cite{gururangan2020dont}.

% ---------------------------------------------------------------
\subsection{Corpus Composition: Bloomberg Archives, EDGAR, Financial News}
\label{sec:ch08-corpus-composition}

The quality of a domain-specific financial LLM depends critically on the
composition, quality, and diversity of its training corpus.  A poorly composed
corpus---too narrow, too noisy, or with systematic gaps---will produce a model
that is brittle outside the narrow distribution it was trained on.

\textit{Bloomberg archives.}  The Bloomberg News archive is the gold standard
for financial news pre-training.  It contains decades of professionally edited
financial news articles with consistent terminology, calibrated tone, and high
factual density.  BloombergGPT's FinPile dataset \cite{wu2023bloomberggpt} draws
on Bloomberg news as one of its major components.  The archive is proprietary,
however, which is a significant barrier for academic and open-source projects.

\textit{SEC EDGAR.}  The SEC's Electronic Data Gathering, Analysis, and Retrieval
system provides free public access to all US public company filings since 1996:
10-K annual reports, 10-Q quarterly reports, 8-K current reports, proxy statements,
and registration statements \cite{sec2009xbrl}.  EDGAR constitutes perhaps the most
valuable open financial corpus: it covers the full universe of US public companies,
it is updated continuously, and its formal regulatory language represents a domain
that is highly distinctive and practically important.  A significant limitation is
that EDGAR text quality is uneven---filings from small companies often contain
boilerplate legal text that is informationally redundant---and the XML structure
requires careful parsing to extract prose sections from financial tables and
metadata.

\textit{Financial news aggregators.}  Open sources such as Reuters, Associated
Press financial feeds, Seeking Alpha, and financial subreddits provide broader
topical coverage but lower editorial quality than Bloomberg archives.  The
Reuters TRC2 corpus \cite{araci2019finbert} is a standard benchmark source.

\textit{Analyst reports.}  Buy-side and sell-side research reports are among the
richest and most technically precise financial texts available, but they are
almost entirely proprietary.  Licensing them at pre-training scale is cost-prohibitive
for all but the largest institutions.

\begin{example}[Corpus composition for a central bank communication model]
\label{ex:ch08-corpus-composition}
Suppose a research team at a central bank wishes to build a model specialised
in monetary policy communication analysis.  Their goal is to classify FOMC
statement paragraphs by their hawkish or dovish orientation---a task that
requires the model to understand the precise vocabulary of monetary policy:
``accommodation,'' ``balance sheet normalisation,'' ``inflation expectations,''
``neutral rate.''

A well-composed pre-training corpus for this application would include:
\begin{enumerate}
  \item All FOMC statements and minutes from 1994 to present (approximately
        50M tokens): the primary target distribution.
  \item All other major central bank publications---ECB introductory statements,
        Bank of England Inflation Reports, Fed Beige Books
        (approximately 200M tokens): broadens the monetary policy register.
  \item Academic papers on monetary policy from SSRN and the NBER working paper
        series (approximately 500M tokens): provides precise economic terminology
        in expository prose.
  \item Financial news articles mentioning monetary policy keywords (1--2B tokens):
        provides context about how policy is discussed and interpreted in practice.
\end{enumerate}
This corpus is deliberately narrow.  The model would not be competitive on general
financial tasks like earnings sentiment or credit scoring.  But for the specific
task of monetary policy communication analysis \cite{hansen2018transparency}, a
model pre-trained on this corpus and fine-tuned on a labelled set of hawkish/dovish
paragraphs will substantially outperform a general-purpose model fine-tuned on the
same labels, because the pre-training corpus aligns closely with both the domain
vocabulary and the pragmatic conventions of central bank language.
\end{example}

Table~\ref{tab:ch08-corpus-summary} summarises the major corpus sources available
for financial LLM pre-training, with approximate scale, quality, and accessibility.

\begin{table}[htbp]
\caption{Summary of major corpus sources for financial LLM pre-training.}
\label{tab:ch08-corpus-summary}
\centering
\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lllll}
\toprule
\textbf{Source} & \textbf{Scale (tokens)} & \textbf{Quality} & \textbf{Access} & \textbf{Coverage} \\
\midrule
Bloomberg News      & $\sim$300B    & Very high & Proprietary & News, markets \\
SEC EDGAR           & $\sim$50B     & High      & Open        & Regulatory, corporate \\
Reuters TRC2        & $\sim$2B      & High      & Licensed    & Financial news \\
Financial PhraseBank & $\sim$5M     & High      & Open        & Sentiment-labelled \\
Seeking Alpha       & $\sim$10B     & Medium    & Open/API    & Retail analysis \\
Earnings transcripts & $\sim$5B     & High      & Commercial  & Corporate comms \\
SSRN / arXiv finance & $\sim$2B    & High      & Open        & Academic research \\
General web (Pile)  & $\sim$800B    & Mixed     & Open        & Broad coverage \\
\bottomrule
\end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Benchmark Comparisons}
\label{sec:ch08-benchmark-comparisons}

Rigorous evaluation of financial LLMs requires benchmarks that test the
capabilities most relevant to financial practice: numerical reasoning,
domain-specific sentiment, information extraction from regulatory text, and
multi-step question answering over financial documents.  The field has converged
on a set of standard benchmarks, though their limitations---narrow task scope,
label quality, and susceptibility to overfitting---must be understood alongside
the numbers they report.

% ---------------------------------------------------------------
\subsection{Financial NLP Benchmarks: FinQA, FLUE, FinBen}
\label{sec:ch08-benchmarks}

\textit{FinQA.}  Introduced by \citet{chen2021finqa}, FinQA is a dataset of
8,281 question-answer pairs grounded in 10-K and 10-Q filings from S\&P~500
companies.  Each question requires a chain of arithmetic operations over one or
more tables and prose passages from the source document.  A representative question
might ask: ``What was the percentage change in operating income from 2018 to 2019,
given the following income statement?''  The answer requires locating the relevant
rows, extracting the numeric values, and performing the division, all without
arithmetic errors.  FinQA is widely considered the most challenging standard
benchmark for financial NLP because it simultaneously tests reading comprehension,
table parsing, and numerical reasoning.  On the public leaderboard, the best
fine-tuned specialist systems achieve execution accuracy around 68--72\%, while
general-purpose zero-shot GPT-4 class models achieve approximately 55--60\%.

\textit{FLUE.}  The Financial Language Understanding Evaluation benchmark
\cite{shah2022flue} is a multi-task suite designed to mirror GLUE/SuperGLUE for
the financial domain.  It includes five tasks: financial sentiment analysis (over
Financial PhraseBank), news headline classification (whether a news headline is
bearish, bullish, or neutral for a named company), named entity recognition over
financial text, question answering (FiQA), and semantic textual similarity of
financial sentences.  FLUE allows direct comparison of models across a range of
NLU tasks without the complexity of multi-step reasoning, making it well suited
to evaluating encoder-only models.  FinBERT and its variants consistently
outperform general BERT and RoBERTa on the FLUE tasks, with the largest margins
on the sentiment and NER sub-tasks.

\textit{FinBen.}  The most comprehensive recent benchmark is FinBen
\cite{xie2024finben}, which assembles 36 datasets spanning 24 financial
tasks across five capability categories: information extraction, text analysis,
question answering, text generation, and risk management.  FinBen is
specifically designed to evaluate the open-ended generation capabilities of
decoder-only models in addition to the classification capabilities tested by FLUE.
It includes stress-test tasks such as credit scoring from textual loan descriptions,
financial forecasting from annual report narratives, and multi-document
summarisation of earnings call transcripts---tasks that require a model to
integrate information across multiple sources and produce coherent structured output.
A complementary benchmarking methodology is demonstrated by \citet{CookKazinnik2023},
who evaluate local (domain-specific) LLMs on financial and economic text using
bank earnings calls as the evaluation domain; their framework is notable for
its practical orientation, covering tasks that matter to central bank researchers
and providing design principles transferable to institutional deployment settings
where existing public benchmarks may not align with the target task distribution.

\begin{definition}[Execution Accuracy]
\label{def:ch08-execution-accuracy}
For a numerical reasoning benchmark such as FinQA, \emph{execution accuracy}
is defined as the fraction of questions for which the model's predicted numerical
answer matches the gold answer within a tolerance $\epsilon$:
\begin{equation}
  \mathrm{EA} = \frac{1}{|\mathcal{Q}|} \sum_{q \in \mathcal{Q}}
  \mathbf{1}\!\left[\,\left|\hat{a}_q - a_q\right| \leq \epsilon\right],
  \label{eq:ch08-execution-accuracy}
\end{equation}
where $\hat{a}_q$ is the model's predicted answer and $a_q$ is the gold answer
for question $q$.  The tolerance $\epsilon$ is typically set to zero for
exact-match evaluation or to a small relative threshold (e.g., $1\%$) to allow
for rounding differences.
\end{definition}

% ---------------------------------------------------------------
\subsection{When Do FinLLMs Outperform General Models? When Do They Lose?}
\label{sec:ch08-finllm-vs-general}

The empirical record from FLUE, FinQA, and FinBen reveals a nuanced pattern:
domain-specific financial LLMs reliably outperform general models on tasks that
are deeply tied to domain vocabulary and syntax, but general models---particularly
at the frontier of scale---are competitive or superior on tasks requiring
broad world knowledge, multi-step reasoning, and instruction following.

\textit{Where FinLLMs win.}  The performance gap is largest on tasks that require
correct handling of domain-specific lexical semantics.  On Financial PhraseBank
sentiment classification, FinBERT \cite{araci2019finbert} achieves F1 scores
of approximately 0.88 compared to 0.72 for general BERT-base and 0.80 for
RoBERTa-base---a gap that persists even when the general models are fine-tuned
on the same training set.  The explanation is straightforward: sentiment labels
in financial text depend on domain vocabulary whose correct representation was
only learned during financial domain pre-training.  Similarly, on EDGAR named
entity recognition, SEC-BERT \cite{loukas2022secbert} substantially outperforms
general BERT because the distribution of entities and their surface forms in
legal-regulatory text differs substantially from news or web text.

\textit{Where general models win or are competitive.}  On tasks requiring broad
world knowledge---such as answering questions about macroeconomic events, regulatory
history, or geopolitical risk factors that appear in financial text---frontier
general models (GPT-4, Claude 3, Llama-3 70B) are competitive with domain-specific
models.  More strikingly, on multi-step numerical reasoning tasks like FinQA,
chain-of-thought prompting \cite{wei2022chain} applied to a large general model
often matches or exceeds the performance of a smaller domain-specific model,
because the reasoning capability at this level of complexity is more a function
of model scale than of domain-specific pre-training.

\textit{The scaling crossover.}  One of the clearest findings to emerge from the
literature is that the performance advantage of domain-specific models is
\emph{inversely proportional to model scale at a fixed task difficulty}.  For small
models (7B parameters and below), DAPT on financial text provides large, consistent
gains across most financial NLP tasks.  For models at the 70B+ scale, the gains
are narrower and task-specific: they persist for lexically sensitive tasks
(sentiment, tone, NER) but largely disappear for tasks whose difficulty is
primarily about reasoning over long documents.  \citet{RahimikiaDrinkall2024} sharpen this
picture by demonstrating that year-specific, domain-pre-trained financial LLMs
outperform general-purpose models up to 50 times their parameter count in return
prediction tasks---a particularly striking result because it controls for the
look-ahead bias that inflates many published performance figures---providing
direct empirical support for the value of targeted domain-adaptive pre-training
even in a period when frontier general models are rapidly improving.
The implication for practitioners is that the correct architecture choice depends
on the available compute budget: a team constrained to a small model should invest
in domain-specific pre-training; a team with access to large-scale inference
should evaluate whether prompting a frontier general model already meets their
accuracy threshold.

\begin{remark}[The contamination problem]
A persistent methodological concern in financial LLM evaluation is \emph{data
contamination}: the possibility that a benchmark dataset has appeared in the
model's pre-training corpus.  Several large general models trained on internet-scale
corpora are suspected to have seen the Financial PhraseBank or FinQA training
splits as part of their pre-training data, inflating their apparent performance.
Rigorous evaluation requires held-out test sets with temporal cutoffs ensuring
that the test data postdates the model's knowledge cutoff, a condition that most
published benchmarks do not currently enforce.
\end{remark}

\begin{remark}[Hallucination in financial contexts]
Domain-specific pre-training does not eliminate hallucination and may in some
cases worsen it by training the model to generate plausible-sounding but
unverified financial statistics.  \citet{kang2023hallucination} document
systematic hallucination in financial LLMs, finding that models trained on
financial news are more likely to fabricate specific numerical claims (earnings
figures, price changes, percentages) than general models given the same prompt.
The mitigation strategies discussed in Chapter~\ref{ch:applications-future}---
retrieval augmentation, constrained generation, and automated factual
verification---are therefore as important for domain-specific models as for
general ones.
\end{remark}

% ---------------------------------------------------------------
\section{Practical Deployment Considerations}
\label{sec:ch08-deployment}

The decision to deploy a domain-specific financial LLM in a production environment
raises a set of practical considerations that are orthogonal to the model's
benchmark performance.  A model that achieves state-of-the-art accuracy on FinQA
but cannot be licensed for commercial use, requires a compute budget that the
organisation cannot sustain, or generates outputs that cannot be traced to source
documents for audit purposes, is not a deployable model.  This section addresses
the three most consequential practical dimensions: licensing and data governance,
hosting strategy, and the cost-versus-performance trade-off that motivates hybrid
deployment architectures.

% ---------------------------------------------------------------
\subsection{Licensing, Hosting, and Data Governance}
\label{sec:ch08-licensing-hosting}

\textit{Model licensing.}  Financial LLMs span a wide range of licensing regimes.
BloombergGPT is proprietary and not publicly released; access is restricted to
Bloomberg terminal subscribers.  FinBERT \cite{araci2019finbert} and the
Yang--Huang FinBERT \cite{yang2020finbert} are released under the Apache 2.0
license, permitting commercial use and modification.  FinGPT \cite{Yang2023}
is released under MIT.  InvestLM \cite{yang2023investlm} and FinMA
\cite{xie2023pixiu} are released under licenses that permit research use but
restrict commercial deployment (following the LLaMA-1 and LLaMA-2 license
structures respectively).  Any deployment decision must begin with a license audit:
does the intended use case fall within the permitted uses of the base model license,
the fine-tuning dataset license, and any third-party libraries used in training?

\textit{Data governance.}  Financial institutions operate under data governance
frameworks---GDPR in Europe, GLBA and CCPA in the United States, and sector-specific
regulations---that constrain how customer data may be used to train or fine-tune
models.  Customer financial data (account balances, transaction histories, credit
scores) may not, in most jurisdictions, be used to train models deployed for other
customers without explicit consent and appropriate data protection agreements.
This constraint is particularly acute when the desired fine-tuning task requires
actual customer data: a model fine-tuned on real loan application text to predict
default probability may not be trainable if the training data was not collected
with model training as a declared purpose.  Synthetic data generation---using a
general model to produce realistic but fictitious financial training examples---is
an increasingly common workaround, though it introduces its own distributional
shift concerns.

\textit{Hosting strategy.}  The choice between cloud-hosted API access and
on-premises deployment is driven by cost, latency, data residency, and the
organisation's risk appetite for vendor dependency.  For encoder-only models
like FinBERT (110M--340M parameters), on-premises deployment on commodity GPU
hardware is straightforward: inference on a single A10G or equivalent GPU provides
sub-50ms latency for classification tasks, and the full model fits comfortably in
memory.  For decoder-only models at the 7--50B parameter scale, on-premises
deployment requires multi-GPU or specialised inference hardware.  Quantisation
techniques---GPTQ \cite{frantar2022gptq}, AWQ \cite{lin2023awq}, and
QLoRA \cite{dettmers2023qlora}---can reduce memory requirements by 2--4$\times$
at modest accuracy cost, making on-premises deployment of 13B--30B parameter
models feasible on a single high-memory GPU server.  For BloombergGPT-class
models (50B+ parameters), on-premises deployment at production throughput
typically requires a cluster-scale investment that is beyond all but the largest
financial institutions.

% ---------------------------------------------------------------
\subsection{Cost vs.\ Performance Trade-offs: Hybrid Strategies}
\label{sec:ch08-cost-performance}

The cost of operating a financial LLM at production scale is a function of three
factors: inference compute (measured in tokens per second per dollar), model
accuracy (the fraction of tasks for which the model produces an acceptable output),
and the cost of human review for cases where the model's confidence falls below
a threshold.  These factors interact in a way that makes simple benchmark
comparisons insufficient: a model that achieves 5\% higher accuracy but costs
10$\times$ more per inference may be inferior to a cheaper model supplemented
by human review for the small fraction of hard cases.

\begin{definition}[Effective cost per correct output]
\label{def:ch08-effective-cost}
Let $c_m$ denote the per-query inference cost of model $m$, $c_h$ the cost
of a human review, $p_m$ the model's probability of producing a correct output
without human review, and $\rho$ the fraction of queries routed to human review.
The \emph{effective cost per correct output} is:
\begin{equation}
  C_{\mathrm{eff}}(m, \rho) = \frac{c_m + \rho \, c_h}{p_m + \rho(1 - p_m)},
  \label{eq:ch08-effective-cost}
\end{equation}
where we assume human review recovers all incorrectly classified queries.
An optimal deployment minimises $C_{\mathrm{eff}}$ subject to a minimum accuracy
constraint $p_m + \rho(1-p_m) \geq p^\star$.
\end{definition}

\textit{The hybrid routing pattern.}  The most cost-effective deployment pattern
for many financial applications is a two-stage system: a small, fast, domain-specific
model handles the majority of queries; a subset of queries deemed hard (by low
confidence score or explicit classifier) is routed to either a larger model or
a human reviewer.  This pattern is particularly effective when the task difficulty
distribution is skewed: most financial text classification tasks have a large
``easy'' majority that a FinBERT-scale model handles accurately, and a small
``hard'' tail that requires deeper reasoning.

\textit{Retrieval-augmented generation as a complement.}  For tasks requiring
factual precision---such as extracting specific figures from a 10-K or answering
questions about a company's disclosed risk factors---retrieval-augmented generation
(RAG) \cite{lewis2020rag} can substitute for a larger domain-specific model.
A compact decoder model (7--13B parameters, potentially a fine-tuned general model
rather than a dedicated FinLLM) grounded in a well-indexed EDGAR document store
can match the accuracy of a larger ungrounded model on document-specific tasks
while operating at a fraction of the inference cost and with full traceability
to source documents---a regulatory requirement in many financial AI deployments
\cite{sr117}.  The modular approach is taken further by
\citet{Liu2024}, who extend the FinGPT framework to retrieval-augmented search
agents tailored separately for individual and institutional users, demonstrating
how domain-specific financial LLMs can be customised for real-world deployment
subject to privacy constraints, latency requirements, and the heterogeneous
information needs of different user classes.

\begin{example}[Cost-optimal architecture for a regulatory filing classifier]
\label{ex:ch08-deployment}
A compliance team at an asset manager needs to classify the ``risk factor'' sections
of 10-K filings into a taxonomy of 15 risk categories (market risk, credit risk,
regulatory risk, litigation risk, etc.) for a portfolio of 2,000 companies updated
annually.  Each filing contains 3--10 risk factor paragraphs.

At a naive per-token pricing of \$0.002 per 1K tokens, using a 70B general model
via API for all 20,000--60,000 paragraphs costs approximately \$200--\$600 per
annual update cycle.  A fine-tuned FinBERT-base classifier ($\sim$110M parameters)
deployed on a single T4 GPU can classify 1,000 paragraphs per second at a monthly
server cost of approximately \$400, amortised over the full year of operation.

If the FinBERT classifier achieves 88\% accuracy and the 70B model achieves 93\%,
the hybrid strategy---use FinBERT for all paragraphs, route the 12\% with
confidence below a threshold to the large model---achieves 92\%+ accuracy at
a blended cost of approximately \$250 per annual update cycle, lower than either
model alone.  The break-even accuracy-cost point is highly sensitive to the shape
of the confidence distribution, which must be measured empirically on a held-out
validation set.
\end{example}

% ---------------------------------------------------------------
\section{Summary}
\label{sec:ch08-summary}

This chapter has developed a systematic account of domain-specific financial LLMs
from the motivation through deployment.  The key takeaways are as follows.

Financial language constitutes a distinct register characterised by specialised
vocabulary, high numeric density, and tabular structure.  These features create
a systematic performance gap between general-purpose models and the financial
distribution.  The LM financial dictionary \cite{loughran2011liability} provides
perhaps the starkest quantification of this gap: hundreds of common English words
have inverted or neutralised polarity in financial text, and any model that has not
learned this inversion will misclassify financial sentiment at a rate that makes
it unsuitable for production use.

The financial LLM landscape divides into two families.  Encoder-only models
(FinBERT, SEC-BERT, FinBERT-tone) are compact, well-understood, and highly
effective for classification and extraction tasks where the output space is fixed.
Decoder-only models (BloombergGPT, FinGPT, FinMA, InvestLM) are substantially
larger but capable of open-ended generation, multi-step numerical reasoning, and
instruction following---capabilities that encoder models cannot support.

Domain-adaptive pre-training \cite{gururangan2020dont} is the most broadly
applicable strategy for building financial LLMs.  By continuing a general model's
pre-training objective on a financial corpus, DAPT shifts the model's representations
toward the financial distribution at a fraction of the cost of training from scratch.
The optimal corpus blends high-quality financial sources (EDGAR, Bloomberg-class
news, earnings transcripts) with enough general text to preserve broad reasoning
capabilities.

Benchmark evaluation reveals that FinLLMs reliably outperform general models on
lexically sensitive tasks (sentiment, tone, NER) but lose their advantage on
large-scale reasoning tasks at sufficient model size.  The FLUE, FinQA, and FinBen
benchmarks provide a useful but imperfect proxy for real-world performance; data
contamination and distributional mismatch between benchmark and deployment data
are persistent concerns.

Production deployment decisions must account for licensing, data governance, hosting
cost, and the accuracy-cost trade-off.  Hybrid architectures that route easy
queries to compact domain-specific models and hard queries to larger models or
human reviewers dominate simple single-model deployments on the Pareto frontier
of accuracy and cost.

% ---------------------------------------------------------------
\section{Further Reading}
\label{sec:ch08-further-reading}

The foundational paper for domain-adaptive pre-training is
\citet{gururangan2020dont}, which should be read in conjunction with the
original BERT paper \cite{devlin2019} and the RoBERTa pretraining improvements
\cite{liu2018} to understand what the domain-adaptive stage modifies and what
it inherits.  The BloombergGPT technical report \cite{wu2023bloomberggpt} is
the most detailed public account of what pre-training a 50B-scale financial model
involves, including corpus design, tokeniser construction, and benchmark evaluation.
For the open-source counterpart, \citet{Yang2023} describe the FinGPT
framework and the LoRA-based fine-tuning pipeline in detail accessible to practitioners
without cluster-scale resources.

For benchmarking, \citet{chen2021finqa} remains the canonical reference for
numerical reasoning evaluation; the FinBen paper \cite{xie2024finben} provides
the most comprehensive recent overview of the full task landscape.  On the
evaluation methodology side, \citet{shah2022flue} discuss the design principles
behind FLUE and the limitations of each sub-task as a proxy for real-world performance.

The Loughran--McDonald dictionary paper \cite{loughran2011liability} and the
subsequent review \cite{loughran2020textual} are essential reading for understanding
the lexical properties of financial language that motivate domain-specific
pre-training.  \citet{huang2023finbert} provide the most thorough accounting
research perspective on FinBERT's capabilities and limitations.

For deployment and governance, the SR~11-7 model risk management guidance
\cite{sr117} remains the authoritative framework for deploying
models in regulated financial contexts.  Chapter~\ref{ch:applications-future}
of this book extends the discussion to the EU AI Act and MiFID~II obligations
arising from LLM-based financial products.  Readers interested in the broader
landscape of financial LLM applications should consult the survey by
\citet{li2023llmfinsurvey}.

% ---------------------------------------------------------------
\section*{Exercises}

\textbf{[B]} 1. Explain in your own words why financial text poses unique tokenisation challenges for general-purpose LLMs. Give two examples of financial terms or expressions that a standard BPE tokeniser trained on Wikipedia would likely segment incorrectly.

\textbf{[I]} 2. BloombergGPT was trained on a 708-billion-token corpus mixing proprietary financial data with public text. Discuss the trade-off between domain specificity and general capability. Under what conditions would you prefer BloombergGPT over GPT-4o for a financial NLP task, and vice versa?

\textbf{[A]} 3. Suppose you are fine-tuning a 7B-parameter decoder-only LLM on a dataset of 10,000 quarterly earnings call transcripts annotated with analyst sentiment scores. Design a complete fine-tuning pipeline: specify the base model, adaptation technique (full fine-tuning, LoRA, QLoRA), training objective, evaluation metric, and one ablation you would run to validate the domain-adaptation benefit.

% ---------------------------------------------------------------
\section{Data Lab}\label{sec:ch08-data-lab}

The following exercises use live data from the SEC EDGAR REST API
(\url{https://data.sec.gov}). Starter code is in
\texttt{code/notebooks/08-domain-specific-llms/exercises.ipynb}, Data Lab section.

\begin{exercise}[Finance-Domain Vocabulary Size]\label{ex:ch08-edgar-b}
\textbf{[B]} Download the MD\&A section from the most recent 10-K for three
companies: AAPL (technology), JPM (banking), and XOM (energy).
For each company, compute:
(\textit{a})~total token count, unique token count, and vocabulary density
(unique/total);
(\textit{b})~the fraction of tokens that appear only once (hapax legomena);
(\textit{c})~a grouped bar chart comparing these metrics across the three companies.
Discuss whether financial filings exhibit greater or lower lexical richness
than general English text.
See \texttt{code/notebooks/08-domain-specific-llms/exercises.ipynb}, Data Lab section.
\end{exercise}

\begin{exercise}[Out-of-Vocabulary Rate vs.\ General English]\label{ex:ch08-edgar-i}
\textbf{[I]} Using a hardcoded set of the 500 most common English words, compute
the out-of-vocabulary (OOV) rate for the MD\&A vocabulary of each company from
Exercise~\ref{ex:ch08-edgar-b}: the fraction of unique finance tokens absent from
the common-English set.
(\textit{a})~Report OOV rates across sectors and plot as a horizontal bar chart.
(\textit{b})~List the 20 most frequent tokens \emph{not} in the common-English set.
(\textit{c})~Discuss the implications for tokeniser vocabulary design in
domain-adapted language models.
See \texttt{code/notebooks/08-domain-specific-llms/exercises.ipynb}, Data Lab section.
\end{exercise}

\begin{exercise}[Domain Vocabulary Drift Across Annual Reports]\label{ex:ch08-edgar-a}
\textbf{[A]} For Microsoft (\texttt{MSFT}), download the five most recent
annual 10-K filings and extract the MD\&A section from each.
(\textit{a})~Compute the Jaccard similarity of the MD\&A vocabulary between
each pair of consecutive years.
(\textit{b})~Plot vocabulary similarity as a time series.
(\textit{c})~Identify the year with the lowest similarity to its predecessor
and discuss what might have driven the shift (new technologies, products,
regulatory language, or macroeconomic conditions).
See \texttt{code/notebooks/08-domain-specific-llms/exercises.ipynb}, Data Lab section.
\end{exercise}
