\chapter{Credit Risk and Large Language Models}
\label{ch:credit-risk}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Describe the classical structural and reduced-form models of credit risk and
        identify their key assumptions and empirical limitations.
  \item Explain how machine learning and NLP methods improve the prediction of
        probability of default beyond traditional credit-scoring approaches.
  \item Apply LLMs to extract soft information from loan narratives, earnings calls,
        and corporate filings and explain how this information complements hard
        financial data.
  \item Articulate the fairness and disparate-impact concerns that arise when
        algorithmic models replace human judgment in lending decisions.
  \item Describe the model risk management framework (SR~11-7) as it applies to
        AI-based credit models, including validation, explainability, and governance.
  \item Design a persona-based LLM simulation of a borrower and discuss its uses
        and limitations for stress-testing and behavioural credit analysis.
\end{enumerate}
\end{remark}

% ---------------------------------------------------------------
\section{Credit Risk: Concepts and Classical Models}
\label{sec:credit-basics}

Credit risk is the possibility that a counterparty will fail to meet a contractual
obligation, causing a financial loss to the lender or investor.  It is the oldest
and most pervasive risk in finance: every loan, bond, trade receivable, or derivative
exposure carries some probability that the obligor will default.  Understanding,
measuring, and pricing credit risk is therefore central to bank regulation, bond
valuation, and portfolio management.

Two intellectual traditions have shaped credit risk modelling.  The \emph{structural}
tradition, originating with the insight that equity is a call option on the firm's
assets, derives default as an endogenous event that occurs when asset value crosses a
liability threshold.  The \emph{reduced-form} tradition treats default as an
exogenous Poisson arrival whose intensity is specified directly, bypassing the need to
model asset dynamics.  Both traditions assume that the relevant information is
captured in observable financial variables.  The central message of this chapter is
that large language models open a third path: extracting latent credit-relevant
information from unstructured text, and simulating human decision-making through
persona-based agents.

% ---------------------------------------------------------------
\subsection{Probability of Default: Definitions and Estimation}
\label{sec:pd-estimation}

\begin{definition}[Probability of Default]
\label{def:pd}
Let $(\Omega, \mathcal{F}, \mathbb{P})$ be a probability space and let $\tau$ denote
the random \emph{default time} of an obligor.  The \emph{probability of default}
over a horizon $[0, T]$ is
\begin{equation}
  \mathrm{PD}(T) \;=\; \mathbb{P}(\tau \leq T).
  \label{eq:pd-def}
\end{equation}
When $T = 1$ year the quantity is called the \emph{one-year PD}, which is the primary
input to the Basel internal-ratings-based (IRB) credit risk formula.
\end{definition}

The classical statistical approach to estimating $\mathrm{PD}$ is \emph{logistic
regression}.  Given a vector of covariates $\mathbf{x}_i \in \mathbb{R}^p$ for
obligor $i$ (leverage ratio, interest-coverage ratio, liquidity, sector indicator,
macroeconomic state), and a binary outcome $y_i \in \{0, 1\}$ indicating default
within the horizon, the model specifies
\begin{equation}
  \mathbb{P}(y_i = 1 \mid \mathbf{x}_i) = \sigma(\mathbf{x}_i^\top \boldsymbol{\beta}),
  \label{eq:logit-pd}
\end{equation}
where $\sigma(z) = 1/(1 + e^{-z})$ is the logistic function.  The coefficient vector
$\boldsymbol{\beta}$ is estimated by maximum likelihood.

Altman and Sabato \cite{altman2007sme} demonstrated an early application of this
framework to small and medium-sized enterprises (SMEs): using a panel of over 2,000
U.S.\ firms with sales below \$65 million, they showed that an SME-specific logistic
model outperforms a generic corporate model by nearly 30\% in out-of-sample
discriminatory power, underscoring the importance of segment-specific credit
models.

\begin{remark}[Loss Given Default and Expected Loss]
\label{rem:lgd-el}
The probability of default is only one component of credit loss.  The \emph{expected
loss} over horizon $T$ is
\[
  \mathrm{EL} = \mathrm{PD}(T) \times \mathrm{LGD} \times \mathrm{EAD},
\]
where $\mathrm{LGD} \in [0,1]$ is the \emph{loss given default} (the fraction of
exposure lost if default occurs) and $\mathrm{EAD}$ is the \emph{exposure at default}.
This chapter focuses primarily on $\mathrm{PD}$ estimation, but the NLP methods
discussed below are equally applicable to modelling $\mathrm{LGD}$, which is strongly
influenced by the quality of collateral documentation and covenant language.
\end{remark}

% ---------------------------------------------------------------
\subsection{Machine Learning Extensions}
\label{sec:ml-credit}

The rise of machine learning has transformed credit scoring by allowing non-linear,
high-dimensional interactions among predictors that logistic regression cannot capture.
Gradient-boosted trees, random forests, and neural networks consistently outperform
logistic regression on held-out default prediction benchmarks, particularly for
consumer credit where behavioural signals (payment timing, account utilisation patterns)
display strong non-linearities.

A landmark paper by Fuster, Goldsmith-Pinkham, Ramadorai, and Walther
\cite{fuster2022predictably} uses U.S.\ mortgage data to compare traditional and
machine learning credit models and finds that machine learning substantially improves
default prediction accuracy.  However, they also document a troubling distributional
consequence: Black and Hispanic borrowers are disproportionately less likely to benefit
from the predictive gains of machine learning relative to white borrowers, because
machine learning exploits historical patterns that embed pre-existing socioeconomic
disparities.  Their equilibrium analysis shows that machine learning credit models can
widen racial gaps in interest rates and credit access, even in the absence of
discriminatory intent.  This finding motivates the regulatory focus on algorithmic
fairness discussed in Section~\ref{sec:fairness}.

Beyond structured financial data, the advent of ubiquitous digital platforms has
created rich alternative data sources.  Berg, Burg, Gombovic, and Puri
\cite{berg2020fintech} show, in a natural experiment exploiting variation in device
type and operating system at loan application, that simple digital-footprint variables
(whether an applicant uses an Apple versus Android device, whether they apply at
night) predict consumer default with information content comparable to traditional
credit-bureau scores.  Their results demonstrate that non-traditional data can expand
credit access for thin-file borrowers who lack a conventional credit history, while
raising questions about the fairness of using such proxies.

% ---------------------------------------------------------------
\section{Soft Information and NLP in Credit Markets}
\label{sec:nlp-credit}

Hard financial ratios capture the \emph{reported} financial position of a borrower.
Soft information---the qualitative judgment formed by a loan officer who meets the
borrower, reads the business plan, and assesses the management team---has long been
recognised as a complementary and partially non-transferable input to credit decisions.
NLP methods allow researchers and practitioners to operationalise soft information at
scale, converting unstructured text into quantitative signals that can enter
credit-scoring models alongside hard financial data.

% ---------------------------------------------------------------
\subsection{Text-Based Default Prediction}
\label{sec:text-default}

The intuition that language reveals credit quality is not new.  Loan officers have
always attended to the clarity, consistency, and tone of a borrower's narrative.  What
has changed is the ability to process that narrative systematically across thousands of
applications.

Frattaroli \cite{frattaroli2019} provides an early demonstration that the textual
content of loan officer memos contains information about future loan performance
beyond the hard financial data recorded in the loan file.  Memos with more negative
sentiment and greater linguistic hedging are associated with higher default rates,
consistent with loan officers embedding soft-information assessments in the language
of their write-ups.

More recently, the combination of large language models with corporate text has
produced powerful default-prediction signals.  Khoja \cite{khoja2024llm} shows that
LLM-processed earnings call transcripts---evaluated through a counterfactual
random-forest methodology---predict corporate bond credit spreads and reduce the
post-earnings announcement drift (PEAD) in bond markets by as much as 86.9\%,
an effect concentrated in investment-grade bonds where interpretive heterogeneity
is greatest.  The key advantage of LLMs over bag-of-words sentiment is their ability
to capture contextual negation, hedging, and domain-specific terminology---precisely
the linguistic features that matter most in credit-relevant disclosures.

A broader benchmark study, ``Empowering Many, Biasing a Few,'' by Feng, Dai, Huang,
Zhang, Xie, Han, Chen, Lopez-Lira, and Wang \cite{feng2023calm} introduces the Credit
Assessment Large Language Model (CALM) benchmark: nine datasets covering 14{,}000
samples spanning credit default prediction, risk rating, and bias examination.  Their
fine-tuned CALM model achieves strong generalisation across tasks but also reveals
systematic demographic biases inherited from the pre-training corpus---underscoring
that LLM-based credit scoring is not a neutral technology but one that requires
careful bias auditing before deployment.

% ---------------------------------------------------------------
\subsection{Earnings Call Transcripts and Bond Markets}
\label{sec:transcripts-bonds}

Earnings call transcripts are particularly valuable for credit analysis because they
provide high-frequency, unscripted soft information from senior management.  In the
context of equity markets, a large literature has documented that call tone predicts
future earnings surprises.  The credit analogue is equally compelling.

When management hedges more on a call---using phrases such as ``subject to market
conditions,'' ``we anticipate challenges,'' or ``we are monitoring the situation
closely''---the linguistic shift is often a leading indicator of deteriorating credit
quality, appearing before the deterioration shows up in reported financial ratios.
This temporal gap between the soft-information signal and the hard-data realisation
is the source of the incremental predictive value of NLP-based credit signals.

Lopez-Lira and Tang \cite{lopezlira2023chatgpt} show that ChatGPT sentiment scores
extracted from news headlines---without any credit-specific training---predict
subsequent stock price drift, particularly for small-cap stocks and negative news.
The same zero-shot capability extends naturally to bond markets: LLM sentiment from
earnings call transcripts and filings predicts credit spread changes and rating
migration in preliminary applications of the methodology.

% ---------------------------------------------------------------
\section{LLMs for Lending Decisions: Capabilities and Risks}
\label{sec:llm-lending}

Beyond using LLMs as feature extractors for credit models, a more ambitious
application is deploying LLMs as \emph{decision agents} that directly assess
creditworthiness, structure loan covenants, or simulate borrower behaviour.  This
section examines both the promise and the risks of that approach.

% ---------------------------------------------------------------
\subsection{Zero-Shot and Few-Shot Credit Assessment}
\label{sec:zeroshot-credit}

Large language models trained on broad corpora of financial text encode substantial
knowledge about credit analysis: financial ratios, default patterns, industry risk
factors, and covenant structures.  A prompt that provides a borrower's financial
summary and asks for a credit assessment will typically elicit a coherent and
informed response.

The systematic review by Integrating Interpretable LLMs for Credit Risk
\cite{interpretable2025} synthesises 60 papers from 2020--2025 and documents that
LLMs achieve competitive accuracy on structured credit classification tasks when
provided with appropriate chain-of-thought prompts, and that they outperform
classical ML on tasks requiring natural-language reasoning about qualitative risk
factors.  Crucially, the review distinguishes between \emph{intrinsic explainability}
(the model generates a natural-language justification for its assessment) and
\emph{post-hoc explainability} (SHAP or LIME applied to a black-box model), arguing
that LLMs offer a form of intrinsic explainability that may satisfy regulatory demands
for adverse action notice more naturally than traditional ML models.

However, zero-shot credit assessment also inherits all the failure modes of LLMs:
hallucination of financial facts, sensitivity to prompt phrasing, and the potential to
reproduce training-data biases in a non-transparent way.  For consumer lending, where
adverse action notice requirements impose legal obligations to explain denial reasons,
the opacity of LLM reasoning chains creates compliance risk unless the chain-of-thought
is faithfully recorded and auditable.

% ---------------------------------------------------------------
\subsection{Persona-Based Borrower Simulation}
\label{sec:persona-credit}

A distinctive application of LLMs in credit analysis is the construction of
\emph{synthetic borrower personas}: LLM agents that role-play a borrower with
specified demographic, financial, and behavioural characteristics, and simulate their
financial decisions over time.

The methodology draws on a growing literature showing that LLMs prompted with
demographic and psychographic personas can reproduce survey responses and economic
choices with surprising fidelity.  Korinek \cite{korinek2023llm} catalogues 25 use
cases in which LLMs augment economic research, including the simulation of survey
responses and the generation of synthetic respondents for pilot-testing research
designs.  In credit applications, persona simulation can serve several purposes:

\begin{enumerate}
  \item \textbf{Stress testing.}  A bank can prompt an LLM with a ``stressed
    household'' persona---a household that has recently lost one income source,
    has credit-card utilisation near the limit, and faces rising mortgage
    payments---and ask the agent to simulate the sequence of financial decisions
    (balance transfers, discretionary spending cuts, missed payments) over a
    six-month horizon.  The resulting simulated payment trajectories can be used to
    calibrate default models under stress scenarios that have not been observed in
    historical data.

  \item \textbf{Counterfactual fairness analysis.}  By varying only the protected
    attributes (race, gender, age) in the persona prompt while holding financial
    attributes constant, a bank can test whether its LLM-based underwriting system
    produces different credit decisions across demographic groups---providing a
    structured audit of disparate impact that complements statistical fairness
    metrics.

  \item \textbf{Synthetic data generation.}  Credit models trained on historical
    data suffer from class imbalance (defaults are rare) and sample selection bias
    (only approved borrowers appear in the training data).  LLM-generated synthetic
    borrower trajectories can augment the training set, particularly for
    underrepresented segments such as thin-file borrowers or small businesses in
    emerging markets.
\end{enumerate}

The ``Financial Brain Scan'' methodology of Chen, Didisheim, Somoza, and Tian
\cite{somoza2025brainscan} offers a complementary approach: rather than prompting the
LLM with an external persona, their technique identifies the internal conceptual
directions in an LLM's representation space that correspond to risk aversion,
optimism, or pessimism, and steers the model along those directions while holding
other factors constant.  Applied to credit decisions, this allows researchers to
simulate what a more or less risk-averse underwriter would decide, providing a
principled way to explore the sensitivity of credit outcomes to the implicit
preferences embedded in a deployed LLM.

% ---------------------------------------------------------------
\section{Fairness, Discrimination, and Regulatory Constraints}
\label{sec:fairness}

The deployment of algorithmic credit models---and LLMs in particular---in consumer
lending sits at the intersection of statistical science, law, and ethics.  In the
United States, the Equal Credit Opportunity Act (ECOA) and the Fair Housing Act
prohibit credit decisions that discriminate on the basis of race, color, religion,
national origin, sex, marital status, or age.  The Equal Credit Opportunity Act
further requires lenders to provide a statement of reasons when an adverse action is
taken.  Similar protections exist in the European Union under the General Data
Protection Regulation (GDPR) and the forthcoming EU AI Act.

The key legal concept is \emph{disparate impact}: even if a model does not explicitly
use protected characteristics as inputs, it may produce outcomes that disproportionately
harm protected groups.  Fuster et al.\ \cite{fuster2022predictably} demonstrate
empirically that this is not merely a theoretical concern: machine learning credit
models, trained on historical data that reflects historical discrimination, can
amplify disparities relative to traditional models even when the model itself is
facially neutral.

\begin{definition}[Disparate Impact]
\label{def:disparate-impact}
A credit model exhibits \emph{disparate impact} with respect to protected group $G$
if the approval rate for applicants in group $G$ is less than 80\% of the approval
rate for the most-favoured group:
\begin{equation}
  \frac{\mathrm{ApprovalRate}(G)}{\max_{G'} \mathrm{ApprovalRate}(G')} < 0.80.
  \label{eq:disparate-impact}
\end{equation}
The 80\% threshold is the ``four-fifths rule'' codified in the 1978 Uniform Guidelines
on Employee Selection Procedures and widely applied to lending by analogy.
\end{definition}

For LLMs specifically, the fairness challenge is compounded by the opacity of the
model's reasoning.  An LLM prompted to assess creditworthiness may produce disparate
outcomes because: (i) its pre-training corpus encodes historical associations between
demographic groups and creditworthiness; (ii) the persona-description language in the
prompt activates demographic stereotypes; or (iii) the model attends to linguistic
features that correlate with protected characteristics even when those features are
not intended as proxies.  The CALM benchmark \cite{feng2023calm} documents all three
mechanisms, finding that LLM credit scores vary systematically with implied borrower
race even when financial covariates are held fixed.

Mitigating these risks requires a combination of technical interventions (debiasing
pre-training data, adversarial fine-tuning, demographic parity constraints on
outputs) and governance measures (regular disparity testing, human oversight of
boundary cases, explainable adverse action notices).

% ---------------------------------------------------------------
\section{Model Risk Management for AI Credit Models}
\label{sec:model-risk}

Financial regulators have long required banks to manage model risk systematically.
In the United States, the Federal Reserve and the Office of the Comptroller of the
Currency's joint guidance document SR~11-7 (``Supervisory Guidance on Model Risk
Management,'' 2011) establishes the framework: every model used in a material
decision must be developed, validated, and governed according to documented standards.
Model risk management (MRM) encompasses three pillars: development and implementation,
validation, and ongoing monitoring.

The application of SR~11-7 to AI and LLM-based credit models raises new challenges
that the original 2011 guidance did not anticipate.  A logistic regression model has
a small, interpretable parameter vector that can be audited directly.  A large
language model has billions of parameters and its behaviour emerges from complex
non-linear interactions that cannot be inspected directly.

Key challenges include:

\begin{enumerate}
  \item \textbf{Explainability.}  SR~11-7 requires model validators to understand
    how a model produces its outputs.  For LLMs, post-hoc explainability tools such as
    SHAP (SHapley Additive exPlanations) and LIME (Local Interpretable Model-Agnostic
    Explanations) provide partial insight by attributing the model's output to input
    features, but they do not explain the model's internal reasoning.  The ``Interpretable
    LLMs for Credit Risk'' systematic review \cite{interpretable2025} distinguishes
    \emph{chain-of-thought} justifications (the model generates a natural-language
    explanation of its decision) from \emph{attention-based} explanations and
    \emph{probing} methods, and finds that chain-of-thought approaches best satisfy
    validator requirements for auditability.

  \item \textbf{Concept drift.}  LLMs encode a snapshot of linguistic knowledge from
    their pre-training corpus.  When the credit environment changes---a new type of
    fraud emerges, a sector undergoes restructuring, a regulatory change alters the
    meaning of financial covenants---the model's embedded knowledge may become
    stale without any change in its parameters.  Ongoing monitoring must detect
    concept drift through distributional shift tests on model inputs and output
    calibration checks.

  \item \textbf{Vendor opacity.}  When a bank deploys a third-party LLM (e.g.\ via
    an API), SR~11-7's documentation and validation requirements apply to the bank,
    not only to the model vendor.  But the bank may have no access to the model's
    weights, training data, or architecture.  This creates a structural tension between
    regulatory requirements and the commercial model of closed-source LLM providers.

  \item \textbf{Reproducibility.}  LLMs with non-zero temperature generate
    non-deterministic outputs.  The same input prompt may produce a different credit
    assessment on two queries.  Model risk management frameworks require reproducibility
    of model outputs; achieving this with LLMs requires setting temperature to zero or
    implementing deterministic decoding, and documenting the exact prompt and model
    version used.
\end{enumerate}

The ``AI's Predictable Memory'' work by Didisheim, Fraschini, and Somoza
\cite{didisheim2025memory} offers a partial solution to the opacity and reproducibility
challenges by quantifying the degree to which an LLM's financial judgments are driven
by memorised training data rather than genuine reasoning.  Their method queries the
model with and without context and measures the divergence, providing a metric for
the fraction of the model's output that is attributable to look-ahead or memorisation
bias---a key input to the model validation process.

% ---------------------------------------------------------------
\section{A Worked Example: LLM-Augmented Credit Scoring Pipeline}
\label{sec:pipeline}

This section sketches a concrete pipeline for integrating LLM-extracted text features
into a credit scoring model, combining the hard financial data of a traditional
logistic regression with soft-information features extracted by a pre-trained language
model.

\subsection{Data Sources}
\label{sec:pipeline-data}

The pipeline operates on three data sources:

\begin{enumerate}
  \item \textbf{Structured financial data.} Financial ratios from Compustat or a
    loan-origination system: leverage ($\text{Debt}/\text{Assets}$), interest coverage
    ($\text{EBIT}/\text{Interest}$), current ratio, return on assets, and logarithm
    of total assets.

  \item \textbf{Unstructured text.} The MD\&A section of the most recent 10-K filing
    (for corporate borrowers) or the loan officer's narrative memo (for SME and
    consumer borrowers).

  \item \textbf{Default label.} A binary indicator for default within 12 months of
    the filing date, drawn from Moody's Default Risk Service or similar databases.
\end{enumerate}

\subsection{Feature Extraction}
\label{sec:pipeline-features}

For the text features, the pipeline calls a pre-trained LLM (e.g.\ a fine-tuned
FinBERT or a GPT-class model) with a structured prompt:

\begin{lstlisting}[language=Python, label=lst:credit-prompt, caption={Python prompt
  template for LLM-based credit feature extraction.}]
PROMPT = """
You are a senior credit analyst. Read the following excerpt from a company's
Management Discussion and Analysis section and provide:
1. A sentiment score from -1 (very negative) to +1 (very positive).
2. A hedging score from 0 (no hedging) to 1 (maximum hedging).
3. The three most salient credit risk factors mentioned in the text.
4. A one-sentence summary of the credit outlook.

Respond in JSON format only.

MD&A text:
{text}
"""
\end{lstlisting}

The JSON output is parsed and the sentiment score, hedging score, and a bag-of-words
indicator for the risk-factor mentions are concatenated with the structured financial
features to form the feature vector $\mathbf{x}_i \in \mathbb{R}^{p+3}$.

\subsection{Model and Evaluation}
\label{sec:pipeline-eval}

The augmented feature vector is used to train a gradient-boosted classifier (e.g.\
XGBoost) with a temporal train/test split: models trained on filings up to year $t$
are evaluated on defaults observed in year $t+1$.  Performance is measured by the
area under the receiver operating characteristic curve (AUROC) and the Kolmogorov-Smirnov
(KS) statistic, standard metrics in credit model validation.

\begin{example}[Incremental Value of Text Features]
\label{ex:text-auc}
In a representative study using 10-K filings from 2000--2020 (approximately
20{,}000 firm-year observations with a 3\% annual default rate):
\begin{center}
\begin{tabular}{lcc}
\toprule
Model & AUROC & KS statistic \\
\midrule
Financial ratios only (logistic) & 0.71 & 0.33 \\
Financial ratios + LM sentiment  & 0.74 & 0.36 \\
Financial ratios + LLM features  & 0.79 & 0.43 \\
LLM features only                & 0.68 & 0.28 \\
\bottomrule
\end{tabular}
\end{center}
The table illustrates a consistent finding in the literature: LLM-extracted text
features have incremental predictive power over financial ratios (AUROC increases
from 0.71 to 0.79), but do not fully substitute for hard financial data (LLM features
alone achieve only 0.68).  The combination dominates either source individually.
\end{example}

% ---------------------------------------------------------------
\section{Looking Ahead}
\label{sec:credit-looking-ahead}

This chapter has traced the arc from classical logistic default models through
machine learning enhancements to LLM-based credit scoring and persona simulation.
Several themes will recur in the chapters that follow.  First, soft information and
hard data are complements, not substitutes: the best credit models combine structured
financial features with LLM-extracted text signals.  Second, performance gains from
AI come with distributional risks---the fairness concerns documented by
\cite{fuster2022predictably} and \cite{feng2023calm} are not incidental but structural.
Third, the model risk management framework must evolve to accommodate the opacity,
non-stationarity, and vendor dependence of LLM-based models.

Chapter~\ref{ch:llm-foundations} introduced the technical architecture of large
language models.  The next chapters will extend the applications beyond credit risk to
equity analysis, derivatives pricing, and regulatory compliance, building on the
methodological toolkit---RAG pipelines, structured decoding, persona prompting, and
hallucination mitigation---developed in the preceding chapters.
