\chapter{Artificial Intelligence, Machine Learning, and Text in Finance}
\label{ch:ai-ml-finance-text}

\begin{remark}[Learning Objectives]
After completing this chapter, the reader should be able to:
\begin{enumerate}
  \item Define artificial intelligence and distinguish it from machine learning
        and deep learning, placing each in the AI--ML--DL--LLM hierarchy.
  \item Contrast symbolic and statistical approaches to AI and explain why
        financial applications favour statistical methods.
  \item Identify at least three ways AI creates value in finance, including
        forecasting, risk management, and regulatory compliance.
  \item Categorise the main sources of financial text and explain why each
        source presents distinct challenges for quantitative analysis.
  \item Trace the evolution of NLP in finance from bag-of-words methods and
        domain-specific dictionaries to transformer-based large language models.
  \item State the scaled dot-product attention formula and explain why the
        transformer architecture is particularly suited to financial text.
\end{enumerate}
\end{remark}

% --- Section: Artificial Intelligence ---
% -----------------------------------------------------------------------
\section{Artificial Intelligence: Definitions and Scope}
\label{sec:ai-definitions}

Few terms in contemporary discourse are used more broadly, or with less
precision, than \emph{artificial intelligence}.  Before treating the
specific techniques that fill the rest of this book, this section
establishes a working definition and traces the intellectual arc that
connects the term's origins in the 1950s to today's large language
models.

\subsection{From Expert Systems to Learning Machines}
\label{subsec:expert-to-learning}

The phrase ``artificial intelligence'' was coined in 1956 by John McCarthy,
who used it to describe the project of making machines perform tasks that
would require intelligence if performed by humans.  For the first three
decades of the field's existence that project was pursued primarily through
\emph{expert systems}: hand-crafted programs encoding domain knowledge as
explicit rules.  A financial expert system of the 1980s might, for instance,
contain hundreds of ``if-then'' rules describing when a loan applicant should
be classified as high-risk.  The appeal was transparency --- every decision
could be traced to a specific rule --- but the brittleness was equally
apparent.  Rules must be written by human experts, they proliferate rapidly
as the domain grows, and they fail without recourse when they encounter
situations their authors did not anticipate.

The limitations of rule-based systems motivated a gradual shift toward
approaches in which intelligence is \emph{induced from data} rather than
\emph{encoded by hand}.  Statistical pattern recognition, which had existed
as a research area since the 1950s, began to attract serious attention from
the AI community in the late 1980s and accelerated through the 1990s as
larger datasets and faster hardware made empirical evaluation feasible.  By
the mid-2000s, data-driven methods had overtaken symbolic approaches in
nearly every application domain where labeled data were available: speech
recognition, image classification, spam filtering, and, eventually, natural
language processing.

\begin{remark}[The shifting definition of AI]
\label{rem:ai-definition-shift}
The boundary of ``artificial intelligence'' has always been socially
constructed as much as technically defined.  Tasks once considered hallmarks
of intelligence --- playing chess, diagnosing disease, translating text ---
tend to be reclassified as ``mere computation'' once machines perform them
reliably.  This ratchet effect, sometimes called the \emph{AI effect}, means
that the field perpetually chases a moving target.  For practical purposes
in this book, we treat AI as the broad enterprise of building systems that
perform cognitive tasks, machine learning as the dominant contemporary
methodology for building such systems, and deep learning and large language
models as particular implementations of that methodology.
\end{remark}

\subsection{Symbolic vs.\ Statistical AI}
\label{subsec:symbolic-vs-statistical}

The tension between \emph{symbolic} and \emph{statistical} approaches
to AI is one of the oldest and most productive fault lines in the field.
Symbolic AI represents knowledge explicitly, manipulating discrete symbols
according to formal rules; logic programming, semantic networks, and
knowledge graphs are its natural home.  Statistical AI, by contrast,
represents knowledge implicitly in the parameters of a model fitted to
data, and makes predictions by computing probabilities rather than
applying rules.

\begin{definition}[Symbolic AI]
\label{def:symbolic-ai}
A \emph{symbolic AI system} represents the world through discrete, human-
readable symbols and reasons by applying explicit, interpretable rules to
those symbols.  Knowledge is encoded by domain experts prior to deployment
and does not change as new data arrive.
\end{definition}

\begin{definition}[Statistical AI]
\label{def:statistical-ai}
A \emph{statistical AI system} represents the world through numerical
parameters estimated from data and makes predictions by evaluating a
learned function.  Knowledge is implicit in the parameter values and
changes whenever the model is retrained on new observations.
\end{definition}

The two paradigms are not mutually exclusive.  Modern systems often combine
a statistical backbone --- a neural network, say --- with symbolic
post-processing such as a grammar checker or a knowledge-graph lookup.
In finance this hybrid character is common: a statistical model may flag
an anomalous transaction, but the final decision about regulatory reporting
follows a deterministic rule specified in compliance policy.  Throughout
this book we lean toward statistical methods because the data richness of
financial markets makes learning from evidence more reliable than encoding
rules by hand, but we return to the symbolic perspective when discussing
interpretability (Chapter~12) and regulatory constraints.

% -----------------------------------------------------------------------
% --- Section: AI-ML-DL Hierarchy ---
\section{The AI--ML--Deep Learning Hierarchy}
\label{sec:ai-ml-dl-hierarchy}

Artificial intelligence, machine learning, and deep learning are nested
concepts, each a proper subset of the one above it.  Understanding the
hierarchy clarifies what distinguishes large language models from earlier
AI systems and why they represent a qualitative advance rather than a
marginal improvement.

\subsection{Machine Learning: Learning from Data}
\label{subsec:ml-from-data}

Machine learning is the subfield of AI concerned with algorithms that
improve their performance at a task through experience, where ``experience''
means exposure to data \parencite{goodfellow2016deep}.  The central
desideratum is \emph{generalization}: a model should perform well not only
on data it has already seen but on new inputs drawn from the same
distribution.

\begin{definition}[Supervised Learning]
\label{def:supervised-learning}
In \emph{supervised learning} an algorithm is given a training set
$\{(\mathbf{x}_i, y_i)\}_{i=1}^n$ of input--output pairs and learns a
function $f_\theta : \mathcal{X} \to \mathcal{Y}$ by minimizing a loss
$\mathcal{L}(\theta) = \frac{1}{n}\sum_{i=1}^n \ell(f_\theta(\mathbf{x}_i), y_i)$
over parameters $\theta$.  The goal is for $f_\theta$ to generalize to
unseen inputs.
\end{definition}

Supervised learning covers classification and regression, the two workhorses
of quantitative finance.  Credit scoring, default prediction, and factor
model estimation are all supervised tasks at their core.  Alongside
supervised learning, \emph{unsupervised} methods discover structure in
unlabeled data --- clustering trades into market regimes, for instance ---
and \emph{reinforcement learning} trains agents to maximize cumulative reward
through sequential interaction, a framework naturally suited to algorithmic
trading and portfolio rebalancing.  \textcite{lopezdeprado2018advances}
provides a thorough treatment of these paradigms in financial applications.

The learning paradigm carries substantial practical implications.
A supervised credit model requires labeled historical defaults and is
constrained to the distribution of the training period.  A model trained
on data from a benign credit cycle may badly underestimate tail risk during
a crisis --- an instance of \emph{distribution shift} that has practical
and regulatory consequences.  These generalization challenges pervade
financial machine learning and motivate much of the methodological care
described in later chapters.

\subsection{Deep Learning: Representation at Scale}
\label{subsec:deep-learning-representation}

Deep learning is machine learning with neural networks of sufficient depth
that the model learns intermediate representations automatically, rather
than requiring the practitioner to specify features by hand.  The key
insight, formalized in the universal approximation theorem for networks
with hidden layers, is that hierarchical compositions of simple nonlinear
functions can represent arbitrarily complex mappings given sufficient
width or depth \parencite{goodfellow2016deep}.

In practice, depth confers a more specific advantage: lower layers tend to
learn general, transferable features (edges in images, phonemes in speech,
common collocations in text) while higher layers combine these into
task-specific representations.  This hierarchical decomposition is
efficient in terms of both data and computation, and it allows a model
pre-trained on a broad dataset to be fine-tuned effectively on a narrow
downstream task with far fewer labeled examples.

\begin{context}
The analogy between deep networks and financial markets is instructive.
Markets aggregate information across many interacting agents operating at
different time scales and with different information sets.  A deep network
aggregates information across layers, each of which operates on the
compressed output of the one below.  Both systems perform hierarchical
computation, and both are difficult to interpret from the outside:
understanding why a market price moved to a particular level is as
challenging as understanding why a deep network produced a particular
output.  This parallel motivates the explainability methods discussed in
Chapter~12.
\end{context}

\subsection{Large Language Models as a Special Case}
\label{subsec:llms-special-case}

Large language models (LLMs) are deep neural networks trained on text at
scale to model the probability distribution over sequences of tokens.
They inherit all the characteristics of deep learning --- learned
representations, transfer learning, gradient-based optimization --- and
add several that are specific to their scale and modality.

\begin{definition}[Large Language Model]
\label{def:llm}
A \emph{large language model} is a neural network with billions of
parameters, trained on a corpus of hundreds of billions of tokens,
whose primary objective is to assign probabilities to token sequences.
Formally, let $x_1, x_2, \dots, x_T$ denote a sequence of $T$ tokens
drawn from a vocabulary $\mathcal{V}$, and let $P_\theta(x_t \mid x_1,
\dots, x_{t-1})$ be the conditional probability the model assigns to
token $x_t$ given the preceding context.  The canonical training
objective is next-token prediction:
\[
  \mathcal{L}_{\mathrm{LM}} = -\sum_{t=1}^{T} \log P_\theta(x_t \mid x_1, \dots, x_{t-1}),
\]
where $\theta$ denotes the model parameters.
\end{definition}

What makes LLMs a \emph{special case} is the combination of scale,
architecture, and emergent capability.  Models trained at sufficient
scale exhibit \emph{emergent abilities} --- capabilities that appear
abruptly as parameters and data increase, not predicted by smooth
extrapolation from smaller models \parencite{wei2022emergent}.  These include multi-step reasoning,
instruction following, and in-context learning, in which the model adapts
to a new task from a handful of examples provided in the prompt without
any parameter update.  For finance, the practical consequence is that a
single LLM can serve as a flexible substrate for tasks as diverse as
earnings call summarization, regulatory document parsing, and quantitative
strategy explanation --- topics explored throughout this book and
exemplified by domain-adapted models such as BloombergGPT
\parencite{wu2023bloomberggpt}.

% -----------------------------------------------------------------------
% --- Section: AI Applications in Finance ---
\section{AI Applications in Finance}
\label{sec:ai-applications-finance}

The financial industry is one of the earliest and most intensive adopters
of data-driven modeling.  Statistical arbitrage, credit scoring, and
algorithmic execution have relied on quantitative methods for decades.
What has changed with modern AI is the scope of problems that can be
addressed, the quality of solutions achievable, and the speed at which
new applications are being discovered.

\subsection{Decision-Making Under Uncertainty}
\label{subsec:decision-making-uncertainty}

Finance is fundamentally a discipline of decision-making under uncertainty.
Every investment, lending, or hedging decision involves forming beliefs about
future states of the world and acting on those beliefs subject to constraints.
AI methods contribute at two distinct levels: they can improve the
\emph{quality of beliefs} (better forecasts, richer models of uncertainty)
and they can improve the \emph{quality of decisions} conditional on those
beliefs (portfolio optimization, dynamic programming, reinforcement learning).

The separation is conceptually clean but empirically entangled.  A model
that produces well-calibrated probability estimates over future returns
enables better portfolio construction, but the portfolio construction
procedure itself may reshape market dynamics and invalidate the forecast
model --- an instance of the \emph{reflexivity} that distinguishes financial
markets from most other prediction targets.  AI systems deployed in
finance must therefore be understood as actors within the market, not
merely observers of it.  This reflexive character adds a strategic
dimension to model design that has no direct analogue in, say, medical
imaging.

\begin{example}[AI in algorithmic trading]
\label{ex:algo-trading}
A market-making algorithm must quote bid and ask prices for a security at
high frequency, earning the spread while managing inventory risk.  A
deep reinforcement learning agent can be trained to optimize this
objective across millions of simulated market episodes, learning execution
strategies that adapt to changing liquidity conditions.  The agent's
actions, however, affect the order book, which in turn affects the signals
on which the agent conditions its decisions.  This feedback loop must be
modeled explicitly in the training environment to avoid policies that
perform well in simulation but destabilize in live markets.
% [CITE: high-frequency trading AI applications, market impact models]
\end{example}

\subsection{Forecasting, Risk, and Alpha Generation}
\label{subsec:forecasting-risk-alpha}

Asset return forecasting, risk estimation, and the identification of
profitable trading signals (``alpha'') are the core quantitative problems
in investment management, and modern machine learning has transformed all
three.  The key challenge in each is the signal-to-noise ratio:
financial returns are nearly unpredictable at short horizons, risk models
must extrapolate into tails observed rarely in historical data, and genuine
alpha erodes quickly once discovered.

Machine learning addresses the forecasting challenge through its capacity
to process high-dimensional feature spaces --- thousands of technical
indicators, macro variables, and cross-sectional characteristics --- without
requiring the analyst to pre-specify which interactions matter.  The
\emph{Advances in Financial Machine Learning} framework
\parencite{lopezdeprado2018advances} articulates the methodological
discipline required to make this work in practice: purging look-ahead
bias from cross-validation, controlling for the multiple comparisons
implicit in large feature sets, and constructing features that are
stationary and informative.  Text-based signals, which are the central
focus of this book, extend this toolkit by incorporating information that
is entirely absent from price and volume data.

Risk modeling has been similarly transformed.  Traditional parametric
approaches such as variance-covariance VaR assume Gaussian returns and
stable correlations, assumptions that fail precisely when they matter
most, during market stress.  Machine learning methods can fit richer
distributional models, incorporate regime-switching dynamics, and learn
from high-dimensional cross-asset data to produce more robust tail risk
estimates.  Language models contribute to risk analysis through their
ability to parse narrative risk disclosures and translate qualitative
language about uncertainty into quantitative signals.

\subsection{Regulatory Compliance and Fraud Detection}
\label{subsec:compliance-fraud}

Beyond investment management, AI plays a critical role in the operational
and compliance functions of financial institutions.  Anti-money laundering
(AML) systems, sanctions screening, Know-Your-Customer (KYC) verification,
and real-time transaction fraud detection all involve identifying rare,
anomalous patterns in high-volume data streams --- tasks for which
statistical classifiers trained on labeled historical incidents have proven
effective.

Language models add particular value in the regulatory domain because
much of the relevant information is textual: suspicious activity reports,
correspondence, legal and regulatory filings, and analyst opinions.  An
LLM can surface unusual phrasing in a client communication, flag
discrepancies between a filing's narrative risk disclosures and its
quantitative exhibits, or assist compliance officers in interpreting
newly issued regulatory guidance.  Chapter~11 treats these applications
in depth.

The regulatory context also introduces obligations that constrain the
design of AI systems.  Explainability requirements under frameworks such
as the European Union's AI Act and GDPR's right to explanation mean that
opaque models may be inadmissible for certain decisions regardless of their
accuracy.  This tension between predictive performance and interpretability
is a recurring theme in financial AI and shapes the architecture choices
discussed throughout the book.

% -----------------------------------------------------------------------
% --- Section: Textual Nature of Financial Data ---
\section{The Textual Nature of Financial Data}
\label{sec:textual-finance}

One of the defining characteristics of financial markets is the
extraordinary diversity of data they generate.  Price feeds, order books,
corporate filings, analyst reports, central bank communications, news
wires, social media, and earnings call transcripts all carry information
relevant to asset prices, yet they differ fundamentally in structure,
volume, and the kind of processing they require.

\subsection{Structured vs.\ Unstructured Data in Finance}
\label{subsec:structured-unstructured}

\begin{definition}[Structured Data]
\label{def:structured-data}
\emph{Structured data} are observations that can be represented as rows
and columns in a relational table, with well-defined data types for each
field.  Each observation corresponds to a known entity (a security, a
trade, a firm) at a specific point in time, and every field carries a
consistent, machine-readable meaning.
\end{definition}

\begin{definition}[Unstructured Data]
\label{def:unstructured-data}
\emph{Unstructured data} lack a pre-defined schema.  Text, audio,
images, and video are canonical examples.  Extracting quantitative
features from unstructured data requires a processing pipeline that
transforms raw content into a representation amenable to statistical
analysis.
\end{definition}

Structured financial data --- daily closing prices, quarterly accounting
items, trade volumes --- have been the primary substrate of quantitative
finance since the 1960s.  They are well-served by conventional statistical
tools: regression, factor models, time series analysis.  Their limitations
are equally well understood: they are backward-looking by construction,
they capture outcomes (prices, earnings) rather than the reasoning and
sentiment that drive those outcomes, and they are observable by many
market participants simultaneously, which limits the informational edge
any single analyst can extract.

Unstructured data, and text in particular, are far more abundant and far
less efficiently processed.  A large institutional investor may have access
to tens of thousands of documents per day --- earnings call transcripts,
analyst notes, regulatory filings, news articles, patent applications ---
that contain decision-relevant information but are too voluminous for
analysts to read systematically.  This processing bottleneck creates a
persistent opportunity for machine learning methods that can scale.

\subsection{Sources of Financial Text: Filings, News, and Earnings Calls}
\label{subsec:financial-text-sources}

The primary sources of financial text can be organized by their
production process, their informational content, and the regulatory
context that governs their disclosure.

Corporate filings submitted to securities regulators are among the most
carefully scrutinized financial documents.  In the United States the SEC's
EDGAR database contains annual reports (10-K), quarterly reports (10-Q),
and current reports (8-K) for all public companies.  These filings include
both quantitative financial statements and mandatory narrative disclosures:
the Management Discussion and Analysis (MD\&A) section, the Risk Factors
section, and forward-looking statements.  Because managers face both legal obligations and reputational incentives
regarding these disclosures, the text carries information complementary
to the financial statements themselves.  \textcite{loughran2011liability}
demonstrated that the sentiment of 10-K filings --- measured using a
purpose-built financial lexicon --- is associated with post-filing stock
returns in ways that general-text dictionaries do not capture.

News and wire services produce high-frequency text that incorporates and
transmits public information.  The relationship between media sentiment
and market returns has been studied at least since \textcite{tetlock2007giving},
who found that pessimistic language in the Wall Street Journal ``Abreast of
the Market'' column predicts downward price pressure followed by
mean-reversion.  Modern news data providers offer real-time, machine-readable
feeds covering equities, fixed income, foreign exchange, and commodities
across dozens of languages and geographies.

Earnings conference calls occupy a position between corporate filings and
market commentary: they are live events in which executives and analysts
exchange views on company performance and outlook.  The transcript of an
earnings call captures not only the scripted prepared remarks but also
the unscripted Q\&A, where analysts probe management on sensitive topics.
Research has consistently found that both the content and the linguistic
properties of these calls --- management tone, the types of questions
asked, hedging language, and even the pace of speech in audio recordings
--- contain incremental information about future firm performance beyond
the reported numbers \parencite{loughran2020textual}.

\begin{example}[Information in financial text]
\label{ex:financial-text-info}
A company reports earnings per share of \$1.20, beating the analyst
consensus of \$1.15.  On the basis of reported numbers alone the result
appears positive.  But the 10-K's Risk Factors section has grown by
30\% year-over-year, three of the six newly added risks concern supply
chain exposure in a politically sensitive region, and the CEO used the
word ``challenging'' eleven times on the earnings call --- more than in
any prior quarter.  A quantitative analyst reading only the headline EPS
figure misses the signal embedded in the text.
\end{example}

\subsection{Why Text is Hard: Noise, Ambiguity, and Context}
\label{subsec:text-challenges}

Despite the informational richness of financial text, extracting
reliable signals from it is technically demanding.  Several properties
of natural language conspire to make this difficult.

First, language is \emph{ambiguous}.  The same sequence of words can
mean different things depending on who wrote it, when, and in what
institutional context.  The word ``uncertainty'' in a Federal Reserve
statement carries a very different connotation than the same word in a
startup founder's shareholder letter.  Domain-specific terminology
compounds the problem: the phrase ``positive outlook'' is bullish in
equity research but cautious in credit rating reports, where ``positive''
denotes a potential upgrade from a currently difficult position.
\textcite{loughran2011liability} showed that approximately three-quarters
of the words classified as negative by the Harvard General Inquirer
dictionary carry neutral or positive connotations in financial filings,
motivating the construction of a domain-specific lexicon.

Second, language is \emph{high-dimensional and sparse}.  A vocabulary of
fifty thousand words --- modest by financial NLP standards --- implies a
document-term matrix with vastly more dimensions than observations,
creating severe statistical challenges for classical regression methods.
Dimensionality reduction through topic modeling, word embeddings, or
language model encoders is necessary before most downstream analyses
can proceed.

Third, financial text is \emph{strategically produced}.  Corporate
disclosures are written by lawyers and communications professionals
who are acutely aware of legal liability and market reaction.  Managers
have incentives to emphasize positive developments, obscure negative
ones through hedging language, and time disclosures strategically.
A model that takes financial text at face value --- treating it as
an unbiased signal of the underlying state of the firm --- will be
systematically misled.  Understanding the incentive structure of document
producers is as important as understanding the statistical properties
of the text itself.

\begin{remark}[Context as a first-class variable]
\label{rem:context-financial-text}
Many of the hardest problems in financial NLP reduce to the challenge
of context.  Determining whether a given passage is bullish or bearish,
legally compliant or potentially misleading, or consistent with prior
disclosures requires the model to maintain a rich representation of the
entity, the time period, the regulatory environment, and the genre of
the document.  This is precisely the kind of long-range contextual
reasoning that transformer-based language models are designed to perform,
and it is a large part of why LLMs represent such a significant advance
over earlier NLP methods for financial applications.
\end{remark}

% -----------------------------------------------------------------------
% --- Section: Bag-of-Words to LLMs ---
\section{From Bag-of-Words to Large Language Models}
\label{sec:bow-to-llms}

The application of quantitative text analysis to finance has a history
stretching back several decades, and the techniques in use have changed
dramatically over that period.  Tracing this evolution clarifies what
the current generation of language models can do that earlier approaches
could not, and situates the rest of the book in a longer intellectual
tradition.

\subsection{Early NLP Approaches in Finance}
\label{subsec:early-nlp-finance}

The first generation of quantitative financial text analysis relied on
methods developed for information retrieval and computational linguistics
that do not require labeled training data at scale.  Bag-of-Words (BoW)
models represent each document as a vector of word counts, discarding
all information about word order and grammar.  Combined with tf-idf
weighting --- which up-weights rare, informative words and down-weights
common ones --- BoW models can classify documents, measure similarity,
and track the frequency of concept-related terms over time.  These
Researchers applied these approaches to financial filings as early as
the 1990s to measure changes in management tone and document readability
\parencite{kearney2014textual}.

Dictionary-based sentiment analysis extends the BoW framework by
mapping words to pre-defined sentiment scores.  The analyst defines
a list of positive words, a list of negative words, and computes the
net sentiment of a document as the difference in their frequencies.
The construction of finance-specific dictionaries was a significant
methodological contribution: \textcite{loughran2011liability} introduced
a lexicon of roughly 2,700 words classified into six categories
(negative, positive, uncertainty, litigious, strong modal, and weak modal)
based on their frequency in financial filings, and demonstrated superior
performance relative to general-purpose dictionaries on a range of
financial prediction tasks.  This lexicon remains widely used as a
benchmark \parencite{loughran2020textual}.

Latent Semantic Analysis (LSA) and Latent Dirichlet Allocation (LDA)
took a different approach, learning document representations from the
co-occurrence structure of words in a corpus rather than from pre-defined
lists.  Topic models identify latent themes in a collection of documents
and represent each document as a mixture over those themes.  Applied to
earnings call transcripts, a topic model might discover that one latent
dimension captures discussions of operational efficiency while another
captures regulatory risk --- themes that no dictionary predicts in advance.
The principal limitation of all these approaches is their shallow treatment
of semantics: words are treated as atomic units, their meanings are context-
independent, and word order carries no information.

\subsection{The Transformer Revolution}
\label{subsec:transformer-revolution}

The publication of \textcite{vaswani2017attention}, ``Attention Is All
You Need,'' in 2017 triggered the most significant transition in the
history of natural language processing.  The Transformer architecture
introduced \emph{self-attention}: a mechanism that allows each position
in a sequence to attend to every other position, computing a weighted
combination of all token representations where the weights are learned
functions of the tokens themselves.  This allows the model to resolve long-range dependencies --- connecting a
pronoun to its antecedent many sentences earlier, or linking a risk factor
in the first paragraph of a filing to a quantitative disclosure in the
financial statements.  Recurrent architectures, which process tokens
sequentially and suffer from vanishing gradients over long contexts,
cannot do this reliably.

\begin{deepdive}
The self-attention mechanism computes, for each position $i$ in a sequence,
a weighted sum of value vectors $\mathbf{v}_j$ at all positions $j$:
\[
  \mathrm{Attention}(\mathbf{Q}, \mathbf{K}, \mathbf{V})
  = \mathrm{softmax}\!\left(\frac{\mathbf{Q}\mathbf{K}^\top}{\sqrt{d_k}}\right)\mathbf{V},
\]
where $\mathbf{Q}$ and $\mathbf{K}$ are linear projections of the input
sequence into query and key spaces of dimension $d_k$, and $\mathbf{V}$
is a linear projection into a value space of dimension $d_v$ (in the
standard single-head presentation $d_v = d_k$ is assumed for notational
simplicity, though multi-head implementations may differ).
The $\sqrt{d_k}$ scaling prevents the dot products from growing large
enough to push the softmax into a saturated regime where gradients
vanish.  Multiple attention heads operating in parallel subspaces allow
the model to jointly attend to different aspects of the context --- one
head might track syntactic agreement while another tracks coreference.
The full derivation and financial NLP implications are developed in
Chapter~2.
\end{deepdive}

Pre-trained Transformer models --- BERT, GPT, RoBERTa, and their
successors --- combined the self-attention mechanism with unsupervised
pre-training on very large text corpora.  The resulting models acquire
rich, context-sensitive representations of words and phrases that can
be fine-tuned on downstream tasks with relatively small labeled datasets.
This transfer-learning paradigm transformed the economics of NLP: tasks
that previously required thousands of labeled examples could now be
solved with hundreds, lowering the barrier to entry for financial
applications where labeled data are expensive to obtain.

\subsection{Why LLMs are Particularly Suited to Finance}
\label{subsec:llms-suited-finance}

The alignment between what large language models do well and what
financial analysis requires is unusually close.  Four properties of
LLMs are especially valuable in the financial domain.

First, LLMs excel at \emph{long-document reasoning}.  Modern models with
context windows of tens or hundreds of thousands of tokens can process
an entire 10-K filing, an earnings call transcript, or a legal agreement
in a single pass, maintaining coherent representations of entities and
relationships across the full document.  This eliminates the
information loss inherent in the chunking strategies required by
earlier architectures.

Second, LLMs support \emph{in-context learning} without parameter
updates.  A compliance analyst can supply the model with three examples
of suspicious transaction patterns and then ask it to evaluate a new
transaction, relying on the model's ability to generalize from the
provided examples.  This few-shot capability is particularly valuable
in finance, where the distribution of interesting events is narrow and
labeled examples are scarce.

Third, the scale of pre-training data ensures broad \emph{world knowledge}
that complements financial domain knowledge.  A model that understands
geopolitical events, supply chain structures, and corporate governance
norms can contextualize a risk disclosure in ways that a purely financial
model cannot.  BloombergGPT \parencite{wu2023bloomberggpt} demonstrated
that further pre-training on a large financial corpus improves performance
on financial NLP benchmarks, but general-domain LLMs already perform
impressively on many tasks, reflecting the degree to which financial
knowledge is embedded in broadly available text.

Fourth, LLMs expose a natural language \emph{interface} that lowers the
barrier to use.  A portfolio manager can query a model in plain English ---
``Summarize the key risks disclosed in this 10-K and flag any that differ
from last year's filing'' --- without writing code or configuring a
classification pipeline.  This democratization of quantitative text
analysis has important implications for how financial research is
conducted and how AI capabilities diffuse across organizational
hierarchies.

\begin{remark}[LLMs are not a panacea]
\label{rem:llm-limitations-preview}
The properties listed above are genuine and consequential, but they
are accompanied by equally genuine limitations.  LLMs hallucinate
--- they generate plausible-sounding but factually incorrect statements
with confidence.  They are sensitive to prompt wording in ways that
make systematic evaluation difficult.  Their reasoning processes are
opaque.  And their behavior on rare, high-stakes financial events may
be poorly calibrated because such events are underrepresented in
pre-training data.  These limitations are not reasons to avoid LLMs
in finance, but they are reasons to deploy them carefully and to
invest seriously in evaluation, monitoring, and human oversight.
Chapter~13 is dedicated to these issues.
\end{remark}

% -----------------------------------------------------------------------
% --- Section: Roadmap ---
\section{Roadmap: What This Book Covers and How to Read It}
\label{sec:book-roadmap}

The remainder of this book is organized in three parts, each building on
the conceptual scaffolding erected in the present chapter.

The first part builds the technical foundations.  The next chapter
develops the architecture of transformer-based language models from the
ground up, deriving the attention mechanism, the pre-training objectives,
and the tokenization schemes that underlie all modern LLMs.  The chapters
that follow cover training and fine-tuning --- how language models are
adapted to specific tasks and domains using supervised fine-tuning,
reinforcement learning from human feedback, and parameter-efficient
methods such as LoRA --- and extend the single-model view to multi-agent
systems, examining how LLMs can be composed into pipelines that plan,
use tools, and interact with external data sources.  The final chapter
of the first part applies these foundations to business valuation,
showing how LLMs can assist in extracting and synthesizing the inputs to
discounted-cash-flow and comparable-company analyses.

The second part covers the primary application domains in finance.
Successive chapters treat credit risk, emerging and frontier applications,
domain-specific language models trained or adapted for finance, financial
NLP and sentiment analysis, portfolio construction and quantitative
trading, regulatory technology and anti-money laundering, and
explainability.  The application chapters are designed to be read
independently once the technical foundations are in place; a reader
with a specific use case --- say, compliance automation or sentiment
analysis --- can proceed directly to the relevant chapter after
completing the foundational part.

The third part turns to evaluation, deployment, and operational concerns.
It provides a systematic treatment of LLM limitations and evaluation
methodology, covers financial text summarization, and examines
privacy-preserving and locally deployed models, addressing the
data-residency and confidentiality concerns that constrain the use of
cloud-hosted models in financial institutions.

Throughout the book, formal definitions and examples are set off from
the main text in the environments the reader has already encountered in
this chapter.  \emph{Context} boxes provide broader perspective --- on
connections to economic theory, institutional practice, or adjacent
fields --- that enriches the main argument without interrupting its
flow.  \emph{Deep Dive} boxes provide technical derivations and
implementation details for readers who want them, and can be skipped
without loss of continuity by those who do not.  Exercises at the end of
each chapter are tagged by difficulty: \textbf{[B]} for beginner,
\textbf{[I]} for intermediate, and \textbf{[A]} for advanced.  Python
code implementing the examples is available in the companion repository
and in the Jupyter notebooks distributed with the book.
